### 2025-04-29 manuscript.tex
Add files via upload

%%CONTEMPORARY%%%
%\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
\documentclass[unnumsec,webpdf,contemporary,large,namedate,figuresright,figuresleft]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}

%%%MODERN%%%
%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}

%%%TRADITIONAL%%%
%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}

%\onecolumn % for one column layouts

%\usepackage{showframe}

% line numbers
%\usepackage[mathlines, switch]{lineno}
\usepackage[switch, columnwise]{lineno}

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

\usepackage{caption}
\usepackage{natbib}
%\usepackage[backend=biber,style=apa,defernumbers=true]{biblatex}
%\addbibresource{./manuscript.bib}

\usepackage{tabularx}
\usepackage{makecell}
\usepackage{placeins}
\usepackage{tikz}
\usetikzlibrary{calc}
%\usepackage[figuresleft]{rotating}

%usepackage[nodisplayskipstretch]{setspace}\setstretch{1.5}

\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
\newcommand{\DrawBox}[4][]{%
    \tikz[overlay,remember picture]{%
        \coordinate (TopLeft)     at ($(#2)+(-0.2em,0.9em)$);
        \coordinate (BottomRight) at ($(#3)+(0.2em,-0.3em)$);
        %
        \path (TopLeft); \pgfgetlastxy{\XCoord}{\IgnoreCoord};
        \path (BottomRight); \pgfgetlastxy{\IgnoreCoord}{\YCoord};
        \coordinate (LabelPoint) at ($(\XCoord,\YCoord)!0.5!(BottomRight)$);
        %
        \draw [red,#1] (TopLeft) rectangle (BottomRight);
        \node [below, #1, fill=none, fill opacity=1] at (LabelPoint) {#4};
    }
}

\DeclareMathOperator{\logit}{\textrm{logit}}

\begin{document}

\journaltitle{Briefings in Bioinformatics}
\DOI{DOI HERE}
\copyrightyear{2019}
\pubyear{2019}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Paper}

\firstpage{1}

%\subtitle{Subject Section}

\title[Detecting Interactions]{Detecting Interactions in Random Forest for Genomic Data}

\author[1,$\ast$]{Hawlader A. Al-Mamun} 
\author[2]{Rob Dunne}
\author[3]{Ross L. Tellam}
\author[4]{Klara Verbyla}

\authormark{Author Name et al.}

\address[1]{\orgdiv{Data61}, \orgname{CSIRO}, \orgaddress{\street{Ecosciences Bldg 41 Boggo Road }, \postcode{Dutton Park 4102}, \state{QLD}, \country{Australia}}}
\address[2]{\orgdiv{Data61}, \orgname{CSIRO}, \orgaddress{\street{Eveleigh Office, 13 Garden Street}, \postcode{2015}, \state{NSW}, \country{Australia}}}
\address[3]{\orgdiv{Agriculture \& Food}, \orgname{CSIRO}, \orgaddress{\street{306 Carmody Rd}, \postcode{St Lucia 4067}, \state{QLD}, \country{Australia}}}
\address[4]{\orgdiv{Agriculture \& Food}, \orgname{CSIRO}, \orgaddress{\street{Street}, \postcode{Postcode}, \state{State}, \country{Australia}}}

\corresp[$\ast$]{Corresponding author. \href{email:hawlader.almamun@data61.csiro.au}{hawlader.almamun@.csiro.au}}

\received{Date}{0}{Year}
\revised{Date}{0}{Year}
\accepted{Date}{0}{Year}

%\editor{Associate Editor: Name}

%\abstract{
%\textbf{Motivation:} .\\
%\textbf{Results:} .\\
%\textbf{Availability:} .\\
%\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Briefings in Bioinformatics}
%online.}

% \abstract{Detecting epistatic interactions is very important because it will help us to understand the genetic mechanism
%   on phenotypic variation. Finding epistatic interactions from high density SNP data is very challenging due to the
%   computational burden induced by the combinatorial number of tests that have to be performed. Random Forest models were
%   widely used in a variety of disciplines including epistatic interactions. Although Random Forest models account for
%   variable interactions, it does not explicitly detect variable interactions. And most of the Random Forest based
%   methods to detect interactions relay on some kind of variable importance measures that suffer when the interacting
%   variables have very small or no marginal effects. We proposed a Random Forest based method that is not based on
%   variable importance and can find epistatic interactions from high density SNP markers in reasonable time. The proposed
%   method was first applied on several simulation datasets representing different data structures with respect to SNP
%   markers and heritibilities. Once the method was validated for finding epistatic interactions from simulated SNP data,
%   it was applied on two different traits (case-control data rheumatoid arthritis, and quantitative traits of mouse body
%   weight) in human and mouse, respectively. In both cases, the method identified several genes closely positioned to the
%   interacting SNPs that potentially have effects on the respective traits.}  \keywords{Epistatic interactions, Random
%   Forest, SNP data}

\abstract{Epistatic interactions can play an important role in the genetic mechanisms that control phenotypic
  variation. However, identifying these interactions in high dimensional genomic data can be very challenging due to the
  large computational burden induced by the high volume of combinatorial tests that have to be performed to explore the
  entire search space.  Random Forests Decision Trees are widely used in a variety of disciplines and are often said to
  detect interactions.  However, Random Forests models do not explicitly detect variable interactions. Most Random
  Forests based methods that claim to detect interactions rely on different forms of variable importance measures that
  suffer when the interacting variables have very small or no marginal effects. The proposed Random Forests based method
  detects interactions using a two-stage approach and is computationally efficient. The approach is demonstrated and
  validated through its application on several simulated datasets representing different data structures with respect to
  genomic data and trait heritabilities. The method is also applied to two high dimensional genomics data sets to
  validate the approach. In both cases, the method results were used to identify several genes closely positioned to the
  interacting markers that showed strong biological potential for contributing to the genetic control for the respective
  traits tested.}

% \boxedtext{
% \begin{itemize}
% \item Key boxed text here.
% \item Key boxed text here.
% \item Key boxed text here.
% \end{itemize}}

\maketitle

\linenumbers
\section{Introduction}
Many phenotypes and disease traits in human, animals and plants are complex and involve many genes and their
interactions.  Quantitative traits such as human height or yeast growth are influenced by many genetic variants each
typically of small effect size.  Numerous genetic variants have been reported in genome-wide association studies (GWAS)
and quantitative trait loci (QTL) studies that are associated with various quantitative traits in different
species. However, for each trait, the reported variants cumulatively can explain only a portion of the total
heritability of the trait.  This phenomenon of unexplained heritability is termed ``missing heritability''
\cite[]{Maher2008}, which is partly due to not considering the genetic variant interactions (epistasis) between the
genetic markers \cite[]{Zuk.et.al.2012}.

Epistasis occurs when there is any non-additive interaction relationship between two or more genes in their combined
effects on a phenotype. It can be potentially identified when the effects of two or more genetic markers associated with
a trait differ from the sum of the individual marker effects. Understanding epistasis is important because it can help
explain the functional mechanisms of the genes that together contribute to disease or trait expression and enable more
targeted and nuanced interventions to be developed. For example, being able to estimate the effect of non-additive
interactions could potentially be exploited to improve the genetic predictions \cite[]{Ansarifar.et.al.2020} for
economically important traits such as wheat yield or carcass weight of cattle.

% The most direct approach to detect epistatic interactions is to evaluate pair-wise linear models (LM) for all
% possible locus pairs (usually SNP pairs) to test for non-additive interactions associated with a complex trait
% \cite[]{Zhang.et.al.2008}. This type of exhaustive search is only feasible when the number of candidate SNPs is
% relatively small. However, as the number of SNP markers in a dataset grows, this approach suffers from the computational
% burden induced by the combinatorial number of tests that have to be performed. For example, The Welcome Trust Case
% Control studies \cite[]{burtonGenomewideAssociationStudy2007brief} has 17,000 individuals (2,000 cases for each of the
% seven diseases and 3,000 combined controls) genotyped with the GeneChip 500K Mapping Array Set (Affymetrix chip), which
% interrogates 500,568 SNPs. Thus more than $1.25\times 10^{13}$ pairwise tests would need to be performed, which is not
% computationally feasible. Moreover, recent technological advancements and reduced costs have led to experimental designs
% utilising more than 1 million SNP markers for hundreds of individuals, which massively confound the ability to identify
% interacting pairs of SNPs \cite[]{Ha.et.al2014, Gholami.et.el2014,1000GenomeProject.2015.brief,utah.edu}. Consequently,
% two step-based methods have been proposed to detect epistatic interactions from genomic data, where in the first step
% SNPs are pre-selected or weighted and then in the second step interaction testing is performed on the pre-selected SNPs
% only. The preselection can be performed based on statistical tests (e.g., GWAS \cite[]{Pecanka.et.al.2017}) or machine
% learning methods with feature selection \cite[]{Jiang.et.al.2009, Meng.et.al.2007}.

The most direct approach to detect epistatic interactions is to use linear models (LM) to evaluate all pair-wise loci
combinations to test for non-additive interactions associated with a complex trait \cite[]{Zhang.et.al.2008}. This type
of exhaustive search is only feasible when the number of candidate markers is relatively small or the computational
power available is extreme. However, as the number of markers in a dataset grows, this approach suffers from the
computational burden induced by the combinatorial number of tests that have to be performed. For example, The Welcome
Trust Case Control studies \cite[]{burtonGenomewideAssociationStudy2007brief} had 17,000 individuals (2,000 cases for
each of seven diseases and 3,000 combined controls) genotyped with the GeneChip 500K Mapping Array Set (Affymetrix
chip), which interrogated 500,568 SNPs. Thus, more than $1.25\times 10^{13}$ pairwise tests would need to be performed, which is
not computationally feasible using standard equipment. Moreover, recent technological advancements and reduced costs
have led to experimental designs utilising more than 1 million variants and markers for hundreds of individuals,
which massively confound the ability to identify interacting pairs of SNPs \cite[]{Ha.et.al2014,
  Gholami.et.el2014,1000GenomeProject.2015.brief,utah.edu}. Consequently, two step-based methods have been proposed to
detect epistatic interactions from genomic data, where in the first step markers were pre-selected or weighted and then
in the second step interaction testing was performed only on the reduced set of markers. The preselection can be
performed based on statistical tests (e.g., GWAS \cite[]{Pecanka.et.al.2017}) or machine learning methods with feature
selection \cite[]{Jiang.et.al.2009, Meng.et.al.2007}.

% Random Forests are an ensemble learning method first proposed by Leo Breiman \cite[]{Breiman2001}. Ensemble learning
% combines multiple learning algorithms/models to improve prediction or classification performance beyond what could be
% obtained from the constituent models.  In a Random Forest, \textit{m} random samples (with replacement) are selected by
% bootstrap sampling. Then multiple decision trees are created based on the bootstrap samples. To break
% down the correlation between features, only a random subset of features is considered at any node of the tree building
% process. For classification problems, the class of the validation data is predicted based on majority voting and for the
% regression problem, the model outputs the mean of the individual trees prediction. There are two main parameters in the
% implementation of Random Forest algorithm namely the number of decision trees (\texttt{num.trees}) and the number of
% randomly selected features (\texttt{mtry}) from which the best split is selected at any given node.

Random Forests (RF) algorithms are an ensemble learning method first proposed by Leo Breiman \cite[]{Breiman2001}. Ensemble
learning combines multiple learning algorithms/models to improve prediction or classification performance beyond what
could be obtained from the constituent models. In Random Forests, \textit{m} random samples (with replacement) are
selected by bootstrap sampling. Then multiple decision trees are created based on the bootstrap samples. To break down
the correlation between features, only a random subset of features is considered at any node of the tree building
process. For classification problems, the class of the validation data is predicted based on majority voting and for the
regression problem, the model outputs the mean of the individual tree's prediction. There are two main parameters in the
implementation of the Random Forests algorithm, namely the number of decision trees (num.trees) and the number of
randomly selected features (\texttt{mtry}) from which the best feature for splitting is selected at any given node.

Random Forest models are widely used \cite[]{lundbergConsistentIndividualizedFeature2019} in a variety of
disciplines. The attraction of the model is based on several attributes, namely:
\begin{itemize} 
\item a high degree of accuracy;
\item a resistance to overfitting, including in the presence of large numbers of redundant variables; 
\item the provision of an out-of-bag error measure; 
\item the provision of a variable ranking at little extra computational cost.
\end{itemize}

In addition it is clear that Random Forests can perform well with correlated and interacting variables, at least in the
sense that the fitted model performs well \cite[]{Ziegler.et.al.2007}. However, most of the Random Forest based
approaches are either based on single variable importance \cite[]{Jiang.et.al.2009, yoshidaSNPInterForestNewMethod2011}
or based on pairwise variable importance \cite[]{Ziegler.et.al.2007}. The Random Forest-based approaches for identifying
interactions based on variable importance measurements cannot distinguish whether a discovered interaction is a true
interaction or simply two variables with strong marginal effects. Additionally, if the interacting variables have small
marginal effects, they will not appear near the top of the variable importance list. This means that detecting
interactions based on variable importance can be problematic. It is also heavily dependent on extensive permutation
tests.  \citeauthor{schmalohrDetectionEpistaticInteractions2018} notes that there is a lack of accurate methods for the
detection of interactions between genomic loci that affect QTLs but that several studies have shown that RF outperforms
other QTL mapping methods at identifying QTL and discrete trait associations. They provided benchmarks on simulated and
real yeast cross data indicating that the RF-based methods outperform other commonly used approaches for detecting
epistasis.

To overcome the challenge of a large number of pairwise SNP interaction tests, we proposed an effective two-step process
based on our observations from using a Random Forest approach. Using simulation data, we observed that ``if two SNPs are
interacting, they appear as parent-child nodes in a decision tree of a random forest more frequently than if they are
not interacting''.

% This observation was used to narrow down the vast search space of pairwise tests to detect SNP
% interactions. First, we built a Random Forest and for each pair of SNPs we count the total number of times any SNP pair
% occurred as parent-child nodes in this forest. Then we selected only the SNP pairs that occurred more times that would
% be expected by chance and exceeded a set threshold of \textit{t}. In our study, we selected all the SNP pairs those have
% at least two (\textit{t}=2) appearance as parent-child nodes in the decision trees of the Random Forest.  Finally, using
% a linear model, each selected SNP pair was tested to determine whether the interaction was significant.

This new method and the validation results obtained using simulated data are presented in this manuscript. In
addition, the results from the use of the method using two real datasets are discussed.

\subsection{The Null Distribution}

Working from the random forest output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the
count of the number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices
and set the diagonal to 0, we form the interaction vector, of length $\frac {p(p-1)} 2 = {p \choose 2}$.
This is likely to be large ($4 \times 10^9$ in the rheumatism example) and contain a large number of zeros (0.999 zeros
in the rheumatism example).

Efron's Empirical Bayes procedure (\cite{efronLargeScaleInferenceEmpirical2010}) is now widely used to determine levels
of significance in high throughput genomics data analysis.
These data sets led to the use of large numbers of parallel tests to determine the significant variables and then
the use of multiple testing corrections \cite{limma.2015}.
Efron suggests that instead of assuming a distribution, such as the $t$, 
a the better approach is to use the large number of variables to model the null distribution, and then do
significance testing.

Applying the insights of Efron's work to this problem we start with the assumption that:
\begin{itemize}
\item most of the counts values come from a null distribution, that is a distribution of interaction
  counts that we would observe if there was no signal in the data and these counts were random values from some null distribution;
\item there is a smaller set of interactions that reflect the underlying biological processes. These are drawn from a
  different distribution to the null counts.
\end{itemize}

We model the data as 
\[
f (j ) = \pi_0 f_0 (j ) + (1 - \pi_0 )f_1 (j )
\]
where $j$ is the count data, $f_0$ is the null distribution and $f_1$ is some contaminating distribution of the
non-null counts. Following \cite{Gauran.et.al.2018} we could make the additional assumption that there is some integer $C$ such that
\[
  f(j ) = \pi_0 f_0 (j ) \textrm{ for } j \leq C.
\]
So we only have the null distribution for counts less than $C.$

As we have count data we model the null distribution $f_0$ as a Poisson or  generalized Poisson ($g$, with parameters
$\lambda$ and $\theta$).
If zero is observed with a significantly higher frequency, as here, we can include a zero-inflation parameter in the
distribution, then $J \sim ZIGP(\eta, \lambda, \theta)$ (zero-inflated generalized Poisson) and the probability that $J = j$, denoted by $f_0 (j)$, 
\[
  f_0(j) = \eta I_{\{0\}} (j) + (1 - \eta) g(j) I_{\{0,1,2,\ldots \}} (j)
\]
where $j$ is a non-negative integer, $0 \leq \eta < 1$, $0 \leq \theta < 1$ and $\lambda > 0$. The indicator function
$I_S (j)$ is equal to 1 if $j \in  S$ and 0 otherwise. Fitting this model involves estimating $\{ \eta, \pi_0,
\lambda$ and $\theta \}$. Having estimated the parameters we choose a cutoff value $D_n$, chosen to control the false
discovery rate, so that counts $\geq  D_n$ are reported as significant.

Using the Rheumatoid data as an example, we have $p= 94485$ and after fitting a random forest with
1000 trees we see about 290 non-terminal splits per tree. We observe 291142 splits with a distribution of frequencies
given in the ``Observed'' column of Table \ref{tab:expected_vs_observed}.

Fitting the ZIGP null model to the Rheumatoid data yields the following parameters:
\begin{itemize}
\item $\eta= 0.4727$    the zero-inflation parameter (proportion of structural zeros);
\item $\lambda= 1.002$   the mean parameter
\item $\theta= 1.1340\times 10^{-12}$   the dispersion parameter. As this is small in this case, we are essentially
  fitting a Poisson distribution;
\item $\pi= 0.9999$  the estimated proportion of null hypotheses
\end{itemize}

See Table \ref{tab:expected_vs_observed} for the expected values for the counts under this model. See also Figure
\ref{figure:plot_zigp_comparison_fixed.png} for a plot of the observed and expected values.

There are several options described in  \cite{Gauran.et.al.2018}.
Using the ``one stage procedure'' and $C_1$ constant (see supplementary information)
we get a $C$ value (ref equation) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.

We note an important caveat here.  \cite{Gauran.et.al.2018} show that the significance level can be
determined and the FDR controlled in a series of simulation examples. For example  Figure 1 (\cite{Gauran.et.al.2018})
shows an example where $\lambda =20$ for the null distribution and the contaminating distribution is a Binomial with
$\mu=50.$
It would appear that these examples all have $\lambda$ and $\mu$ (the mean of the contaminating distribution) with a
greater separation than we have in the Rheumatoid data.
Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forest with more
trees and we will observe counts.

It appears that there is a trade-off  here. And as we are sampling a space of $4 \times 10^{9}$ possible interactions,
we might have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
distribution.
In this case we have followed a simple strategy:
\begin{itemize}
\item threshold the observed count data. We do this a number of ways to do this (list them) and then take the median. IN
  the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
\item test each interaction with a linear model (see below).
\end{itemize}

This, despite some processing of the linear models, is less time consuming than training a much larger Random Forest.

\begin{table}[htbp]
\centering
\caption{Comparison of Observed vs. Expected Counts Under ZIGP Model}
\label{tab:expected_vs_observed}
\begin{tabular}{lrrr}
\hline
Count  & Observed      & Expected      & Difference  \\
\hline
0      & 4,463,369,228 & 4,463,369,228 & 0    \\
1      & 290,502       & 290,490       & 12   \\
2      & 418           & 427           & -9   \\
3      & 89            & 85            & 4    \\
4      & 29            & 21            & 8    \\
5      & 22            & 5             & 17   \\
6      & 14            & 1             & 13  \\
7      & 11            & 0             & 11   \\
8      & 7             & 0             & 7    \\
9      & 5             & 0             & 5    \\
10     & 4             & 0             & 4    \\
11     & 3             & 0             & 3    \\
12     & 3             & 0             & 3    \\
13     & 3             & 0             & 3    \\
14     & 4             & 0             & 4    \\
15     & 1             & 0             & 1    \\
16     & 1             & 0             & 1    \\
17     & 1             & 0             & 1    \\
18     & 1             & 0             & 1    \\
21-30  & 7             & 0             & 7    \\
31-50  & 3             & 0             & 3    \\
51-100 & 9             & 0             & 9    \\
>100   & 3             & 0             & 3    \\
\hline
Total  & 4,463,660,370 & 4,463,660,257 & 113 \\
\hline
\end{tabular}
\end{table}

\begin{figure}[thbp]
  \begin{center}
    \centering
    \includegraphics[totalheight=8cm]{./figs/plot_zigp_comparison_fixed.png}
    \caption{The observed and expected values. The 0 counts are not shown, and the missing bars are count values that
      were not observed.}
    \label{figure:plot_zigp_comparison_fixed.png} 
  \end{center}
\end{figure}

\section{Methods}
% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forest model was created using all avaiable data. Then for each pair 
% of SNP, how many times they appeared as parent-child node in each of the decision trees were recorded. Then the SNP pairs were ranked based on
% the count and only the pairs having count >= 2 were retained. Finally, using the linier equation the interaction term was tested. In our
% studies, we used the ranger R package \cite[]{Wright.and.Ziegler.2017} and we selected \texttt{num.tree} = 1000 and \texttt{mtry} =
% \#SNPs/3.
% For a linear model, an interaction occurs when an independent variable has a different effect on the outcome depending
% on the values of another independent variable. So our variables in the model can be divided between main effects and
% interaction effects. A multiplicative epistasis model was used, where the interaction between two SNPs, $i$ and $j$,  were modelled as the
% product of the two genotype values, using the following linear regression model:

The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model is created using all
available genomic data. For both the simulated and real data examples, the R package ranger
\cite[]{Wright.and.Ziegler.2017} was used with the parameters \texttt{num.tree} = 1000 and \texttt{mtry} = \#SNPs/3.
Once the Random Forests model is built, the number of times each pair of markers was found as parent-child nodes in all
the decision trees were recorded. Then the marker pairs were ranked based on the count and only the pairs having count
more than 2 were retained. A threshold of at least two ($t = 2$) appearances as parent-child nodes in the decision trees
of the Random Forests was used because the use of one ($t = 1$) resulted in a substantially higher number of SNP pairs,
often random occurrences, that did not improve the identification of interactions.  Finally, using linear models the
significance of the interactions are tested. For a linear model, an interaction occurs when an independent variable has
a different effect on the outcome depending on the values of another independent variable. The variables in the model
are divided between main and interaction effects. A multiplicative epistasis model was used, where the interaction
between two markers, $i$ and $j$, are modelled as the product of the two genotype values, using the following linear
regression model:

\begin{equation}
  y = \beta_0 + \beta_i x_i +  \beta_j x_j + \beta_{i,j} x_ix_j + \epsilon.
  \label{eqn:linearmmodel}
\end{equation}

Here $y$ is the qualitative measure of phenotype, and $x$ are the genotype values \{0, 1, 2\}  for SNP $i$ and $j$ respectively,
$\beta_0$ is the mean phenotype, and  $\{\beta_i, \beta_j\}$ are the marginal effects for SNP $i$ and $j$, $\beta_{i,j}$
is the interaction effect between SNP $i$  and  $j$, and $\epsilon$ is the residual with $N(0,1)$. For case-control disease data, the interaction was modelled
through a logit function as follows:

%Here $y$ is the qualitative measure of phenotype, and $x$ are the genotype values \{0, 1, 2\} for SNP $i$ and $j$
%respectively, $\beta_0$ is the mean phenotype, and $\{\beta_i, \beta_j\}$ are the marginal effects for SNP $i$ and $j$,
%$\beta_{i,j}$ is the interaction effect between SNP $i$ and $j$, and $\epsilon$ is the residual with N(0,1).  For
%case-control disease data, the interaction was modelled through a logit function as follows:

\begin{equation}
  \logit (P(y =1)) =  \beta_0 + \beta_i x_i +  \beta_j x_j + \beta_{i,j} x_ix_j + \epsilon,
  \label{eqn:logitmodel}
\end{equation}
where '1' is the disease state and '0' is the control state.  An interaction is considered identified when the p-value of $\beta_{i,j}$ is significant.

\begin{figure}[thbp]
  \begin{center}
    \centering
    \includegraphics[totalheight=5cm]{./figs/Figure_1.png}
    \caption{The stages of the proposed method. (a) A Random Forests model is trained with all genomic markers. (b) A
      parent-child node count matrix is generated. (c) Marker pairs are ranked based on the number of appearances as
      parent-child pairs. (d) Using linear model, interaction tests are performed only for the selected marker pairs. }
    \label{figure:RF_Method} 
  \end{center}
\end{figure}

\subsection{Simulation data}
% We considered two-way interactions between two SNPs based on three interaction models. In the ``base model'' neither the
%  SNPs nor their interaction have any effect on the phenotype; in the interaction only model both SNP have no marginal
% effects but have interaction effect, and in the ``no interaction'' model both SNPs have their marginal effect but there is
% no additional interaction effect. We used publicly available genotype data of 3,534 animals from a single PIC nucleus
% pig line genotyped using the Illumina ProcineSNP60 chip \cite[]{clevelandCommonDatasetGenomic2012}.

% The original dataset contained 52,843 SNPs, but in our studies, the non-informative SNPs were removed leaving 48,015
% SNPs. From these genotype data, the simulated data was created by randomly selecting 5,000, 10,000 and 48,000
% SNPs. Using these genotypes, simulated phenotypes were then created having different heritabilities (h2 = 0.33, 0.50,
% 0.66 and 1.00). Data were generated using 3 different models:
% \begin{itemize}
% \item \textbf{Base model} phenotype = sum of ten QTL effects + noise (depending on heritability);
% \item \textbf{no-interaction model} we added the effect of another 6 SNPs marginal effects on the top of 10 QTL
%   effects, but there were no interaction effects between these 3 SNP pairs. phenotype = sum of ten QTL effects + sum of
%   marginal effect of epistatic SNPs + noise (depending on heritability);
% \item \textbf{interaction model} where phenotypes were the sum of 10 QTL effects plus pairwise interaction effects of
%   the 6 SNPs. phenotype = sum of ten QTL effects + sum of interaction effect + noise (depending on
%   heritability).
% \end{itemize}

Three different interaction scenarios were considered to access the performance of the proposed method in detecting
pairwise interactions.  A simulated dataset was generated using publicly available genotype data containing of 3,534
animals from a single PIC nucleus pig line genotyped using the Illumina ProcineSNP60 chip \cite[]{clevelandCommonDatasetGenomic2012}. The
original dataset contained 52,843 single nucleotide polymorphisms (SNP) but in this study, all non-informative SNPs were
removed leaving 48,015 markers.  The simulated data was created by randomly selecting 5,000; 10,000 and 48,000 markers
from the dataset.  Using these three datasets, simulated phenotypes were then created having different heritabilities
$(h2 = 0.25, 0.33, 0.50, 0.66, 0.75 \textrm{ and } 1.00)$. Data was generated using three different models, a ``Base model'' where none
of the markers nor their interactions have any effect on the phenotype; a ``No Interaction model'' where markers have
marginal effects but there is no additional interaction effect, and an ``Interaction model'' where three pairs of markers have no marginal effects but have interaction
effects. The scenarios can be summarised as:
\begin{itemize}
\item  Base model phenotype = sum of ten QTL effects + noise (depending on heritability);
\item No Interaction model phenotype = sum of ten QTL effects + sum of marginal effect of epistatic SNPs (no interactions) + noise (depending on heritability);
\item  Interaction model phenotype = sum of ten QTL effects + sum of interaction effects for the three pairs (no   marginal effects) + noise (depending on heritability).
\end{itemize}

\subsection{Real datasets -- Wellcome Trust Case Control Consortium data for rheumatoid arthritis}
The first dataset was obtained from the Wellcome Trust Case Control Consortium (WTCCC)
\cite[]{burtonGenomewideAssociationStudy2007brief}. The original dataset contains 14,000 cases for seven complex human
diseases and 3,000 shared controls. The control individuals were from two different sources: 1958 British Birth Cohort
(58C) and the UK Blood Service Control Group (UKBS). Each of the control group has 1,500 individuals. Samples of DNA
from all 17,000 individuals were genotyped with the GeneChip 500K Mapping Array Set (Affymetrix chip).  The current
analysis only used the data for rheumatoid arthritis (RA) that consisted of 4,798 individuals with 1,860 cases and 2,938
controls. Variants having $>10\%$ missing values and MAFs$<0.05$ were removed. SNP markers that failed the
Hardy-Weinberg test (p $\leq$ 0.05) were also removed. As linkage disequilibrium (LD) could affect how the Random Forest
(RF) algorithm selects a SNP at a particular node (for more detail how LD can affect RF, please see
\cite{mengPerformanceRandomForest2009}), the dataset was also filtered based on LD. Using PLINK 1.9
\cite[]{purcellPLINKToolSet2007} and a sliding window of 200 kb, one SNP from a pair of SNPs in that window was removed
if pairwise $r^2 >=0.3$.

\subsection{Real datasets -- Mouse data}
The second real data set we used was published by \cite{solbergProtocolHighthroughputPhenotyping2006};
\cite{valdarGeneticEnvironmentalEffects2006}. A curated version of this dataset was downloaded (Additional file 1 in
\cite{martiniGenomicPredictionEpistasis2017}). The curated mouse dataset contained 9,265 SNPs for each of 1,298
individuals. Pre-corrected residuals of thirteen phenotypes were available as phenotypes. In the current analysis, the
phenotype ``weight at six weeks'' (W6W) was used.

% \subsection{The Proposed Method -- count co-occurrences}
% To overcome the challenge of a large number of pairwise interaction tests, we proposed an effective two-step process
% based on our observations from using Random Forest. Using different simulation data, we observed that ``if two SNPs are
% interacting, they appear as parent-child nodes in a decision tree of a random forest more frequently than if they are
% not interacting''.  So, we utilised this observation to narrow down the vast search space of pairwise tests to detect
% interactions. First, we build a random forest and for each pair of SNPs we count the total number of times any SNP pair
% occurred as parent-child nodes in this forest. Then we select only the SNP pairs that occur more times that would be
% expected by chance and exceed the set threshold. Finally, using a linear model, each selected SNP pair is tested to see
% if the interaction is significant.

\section{Results and Discussion}
\subsection{simulated data}
% Based on publicly available pig genotype data, simulated phenotypes using different scenarios with various combinations
% of number of SNPs, number of interacting SNP pairs, heritability, QTL effect size and interaction effect size were used.
% The performance of the method was assessed based on the proportion of time the method selected the interacting SNP
% pairs for pairwise interaction tests. Table \ref{table:ranking10qtl} shows the importance ranking of three interacting
% SNP pairs based on the number of times they appear as parent-child in the Random Forest decision trees. Higher ranking,
% represented by a smaller number, means a higher count. For example, a value of 1 for pair P3 means that SNP pair 3
% appeared as parent-child more than any other SNP pair in that Random Forest. Since in the base model three pairs of
% interacting SNPs do not have any marginal effects or any interaction effects, they did not appear as parent-child in any
% decision tree of the Random Forest. In the no interaction model, the three interacting SNP pairs appeared as
% parent-child in some of the scenarios. This is because in the no interaction model, although the three pairs do not
% have any interaction effect, they have their own marginal effects. Hence, they were selected at different nodes of the
% decision trees. For the interaction model, rankings of the three interacting SNP pairs are relatively higher than the
% other two models. This ranking table demonstrates that if we rank the SNP pairs based on the count of their appearance
% as parent-child, we only need to test pairwise interaction for a small amount of SNP pairs from the vast number of total
% possible combinations of SNP.

The method's performance was assessed by determining the number of interactions that were correctly identified and
through evaluating the repeatability and false discovery rate for the approach.  The evaluation of the method's ability
to correctly identify interacting markers was assessed based on the proportion of times, the method selected the
interacting marker pairs for pairwise interaction tests. Table  \ref{table:ranking10qtl} shows the importance ranking of three interacting SNP
pairs based on the number of times they appeared as parent-child nodes in the Random Forests decision trees. Higher
ranking, represented by a smaller number, meant a higher count.  For example, a value of 1 for pair P3 means that SNP
pair 3 appeared as parent-child nodes more than any other pair of markers in that Random Forests analysis.

As was expected in the ``Base model'', three pairs of interacting SNP do not appear as parent-child nodes in any decision
tree of the Random Forests. This reflected that these SNP did not have any marginal or interaction effects. In the ``No
interaction model'', the three interacting SNP pairs appeared as parent-child nodes in some of the scenarios. This is
expected as in the ``No Interaction model'' all the individual SNP have their own marginal effects while not
interacting. Hence, they were correctly selected at different nodes of the decision trees.  For the ``Interaction model'',
rankings of the three interacting SNP pairs are relatively higher than the other two models. The rankings observed
demonstrated that if the SNP pairs are ranked based on the count of their appearance as parent-child nodes, it should be
possible to only test pairwise interaction for a subset of marker pairs from the vast number of total possible combina-
tions of markers.  To quantify the efficiency of the method in reducing the search space, the number of tests performed
was compared to the number of markers and potential pairwise combinations. For this test, a threshold of $t = 2$ was
used, meaning that if any of the marker pairs appeared as parent-child nodes in the decision trees of the Random Forests
for at least twice, they were then tested for interaction significance using Equation \ref{eqn:linearmmodel}. Table  \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of
pair-wise interaction tests performed for the ``Interaction model'' alongside the number of possible pairwise combinations
for specified sets of heritability and marker numbers. The final column of Table  \ref{table:Number.of.tests.performed.to.detect.interactions} shows that for all scenarios, less
than one percentage of the total marker pairs were tested for interaction.  In all cases, including for the largest
simulated dataset with 48,000 markers, the three interacting pairs (P1, P2 and P3) were selected for pairwise
interaction tests.

Table  \ref{table:Number.of.tests.performed.to.detect.interactions} demonstrates that, as the number of markers increased the total number of tests performed decreased. This result
is due to the fact that the number of trees (num.tree = 1000) and a threshold of $t = 2$ remained the same for all
simulations. Thus, as combinatorial search space increases due to increasing numbers of markers, the number of pairs of
markers that exceed the parent-child node threshold ($t = 2$) is reduced. This suggests that for smaller marker sets,
the number of decision trees could be reduced to minimise computation and time demands. One strategy could be to
adjust the number of decision trees dependent on the number of markers.

  \begin{table}
\begin{center}
	\begin{tabular}{|p{0.7cm}|c|p{0.375cm}p{0.375cm}p{0.375cm}|p{0.375cm}p{0.375cm}p{0.375cm}|p{0.375cm}p{0.375cm}p{0.375cm}|}
 \hline
Num SNP  & H2     &   \multicolumn{3}{|c|}{Base Model}&    \multicolumn{3}{|c|}{No Interaction}  &
                                                                                               \multicolumn{3}{|c|}{Interaction}\\
          \hline
          &    & P1 & P2 & P3 & P1   & P2  & P3   & P1  & P2 & P3 \\
          \hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{5,000}}}  & 0.33 & NC & NC & NC & 323  & 2   & 287  & 10  & 5  & 7  \\
         & 0.50 & NC & NC & NC & 151  & 1   & 3373 & 6   & 4  & 10 \\
         & 0.66 & NC & NC & NC & 730  & 3   & 699  & 8   & 4  & 9  \\
          & 1.00 & NC & NC & NC & 230  & 1   & 315  & 7   & 3  & 11 \\
          \hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{10,000}}}  & 0.33 & NC & NC & NC & NC   & 20  & NC   & 132 & 9  & 3  \\
         & 0.50 & NC & NC & NC & NC   & 40  & NC   & 293 & 7  & 1  \\
         & 0.66 & NC & NC & NC & NC   & 59  & NC   & 88  & 5  & 1  \\
          & 1.00 & NC & NC & NC & 1077 & 32  & NC   & 86  & 6  & 1  \\
          \hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{48,000}}}  & 0.33 & NC & NC & NC & NC   & NC  & NC   & 4   & 50 & 63 \\
         & 0.50 & NC & NC & NC & 204  & 412 & NC   & 5   & 27 & 12 \\
         & 0.66 & NC & NC & NC & 806  & 224 & NC   & 2   & 30 & 13 \\
          & 1.00 & NC & NC & NC & 663  & 118 & 909  & 13  & 45 & 9  \\
          \hline
\end{tabular}
\end{center}
\caption{Ranking of the SNP pairs based on counts. Simulated data for 10 QTL and 3 pairs of interactions.
  NC = No count. P1, P2 and P3 are three interacting SNP pairs.}
\label{table:ranking10qtl}
\end{table}

% In the proposed two step method for detecting SNP-SNP interactions, a random forest model was built and the count
% determined for how many times a given pair of SNPs appears as parent-child. In the second step, only those SNP pairs
% having the counts higher than a certain number t were tested. In this simulation, we set t = 2, meaning that if any of
% the SNP pairs appeared as parent-child in the decision trees of the Random Forest for at least 2 times, we selected them
% to test for pairwise interaction using Equation 1. The third column of Table
% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of pair-wise interaction tests performed
% with the interaction model for different combinations of heritability and number of SNPs. The fourth column of Table
% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of possible pairwise tests for a given
% number of SNPs. From the last column of the Table \ref{table:Number.of.tests.performed.to.detect.interactions} we can
% see that, using our method we were able to narrow down the search space of pairwise test significantly. For all
% scenarios, less than one percentage of SNP pairs were tested for interaction.

 \begin{table}
\begin{center}
	\begin{tabular}{|c|c|p{1.6cm}|p{1.6cm}|p{1.6cm}|}
 \hline
          Num SNP & H2   & Num test performed & Num of possible combination & Percentage  \\
          \hline
5,000             & 0.33 & 61,346             & 12,497,50                   & 0.004909        \\
                  & 0.50 & 62,610             & 12,497,50                   & 0.00501         \\
                  & 0.66 & 64,111             & 12,497,50                   & 0.00513         \\
                  & 1.00 & 57,747             & 12,497,50                   & 0.004621        \\
      \hline
10,000            & 0.33 & 18226              & 49,995,00                   & 0.000365        \\
                  & 0.50 & 18938              & 49,995,00                   & 0.000379        \\
                  & 0.66 & 20324              & 49,995,00                   & 0.000407        \\
                  & 1.00 & 21204              & 49,995,00                   & 0.000424        \\
       \hline
48,000            & 0.33 & 1742               & 1.15E+09                    & 1.51E-06        \\
                  & 0.50 & 2400               & 1.15E+09                    & 2.08E-06        \\
                  & 0.66 & 3064               & 1.15E+09                    & 2.66E-06        \\
                  & 1.00 & 5531               & 1.15E+09                    & 4.80E-06        \\
        \hline
\end{tabular}
\end{center}
\caption{Number of tests performed to detect interactions.}
\label{table:Number.of.tests.performed.to.detect.interactions}
\end{table}

% \subsection{Effect of heritability and interactions p-value on the proposed method to detect candidate SNP pairs}
% Figure \ref{figure:Eff_Trait_h2_simulated_data.png} shows the pairwise counts for each of the interacting SNP pairs in
% each Random Forest at different heritabilities (h = 0.33, 0.50, 0.66, and 1.00) for 10,000 SNP with 10 QTL and 3 pairs
% of interacting SNPs. At low heritability, the pairwise count is smaller and as heritability increases, the pairwise
% count also increases. Although the three interacting SNP pairs have very similar interaction effects (ranged between 0.3
% to 0.4), they have very different numbers of counts. There is a direct relationship between the $p$-value of the
% interaction term with the pairwise count. Lower $p$-values of the interaction terms have higher pairwise count
% (correlation coefficient = 0.84). This is particularly interesting as even very small interaction effects with high
% significance will not be overlooked because of higher pairwise count.

\subsection{heritability}

The effect of heritability on the method's performance was explored with the pairwise counts for each of the
interacting SNP pairs in each Random Forests at different heritabilities (h2 = 0.33, 0.50, 0.66, and 1.00) for 10,000
SNP with 10 QTL and 3 pairs of interacting SNPs recorded. The results are shown in Figure \ref{figure:Eff_Trait_h2_simulated_data.png}. At lower heritabilities,
the pairwise count is small and as heritability increases, the pairwise count also increases. Although the three
interacting SNP pairs have very similar interaction effects (ranging between 0.3 to 0.4), they have very different
numbers of counts. There is a direct relationship between the $p$-value of the interaction term with the pairwise
count. Lower $p$-values of the interaction terms have higher pairwise count (correlation coefficient = 0.84, $p$-values
and correlations are not shown). This is particularly interesting as even very small interaction effects with high
significance will not be overlooked because of higher pairwise count.

\begin{figure}[thbp]
    \begin{center}
  \centering
\includegraphics[totalheight=5cm]{./figs/Figure_2.png}
\caption{ The effect of trait heritability on pairwise counts. P1, P2 and P3 are interacting SNP pair 1, 2 and 3 respectively.
The Y- axis shows the number of pairwise counts for each interacting SNP pair in each Random Forests.}
  \label{figure:Eff_Trait_h2_simulated_data.png} 
    \end{center}
\end{figure}

\subsection{Repeatability and false discovery rate}
% To test repeatability of our proposed method, we ran the process at different heritabilities (0.75, 0.66, 0.50, 0.33,
% and 0.25) with ten repeats for each heritability. We evaluated the performance of the process based on how many times it
% was successful to select any interacting SNP pair. Table \ref{table:Repeatability_of_the_proposed_method} shows the
% results for 10,000 SNPs with 10 QTLs and 3 pairs of interacting SNPs. In most of the scenarios tested, the method was
% able to successfully select the interacting SNP pairs except for interacting pair one at low heritability.  This is
% expected as at low heritability, there was more noise induced in the data.

To test repeatability of the proposed method, the ``Interaction model'' was generated ten times at each heritability
(0.25, 0.33, 0.50, 0.66, and 0.75) with the correct identification of the interacting marker pairs assessed. Table
\ref{table:Repeatability_of_the_proposed_method} shows the results for 10,000 SNPs with 10 QTLs and 3 pairs of
interacting SNPs. In most of the scenarios tested, the method was able to successfully identify the interacting pairs
except for pair P1 at lower heritabilities. This result appears to be because pair P1 has a lower interaction effect
when compared to pairs P2 and P3. Pair P1 appeared on the candidate lists in all instances but did not pass multiple
test corrections several times when the heritabilities were low (ie 0.25, 0.33).  To assess the false discovery rate of
the method, the number of false positive interactions were counted across the 10 replicates at the different
heritability levels. A false positive was declared when a pair of SNP were not in the original data and when these SNP
were not correlated or in high LD with one of the three simulated pairs. This is because there were multiple
interactions identified where each marker in the pair correlated highly with corresponding markers in one of the
simulated interaction pairs.  Only when this was not the case was a false positive declared. A Pearson correlation
coefficient of less than 0.25 was used as the threshold. This threshold was chosen as using a 0.5 threshold revealed a
number of pairs incorrectly identified as false positives where each marker in the pair was correlated, normally with a
coefficient above 0.4, with one of the simulated three pairs. The mean false positive results for each heritability
calculated using the ten replicates, are shown in Table \ref{table:Repeatability_of_the_proposed_method}. The number of
false positives increased in the data where the heritability was higher. This was understandable as the variance
explained by the marker data (genetic signal) increases at higher heritabilities.

The number of false positives detected by the proposed method was low across all heritabilities.

% \begin{table}
% \begin{center}
%   \begin{tabular}{|cccc|}
%     \hline
%     &P& P2& P\\
%     H2=0.75&10& 10& 10\\
%     H2=0.66&10& 10& 10\\
%     H2=0.50&9&10 &10\\
%     H2=0.33&7&10 &10\\
%     H2=0.25&4&10 &10\\
%     \hline          
%   \end{tabular}
% \end{center}
% \caption{Repeatability of the proposed method. The data contains 10,000 with 10 QTLS and 3 pairs of interacting SNP. The
%   process was repeated 10 times for each heritability times for each heritability.}
% \label{table:Repeatability_of_the_proposed_method}
% \end{table}

\begin{table}
  \begin{center}
    \begin{tabular}{|ccccc|}
      \hline
      $h^2$ & P1 & P2 & P3   &  False + \\
      \hline                
      0.75  & 10 & 10 & 10   &  0\\
      0.66  & 10 & 10 & 10   &  0.1 \\
      0.50  & 9  & 10 & 10   &  0.3 \\
      0.33  & 7  & 10 & 10   &  0.8 \\
      0.25  & 4  & 10 & 10   &  1.9 \\
      \hline          
    \end{tabular}
  \end{center}
  \caption{Repeatability and number of false positives for the proposed method. The data contains 10,000 with 10 QTLS and 3 pairs of interacting SNP.
    The process was repeated 10 times for each heritability with the number of times each pair (P1, P2 or P3) found and the mean number of false positive
    interactions identified indicated.}
  \label{table:Repeatability_of_the_proposed_method}
\end{table}

\section{Results for real datasets}
% Once the method demonstrated that pairwise SNP-SNP interactions could be identified in the simulated data, we then applied
% the method to two different real datasets.

To evaluate the method's performance on real data, two real datasets were assessed for the presence of interacting SNP
pairs.

\subsubsection{Rheumatoid arthritis data}
% After pre-processing and filtering, the RA dataset contained 94,486 SNPs (~4.46 billion possible pairwise interaction
% tests) and 4,798 individuals.  Using random forest, we have selected only 654 SNP pairs (out of more than 4.46 billion
% possible pairs) for pairwise interaction tests (Equation 2). Among the 654 SNP pairs, 26 pairs have significant
% Bonferroni corrected interaction $p$-values ($p < 0.05$). In detail, 24 SNPs were involved in 26 significant interactions
% among which 22 SNPs were from chromosome 6 and two other from chromosome 1 and chromosome 7. Furthermore, several SNPs
% from chromosome 6 were involved in multiple interactions. All the 22 SNPs in chromosome 6 are in the HLA (human
% leukocyte antigen) region spanning from 31.35 MB to 32.80 MB. The HLA region in human genome is known to contain very
% high linkage disequilibrium (LD). If interacting SNP pairs are in the same haplotype or LD block, the significant
% interaction between the pairs might be a haplotype effect on RA, rather than true interaction. Out of 26 pairs of
% interacting SNPs, 12 pairs (all are in chromosome 6) have moderate to high LD (D’ = (0.26 to 0.55)) between the
% pairs. Hence, we retained only 14 SNP pairs which has low LD (D’ $\leq$ 0.2).

% Table 10 lists 14 interacting SNP pairs that are associated with rheumatoid arthritis in humans. Genes
% containing any of the interacting SNPs as well as flanking genes for intergenic SNP are also listed. All 14 significant
% SNP pair interactions had at least one SNP of each pair positioned in the major histocompatibility complex (MHC) class
% II centred on the HLA-DQ to HLA-DR regions located in chromosome 6p21. The HLA region in human chromosome is a 3.78 Mb
% region that is the most gene dense region withing the human genome encoding about 253 genes including several key immune
% response genes \cite[]{shiinaHLAGenomicLoci2009}. The genes in this region predominantly have immune response
% functions. The dominance of the HLA-DQ to HLA-DR region in the interacting SNPs may be due to several factors including
% the clustering of genes with key roles in immune response and its regulation, high gene density including some
% overlapping and anti-sense genes, and a high rate of polymorphism and therefore high linkage disequilibrium. After
% filtering of the SNP pairs for LD less than xx, eight SNP pairs remained. This result suggests that the dominance of the
% identified SNP interactions in this subregion of the MHC class II region reflects strong association of SNPs with the
% rheumatoid arthritis phenotype and not a consequence of the relatively high level of LD in the broader region. Hence,
% all 14 SNP pair interactions were retained for subsequent analyses of gene and potentially SNP functions. Future fine
% mapping of this region in chromosome 6p21 to localise causal SNP interaction pairs associated with rheumatoid arthritis
% is likely to be difficult due to the reasons listed above. Therefore, other approaches are required to understand the
% abundance of interactive SNPs in this subregion of the MHC class II region that are associated with rheumatoid
% arthritis.  The normal function of the MHC class II cell surface receptors encoded by the HLA-DQ and HLA-DR genes is to
% bind and present foreign antigens to T-cells, which then proliferate and signal to B-cells to produce antibodies against
% potential threats. The antigens can be foreign and self-antigens. In the latter instance, immune tolerance to
% self-antigens is normally developed at a very young age. Thus, activation of the MHC class II receptors can elicit or
% suppress T-(helper)-cell responses. In some autoimmune conditions, however, self-antigens activate T-cells and generate
% an inappropriate immune response resulting in chronic inflammation. The HLA-DQ and HLA-DR genes and genes within the
% wider MHC class II region are associated with a variety of inflammatory autoimmune diseases including rheumatoid
% arthritis, type 1 diabetes, systemic sclerosis, psoriasis, amkylosing spondylitis, systemic lupus erythematosus, Crohn’s
% disease, dermatomyositis and Graves’ \cite{Kennedy.et.al.2017};
% \cite{liuGenomewideScreenGenegene2011}; \cite{matzarakiMHCLocusGenetic2017}.

% The current investigation has identified 14 pairs of interacting SNPs in the HLA-DQ - HLA-DR region on chromosome
% 6p21. This finding suggests extensive cis-acting positive and negative interactions between SNPs in this region and
% hence potential for positive and negative interactions between genes in this gene-dense region. How these interactions
% occur is unclear. The identified SNPs may simply be genetic markers of closely linked causal genetic variants. However,
% it is also possible that some of the interacting SNPs may be causal. None of the unique SNPs are present in gene coding
% sequences, 6 are intronic and 14 are intergenic. Examination of the UCSC human genome (hg38 assembly) browser 
% and its Regulation tracks for this region identified several of the interacting SNPs located
% within potential long range gene regulatory elements defined by a variety signatures (i.e.,
% \begin{itemize}
% \item rs2734573, H3K4me1 and H3K27Ac (ENCODE Regulation track);
% \item rs9296009, distal enhancer-like signature (enhD) (ENCODE Cis Regulatory Elements track);
% \item   rs427037, enhD; rs910050, enhD; rs9268831, enhD;
%   \item rs615672, regulatory elements and gene-gene interactions (GeneHancer Regulatory Elements and Gene Interactions
%     track);
%   \item rs9273363, transcription factor cluster and H3K4me1
% (ENCODE Transcription Factor ChIP-seq Clusters track and ENCODE Regulation track, respectively)
% \end{itemize}

% % rs2734573, H3K4me1 and
% % H3K27Ac (ENCODE Regulation track); rs9296009, distal enhancer-like signature (enhD) (ENCODE Cis Regulatory Elements
% % track); rs427037, enhD; rs910050, enhD; rs9268831, enhD; rs615672, regulatory elements and gene-gene interactions
% % (GeneHancer Regulatory Elements and Gene Interactions track); rs9273363, transcription factor cluster and H3K4me1
% % (ENCODE Transcription Factor ChIP-seq Clusters track and ENCODE Regulation track, respectively).

% Two pairs of interactive SNP (rs615672/rs9268831 and rs92688831/rs615672) contained both interactive SNP in potential
% genome regulatory elements. Collectively, this information suggests that local gene-gene interactions in this region
% could be mediated by cis-acting regulatory mechanisms involving short and long range DNA loops. Some of identified
% interactive SNPs were located near or within introns of genes unrelated to the HLA-DQ and HLA-DR genes but interspersed
% within the same region on chromosome 6p21. These genes included NOTCH4 (notch receptor 4), TNXB (tenascin XB) and BTNL2
% (butryophilin-like 2). NOTCH4 regulates cell fate determination, proliferation and apoptosis programs and has a
% regulatory role in branching morphogenesis in the developing vascular system. It is also implicated in the regulation of
% inflammation \cite{Harb.et.el.2020}\cite{Zheng.et.el.2018}, a characteristic feature of autoimmune diseases such as
% rheumatoid arthritis. TNXB (tenascin XB) encodes a tenascin family member of extracellular matrix proteins. It has
% anti-adhesive effects and is involved in extracellular matrix maturation following wound healing \cite{GeneCards}. Thus,
% the protein encoded by TNXB may potentially be involved in tissue repair after damage caused by chronic
% inflammation. BTNL2 belongs to a family of immune regulators. It decreases T-cell proliferation and cytokine release
% during an inflammatory response and has been implicated in a number autoimmune diseases
% \cite{arnettBTNL2ButyrophilinB7Like2007}; \cite{GeneCards};
% % Arnett et al., 2007; "GeneCards - Human Genes | Gene Database | GeneSearch," ;
% \cite{lebrero-fernandezAlteredExpressionButyrophilin2016}. The SNP interaction partner to the SNP identified in BTNL2 on
% 6p21 is located near NRF1 (nuclear respiratory facto 1) on chromosome 7. NRF1 encodes a transcription factor activating
% key metabolic and cell growth genes as well as mitochondrial DNA transcription and replication. NRF1 has also been
% implicated in the regulation of inflammation, \cite{GeneCards}; \cite{vantienenProlongedNrf1Overexpression2010};
% \cite{yangNglycanaseNGLY1Regulates2018}. One possibility is that NRF1 acts in trans directly or indirectly via its
% encoded transcription factor to regulate BTNL2 expression on chromosome 6p21 and thus modulate T-cell mediated immune
% activation and inflammation.

% Notably, multiple genome wide association studies (GWAS) identified some of the interactive SNPs as associated with
% various autoimmune diseases i.e., rs9273363, rheumatoid arthritis; type 1 diabetes; rs6457617, multiple sclerosis,
% Graves disease, systemic sclerosis; rs615672, rheumatoid arthritis ("UCSC Genome Browser Home,").

After pre-processing and filtering, the Rheumatoid arthritis (RA) dataset contained 94,486 SNPs (4.46 billion possible
pairwise interaction tests) and 4,798 individuals. The method, after building a Random Forests analysis, selected only
654 SNP pairs for the second stage of testing to identify significant pairwise interactions (Equation \ref{eqn:logitmodel}). Among the
654 SNP pairs, 26 pairs were found significant using a Bonferroni corrected $p$-value ($p < 0.05$). Specifically, a total of
24 SNP were involved in 26 significant interactions among which 22 of these SNP were from chromosome 6 with two
markers from chromosome 1 and chromosome 7. Furthermore, several SNPs from chromosome 6 were involved in multiple
interactions. All 22 SNP on chromosome 6 were located in the HLA (human leukocyte antigen) region spanning from 31.35 MB
to 32.80 MB. The HLA region in the human genome is known to be highly conserved with high linkage disequilibrium
between markers in the region. If interacting SNP pairs are in the same haplotype or LD block, the significant
interaction between the pairs might be a result of a haplotype effect on RA, rather than true interaction. Out of the 26
pairs of interacting SNPs, 12 pairs (all markers on chromosome 6) have moderate to high LD (D' = (0.26 to 0.55)) between
the pairs.  These were deemed to likely be significant due to being part of the same haplotype. Thus, only 14 SNP pairs
with low LD (D' $\leq$ 0.2) were retained for further investigation.

Table \ref{table:eight.pairs} lists the 14 interacting SNP pairs that were found using the method to be associated with rheumatoid arthritis
in humans.  Genes containing any of the interacting SNPs as well as flanking genes for intergenic SNP are listed in
Supplementary Table S1.

All 14 significant SNP pair interactions had at least one SNP of each pair positioned in the
major histocompatibility complex (MHC) class II centred on the HLA-DQ to HLA-DR regions located in chromosome 6p21. The
HLA region in human chromosome is a 3.78 Mb region that is the most gene dense region within the human genome,
encoding about 253 genes including several key immune response genes \cite[]{shiinaHLAGenomicLoci2009a}. The genes in this region
predominantly have immune response functions. The dominance of the HLA-DQ to HLA-DR region in the interacting SNPs may
be due to several factors including the clustering of genes with key roles in immune response and its regulation, high
gene density including some overlapping and anti-sense genes, and a high rate of polymorphism and therefore high linkage
disequilibrium.  The identification of 12 pairs with both SNP in the HLA-DQ - HLA-DR region on chromosome 6p21
demonstrates that the proposed method is able to identify cis-acting positive and negative SNP interactions
identifying potential gene interactions in this gene-dense HLA-DQ - HLA-DR region.  The identified SNPs may be genetic
markers closely linked to causal genetic variants or it is also possible that some of the interacting SNPs may be
causal. Two identified interacting pairs of SNP (rs615672/rs9268831 and rs92688831/rs615672) were found to have both
markers in potential genome regulatory elements. Collectively, this information suggests that local gene-gene
interactions in this region could be mediated by cis-acting regulatory mechanisms involving short and long range DNA
loops. Some of the identified interactive SNPs were located near or within introns of genes unrelated to the HLA-DQ and
HLA-DR
genes but interspersed within the same region on chromosome 6p21. These genes included NOTCH4 (notch receptor 4),
TNXB (tenascin XB) and BTNL2 (butryophilin-like 2). NOTCH4
regulates cell fate determination, proliferation and apoptosis
programs and has a regulatory role in branching morphogenesis
in the developing vascular system. It is also implicated in the
regulation of inflammation
\cite[]{harbRegulatoryCellNotch4GDF152020,zhengNotch4NegativelyRegulates2018}
, a characteristic feature of autoimmune diseases such as
rheumatoid arthritis. TNXB encodes a tenascin family member of
extracellular matrix proteins. It has anti-adhesive effects and is
involved in extracellular matrix maturation following wound
healing \cite[]{stelzerGeneCardsSuiteGene2016}
. Thus, the protein encoded by
TNXB may potentially be involved in tissue repair after damage
caused by chronic inflammation. BTNL2 belongs to a family of
immune regulators. It decreases T-cell proliferation and cytokine
release during an inflammatory response and has been implicated in a number of autoimmune diseases
\cite[]{arnettBTNL2ButyrophilinB7Like2007,stelzerGeneCardsSuiteGene2016,lebrero-fernandezAlteredExpressionButyrophilin2016a}.

The SNP interaction partner to the SNP identified in BTNL2 on 6p21 is located near NRF1 (nuclear respiratory facto 1) on
chromosome 7. NRF1 encodes a transcription factor activating key metabolic and cell growth genes as well as
mitochondrial DNA transcription and replication. NRF1 has also been implicated in the regulation of inflammation
\cite[]{stelzerGeneCardsSuiteGene2016,vantienenProlongedNrf1Overexpression2010, yangNglycanaseNGLY1Regulates2018}.

One possibility is that NRF1 acts in trans directly or indirectly via its encoded transcription factor to regulate BTNL2
expression on chromosome 6p21 and thus modulate T-cell mediated immune activation and inflammation.

Notably, multiple genome wide association studies (GWAS) identified a subset of the interactive SNPs as associated with
various autoimmune diseases i.e., rs9273363, rheumatoid arthritis; type 1 diabetes; rs6457617, multiple sclerosis,
Graves disease, systemic sclerosis; rs615672, rheumatoid arthritis ("UCSC Genome Browser Home").
\marginpar{is UCSC Genome Browser Home a reference?}
The results fromusing the proposed method on the human rheumatoid arthritis data revealed a number of promising results supporting the
potential value of the proposed method for identifying epistatic interactions in case control data including the
ability to find positive and negative cis-acting interactions.

The results from using the proposed method on the human rheumatoid arthritis data revealed a number of promising results
supporting the potential value of the proposed method for identifying epistatic interactions in case control data
including the ability to find positive and negative cis-acting interactions.

\begin{sidewaystable}
  \tiny
 \begin{center}
   \begin{tabularx}{\textwidth}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c||}
     \hline
  SNP pair & D'         & r$^2$ & SNP1 location & \makecell{SNP1 odds\\ ratio} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} &
  \makecell{SNP1 \\$p$-value$^d$} & SNP2 location & \makecell{SNP2 odds\\ ratio} & \makecell{SNP2 odds \\ ratio (95\% CI)$^b$ } &
  \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction\\ Odds ratio} &\makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} \\
   rs3134926:rs910050     & 0.10 & 0.01 & 6:32232370  & 0.61 & 0.55,0.6  & 1.99E-20 & 6:32347877 & 0.79 & 0.72,0.86 &  7.11E-05& 1.77 & 1.53,2.06 & 3.38E-11 \\
 rs3134926:rs4530903      & 0.14 & 0.01 & 6:32232370  & 0.61 & 0.55,0.67 & 1.99E-20 & 6:32614112 & 1.41 & 1.24,1.59 &  6.59E-05& 1.64 & 1.34,2.01 & 9.90E-04 \\
 rs9431614:rs9296009      & 0.01 & 0.00 & 1:230944623 & 1.00 & 0.89,1.11 & 5.17E+02 & 6:32146738 & 1.80 & 1.64,1.97 &  2.51E-33& 1.62 & 1.35,1.96 & 2.91E-04 \\
 rs3134926:rs9391858      & 0.06 & 0.00 & 6:32232370  & 0.61 & 0.55,0.67 & 1.99E-20 & 6:32373621 & 1.24 & 1.11,1.39 &  5.76E-02& 1.55 & 1.28,1.88 & 4.22E-03 \\
 rs10456058:rs9273363$^a$ & 0.08 & 0.00 & 6:31526961  & 1.11 & 1.03,1.21 & 5.97E+00 & 6:32658495 & 1.34 & 1.23,1.46 &  3.39E-08& 1.50 & 1.32,1.7  & 2.73E-07 \\
 rs910050:rs6457617       & 0.17 & 0.02 & 6:32347877  & 0.79 & 0.72,0.86 & 7.11E-05 & 6:32696074 & 0.44 & 0.4,0.48  &  1.06E-69& 1.39 & 1.22,1.59 & 4.32E-04 \\
 rs9268831:rs615672       & 0.15 & 0.01 & 6:32459971  & 0.69 & 0.63,0.75 & 3.37E-16 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 1.34 & 1.17,1.53 & 9.12E-03 \\
 rs2076533:rs1270689      & 0.03 & 0.00 & 6:32395750  & 0.49 & 0.45,0.54 & 4.60E-55 & 7:12961515 & 1.06 & 0.98,1.16 &  8.23E+01& 1.31 & 1.16,1.48 & 1.48E-02 \\
 rs17421624:rs615672      & 0.18 & 0.01 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 0.73 & 0.64,0.84 & 7.58E-03 \\
 rs6457617:rs2621382$^e$      & 0.17 & 0.03 & 6:32696074  & 0.44 & 0.4,0.48  & 1.06E-69 & 6:32792668 & 1.02 & 0.94,1.11 &  3.44E+02& 0.73 & 0.64,0.83 & 6.46E-04 \\
 rs17421624:rs3134926     & 0.08 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.72 & 0.62,0.83 & 7.12E-03 \\
 rs17421624:rs427037      & 0.03 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:3224448  & 0.70 & 0.62,0.78 &  2.70E-07& 0.69 & 0.59,0.82 & 1.11E-0  \\
rs9296009:rs4530903       & 0.10 & 0.00 & 6:32146738  & 1.80 & 1.64,1.97 & 2.51E-33 & 6:3261411  & 1.41 & 1.24,1.59 &  6.59E-05& 0.67 & 0.55,0.81 & 3.16E-0  \\
rs206015:rs3134926        & 0.06 & 0.00 & 6:32214982  & 1.49 & 1.33,1.67 & 2.90E-09 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.67 & 0.55,0.82 & 4.66E-0  \\
\hline
\end{tabularx}
% \end{tabular}
 \end{center}
 \caption{Eight pairs of interacting SNPs have significant odds ratio after pairwise LD evaluation}
 \label{table:eight.pairs}
 \begin{tablenotes}%
 \item[$^{a}$] rs10456058 has merged into rs2734573.
 \item[$^{b}$]  Odds ratio (OR) of having at least one copy of the risk allele compared to having 0 copies of the same allele for the given SNP.
 \item[$^{c}$]  Odds ratio (OR) of having at least one copy of each risk allele compared to having 0 copies of the same allele for the given SNP pair.
 \item[$^{d}$]  Bonferroni adjusted p-value.
 \item[$^e$]   The pair of SNPs used as an example in section \label{section:A_linear_model_example}.
 \end{tablenotes}
\end{sidewaystable}

\subsection{Mouse  data}
% As the curated mouse data contains 9,265 SNPs there were nearly 43 million possible pairwise comparisons to test for 
% interactions. Using our method, 273,391 pairs of SNP (~0.0064\%) were tested for interactions. The approach identified 
% 59 pairs of interacting SNPs after multiple testing using the Bonferroni correction. These 59 pairs of interactions were 
% from 63 unique SNPs mainly located on MMU chromosomes 2, 8 and 13. We then identified the genes associated with each of 
% these SNPs using the Mouse Genome Informatics \cite[]{mgi}.

The curated mouse data contained 9,265 SNPs resulting in nearly 43 million possible pairwise comparisons to test for the
presence of significant interactions contributing to the phenotype ``weight at 6 weeks''. The proposed method, identified
273,391 pairs of SNP (0.0064\%) for the second stage of interaction testing. After using the Bonferroni multiple testing
correction, 59 pairs of interacting SNPs remained as significant. These 59 pairs of interactions were from 63 unique
SNPs mainly located on MMU chromosomes 2, 8 and 13. To check the validity of the results and proposed method, the genes
associated with each of these SNP were identified using the Mouse Genome Informatics \cite[]{mgi}.

Supplementary Table S2  lists the significant interacting pairs of SNPs that were found associated with the
murine growth trait W6W. The nearest genes (mean distance = 36.7 Kbp; range = 42 bp - 286 Kbp) associated with these SNP
are also listed. The associated MGI ID (Mouse Genome Informatics ID), Entrez ID and GO terms (Gene Ontology) are listed
in Supplementary Table S3. 
The proposed method identified four genes and two functionally
anonymous genes (miR8108 and Gm31784) associated with two additional SNPs, in a 3.8 Mbp interval on MMU Chr8qA2.  Genes
Tm2d2, Tacc1, Slc20a2 and Fgfr1 are contained within a small region of 0.5 Mbp of this larger segment and have a variety
of functions that potentially impact a phenotype associated with growth and development. Fgfr1 (fibroblast growth factor
receptor 1) has wide ranging biological functions associated with regulation of cell division, cell growth and
maturation, embryonic development, wound healing, and formation of blood vessels. In humans, the cellular signalling
mediated by this receptor has roles in the development and growth of bones, particularly bones in the head, face, hand,
feet, and long bones of the arms and legs. Mutations in the gene are associated with human dwarfism (National Institutes
of Health, 2021). Tacc1 (transforming acidic coiled-coil containing protein 1) is involved in processes that promote
cell division before the formation of differentiated tissues \cite[]{stelzerGeneCardsSuiteGene2016}. Tacc1 encodes a protein that
regulates the localization of nuclear receptors, including T3 thyroid hormone and retinoic pathways involved in regula-
tion cell growth and differentiation. Tm2d2 (TM2 domain containing 2) encodes a G-coupled receptor that regulates cell
death and proliferation cascades \cite[]{stelzerGeneCardsSuiteGene2016}. Slc20a2 (solute carrier family 20 member 2) encodes an
inorganic phosphate transporter important in maintaining phosphate homeostasis.  This function is essential for all
intracellular function. The encoded protein also has roles in extracellular matrix and cartilage calcification, which
is important for tissue function \cite[]{stelzerGeneCardsSuiteGene2016}. It is noted that growing bones have high demands for phosphate
(Goretti Penido and Alon 2012).  Four genes identified as potentially involved in interactions were located in a 0.25
Mbp region on the MMU Chr2qA3- 2qB boundary. They were (Mrps2 (mammalian mitochondrial ribosomal protein), Ralgds (ral
guanine nucleotide dissociation stimulator), Ddx31 (dead-box helicase 31) and Ttf1 (transcription termination factor
1). These genes have a variety of functions

including mitochondrial protein synthesis, intracellular signalling, cell growth and division in embryogenesis, and
transcription, respectively \cite[]{stelzerGeneCardsSuiteGene2016}. At present, only Ddx31 is directly linked to
growth related traits, although the molecular functions of the remaining three genes may have this potential.  The
remaining genes identified from the analyses, listed in Tables S2 and S3, map to various chromosomal regions. Notable
gene functions or mutational effects that could influence growth traits include: regulation of cell proliferation,
differentiation and survival (Klf7, Kruppel like factor 7); osteogenesis (Stmn2, Stathmin 2); short stature and skeletal
abnormalities (Wipi2, WD40 repeat protein interacting with phosphoinositides 2); increased blood testosterone,
precocious sexual maturation, increased fecundity, and clearance of luteinizing hormone (Chst8, carbohydrate
sulfo-transferase 8) \cite[]{stelzerGeneCardsSuiteGene2016}. In all, there were several genes identified close to
interacting SNPs listed in Tables S2 and S3 that have effects on murine growth and developmental traits, particularly
osteogenesis and body stature.  Analysis of the mouse data results revealed strong gene candidates for involvement in
growth and development pathways in mice, including genes known to have relevant biological function as well as novel
candidates. This finding indicates that the function of these genes could be influenced by interactions with other genes
and that the proposed method is able to identify multiple promising regions for a quantitative trait for further
exploration.

% \subsection{Gene functions potentially affected by interacting SNPs}
% Supplementary Table S2 lists significant interacting pairs of SNPs that are associated with murine growth traits. The
% nearest genes (mean distance = 36.7 Kbp; range = 42 bp – 286 Kbp) associated with these SNPs are also listed. The
% associated MGI ID (Mouse Genome Informatics ID), Entrez ID and GO terms (Gene Ontology) are listed in Supplementary
% Table S3. Using GSEA (C1 positional gene sets; \cite{Liberzon.et.al.2011}), genes with official gene symbols from Table
% S2 were enriched at two human cytogenic bands equivalent to two syntenic murine cytogenetic band locations (mmu Chr8qA2,
% FDR q = 1.5E-4, Tm2d2, Tacc1, Slc20a2, Fgfr1; mmu Chr2qA3-2qB boundary, FDR q = 8.2E-3, Mrps2, Ralgds, Ddx31, Ttf1).

% The four genes located in mmu Chr8qA2 and two functionally anonymous genes (miR8108 and Gm31784) associated with two
% additional SNPs are all present in a 3.8 Mbp interval. Tm2d2, Tacc1 and Fgfr1 are contained within a small region of
% ~0.5 Mbp of this larger segment. The microRNA miR8108 is located within an intron of Tacc1 and Gm31784 is a poorly
% defined ncRNA located 0.8 Mbp downstream of Fgfr1. Using syntenic human genes to the murine genes near all significant
% SNPs in conjunction with the GSEA CGP database, there was enrichment (q = 2.1E-4) for TM2D2, TACC1, SLC20A2, FGFR1 in
% the Nikolsky Breast Cancer dataset for genes in an amplicon from the human 8p12\_P11 chromosomal region susceptible to
% copy number variation (CNV) in breast cancer samples.  Thus, while the six SNPs in the mmu Chr8qA2 region likely reflect
% SNPs in linkage disequilibrium, another possibility is that the identified SNPs could be contained within or near a CNV
% in this chromosomal region. Notably, a 96,111 bp CNV (1810011O10Rik) has been reported in this region in some mouse
% strains, although it only incorporates Fgf1 \cite{lockeGenomicCopyNumber2015}.

% The four identified and well characterised genes in the mmu Chr8qA2 region have a variety of functions that potentially impact a phenotype associated
% with growth and development. Fgfr1 (fibroblast growth factor receptor 1) has wide ranging biological functions
% associated with regulation of cell division, cell growth and maturation, embryonic development, wound healing, and
% formation of blood vessels. In humans, the cellular signalling mediated by this receptor has roles in the development
% and growth of bones, particularly bones in the head, face, hand, feet, and long bones of the arms and legs.  Mutations
% in the gene are associated with human dwarfism (Genetics Home Reference, NIH US National Library of Medicine
% \cite[]{medline}. Tacc1 (transforming acidic coiled-coil containing protein 1) is involved in processes
% that promote cell division before the formation of differentiated tissues \cite{GeneCards}. Tacc1 encodes a protein that
% regulates the localization of nuclear receptors, including T3 thyroid hormone and retinoic pathways involved in
% regulation cell growth and differentiation. Tm2d2 (TM2 domain containing 2) encodes a G-coupled receptor that regulates
% cell death and proliferation cascades \cite{GeneCards}. Slc20a2 (solute carrier family 20 member 2) encodes an inorganic
% phosphate transporter important in maintaining phosphate homeostasis. This function is essential for all intracellular
% function. The encoded protein also has roles in extracellular matrix and cartilage calcification, which is important for
% tissue function \cite{GeneCards}. It is noted that growing bones have high demands for phosphate
% \cite{Goretti.et.el.2012}.

% The four genes enriched in a localised 0.25 Mbp region on the mmu chr 2qA3-2qB boundary
% (Mrps2 (mammalian mitochondrial ribosomal protein), Ralgds (ral guanine nucleotide dissociation stimulator), Ddx31
% (dead-box helicase 31), Ttf1 (transcription termination factor 1) have a variety of functions \cite{GeneCards} including
% mitochondrial protein synthesis, intracellular signalling, cell growth and division in embryogenesis, and transcription,
% respectively. At present, only Ddx31 is directly linked to growth related traits, although the molecular functions of
% the remaining three genes may have this potential.

% The remaining genes listed in Tables S2 and S3 map to various
% chromosomal regions. Notable gene functions or mutational effects that could influence growth traits include: regulation
% of cell proliferation, differentiation and survival (Klf7, Kruppel like factor 7); osteogenesis (Stmn2, Stathmin 2);
% short stature and skeletal abnormalities (Wipi2, WD40 repeat protein interacting with phosphoinositides 2); increased
% blood testosterone, precocious sexual maturation, increased fecundity, and clearance of luteinizing hormone (Chst8,
% carbohydrate sulfo-transferase 8) \cite{GeneCards}). In all, there were several genes identified close to interacting
% SNPs listed in Tables S2 and S3 that have effects on murine growth and developmental traits, particularly osteogenesis
% and body stature.

\section{Conclusion}
% In this study, we proposed a simple and easy method to select candidate SNP pairs to test for interactions in large
% genomic data. We used a Random Forest algorithm to select a subset of SNP pairs from a vast search space so that the
% selected SNP pairs can be tested for interaction in a reasonable computational time. The method was validated with
% several simulation datasets representing different data structures (with respect to SNP marker density and
% heritabilities). We then used the method to identify SNP interactions for two different traits (case-control data
% rheumatoid arthritis, and quantitative traits of mouse body weight) in human and mouse, respectively. In both cases, the
% method identified several genes closely positioned to the interacting SNPs that potentially have effects on the
% respective traits. Although the approach requires further biological validation, candidate SNP pair markers identified
% by this method will provide valuable information and a genetic basis for understanding a complex trait that is the
% result of effects from multiple genes and their complex interactions. We only used SNP data to demonstrate our proposed
% method. However, the method is equally applicable when the datasets contain both SNP data and environmental data.

In this study, a novel efficient method to identify potentially interacting pairs of markers in genomic data for a
specific trait was proposed and validated. A Random Forests decision tree algorithm was utilised to select a subset of
markers pairs from, the often, vast search space to enable the efficient computational testing of these pairs for
evidence of significant interaction. The method was validated with several simulation datasets representing different
data structures and characteristics including marker density and trait heritability. The method was then used with two
real data sets to identify SNP interactions for two different type of traits, case-control data for the presence of
rheumatoid arthritis in humans, and quantitative trait data for the body weight of mice. In both cases, the method
identified several functionally relevant genes positioned close to the markers shown to be potentially interacting for
the respective traits. The method is demonstrated as having potential in providing new insights into the genetic basis
of complex traits controlled by multiple genes and their interactions. The method can be extended in a biological
context beyond purely genomic data to other 'omic data types or interactions between diverse types of data such as
marker and environmental data to further provide biological insights.

Key points:
\begin{itemize}
\item Detecting epistatic interactions from high dimensional genomic data is very challenging due to the computational
  burden induced by the large number of required combinatorial test.
\item Using a Random Forests algorithm, we provided a computationally simple way to detect pairwise interactions.
\item The method produced reliable results in the case of simulated and interpretable results in several real-world
  data sets
\end{itemize}

\section{Data availability}
PIC nucleus pig line genotype data is available as supporting information in Cleveland et al., 2012. WTCCC genotype data
is available by application to the Wellcome Trust Case Control Consortium Data Access Committee. The mouse dataset is
available as Additional file 1 in \cite{martiniGenomicPredictionEpistasis2017}.

%%%%%%%%%%%%%%

\begin{appendices}

\FloatBarrier
\section{Section title of first appendix}\label{sec11}

\end{appendices}

\section{Competing interests}
There is NO Competing Interest.

\section{Author contributions statement}
H.A., R.D and K.V. conceived the experiment(s), H.A. conducted
the experiment(s), H.A., R.D. and R.T. analysed the results.  H.A., R.D and K.V. and R.T.  wrote and reviewed the manuscript.

\section{Acknowledgements}
The authors thank the anonymous reviewers for their valuable suggestions. This work is supported by CSIRO MLAI Future
Science Platform. This study makes use of data generated by the Wellcome Trust Case Control Consortium. A full list of
the investigators who contributed to the generation of the data is available from www.wtccc.org.uk. Funding for the
project was provided by the Wellcome Trust under award 076113.

%\bibliographystyle{plain}
%\bibliography{reference}

%USE THE BELOW OPTIONS IN CASE YOU NEED AUTHOR YEAR FORMAT.
%\bibliographystyle{abbrvnat}
%\bibliography{reference}

\bibliographystyle{apalike}
\bibliography{./manuscript.bib}
%\printbibliography

\end{document}

% \begin{sidewaystable}
% \begin{center}
%   \begin{tabularx}{\textwidth}{|c|c|c|c|c|c|c|c|c|c|c|c|}
%     \hline
%                 SNP pair & D’         & r$^2$ & SNP1 location & \makecell{SNP1 \\$p$-value$^d$} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} & SNP2 location & \makecell{SNP2 odds \\ ratio (95\% CI) $^b$ } & \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} & \makecell{Interaction \\odds ratio} \\
% \hline
% rs3134926:rs910050       & 0.10       & 0.01  & 6:32232370    & 0.61                            & 0.55,0.6                                    & 2.35E-20      & 6:32347877                                    & 0.79                            & 0.72,0.86                                         & 8.40E-05                               & 1.77 \\
% rs3134926:rs4530903      & 0.14       & 0.01  & 6:32232370    & 0.61                            & 0.55,0.67                                   & 2.35E-20      & 6:32614112                                    & 1.41                            & 1.24,1.59                                         & 7.78E-05                               & 1.64 \\
% rs9431614:rs9296009      & 0.01       & 0.00  & 1:230944623   & 1.00                            & 0.89,1.11                                   & 1             & 6:32146738                                    & 1.80                            & 1.64,1.97                                         & 2.96E-33                               & 1.62 \\
% rs3134926:rs9391858      & 0.06       & 0.00  & 6:32232370    & 0.61                            & 0.55,0.67                                   & 2.35E-20      & 6:32373621                                    & 1.24                            & 1.11,1.39                                         & 0.07                                   & 1.55 \\
% rs10456058:rs9273363$^a$ & 0.08       & 0.00  & 6:31526961    & 1.11                            & 1.03,1.21                                   & 1             & 6:32658495                                    & 1.34                            & 1.23,1.46                                         & 4.01E-08                               & 1.50 \\
% rs910050:rs6457617       & 0.17       & 0.02  & 6:32347877    & 0.79                            & 0.72,0.86                                   & 8.40E-05      & 6:32696074                                    & 0.44                            & 0.4,0.48                                          & 1.25E-69                               & 1.39 \\
% rs9268831:rs615672       & 0.15       & 0.01  & 6:32459971    & 0.69                            & 0.63,0.75                                   & 3.98E-16      & 6:32606394                                    & 0.60                            & 0.54,0.66                                         & 9.19E-24                               & 1.34 \\
% \hline
% \end{tabularx}
% \end{center}
% \caption{Eight pairs of interacting SNPs have significant odds ratio after pairwise LD evaluation}
% \label{table:eight.pairs}
% \begin{tablenotes}%
% \item[$^{a}$] rs10456058 has merged into rs2734573
% \item[$^{b}$]  Odds ratio (OR) of having at least one copy of the risk allele compared to having 0 copies of the same allele for the given SNP
% \item[$^{c}$]  Odds ratio (OR) of having at least one copy of each risk allele compared to having 0 copies of the same allele for the given SNP pair.
% \item[$^{d}$]  Bonferroni adjusted p-value
% \end{tablenotes}
% \end{sidewaystable}

%   \begin{tabularx}{\textwidth}{|c|c|c|c|c|c|c|c|c|c|c|c|}
%     \hline
%                 SNP pair & D’         & r$^2$ & SNP1 location & \makecell{SNP1 \\$p$-value$^d$} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} & SNP2 location & \makecell{SNP2 odds \\ ratio (95\% CI) $^b$ } & \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} & \makecell{Interaction \\odds ratio} \\
% \hline
%   \begin{tabular}{|c|p{1cm}|p{1cm}|p{2cm}|p{2cm}|p{1.5cm}| p{2cm}|p{2cm}|p{1.5cm}| p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
%     \hline
%     \hline
%                 SNP pair & D’ & r$^2$   & SNP1 location & SNP1 odds ratio & SNP1 odds ratio (95\% CI)$^b$ & SNP1 p  value$^d$ &  SNP2 location & SNP2 odds ratio & SNP2 odds ratio (95\% CI) $^b$ & SNP2 pvalue$^d$   & Odds ratio for interaction (95\% CI) $^c$ & interaction p value $^d$   \\
%                SNP pair & D’         & r$^2$ & SNP1 location & \makecell{SNP1 \\$p$-value$^d$} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} & SNP2 location & \makecell{SNP2 odds \\ ratio (95\% CI) $^b$ } & \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} & \makecell{Interaction \\odds ratio} \\

%\onecolumn

### 2025-04-30 manuscript.tex
changes for elsarticle style

@@
-%%CONTEMPORARY%%%
-%\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
-\documentclass[unnumsec,webpdf,contemporary,large,namedate,figuresright,figuresleft]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}
-
-%%%MODERN%%%
-%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}
-
-%%%TRADITIONAL%%%
-%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
-%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}
-
-%\onecolumn % for one column layouts
-
-%\usepackage{showframe}
-
-% line numbers
-%\usepackage[mathlines, switch]{lineno}
-\usepackage[switch, columnwise]{lineno}
-
-\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}
-
-\usepackage{caption}
-\usepackage{natbib}
-%\usepackage[backend=biber,style=apa,defernumbers=true]{biblatex}
-%\addbibresource{./manuscript.bib}
-
+%% 
+%% Copyright 2007-2025 Elsevier Ltd
+%% 
+%% This file is part of the 'Elsarticle Bundle'.
+%% ---------------------------------------------
+%% 
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.3 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.3 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%% 
+%% The list of all files belonging to the 'Elsarticle Bundle' is
+%% given in the file `manifest.txt'.
+%% 
+%% Template article for Elsevier's document class `elsarticle'
+%% with numbered style bibliographic references
+%% SP 2008/03/01
+%% $Id: elsarticle-template-num.tex 272 2025-01-09 17:36:26Z rishi $
+%%
+%%\documentclass[preprint,12pt]{elsarticle}
+
+%% Use the option review to obtain double line spacing
+\documentclass[authoryear,preprint,review,12pt]{elsarticle}
+
+\usepackage{threeparttable}
+\usepackage{url}
+\usepackage{rotating}
+\usepackage{multirow}
 \usepackage{tabularx}
@@
 %usepackage[nodisplayskipstretch]{setspace}\setstretch{1.5}
+
+\usepackage{amsmath}
 
@@
 
+%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
+%% for a journal layout:
+%% \documentclass[final,1p,times]{elsarticle}
+%% \documentclass[final,1p,times,twocolumn]{elsarticle}
+%% \documentclass[final,3p,times]{elsarticle}
+%% \documentclass[final,3p,times,twocolumn]{elsarticle}
+%% \documentclass[final,5p,times]{elsarticle}
+%% \documentclass[final,5p,times,twocolumn]{elsarticle}
+
+%% For including figures, graphicx.sty has been loaded in
+%% elsarticle.cls. If you prefer to use the old commands
+%% please give \usepackage{epsfig}
+
+%% The amssymb package provides various useful mathematical symbols
+\usepackage{amssymb}
+%% The amsmath package provides various useful equation environments.
+\usepackage{amsmath}
+%% The amsthm package provides extended theorem environments
+%% \usepackage{amsthm}
+
+%% The lineno packages adds line numbers. Start line numbering with
+%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
+%% for the whole article with \linenumbers.
+\usepackage[switch, columnwise]{lineno}
+
+\journal{Nuclear Physics B}
+
 \begin{document}
 
-\journaltitle{Briefings in Bioinformatics}
-\DOI{DOI HERE}
-\copyrightyear{2019}
-\pubyear{2019}
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Paper}
-
-\firstpage{1}
+\begin{frontmatter}
+
+%% Title, authors and addresses
+
+%% use the tnoteref command within \title for footnotes;
+%% use the tnotetext command for theassociated footnote;
+%% use the fnref command within \author or \affiliation for footnotes;
+%% use the fntext command for theassociated footnote;
+%% use the corref command within \author for corresponding author footnotes;
+%% use the cortext command for theassociated footnote;
+%% use the ead command for the email address,
+%% and the form \ead[url] for the home page:
+\title{Detecting Interactions in Random Forest for Genomic Data}
+\author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
+\ead{halmamun@intergrain.com}
+%% \ead[url]{home page}
+%% \fntext[label2]{}
+%% \cortext[cor1]{}
+%% \affiliation{organization={},
+%%             addressline={},
+%%             city={},
+%%             postcode={},
+%%             state={},
+%%             country={}}
+%% \fntext[label3]{}
+
+\title{Detecting Interactions in Random Forest for Genomic Data}
+
+%% use optional labels to link authors explicitly to addresses:
+%% \author[label1,label2]{}
+%% \affiliation[label1]{organization={},
+%%             addressline={},
+%%             city={},
+%%             postcode={},
+%%             state={},
+%%             country={}}
+%%
+%% \affiliation[label2]{organization={},
+%%             addressline={},
+%%             city={},
+%%             postcode={},
+%%             state={},
+%%             country={}}
+
+\author{} %% Author name
+
+%% Author affiliation
+\affiliation{organization={},%Department and Organization
+            addressline={}, 
+            city={},
+            postcode={}, 
+            state={},
+            country={}}
+
+%% Abstract
+\begin{abstract}
+Epistatic interactions can play an important role in the genetic mechanisms that control phenotypic
+  variation. However, identifying these interactions in high dimensional genomic data can be very challenging due to the
+  large computational burden induced by the high volume of combinatorial tests that have to be performed to explore the
+  entire search space.  Random Forests Decision Trees are widely used in a variety of disciplines and are often said to
+  detect interactions.  However, Random Forests models do not explicitly detect variable interactions. Most Random
+  Forests based methods that claim to detect interactions rely on different forms of variable importance measures that
+  suffer when the interacting variables have very small or no marginal effects. The proposed Random Forests based method
+  detects interactions using a two-stage approach and is computationally efficient. The approach is demonstrated and
+  validated through its application on several simulated datasets representing different data structures with respect to
+  genomic data and trait heritabilities. The method is also applied to two high dimensional genomics data sets to
+  validate the approach. In both cases, the method results were used to identify several genes closely positioned to the
+  interacting markers that showed strong biological potential for contributing to the genetic control for the respective
+  traits tested.
+\end{abstract}
+
+%%Graphical abstract
+\begin{graphicalabstract}
+%\includegraphics{grabs}
+\end{graphicalabstract}
+
+%%Research highlights
+\begin{highlights}
+\item Research highlight 1
+\item Research highlight 2
+\end{highlights}
+
+%% Keywords
+\begin{keyword}
+%% keywords here, in the form: keyword \sep keyword
+
+%% PACS codes here, in the form: \PACS code \sep code
+
+%% MSC codes here, in the form: \MSC code \sep code
+%% or \MSC[2008] code \sep code (2000 is the default)
+
+\end{keyword}
+
+\end{frontmatter}
+
+%% Add \usepackage{lineno} before \begin{document} and uncomment 
+%% following line to enable line numbers
+%% \linenumbers
+
+%% main text
+%%
+
+%% Use \section commands to start a section
+\section{Example Section}
+\label{sec1}
+%% Labels are used to cross-reference an item using \ref command.
+
+Section text. See Subsection \ref{subsec1}.
+
+%% Use \subsection commands to start a subsection.
+\subsection{Example Subsection}
+\label{subsec1}
+
+Subsection text.
+
+%% Use \subsubsection, \paragraph, \subparagraph commands to 
+%% start 3rd, 4th and 5th level sections.
+%% Refer following link for more details.
+%% https://en.wikibooks.org/wiki/LaTeX/Document_Structure#Sectioning_commands
+
+\subsubsection{Mathematics}
+%% Inline mathematics is tagged between $ symbols.
+This is an example for the symbol $\alpha$ tagged as inline mathematics.
+
+%% Displayed equations can be tagged using various environments. 
+%% Single line equations can be tagged using the equation environment.
+\begin{equation}
+f(x) = (x+a)(x+b)
+\end{equation}
+
+%% Unnumbered equations are tagged using starred versions of the environment.
+%% amsmath package needs to be loaded for the starred version of equation environment.
+\begin{equation*}
+f(x) = (x+a)(x+b)
+\end{equation*}
+
+%% align or eqnarray environments can be used for multi line equations.
+%% & is used to mark alignment points in equations.
+%% \\ is used to end a row in a multiline equation.
+\begin{align}
+ f(x) &= (x+a)(x+b) \\
+      &= x^2 + (a+b)x + ab
+\end{align}
+
+\begin{eqnarray}
+ f(x) &=& (x+a)(x+b) \nonumber\\ %% If equation numbering is not needed for a row use \nonumber.
+      &=& x^2 + (a+b)x + ab
+\end{eqnarray}
+
+%% Unnumbered versions of align and eqnarray
+\begin{align*}
+ f(x) &= (x+a)(x+b) \\
+      &= x^2 + (a+b)x + ab
+\end{align*}
+
+\begin{eqnarray*}
+ f(x)&=& (x+a)(x+b) \\
+     &=& x^2 + (a+b)x + ab
+\end{eqnarray*}
+
+%% Refer following link for more details.
+%% https://en.wikibooks.org/wiki/LaTeX/Mathematics
+%% https://en.wikibooks.org/wiki/LaTeX/Advanced_Mathematics
+
+%% Use a table environment to create tables.
+%% Refer following link for more details.
+%% https://en.wikibooks.org/wiki/LaTeX/Tables
+\begin{table}[t]%% placement specifier
+%% Use tabular environment to tag the tabular data.
+%% https://en.wikibooks.org/wiki/LaTeX/Tables#The_tabular_environment
+\centering%% For centre alignment of tabular.
+\begin{tabular}{l c r}%% Table column specifiers
+%% Tabular cells are separated by &
+  1 & 2 & 3 \\ %% A tabular row ends with \\
+  4 & 5 & 6 \\
+  7 & 8 & 9 \\
+\end{tabular}
+%% Use \caption command for table caption and label.
+\caption{Table Caption}\label{fig1}
+\end{table}
+
+%% Use figure environment to create figures
+%% Refer following link for more details.
+%% https://en.wikibooks.org/wiki/LaTeX/Floats,_Figures_and_Captions
+\begin{figure}[t]%% placement specifier
+%% Use \includegraphics command to insert graphic files. Place graphics files in 
+%% working directory.
+\centering%% For centre alignment of image.
+\includegraphics{example-image-a}
+%% Use \caption command for figure caption and label.
+\caption{Figure Caption}\label{fig1}
+%% https://en.wikibooks.org/wiki/LaTeX/Importing_Graphics#Importing_external_graphics
+\end{figure}
+
+%% The Appendices part is started with the command \appendix;
+%% appendix sections are then done as normal sections
+\appendix
+\section{Example Appendix Section}
+\label{app1}
+
+Appendix text.
+
+%% For citations use: 
+%%       \cite{<label>} ==> [1]
+
+%%
+%%Example citation, See \cite{lamport94}.
+
+%% If you have bib database file and want bibtex to generate the
+%% bibitems, please use
+%%
+%%  \bibliographystyle{elsarticle-num} 
+%%  \bibliography{<your bibdatabase>}
+
+%% else use the following coding to input the bibitems directly in the
+%% TeX file.
+
+%% Refer following link for more details about bibliography and citations.
+%% https://en.wikibooks.org/wiki/LaTeX/Bibliography_Management
+
+%\endinput
+%%
+%% End of file `elsarticle-template-num.tex'.
+
+%\onecolumn % for one column layouts
+
+%\usepackage{showframe}
+
+% line numbers
+%\usepackage[mathlines, switch]{lineno}
+%\usepackage[switch, columnwise]{lineno}
+
+%\theoremstyle{thmstyleone}%
+%\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+%\newtheorem{proposition}[theorem]{Proposition}%
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
+%\theoremstyle{thmstyletwo}%
+%\newtheorem{example}{Example}%
+%\newtheorem{remark}{Remark}%
+%\theoremstyle{thmstylethree}%
+%\newtheorem{definition}{Definition}
+
+%\usepackage{caption}
+%\usepackage{natbib}
+%\usepackage[backend=biber,style=apa,defernumbers=true]{biblatex}
+%\addbibresource{./manuscript.bib}
+%\journal{Nuclear Physics B}
+
+%\begin{document}
+
+%\journaltitle{Briefings in Bioinformatics}
+%\DOI{DOI HERE}
+%\copyrightyear{2019}
+%\pubyear{2019}
+%\access{Advance Access Publication Date: Day Month Year}
+%\appnotes{Paper}
+
+%\firstpage{1}
 
@@
 
-\authormark{Author Name et al.}
+%\authormark{Author Name et al.}
 
@@
 
-\corresp[$\ast$]{Corresponding author. \href{email:hawlader.almamun@data61.csiro.au}{hawlader.almamun@.csiro.au}}
-
-\received{Date}{0}{Year}
-\revised{Date}{0}{Year}
-\accepted{Date}{0}{Year}
+%\corresp[$\ast$]{Corresponding author. \href{email:hawlader.almamun@data61.csiro.au}{hawlader.almamun@.csiro.au}}
+
+%\received{Date}{0}{Year}
+%\revised{Date}{0}{Year}
+%\accepted{Date}{0}{Year}
 
@@
 
-\abstract{Epistatic interactions can play an important role in the genetic mechanisms that control phenotypic
-  variation. However, identifying these interactions in high dimensional genomic data can be very challenging due to the
-  large computational burden induced by the high volume of combinatorial tests that have to be performed to explore the
-  entire search space.  Random Forests Decision Trees are widely used in a variety of disciplines and are often said to
-  detect interactions.  However, Random Forests models do not explicitly detect variable interactions. Most Random
-  Forests based methods that claim to detect interactions rely on different forms of variable importance measures that
-  suffer when the interacting variables have very small or no marginal effects. The proposed Random Forests based method
-  detects interactions using a two-stage approach and is computationally efficient. The approach is demonstrated and
-  validated through its application on several simulated datasets representing different data structures with respect to
-  genomic data and trait heritabilities. The method is also applied to two high dimensional genomics data sets to
-  validate the approach. In both cases, the method results were used to identify several genes closely positioned to the
-  interacting markers that showed strong biological potential for contributing to the genetic control for the respective
-  traits tested.}
-
 % \boxedtext{
@@
 
-\maketitle
+%\maketitle : Missing \endcsname inserted.
 
@@
 
-\begin{appendices}
+%\begin{appendices}
 
@@
 
-\end{appendices}
+%\end{appendices}
 
@@
 
-\bibliographystyle{apalike}
+\bibliographystyle{elsarticle-harv}
 \bibliography{./manuscript.bib}
-%\printbibliography
 

### 2025-04-30 manuscript.tex
changes for elsarticle style

@@
 %% Use the option review to obtain double line spacing
-\documentclass[authoryear,preprint,review,12pt]{elsarticle}
-
-\usepackage{threeparttable}
-\usepackage{url}
-\usepackage{rotating}
-\usepackage{multirow}
-\usepackage{tabularx}
-\usepackage{makecell}
-\usepackage{placeins}
-\usepackage{tikz}
-\usetikzlibrary{calc}
-%\usepackage[figuresleft]{rotating}
-
-%usepackage[nodisplayskipstretch]{setspace}\setstretch{1.5}
-
-\usepackage{amsmath}
-
-\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
-\newcommand{\DrawBox}[4][]{%
-    \tikz[overlay,remember picture]{%
-        \coordinate (TopLeft)     at ($(#2)+(-0.2em,0.9em)$);
-        \coordinate (BottomRight) at ($(#3)+(0.2em,-0.3em)$);
-        %
-        \path (TopLeft); \pgfgetlastxy{\XCoord}{\IgnoreCoord};
-        \path (BottomRight); \pgfgetlastxy{\IgnoreCoord}{\YCoord};
-        \coordinate (LabelPoint) at ($(\XCoord,\YCoord)!0.5!(BottomRight)$);
-        %
-        \draw [red,#1] (TopLeft) rectangle (BottomRight);
-        \node [below, #1, fill=none, fill opacity=1] at (LabelPoint) {#4};
-    }
-}
-
-\DeclareMathOperator{\logit}{\textrm{logit}}
+\documentclass[authoryear,preprint,12pt,3p]{elsarticle}
 
@@
 %% \documentclass[final,5p,times,twocolumn]{elsarticle}
+%% https://tex.stackexchange.com/questions/261864/how-do-i-determine-the-attribute-of-elsarticle-document
+
+\usepackage{threeparttable}
+\usepackage{url}
+\usepackage{multirow}
+\usepackage{tabularx}
+\usepackage{makecell}
+\usepackage{placeins}
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\usepackage[figuresleft]{rotating}
+%\usepackage{rotating}
+
+%usepackage[nodisplayskipstretch]{setspace}\setstretch{1.5}
+
+\usepackage{amsmath}
+
+% \newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
+% \newcommand{\DrawBox}[4][]{%
+%     \tikz[overlay,remember picture]{%
+%         \coordinate (TopLeft)     at ($(#2)+(-0.2em,0.9em)$);
+%         \coordinate (BottomRight) at ($(#3)+(0.2em,-0.3em)$);
+%         %
+%         \path (TopLeft); \pgfgetlastxy{\XCoord}{\IgnoreCoord};
+%         \path (BottomRight); \pgfgetlastxy{\IgnoreCoord}{\YCoord};
+%         \coordinate (LabelPoint) at ($(\XCoord,\YCoord)!0.5!(BottomRight)$);
+%         %
+%         \draw [red,#1] (TopLeft) rectangle (BottomRight);
+%         \node [below, #1, fill=none, fill opacity=1] at (LabelPoint) {#4};
+%     }
+% }
+
+\DeclareMathOperator{\logit}{\textrm{logit}}
 
@@
 %% for the whole article with \linenumbers.
-\usepackage[switch, columnwise]{lineno}
-
-\journal{Nuclear Physics B}
+\usepackage[mathlines, switch, columnwise]{lineno}
+
+\journal{Computational and Structural Biotechnology Reports}
 
@@
 \title{Detecting Interactions in Random Forest for Genomic Data}
+
 \author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
@@
 %% \cortext[cor1]{}
-%% \affiliation{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
+\affiliation{organization={},
+             addressline={},
+             city={},
+             postcode={},
+             state={},
+             country={}}
 %% \fntext[label3]{}
-
-\title{Detecting Interactions in Random Forest for Genomic Data}
 
@@
             country={}}
+
+\author{Rob Dunne} %% Author name
+\affiliation{organization={Data61, CSIRO},%Department and Organization
+  addressline={13 Garden Street}, 
+  city={Eveleigh},
+  postcode={2015}, 
+  state={NSW},
+  country={Australia}}
+
+\author{Ross L. Tellam}
+\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
+  addressline={306 Carmody Rd}, 
+  city={St Lucia},
+  postcode={4067}, 
+  state={QLD},
+  country={Australia}}
+
+\author{Klara Verbyla}
+\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
+  addressline={306 Carmody Rd}, 
+  city={St Lucia},
+  postcode={4067}, 
+  state={QLD},
+  country={Australia}}
 
@@
 
-%%Graphical abstract
-\begin{graphicalabstract}
-%\includegraphics{grabs}
-\end{graphicalabstract}
+% %%Graphical abstract
+% \begin{graphicalabstract}
+% %\includegraphics{grabs}
+% \end{graphicalabstract}
 
@@
 \begin{keyword}
+Epistatic interactions \sep  Random Forest \sep SNP data
+
 %% keywords here, in the form: keyword \sep keyword
@@
 \end{frontmatter}
-
-%% Add \usepackage{lineno} before \begin{document} and uncomment 
-%% following line to enable line numbers
-%% \linenumbers
-
-%% main text
-%%
-
-%% Use \section commands to start a section
-\section{Example Section}
-\label{sec1}
-%% Labels are used to cross-reference an item using \ref command.
-
-Section text. See Subsection \ref{subsec1}.
-
-%% Use \subsection commands to start a subsection.
-\subsection{Example Subsection}
-\label{subsec1}
-
-Subsection text.
-
-%% Use \subsubsection, \paragraph, \subparagraph commands to 
-%% start 3rd, 4th and 5th level sections.
-%% Refer following link for more details.
-%% https://en.wikibooks.org/wiki/LaTeX/Document_Structure#Sectioning_commands
-
-\subsubsection{Mathematics}
-%% Inline mathematics is tagged between $ symbols.
-This is an example for the symbol $\alpha$ tagged as inline mathematics.
-
-%% Displayed equations can be tagged using various environments. 
-%% Single line equations can be tagged using the equation environment.
-\begin{equation}
-f(x) = (x+a)(x+b)
-\end{equation}
-
-%% Unnumbered equations are tagged using starred versions of the environment.
-%% amsmath package needs to be loaded for the starred version of equation environment.
-\begin{equation*}
-f(x) = (x+a)(x+b)
-\end{equation*}
-
-%% align or eqnarray environments can be used for multi line equations.
-%% & is used to mark alignment points in equations.
-%% \\ is used to end a row in a multiline equation.
-\begin{align}
- f(x) &= (x+a)(x+b) \\
-      &= x^2 + (a+b)x + ab
-\end{align}
-
-\begin{eqnarray}
- f(x) &=& (x+a)(x+b) \nonumber\\ %% If equation numbering is not needed for a row use \nonumber.
-      &=& x^2 + (a+b)x + ab
-\end{eqnarray}
-
-%% Unnumbered versions of align and eqnarray
-\begin{align*}
- f(x) &= (x+a)(x+b) \\
-      &= x^2 + (a+b)x + ab
-\end{align*}
-
-\begin{eqnarray*}
- f(x)&=& (x+a)(x+b) \\
-     &=& x^2 + (a+b)x + ab
-\end{eqnarray*}
-
-%% Refer following link for more details.
-%% https://en.wikibooks.org/wiki/LaTeX/Mathematics
-%% https://en.wikibooks.org/wiki/LaTeX/Advanced_Mathematics
 
@@
 %% https://en.wikibooks.org/wiki/LaTeX/Tables
-\begin{table}[t]%% placement specifier
+%\begin{table}[t]%% placement specifier
 %% Use tabular environment to tag the tabular data.
 %% https://en.wikibooks.org/wiki/LaTeX/Tables#The_tabular_environment
-\centering%% For centre alignment of tabular.
-\begin{tabular}{l c r}%% Table column specifiers
+%\centering%% For centre alignment of tabular.
+%\begin{tabular}{l c r}%% Table column specifiers
 %% Tabular cells are separated by &
-  1 & 2 & 3 \\ %% A tabular row ends with \\
-  4 & 5 & 6 \\
-  7 & 8 & 9 \\
-\end{tabular}
+%  1 & 2 & 3 \\ %% A tabular row ends with \\
+%  4 & 5 & 6 \\
+%  7 & 8 & 9 \\
+%\end{tabular}
 %% Use \caption command for table caption and label.
-\caption{Table Caption}\label{fig1}
-\end{table}
+%\caption{Table Caption}\label{fig1}
+%\end{table}
 
@@
 %% https://en.wikibooks.org/wiki/LaTeX/Floats,_Figures_and_Captions
-\begin{figure}[t]%% placement specifier
+%\begin{figure}[t]%% placement specifier
 %% Use \includegraphics command to insert graphic files. Place graphics files in 
 %% working directory.
-\centering%% For centre alignment of image.
-\includegraphics{example-image-a}
+%\centering%% For centre alignment of image.
+%\includegraphics{example-image-a}
 %% Use \caption command for figure caption and label.
-\caption{Figure Caption}\label{fig1}
+%\caption{Figure Caption}\label{fig1}
 %% https://en.wikibooks.org/wiki/LaTeX/Importing_Graphics#Importing_external_graphics
-\end{figure}
-
-%% The Appendices part is started with the command \appendix;
-%% appendix sections are then done as normal sections
-\appendix
-\section{Example Appendix Section}
-\label{app1}
-
-Appendix text.
+%\end{figure}
 
@@
 %\usepackage{showframe}
-
-% line numbers
-%\usepackage[mathlines, switch]{lineno}
-%\usepackage[switch, columnwise]{lineno}
 
@@
 %\subtitle{Subject Section}
-
-\title[Detecting Interactions]{Detecting Interactions in Random Forest for Genomic Data}
-
-\author[1,$\ast$]{Hawlader A. Al-Mamun} 
-\author[2]{Rob Dunne}
-\author[3]{Ross L. Tellam}
-\author[4]{Klara Verbyla}
-
-%\authormark{Author Name et al.}
-
-\address[1]{\orgdiv{Data61}, \orgname{CSIRO}, \orgaddress{\street{Ecosciences Bldg 41 Boggo Road }, \postcode{Dutton Park 4102}, \state{QLD}, \country{Australia}}}
-\address[2]{\orgdiv{Data61}, \orgname{CSIRO}, \orgaddress{\street{Eveleigh Office, 13 Garden Street}, \postcode{2015}, \state{NSW}, \country{Australia}}}
-\address[3]{\orgdiv{Agriculture \& Food}, \orgname{CSIRO}, \orgaddress{\street{306 Carmody Rd}, \postcode{St Lucia 4067}, \state{QLD}, \country{Australia}}}
-\address[4]{\orgdiv{Agriculture \& Food}, \orgname{CSIRO}, \orgaddress{\street{Street}, \postcode{Postcode}, \state{State}, \country{Australia}}}
 
@@
 %%%%%%%%%%%%%%
-
-%\begin{appendices}
+\appendix
 
 \FloatBarrier
-\section{Section title of first appendix}\label{sec11}
-
-%\end{appendices}
+\section{Section title of first appendix}
+\label{sec11}
 
@@
 project was provided by the Wellcome Trust under award 076113.
-
-%\bibliographystyle{plain}
-%\bibliography{reference}
 

### 2025-04-30 manuscript.tex
changes for elsarticle style

@@
 
+\tableofcontents 
+\clearpage
+
 \begin{frontmatter}
@@
 interactions based on variable importance can be problematic. It is also heavily dependent on extensive permutation
-tests.  \citeauthor{schmalohrDetectionEpistaticInteractions2018} notes that there is a lack of accurate methods for the
+tests.  \cite{schmalohrDetectionEpistaticInteractions2018} notes that there is a lack of accurate methods for the
 detection of interactions between genomic loci that affect QTLs but that several studies have shown that RF outperforms
@@
 
+\section{Methods}
+
+The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model is created using all
+available genomic data. For both the simulated and real data examples, the R package ranger
+\cite[]{Wright.and.Ziegler.2017} was used with the parameters \texttt{num.tree} = 1000 and \texttt{mtry} = \#SNPs/3.
+%Once the Random Forests model is built, the number of times each pair of markers was found as parent-child nodes in all
+%the decision trees were recorded,
+Working from the random forest output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the count of the
+number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices and set the diagonal to 0,
+we form the interaction vector, of length $\frac {p(p-1)} 2 = {p \choose 2}$.  This is likely to be large and contain a large number of
+zeros ($4 \times 10^9$ entries and 0.999 zero in the rheumatoid example, section \ref{subsection: rheumatoid}).
+
 \subsection{The Null Distribution}
 
-Working from the random forest output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the
-count of the number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices
-and set the diagonal to 0, we form the interaction vector, of length $\frac {p(p-1)} 2 = {p \choose 2}$.
-This is likely to be large ($4 \times 10^9$ in the rheumatism example) and contain a large number of zeros (0.999 zeros
-in the rheumatism example).
-
-Efron's Empirical Bayes procedure (\cite{efronLargeScaleInferenceEmpirical2010}) is now widely used to determine levels
-of significance in high throughput genomics data analysis.
-These data sets led to the use of large numbers of parallel tests to determine the significant variables and then
-the use of multiple testing corrections \cite{limma.2015}.
-Efron suggests that instead of assuming a distribution, such as the $t$, 
-a the better approach is to use the large number of variables to model the null distribution, and then do
-significance testing.
+Efron's Empirical Bayes procedure \cite[]{efronLargeScaleInferenceEmpirical2010} is now widely used to determine levels of significance in
+high throughput genomics data analysis.  These data sets led to the use of large numbers of parallel tests to determine the significant
+variables and the use of multiple testing corrections \cite{limma.2015}.  Efron suggests that instead of assuming a distribution, such as
+the $t$, the better approach is to use the large number of variables to model the null distribution, and then do significance testing.
 
@@
 non-null counts. Following \cite{Gauran.et.al.2018} we could make the additional assumption that there is some integer $C$ such that
-\[
+\begin{equation}
   f(j ) = \pi_0 f_0 (j ) \textrm{ for } j \leq C.
-\]
+\label{eqn:C}  
+\end{equation}
 So we only have the null distribution for counts less than $C.$
@@
 \]
-where $j$ is a non-negative integer, $0 \leq \eta < 1$, $0 \leq \theta < 1$ and $\lambda > 0$. The indicator function
+where $0 \leq \eta < 1$, $0 \leq \theta < 1$ and $\lambda > 0$. The indicator function
 $I_S (j)$ is equal to 1 if $j \in  S$ and 0 otherwise. Fitting this model involves estimating $\{ \eta, \pi_0,
@@
 Using the ``one stage procedure'' and $C_1$ constant (see supplementary information)
-we get a $C$ value (ref equation) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.
+we get a $C$ value (equation \ref{eqn:C}) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.
 
 We note an important caveat here.  \cite{Gauran.et.al.2018} show that the significance level can be
-determined and the FDR controlled in a series of simulation examples. For example  Figure 1 (\cite{Gauran.et.al.2018})
+determined and the FDR controlled in a series of simulation examples. For example  Figure 1 \cite[]{Gauran.et.al.2018}
 shows an example where $\lambda =20$ for the null distribution and the contaminating distribution is a Binomial with
-$\mu=50.$
-It would appear that these examples all have $\lambda$ and $\mu$ (the mean of the contaminating distribution) with a
-greater separation than we have in the Rheumatoid data.
+$\mu=50.$ It would appear that these examples all have $\lambda$ and $\mu$ (the mean of the contaminating distribution) with a
+greater separation than we have in the Rheumatoid data. So we may have a more inteactable problem here.
+
 Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forest with more
-trees and we will observe counts.
-
-It appears that there is a trade-off  here. And as we are sampling a space of $4 \times 10^{9}$ possible interactions,
-we might have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
-distribution.
+trees and we will observe more counts.
+There is a trade-off here and as we are sampling a space with typically more than $10^{9}$ possible interactions,
+we have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
+distribution. Because of this we resort to a threshholding stratey that is both computationally easier and controls the FDR rate by testing
+each variable.
+
+\subsection{threshholding the data}
 In this case we have followed a simple strategy:
 \begin{itemize}
-\item threshold the observed count data. We do this a number of ways to do this (list them) and then take the median. IN
+\item threshold the observed count data. We do this a number of ways to do this (list them) and then take the median. In
   the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
@@
 
-\section{Methods}
 % The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forest model was created using all avaiable data. Then for each pair 
@@
 
-The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model is created using all
-available genomic data. For both the simulated and real data examples, the R package ranger
-\cite[]{Wright.and.Ziegler.2017} was used with the parameters \texttt{num.tree} = 1000 and \texttt{mtry} = \#SNPs/3.
-Once the Random Forests model is built, the number of times each pair of markers was found as parent-child nodes in all
-the decision trees were recorded. Then the marker pairs were ranked based on the count and only the pairs having count
+Then the marker pairs were ranked based on the count and only the pairs having count
 more than 2 were retained. A threshold of at least two ($t = 2$) appearances as parent-child nodes in the decision trees
@@
 \subsection{Real datasets -- Wellcome Trust Case Control Consortium data for rheumatoid arthritis}
+\label{subsection: rheumatoid}
 The first dataset was obtained from the Wellcome Trust Case Control Consortium (WTCCC)

### 2025-05-01 manuscript.tex
Mamuns changes

@@
 %% and the form \ead[url] for the home page:
-\title{Detecting Interactions in Random Forest for Genomic Data}
+\title{Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests}
 
 \author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
-\ead{halmamun@intergrain.com}
+\ead{hawlader.almamun@uwa.edu.au}
 %% \ead[url]{home page}
@@
 %% \cortext[cor1]{}
-\affiliation{organization={},
-             addressline={},
-             city={},
-             postcode={},
-             state={},
-             country={}}
+\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
+  addressline={2 - 40 Clunies Ross Street}, 
+  city={Acton},
+  postcode={2601}, 
+  state={ACT},
+  country={Australia}}
+
+\affiliation{organization={InterGrain Pty Ltd.},%Department and Organization
+  addressline={19 Ambitious Link}, 
+  city={Bibra Lake},
+  postcode={6163}, 
+  state={WA},
+  country={Australia}}
+
+\affiliation{organization={School of Biological Sciences},%Department and Organization
+  addressline={The University of Western Australia}, 
+  city={Crawley},
+  postcode={6009}, 
+  state={WA},
+  country={Australia}}
+
 %% \fntext[label3]{}
@@
 
-\author{} %% Author name
-
-%% Author affiliation
-\affiliation{organization={},%Department and Organization
-            addressline={}, 
-            city={},
-            postcode={}, 
-            state={},
-            country={}}
-
 \author{Rob Dunne} %% Author name
@@
 \affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
-  addressline={306 Carmody Rd}, 
-  city={St Lucia},
-  postcode={4067}, 
-  state={QLD},
+  addressline={2 - 40 Clunies Ross Street}, 
+  city={Acton},
+  postcode={2601}, 
+  state={ACT},
   country={Australia}}
@@
 \begin{keyword}
-Epistatic interactions \sep  Random Forest \sep SNP data
+Epistatic interactions \sep  Random Forests \sep SNP data
 
@@
 %   on phenotypic variation. Finding epistatic interactions from high density SNP data is very challenging due to the
-%   computational burden induced by the combinatorial number of tests that have to be performed. Random Forest models were
-%   widely used in a variety of disciplines including epistatic interactions. Although Random Forest models account for
-%   variable interactions, it does not explicitly detect variable interactions. And most of the Random Forest based
+%   computational burden induced by the combinatorial number of tests that have to be performed. Random Forests models were
+%   widely used in a variety of disciplines including epistatic interactions. Although Random Forests models account for
+%   variable interactions, it does not explicitly detect variable interactions. And most of the Random Forests based
 %   methods to detect interactions relay on some kind of variable importance measures that suffer when the interacting
-%   variables have very small or no marginal effects. We proposed a Random Forest based method that is not based on
+%   variables have very small or no marginal effects. We proposed a Random Forests based method that is not based on
 %   variable importance and can find epistatic interactions from high density SNP markers in reasonable time. The proposed
@@
 % combines multiple learning algorithms/models to improve prediction or classification performance beyond what could be
-% obtained from the constituent models.  In a Random Forest, \textit{m} random samples (with replacement) are selected by
+% obtained from the constituent models.  In a Random Forests, \textit{m} random samples (with replacement) are selected by
 % bootstrap sampling. Then multiple decision trees are created based on the bootstrap samples. To break
@@
 % regression problem, the model outputs the mean of the individual trees prediction. There are two main parameters in the
-% implementation of Random Forest algorithm namely the number of decision trees (\texttt{num.trees}) and the number of
+% implementation of Random Forests algorithm namely the number of decision trees (\texttt{num.trees}) and the number of
 % randomly selected features (\texttt{mtry}) from which the best split is selected at any given node.
@@
 
-Random Forest models are widely used \cite[]{lundbergConsistentIndividualizedFeature2019} in a variety of
+Random Forests models are widely used \cite[]{lundbergConsistentIndividualizedFeature2019} in a variety of
 disciplines. The attraction of the model is based on several attributes, namely:
@@
 In addition it is clear that Random Forests can perform well with correlated and interacting variables, at least in the
-sense that the fitted model performs well \cite[]{Ziegler.et.al.2007}. However, most of the Random Forest based
+sense that the fitted model performs well \cite[]{Ziegler.et.al.2007}. However, most of the Random Forests based
 approaches are either based on single variable importance \cite[]{Jiang.et.al.2009, yoshidaSNPInterForestNewMethod2011}
-or based on pairwise variable importance \cite[]{Ziegler.et.al.2007}. The Random Forest-based approaches for identifying
+or based on pairwise variable importance \cite[]{Ziegler.et.al.2007}. The Random Forests-based approaches for identifying
 interactions based on variable importance measurements cannot distinguish whether a discovered interaction is a true
@@
 To overcome the challenge of a large number of pairwise SNP interaction tests, we proposed an effective two-step process
-based on our observations from using a Random Forest approach. Using simulation data, we observed that ``if two SNPs are
-interacting, they appear as parent-child nodes in a decision tree of a random forest more frequently than if they are
-not interacting''.
+based on our observations from using a Random Forests approach. Using simulation data, we observed that \textit{``if two SNPs are
+interacting, they appear as parent-child nodes in a decision tree of a Random Forests more frequently than if they are
+not interacting''}.
 
 % This observation was used to narrow down the vast search space of pairwise tests to detect SNP
-% interactions. First, we built a Random Forest and for each pair of SNPs we count the total number of times any SNP pair
+% interactions. First, we built a Random Forests and for each pair of SNPs we count the total number of times any SNP pair
 % occurred as parent-child nodes in this forest. Then we selected only the SNP pairs that occurred more times that would
 % be expected by chance and exceeded a set threshold of \textit{t}. In our study, we selected all the SNP pairs those have
-% at least two (\textit{t}=2) appearance as parent-child nodes in the decision trees of the Random Forest.  Finally, using
+% at least two (\textit{t}=2) appearance as parent-child nodes in the decision trees of the Random Forests.  Finally, using
 % a linear model, each selected SNP pair was tested to determine whether the interaction was significant.
@@
 %the decision trees were recorded,
-Working from the random forest output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the count of the
+Working from the Random Forests output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the count of the
 number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices and set the diagonal to 0,
@@
 
-Using the Rheumatoid data as an example, we have $p= 94485$ and after fitting a random forest with
+Using the Rheumatoid data as an example, we have $p= 94485$ and after fitting a Random Forests with
 1000 trees we see about 290 non-terminal splits per tree. We observe 291142 splits with a distribution of frequencies
@@
 
-Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forest with more
+Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forests with more
 trees and we will observe more counts.
@@
 
-This, despite some processing of the linear models, is less time consuming than training a much larger Random Forest.
+This, despite some processing of the linear models, is less time consuming than training a much larger Random Forests.
 
@@
 
-% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forest model was created using all avaiable data. Then for each pair 
+% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model was created using all avaiable data. Then for each pair 
 % of SNP, how many times they appeared as parent-child node in each of the decision trees were recorded. Then the SNP pairs were ranked based on
@@
 controls. Variants having $>10\%$ missing values and MAFs$<0.05$ were removed. SNP markers that failed the
-Hardy-Weinberg test (p $\leq$ 0.05) were also removed. As linkage disequilibrium (LD) could affect how the Random Forest
+Hardy-Weinberg test (p $\leq$ 0.05) were also removed. As linkage disequilibrium (LD) could affect how the Random Forests
 (RF) algorithm selects a SNP at a particular node (for more detail how LD can affect RF, please see
@@
 % To overcome the challenge of a large number of pairwise interaction tests, we proposed an effective two-step process
-% based on our observations from using Random Forest. Using different simulation data, we observed that ``if two SNPs are
-% interacting, they appear as parent-child nodes in a decision tree of a random forest more frequently than if they are
+% based on our observations from using Random Forests. Using different simulation data, we observed that ``if two SNPs are
+% interacting, they appear as parent-child nodes in a decision tree of a Random Forests more frequently than if they are
 % not interacting''.  So, we utilised this observation to narrow down the vast search space of pairwise tests to detect
-% interactions. First, we build a random forest and for each pair of SNPs we count the total number of times any SNP pair
+% interactions. First, we build a Random Forests and for each pair of SNPs we count the total number of times any SNP pair
 % occurred as parent-child nodes in this forest. Then we select only the SNP pairs that occur more times that would be
@@
 \section{Results and Discussion}
-\subsection{simulated data}
+\subsection{Simulated data}
 % Based on publicly available pig genotype data, simulated phenotypes using different scenarios with various combinations
@@
 % pairs for pairwise interaction tests. Table \ref{table:ranking10qtl} shows the importance ranking of three interacting
-% SNP pairs based on the number of times they appear as parent-child in the Random Forest decision trees. Higher ranking,
+% SNP pairs based on the number of times they appear as parent-child in the Random Forests decision trees. Higher ranking,
 % represented by a smaller number, means a higher count. For example, a value of 1 for pair P3 means that SNP pair 3
-% appeared as parent-child more than any other SNP pair in that Random Forest. Since in the base model three pairs of
+% appeared as parent-child more than any other SNP pair in that Random Forests. Since in the base model three pairs of
 % interacting SNPs do not have any marginal effects or any interaction effects, they did not appear as parent-child in any
-% decision tree of the Random Forest. In the no interaction model, the three interacting SNP pairs appeared as
+% decision tree of the Random Forests. In the no interaction model, the three interacting SNP pairs appeared as
 % parent-child in some of the scenarios. This is because in the no interaction model, although the three pairs do not
@@
 
-% In the proposed two step method for detecting SNP-SNP interactions, a random forest model was built and the count
+% In the proposed two step method for detecting SNP-SNP interactions, a random forests model was built and the count
 % determined for how many times a given pair of SNPs appears as parent-child. In the second step, only those SNP pairs
 % having the counts higher than a certain number t were tested. In this simulation, we set t = 2, meaning that if any of
-% the SNP pairs appeared as parent-child in the decision trees of the Random Forest for at least 2 times, we selected them
+% the SNP pairs appeared as parent-child in the decision trees of the Random Forests for at least 2 times, we selected them
 % to test for pairwise interaction using Equation 1. The third column of Table
@@
 % Figure \ref{figure:Eff_Trait_h2_simulated_data.png} shows the pairwise counts for each of the interacting SNP pairs in
-% each Random Forest at different heritabilities (h = 0.33, 0.50, 0.66, and 1.00) for 10,000 SNP with 10 QTL and 3 pairs
+% each Random Forests at different heritabilities (h = 0.33, 0.50, 0.66, and 1.00) for 10,000 SNP with 10 QTL and 3 pairs
 % of interacting SNPs. At low heritability, the pairwise count is smaller and as heritability increases, the pairwise
@@
 
-\subsection{heritability}
+\subsection{Heritability}
 
@@
 % After pre-processing and filtering, the RA dataset contained 94,486 SNPs (~4.46 billion possible pairwise interaction
-% tests) and 4,798 individuals.  Using random forest, we have selected only 654 SNP pairs (out of more than 4.46 billion
+% tests) and 4,798 individuals.  Using Random Forests, we have selected only 654 SNP pairs (out of more than 4.46 billion
 % possible pairs) for pairwise interaction tests (Equation 2). Among the 654 SNP pairs, 26 pairs have significant
@@
 % In this study, we proposed a simple and easy method to select candidate SNP pairs to test for interactions in large
-% genomic data. We used a Random Forest algorithm to select a subset of SNP pairs from a vast search space so that the
+% genomic data. We used a Random Forests algorithm to select a subset of SNP pairs from a vast search space so that the
 % selected SNP pairs can be tested for interaction in a reasonable computational time. The method was validated with

### 2025-05-02 Supplementary_File_Rheumatoid_Data.Rmd
added Supplementary_File_Rheumatoid_Data.Rmd, other changes to manuscript.tex

---
title: "The Rheumatoid Arthritis data  \nSupplementary File for "Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests"
author: "Hawlader Al-Mamun,  Rob Dunne"
date: "Wednesday, September 25, 2024"
output: 
    bookdown::pdf_document2:
     toc: true
     toc_depth: 3
     fig_caption: yes
     citation_package: biblatex
bibliography: 
header-includes:
  \usepackage{dcolumn}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage{biblatex}
  \usepackage{caption}
  \usepackage{subcaption}
---

We work through the analysis of Rheumatoid Arthritis data presented in "Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests"

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, dpi=300,echo = FALSE)
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

```
\clearpage
# Notes

- is the data available? What about the preprocessing? Can we include this in a vignette?
- dim(dt)   4798 94485 but the paper says 94,486
- looked at variable importance. Variables that have more splits as they are important are likely to have more consecutive splits. Should we take this into account?
- I removed splits from AllInteractions where the split is on the same variable again. This does not matter for the regression method but might for other methods (Empirical Bayes)
- There is a hotspot in chromosome 6. Did we know about this? Does it get dropped when you filter on LD
- we are looking at the interaction as the product of the two SNP variables, so it takes the values $\{0,1,2,4\}$. so we are threating it as a
  regression but it does not take the value 3.  If we treat them as factors we get levels(d1:d2) "0:0" "0:1" "0:2" "1:0" "1:1" "1:2" "2:0" "2:1" "2:2"
  9 levels. Which is a more reasonable model? 
- I look at using the **empiricalzip** (Empirical Bayes zero inflated Poisson) package to model the interaction counts and set a significance
level. The agreement with the regression testing is not great.
- we have the file "Cleaned_CaseControl_RAauto_pruned200kb_.3.bim"

# load data

- 4798 samples
- 94485 SNPs

```{r, echo=FALSE, eval=FALSE}
library(lmtest)
library(ranger)
#library(RCy3)

#Noise <- 0.1
numIte  <-  1
t <- numIte
ntree <- 1000

dirOut  <-  paste0("Results")
if(!dir.exists(dirOut)){
  dir.create(dirOut)
}
#dirOut  <-  paste0("Results/pruned200K.3r2_geno_Missing_Averaged")
dt  <-  readRDS("pruned200K.3r2_geno_Missing_Averaged.rds")
pheno <- read.table("Cleaned_CaseControl_RAauto.fam",header=F)
pheno$ID <- paste0(pheno$V1,"_",pheno$V2)
all.equal(row.names(dt),pheno$ID)
dt <- data.frame(pheno=as.factor(pheno$V6),dt)
dim(dt) #[1]  4798 94485

#94,486 SNPs (4.46 billion possible pairwise interaction tests) and 4,798 individuals.

dt$pheno <- as.factor(dt$pheno)
#dt <- dt[which(dt$treatment=="short"),-2]

# quick load of results etc
dirOut  <-  paste0("Results")
load(file="Results/var_importance_gini.Rdata")
load(file=paste0(dirOut,"/All_Interactions.Rdata"))
load(file="./Results/interaction_snps_counts_fixed.Rdata")
load(file=paste0(dirOut,"/lm_results_fixed.Rdata"))

```

# Fit a Random Forest

```{r, echo=FALSE, eval=FALSE}
##Number of trees:                  1000 
#set.seed(sample(1:1000,1))
set.seed(100)
#system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
#     user    system   elapsed 
#89141.264     3.601  1568.711 
#save(model.rf, file="model.rf.Rdata")

##Number of trees:                  5000 
#system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = 5000, mtry = ncol(dt)/3))
# user     system    elapsed 
#443201.110      8.132   7185.540 
#Number of trees:                  5000 
#Sample size:                      4798 
#Number of independent variables:  94484 
#Mtry:                             31495 
#Target node size:                 1 
#Variable importance mode:         none 
#Splitrule:                        gini 
#OOB prediction error:             33.18 % 
#save(model.rf, file="model.rf_temp2.Rdata")

##Number of trees:   1000, importance = 'impurity'
set.seed(100)
#petrichor-i1: salloc -A OD-217714 --mem=100GB --nodes=1 --ntasks-per-node=64    -J interactive -t 6:00:00 srun --pty /bin/bash -l
#etrichor-i1: sacct -j 60755241  --format="jobid,ntasks,Elapsed,TotalCPU,MaxRSS,AveVMSize,MaxVMSize"
#JobID          NTasks    Elapsed   TotalCPU     MaxRSS  AveVMSize  MaxVMSize 
#60755241                00:13:56  00:06.682                                  
#60755241.ex+        1   00:13:56  00:00.001       995K    124080K    124080K 
#60755241.0         64   00:13:53  00:06.680 104545713K 3274946112 192422568K

system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
#     user    system   elapsed 
#89397.171     4.807  1591.941 
#save(model.rf, file="model.rf_temp3.Rdata")
var_importance <- model.rf$variable.importance
head(rev(sort(var_importance)))
t2 <-count_variables(model.rf)
names(t2) <-names(var_importance)

#save(var_importance , t2, file="Results/var_importance_gini.Rdata")
#load(file="Results/var_importance_gini.Rdata")

```

# variable importance

we use the pacakge RFlocalfdr to give an empirical Bayes estimate of the number of significant SNPs.
It returns 74 SNPs as significant. See figure \@ref(fig:RFlocalfdr). Note that this is just settng a significance level. These
are the 74 SNPs with maximum variable importance from the random forest fit.

```{r RFlocalfdr, eval=TRUE, echo=FALSE, fig.cap='RFlocalfdr',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./script3/RFlocalfdr.png")

```

```{R, echo=FALSE, eval=FALSE}
library(RFlocalfdr)
vignette(package="RFlocalfdr")
#no vignettes found -- why is this?
imp<-log(var_importance)  #6275 var_importance vales =0 
plot(density(imp),xlab="log importances",main="")

cutoffs <- c(0,1,2,3)
res.con<- determine_cutoff(imp,t2 ,cutoff=cutoffs,plot=c(0,1,2,3))
plot(cutoffs,res.con[,3],pch=15,col="red",cex=1.5,ylab="max(abs(y - t1))")

cutoffs <- c(4,5,6)
res.con2 <- determine_cutoff(imp,t2 ,cutoff=cutoffs,plot=c(4,5,6))
plot(c(0,1,2,3,4,5,6),c(res.con[,3],res.con2[,3]),pch=15,col="red",cex=1.5,ylab="max(abs(y - t1))")

temp<-imp[t2 > 3]
temp <- temp - min(temp) + .Machine$double.eps
qq <- plotQ(temp,debug.flag = 1)
ppp<-run.it.importances(qq,temp,debug.flag = 0)

png("./script3/RFlocalfdr.png")
aa<-significant.genes(ppp,temp,cutoff=0.05,debug.flag=0,do.plot=TRUE,use_95_q=TRUE)
dev.off()
length(aa$probabilities) # 74

aa<-significant.genes(ppp,temp,cutoff=0.05,debug.flag=0,do.plot=TRUE,use_95_q=FALSE)
length(aa$probabilities) #  74

length(intersect(names(aa$probabilities) ,names(rev(sort(var_importance)))[1:74])) #74
#but why is the order different

var_importance <- var_importance[match(names(aa$probabilities),names(var_importance))]
#now in the same order as aa$probabilities
cbind(aa$probabilities,var_importance)

```

# look at some SNP counts

- this is the counts of how often a variable was used for a split in RF
- the varaible importance has a very high correlation with this
- There is a hotspot in chromosome 6. See figure \@ref(fig:hotspot)

```{r hotspot, eval=TRUE, echo=FALSE, fig.cap='There is a hotspot in chromosome 6',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./figs/hotspot.png")

```

```{R, echo=FALSE, eval=FALSE}
t2[grep("X6_32471505",names(t2))] #993 
t2[grep("X6_32771829",names(t2))] # 1532
sum(t2) #[1] 295310  all splits

# how are the SNP numbers organized 
plot(as.numeric(gsub("X[1-9]*\\_([0-9]*)","\\1", names(t2))))
# it looks like they are sequential within chromosome

cols<- as.numeric((gsub("X([0-9]*)\\_[0-9]*","\\1",  names(t2))))

#png("./figs/hotspot.png")
# Set up the main plot without x-axis ticks and labels
par(mar=c(5, 4, 4, 2) + 0.1)  # Adjust margins
plot(t2, col=cols, xaxt='n', main="SNP Counts Across Genome",xlab="counts")

# Add an inset plot for chromosome 6
# Define the plotting region for the inset
par(fig=c(0.5, 0.9, 0.5, 0.9), new=TRUE, mar=c(2, 2, 1, 1))  # Adjust margins for inset

# Extract data for chromosome 6
chrom6_data <- t2[cols == 6]

# Plot the inset
plot(chrom6_data[1350:1500], col=6, main="part of Chromosome 6", xaxt='n', yaxt='n')
#dev.off()
# there is a hotspot in this region!

t2[which(t2> 800)]
#X6_32471505 X6_32771829 
#        993        1532

```

# count the interactions in actual data

```{R, echo=FALSE,eval=FALSE}
#t2 <- system.time({

  # countMatrix <- matrix(0,ncol = ncol(dt)-1,nrow=ncol(dt)-1)
  # row.names(countMatrix) <- colnames(dt)[-1]
  # colnames(countMatrix) <- colnames(dt)[-1]
  #

  for(i in 1:ntree){
    #tree2 <- getTree(model.rf2,i,T)
    tree <- treeInfo(model.rf,i)
    #tree <- treeInfo(model.n,5)

    LeftinteractionTable  <-  data.frame(matrix(0,nrow = nrow(tree),ncol=2))
    LeftinteractionTable[,1] <- tree$splitvarName

    for(l in 1:nrow(tree)){
      if(is.na(tree[l,5])==F){
        leftChildID = tree[l,2]
        leftChildIndex = which(tree$nodeID==leftChildID)
        LeftinteractionTable[l,2]=tree$splitvarName[leftChildIndex]
      }
    }
    RightinteractionTable = data.frame(matrix(0,nrow = nrow(tree),ncol=2))
    RightinteractionTable[,1]=tree$splitvarName

    for(l in 1:nrow(tree)){
      if(is.na(tree[l,5])==F){

        rightChildID = tree[l,3]
        rightChildIndex = which(tree$nodeID==rightChildID)
        RightinteractionTable[l,2]=tree$splitvarName[rightChildIndex]
      }
    }

    LeftinteractionTable=na.omit(LeftinteractionTable)
    RightinteractionTable=na.omit(RightinteractionTable)

    allInteractions = rbind(LeftinteractionTable,RightinteractionTable)
    allInteractions[,1] = as.character(allInteractions[,1])
    allInteractions[,2] = as.character(allInteractions[,2])

    if(i ==1){
      index=which(allInteractions$X1> allInteractions$X2)
      tmp=allInteractions$X2
      tmp[index]=allInteractions$X1[index]
      allInteractions$X1[index]= allInteractions$X2[index]
      allInteractions$X2= tmp
      All_Interactions=allInteractions
      #allInteractions$X1_X2=paste0(allInteractions$X1," ",allInteractions$X2)
      #countTable=table(allInteractions$X1_X2)
      #All_count_dtf=data.frame(countTable)
    }else{
      index=which(allInteractions$X1> allInteractions$X2)
      tmp=allInteractions$X2
      tmp[index]=allInteractions$X1[index]
      allInteractions$X1[index]= allInteractions$X2[index]
      allInteractions$X2= tmp
      All_Interactions=rbind(All_Interactions,allInteractions)

    }
  }

#saveRDS(All_Interactions,paste0(dirOut,"/All_Interactions.rds"))
#All_Interactions <- readRDS(paste0(dirOut,"/All_Interactions.rds"))
#save(All_Interactions,file=paste0(dirOut,"/All_Interactions.Rdata"))
#dim(All_Interactions) #[1] 294310      2

```
```{R, echo=FALSE, eval=FALSE}
# just some checks on how  All_Interactions is organized. 
head(All_Interactions)
#            X1           X2      294310      2
#1  X6_32471505  X6_32771829
#2 X21_44075267  X6_32471505
#3 X14_84754343  X4_67191986

#X6_32471505  X6_32771829 are often consecutive splits. We count occurrences

length(grep("X6_32471505", All_Interactions[grep("X6_32771829",All_Interactions$X1),2])) #0
length(grep("X6_32471505", All_Interactions[grep("X6_32771829",All_Interactions$X2),1])) #466

# and for some random selected splits
for (i in  sample(294310,100) ){
    a1<- length(grep(All_Interactions$X1[i], All_Interactions[grep(All_Interactions$X2[i],All_Interactions$X1),2])) #0
    a2<- length(grep(All_Interactions$X2[i], All_Interactions[grep(All_Interactions$X1[i],All_Interactions$X1),2])) #0
    cat(a1, a2, "\n", sep=" ")
}
#so the splits are lexigraphically ordered i.e. where "X6_32771829 X6_32471505" occurs, it is recorded as  "X6_32471505  X6_32771829" 
#never the other way around. This makes things easier to track

```

## expected numbers of consecutive splits

- if the consecutive splits were uniformly random, then variables with a lot of splits would have a higher chance of occuring together
- consider X6_32471505 and  X6_32771829. If random these would have  50 consecutive splits, they have 466

```{R, echo=FALSE, eval=FALSE}

#so how often would we expect the consecutive splits "X6_32471505 X6_32771829"
#(prob X6_32471505)*prob( X6_32771829 )* (total number of split pairs)
#splits on X6_32771829 1532 times, mtry is 31495 , total numer of splits 295310, 94484 varaibles
(31495/ 94484 ) *(993 / 31495)* (31495/ 94484 ) *(1532 / 31495)*294120
#so 50.12057 versus 466

#what about  X6_32481210 X6_32682149
#281391  X6_32481210 X6_32682149    5
t2[grep("X6_32481210",names(t2))] # 121  
t2[grep("X6_32682149",names(t2))] #  277 
(31495/ 94484 ) *(121  / 31495)* (31495/ 94484 ) *( 277  / 31495)*294120
# so 1.1 versus 5

```

## remove "self splits"

1. how often are there self-splits i.e. consecutive splits on the same variable

  - 190 are self-splits (so not many). These are all due to two transcripts "X6_32771829",  "X6_32471505" which are 1 and 2 in importance
  - we  remove these from **All_Interactions**
  - save **interaction_snps_counts_fixed.Rdata**

b

```{R, echo=FALSE, eval=FALSE}

length(which(All_Interactions$X1==All_Interactions$X2))
# of 294310 splits, 190 are self-splits

unique(All_Interactions$X1[which(All_Interactions$X1==All_Interactions$X2)])
# "X6_32771829" "X6_32471505" 2 splits these are 1 and 2 in importance

imp <- names(rev(sort(var_importance)))
grep("X6_32771829",imp)
head(match( imp, All_Interactions$X1))
#  [1]     11      1    882   2916    772   2623   2337    290   1180  61812
# [11]   3524   5620   3520   7688   9321   9917    291   1468   1074   6360

All_Interactions <- All_Interactions[-c(which(All_Interactions$X1==All_Interactions$X2)),]  #this is where we drop the self splits
dim(All_Interactions) #[1] 294120      2
tt<- table(c(All_Interactions$X1,All_Interactions$X2))
tt<-rev(sort(tt))
plot(1:74,match(imp,names(tt))[1:74])
abline(0,1)
# this proves nothing. Counting the number of times a variable is in a split is a reasonable surrogate for importance

interaction_snps <- paste0(All_Interactions$X1," ",All_Interactions$X2)
interaction_snps_counts <- table(interaction_snps)
interaction_snps_counts <- data.frame(interaction_snps_counts)
interaction_snps_counts <- interaction_snps_counts[order(interaction_snps_counts$Freq,decreasing = T),]
head(interaction_snps_counts )
#              interaction_snps Freq   #most are 1
#              interaction_snps Freq
#281122 X6_32471505 X6_32771829  466
#281468 X6_32682149 X6_32771829  144
#280884 X6_32222493 X6_32771829  113

#was                                     before we dropped the repeated splits
#281123 X6_32471505 X6_32771829  466
#281469 X6_32682149 X6_32771829  144
#281616 X6_32771829 X6_32771829  128
# see the repeated split on X6_32771829

aa<-rep(NA, 74)
for ( i in 1:74){
    aa[i] <- length(grep(imp[i],interaction_snps_counts$interaction_snps[1:640]))
    }
aa
# [1] 301 154  43  15  15  19  16  17  13   9   7   7   6   9   8   3   6   4   7
#[20]   3   5   5   4   3   4   4   2   2   2   3   3   3   3   4   3   2   3   4
#[39]   4   1   3   2   2   1   1   4   1   2   1   2   2   3   1   4   1   3   2
#[58]   1   2   3   0   1   1   2   2   3   3   1   0   1   1   1   0   0

sum(interaction_snps_counts$Freq) # 294120

#saveRDS(interaction_snps_counts,paste0(dirOut,"/interaction_snps_counts_fixed.rds"))
#interaction_snps_counts <- readRDS("./Results/interaction_snps_counts_fixed.rds")
#save(interaction_snps_counts, file="./Results/interaction_snps_counts_fixed.Rdata")
#load(file="./Results/interaction_snps_counts_fixed.Rdata")

```

# run linear models

- we consider all interaction_snps where the frequency is greater than 2. That is 640 interactions 
- we fit generalized linear models with an interaction term and record the term and the $p$-value
- we fit models treating the SNP levels as factors, we fit a models with and witout the interaction term and do a likelihood ratio test
- we have Bonferrroni, and FDR adjusted p-values for glm regression with numeric SNP and FDR adjusted p-values for LRT on factor interactions
- the resulting data frame contains

   - interaction_snps  -- the interaction under consideration e.g. "X6_32308125 X6_32535726"
   - Freq              -- the frequency of that interaction 
   - SNP1_effect       -- Estimates
   - SNP1_Pvalue       -- Pr(>|z|)
   - SNP2_effect       -- Estimates
   - SNP2_Pvalue       -- Pr(>|z|)
   - int_effect   interaction term
   - int_Pvalue
   - correlations  cor(dt[,ind1],dt[,ind2])
   - Ch2          lrtest(nested.model, s2) d1 d2 factors  Pr(>Chisq)
   - splits_SNP1   number of times this varaible is the split varaible 
   - splits_SNP2   number of times this varaible is the split varaible 
   - adj_p_value_bonferrroni correction of int_Pvalue
   - adj_p_value_fdr          fdr correction of int_Pvalue
   - adj_p_value_Ch2          fdr correction of Ch2

so

- Bonferrroni on numeric SNP regression picks 21
- fdr on numeric SNP regression picks 88 (containing the 21 selected  by Bonferrroni )
- $\chi^2$ test on SNP factor variables picks  121 (containing 68 of the 88 selected by fdr)

```{R, echo=FALSE, eval=FALSE}

index <- which(interaction_snps_counts$Freq>=2)  #640

lm_results  <-  data.frame(interaction_snps_counts[index,],SNP1_effect=0,SNP1_Pvalue=1,SNP2_effect=0,SNP2_Pvalue=1,int_effect=0,int_Pvalue=1,Ch2=1)
dim(lm_results)
lm_results$correlations <- 0
selectedSNPs <- "pheno"
if(class(dt$pheno)=='factor'){
  for(l in 1:nrow(lm_results)){
    SNP1 <- strsplit(as.character(lm_results[l,1])," ")[[1]][1]
    SNP2 <- strsplit(as.character(lm_results[l,1])," ")[[1]][2]
    selectedSNPs <- union(selectedSNPs,SNP1)
    selectedSNPs <- union(selectedSNPs,SNP2)
    ind1  <-  which(colnames(dt)==SNP1)
    ind2  <-  which(colnames(dt)==SNP2)
    s  <-  summary(glm(dt$pheno ~ dt[,ind1]*dt[,ind2],family = binomial(link="logit")))$coeff
#l<-31 
#                        Estimate Std. Error    z value     Pr(>|z|)
#(Intercept)            0.4202587 0.06693247   6.278846 3.410951e-10
#dt[, ind1]            -0.2898685 0.06767282  -4.283382 1.840739e-05
#dt[, ind2]            -1.0043861 0.07960751 -12.616726 1.707751e-36
#dt[, ind1]:dt[, ind2]  0.2321571 0.06325044   3.670443 2.421308e-04
   lm_results$correlations[l] <- cor(dt[,ind1],dt[,ind2])
    #same as 
    #s2  <-  summary(glm(dt$pheno ~ dt[,ind1] + dt[,ind2] + dt[,ind1]:dt[,ind2],family = binomial(link="logit")))
    #s2$coeff

    #as factors
    d1<- factor(dt[,ind1])
    d2<- factor(dt[,ind2])

    s2  <-  glm(dt$pheno ~ d1 +d2 +d1:d2,family = binomial(link="logit"))
    nested.model <- glm(dt$pheno ~ d1+d2, family= binomial(link="logit"))
    #anova(nested.model, s2, test="LRT") #or
    aa<- lrtest(nested.model, s2)
#Likelihood ratio test
#
#Model 1: dt$pheno ~ d1 + d2
#Model 2: dt$pheno ~ d1 + d2 + d2:d2
#  #Df  LogLik Df Chisq Pr(>Chisq)
#1   5 -2985.5                    
#2   5 -2985.5  0     0          1
# 
#H0: You should use the nested model.
#Ha: You should use the complex model.
#Pr(>Chisq)=1  so use the simple model

    lm_results$splits_SNP1  <- t2[SNP1]
    lm_results$splits_SNP2  <- t2[SNP2]
    lm_results$Ch2[l] <- aa$"Pr(>Chisq)"[2]
    lm_results$SNP1_effect[l]  <-  s[2,1]
    lm_results$SNP1_Pvalue[l]  <-  s[2,4]
    lm_results$SNP2_effect[l]  <-  s[3,1]
    lm_results$SNP2_Pvalue[l]  <-  s[3,4]
    lm_results$int_effect[l]  <-  s[4,1]
    lm_results$int_Pvalue[l]  <-  s[4,4]

  }
}

##############
names(dt)[grep("X6_3",names(dt))]
SNP1 #[1] "X6_32912392" #just the last ones to come out of  the calculation above
SNP2 #[1] "X7_25326942"
dim(All_Interactions[grep(SNP2,All_Interactions$X1),]) #2 2
dim(All_Interactions[grep(SNP2,All_Interactions$X2),]) #35  2
#interaction_snps_counts[grep("X7_25326942", interaction_snps_counts$interaction_snps),]
aa<- interaction_snps_counts[grep("X6_32912392", interaction_snps_counts$interaction_snps),]
aa[grep("X7_25326942", aa$interaction_snps),]
#              interaction_snps Freq
#282088 X6_32912392 X7_25326942    2

grep("X6_32912392", lm_results[,1])
lm_results[ 640,]
#             interaction_snps Freq SNP1_effect  SNP1_Pvalue SNP2_effect SNP2_Pvalue int_effect int_Pvalue       Ch2 correlations splits_SNP1 splits_SNP2
#282086 X6_32912392 X7_25326942    2  -0.3853136 2.289531e-12 -0.02027243   0.7972829  0.1774552 0.04946914 0.1475998  0.004923941          18          14

ind1 <- grep("X6_32912392",names(dt))
ind2 <- grep("X7_25326942",names(dt))

s  <-  summary(glm(dt$pheno ~ dt[, ind1]*dt[, ind2],family = binomial(link="logit")))$coeff
#                                                                             Estimate Std. Error    z value     Pr(>|z|)
#(Intercept)                                                               -0.25152373 0.04720520 -5.3283057 9.913319e-08
#dt[, grep("X6_32912392", names(dt))]                                      -0.38531361 0.05492235 -7.0156068 2.289531e-12
#dt[, grep("X7_25326942", names(dt))]                                      -0.02027243 0.07892247 -0.2568651 7.972829e-01
#dt[, grep("X6_32912392", names(dt))]:dt[, grep("X7_25326942", names(dt))]  0.17745517 0.09032977  1.9645259 4.946914e-02

d1<- factor(dt[,ind1])
d2<- factor(dt[,ind2])
s2  <-  glm(dt$pheno ~ d1 +d2 +d1:d2,family = binomial(link="logit"))
nested.model <- glm(dt$pheno ~ d1+d2, family= binomial(link="logit"))
aa<- lrtest(nested.model, s2)
#Likelihood ratio test
#Model 1: dt$pheno ~ d1 + d2
#Model 2: dt$pheno ~ d1 + d2 + d1:d2
#  #Df  LogLik Df  Chisq Pr(>Chisq)
#1   5 -3174.1                     
#2   9 -3170.7  4 6.7867     0.1476
#all looks right

##########
lm_results <- lm_results[order(lm_results$int_Pvalue, decreasing = F),]
lm_results$adj_p_value_bonferrroni <- lm_results$int_Pvalue*nrow(lm_results) # a bonferrroni correction?
sum(lm_results$adj_p_value_bonferrroni < 0.05) #[1] 21

lm_results$adj_p_value_fdr <- p.adjust(lm_results$int_Pvalue, method = "fdr")
sum(lm_results$adj_p_value_fdr < 0.05) #[1] 88

lm_results$adj_p_value_Ch2 <- p.adjust(lm_results$Ch2 , method = "fdr")
sum(lm_results$adj_p_value_Ch2 < 0.05) #[1] 121
length(intersect( which(lm_results$adj_p_value_fdr < 0.05), which(lm_results$adj_p_value_Ch2 < 0.05)))# [1] 68
#save(lm_results, file=paste0(dirOut,"/lm_results_fixed.Rdata"))# fixed in the sense that the self-split pairs have been removed
#load(file=paste0(dirOut,"/lm_results_fixed.Rdata"))

names(lm_results)
# [1] "interaction_snps"        "Freq"                    "SNP1_effect"             "SNP1_Pvalue"             "SNP2_effect"             "SNP2_Pvalue"             "int_effect"             
# [8] "int_Pvalue"              "Ch2"                     "correlations"            "splits_SNP1"             "splits_SNP2"             "adj_p_value_bonferrroni" "adj_p_value_fdr"        
#[15] "adj_p_value_Ch2"        
lm_results[1:2,]
#              interaction_snps Freq SNP1_effect  SNP1_Pvalue SNP2_effect  SNP2_Pvalue int_effect   int_Pvalue          Ch2 correlations splits_SNP1 splits_SNP2 adj_p_value_bonferrroni
#280954 X6_32308125 X6_32535726   14  -1.0932552 5.783667e-39  -0.7248879 5.844399e-39  0.6083855 3.488304e-16 1.303556e-15  -0.14839455          18          14            2.232514e-13
#280951 X6_32308125 X6_32423632    3  -0.9145729 2.243391e-31  -0.4717926 3.387561e-16  0.5723224 5.161507e-14 4.721907e-12   0.07261602          18          14            3.303364e-11
#       adj_p_value_fdr adj_p_value_Ch2
#280954    2.232514e-13    8.342758e-13
#280951    1.651682e-11    1.007340e-09

#so
#
#- Bonferrroni on numeric SNP regression picks 21
#- fdr on numeric SNP regression picks 88
#- $\chi^2$ test on SNP factor variables picks  121

```

```{R, echo=FALSE, eval=FALSE}
# some plots of the  p-values for the different methods
plot(lm_results$int_Pvalue,pch=16)
points(lm_results$adj_p_value_fdr, col="red",pch=16)
points(lm_results$adj_p_value_bonferrroni , col="green",pch=16)
points(lm_results$adj_p_value_Ch2 , col="blue",pch=16)
abline(h=0.05)
legend(200,1,c("p-values, ordered", "fdr corrected", "Bonferroni corrected", "Chi2 p-values" ),pch=16,col=c("black", "red", "green","blue"))
#some concordance between the regression p-values and the Chi^2 p-values
#Bonferrroni finds 21, "fdr" finds 88 and \Chi^2 with fdr finds 121
length(intersect(which(lm_results$adj_p_value_Ch2  < 0.05), which(lm_results$adj_p_value_fdr  < 0.05))) # 68

lm_results.temp <- lm_results[order(lm_results$adj_p_value_Ch2, decreasing = F),]
plot(lm_results.temp$int_Pvalue, col="black")
points(lm_results.temp$adj_p_value_fdr, col="red")
points(lm_results.temp$adj_p_value_bonferrroni , col="green")
points(lm_results.temp$adj_p_value_Ch2 , col="blue")
abline(h=0.05)
#same thing

lm_results.temp[22,]
 #              interaction_snps Freq SNP1_effect SNP1_Pvalue SNP2_effect  SNP2_Pvalue int_effect int_Pvalue          Ch2 correlations splits_SNP1 splits_SNP2
#180240 X1_161191571 X6_32771829    2  -0.0257612   0.6986353  -0.8204181 3.832185e-29 0.00295658  0.9636916 4.178291e-05  0.001961207          18          14
#       adj_p_value_bonferrroni adj_p_value_fdr adj_p_value_Ch2
#180240                616.7626       0.9821061     0.001215503

ind1  <-  which(colnames(dt)=="X1_161191571")
ind2  <-  which(colnames(dt)=="X6_32771829")
s  <-  summary(glm(dt$pheno ~ dt[,ind1]*dt[,ind2],family = binomial(link="logit")))
#as factors
d1<- factor(dt[,ind1])
d2<- factor(dt[,ind2])
s2  <-  glm(dt$pheno ~ d1 +d2 +d1:d2,family = binomial(link="logit"))
nested.model <- glm(dt$pheno ~ d1+d2, family= binomial(link="logit"))
aa <- lrtest(nested.model, s2)
#Likelihood ratio test  aa$"Pr(>Chisq)" 4.178291e-05
# so the adj_p_value_fdr  is not significant but the adj_p_value_Ch2 is

options(width=195)
lm_results[1:21,]

# the top imp SNPs (selected by RFlocafdr) and how often they appear in the top interactions (glm, selected by FDR)
aa<-rep(NA, 74)
for ( i in 1:74){
    aa[i] <- length(grep(names(var_importance)[i] ,lm_results$interaction_snps[1:88]))
    }
# proving what? The SNPs selected by RFlocafdr are not selected as interactions

aa<-rep(NA, 74)
for ( i in 1:74){
    aa[i] <- length(grep(names(var_importance)[i] ,lm_results$interaction_snps[1:88]))
    }
# proving what? The SNPs selected by RFlocafdr are not selected as interactions

for ( i in 1:74){
    aa[i] <- length(grep(imp[i],lm_results$interaction_snps[1:88]))
    }
aa
#[1] 34 22 11  3  6  4  6  8  4  2  3  2  3  0  3  2  2  0  2  1  0  0  0  1  3
#[26]  2  0  1  0  0  1  0  1  1  0  0  0  1  0  0  0  0  1  0  0  0  0  0  1  0
#[51]  0  0  0  0  0  1  0  0  0  0  0  0  0  0  0  1  0  0  0  0  0  0  0  0

```

# data set up

```{R, echo=FALSE, eval=FALSE}

# Assuming your count data is in a vector 'x'
x <- interaction_snps_counts$Freq

dim(interaction_snps_counts)[1] #[1] 291142
sum(interaction_snps_counts$Freq) # 294120

94485*(94485 -1)/2 - 291142
x2<-  c(rep(0,4463369228),x)

#table(x2) long run time!
#    0         1         2      3      4       5      6      7      8      9     10     11     12     13     14     15     16     17     18     22     23     25     26     27     31     34     41  51     52     54     62     63     64     78     79     90    100    113    144    466 
#4463369228  290502    418     89     29     22     14     11      7      5      4      3      3      3      4     1      1      1      1      2      1      1      1      2      1      1      1   1      1      1      1      2      1      1      1      1      1      1      1      1

# First, prepare your data
counts <- as.numeric(names(table(x)))  # Unique count values (1, 2, 3, ...)
freqs <- as.numeric(table(x))          # Frequencies of each count
n_zeros <- 4463369228                   # Number of zeros

# Recreate the data based on your table x2
counts0 <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14,15,  16,  17,  18,  22,  23,  25,  26,  27,  31,  34,  41, 51,  52,  54,  62,  63,  64,  78,  79,  90, 100, 113, 144, 466) 
freqs0 <- c(4463369228, 290502, 418, 89, 29, 22, 14, 11, 7, 5, 4, 3, 3, 3, 4, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1, 1, 1, 1,   1,   1,   1,   2,   1,   1,   1,   1,   1,   1,   1,   1)
# Recreate the data based on your table x
counts <- c( 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14,15,  16,  17,  18,  22,  23,  25,  26,  27,  31,  34,  41, 51,  52,  54,  62,  63,  64,  78,  79,  90, 100, 113, 144, 466) 
freqs <- c( 290502, 418, 89, 29, 22, 14, 11, 7, 5, 4, 3, 3, 3, 4, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1, 1, 1, 1,   1,   1,   1,   2,   1,   1,   1,   1,   1,   1,   1,   1)

# Create a representative sample of this distribution (for computational efficiency)
set.seed(123)
sample_size <- 100000000
sample_probs <- freqs / sum(freqs)
x_sample <- sample(counts, size = sample_size, replace = TRUE, prob = sample_probs)

```

# Null distribution

What would the distribution of counts look like if 291142 interactions were assigned randomly?
Under a random assignment, this follows a binomial distribution that can be approximated by a Poisson distribution with:
$\lambda= 291,142 / 4,463,565,886 \approx 0.000065$

-  Count = 1: approximately 290,849 pairs
-  Count = 2: approximately 9.5 pairs
-  Count $\geq 3$: effectively 0 pairs

The observed data shows:

- Count = 1: 290502  pairs
-  Count = 2: 418   pairs
-  Count = 3: 89  pairs
-  Count $\geq 4$: 133 pairs

\clearpage
\FloatBarrier

# empiricalzip

- There are  4463613128 potential interactions. We have non-zero values for 294120, so there are a lot of 0 values.
- what to do to handle this? 
- **gpEMAlgorithm** fails -- set which estimates to calculate in the function call
- better name than **RealDataFDR**, what other data would we be running it on
- docuement output better
- "X6_32222493 X6_32771829" see below, how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113

   - it doesnt t2["X6_32222493"] =202 t2["X6_32771829"]= 1532 -- lm_result has a problem

## some investigations of empiricalzip. 
```{R, echo=FALSE, eval=FALSE}
#devtools:::install_github("parsifal9/empiricalzip")
estimate<-  zigpEMAlgorithm(x2) #works
estimate<- zipEMAlgorithm(x2) #works
system.time(estimate<- pEMAlgorithm(x2)) ##works
estimate<- gpEMAlgorithm(x2) #Error in while (((abs(hold.lambda - lambdaEstimate)) > epsilon) || ((abs(hold.theta -  :   missing value where TRUE/FALSE needed

aa <- my.RealDataFDR(x2, 0.1) #has gpEMAlgorithm commented out, accepts data ratehr than a file
# Rejection counts
#       ZIGP ZIP   GP  P         ZIGP ZIP   GP  P        ZIGP  ZIP     GP  P
#   one stage procedure           two stage             Storey's FDR
#[1,]   50   82   82   82      57   82  104   82        35    53    53    53
#[2,]   50   82   82   82      57   82  104   82        35    53    53    53

#[[2]]Parameter estimates using cutoff method C_0

    ## eta: Zero-inflation parameter (proportion of structural zeros)
    ## lambda: Mean parameter
    ## theta: Dispersion parameter (NA for models that don't use it)
    ## pi: Estimated proportion of null hypotheses ($\pi_0$)
    ## C: The estimated cutoff value (\hat{C}) - positions with counts $\leq$ C are assumed to be from null
    ## AD: The truncation point (D_N) - positions with counts $\geq$ AD are automatically considered significant

#        eta         lambda    theta      pi          C0   AD
#[1,] 0.472686094 1.0020464 1.133941e-12 0.9999401    7    8   zigp
#[2,] 0.001277773 0.3350806           NA 0.9999075    5   14   zip
#[3,]          NA 0.3346515 0.000000e+00 0.9999072    4    5   gp
#[4,]          NA 0.3346515           NA 0.9999072    4   14    p
#
#[[3]]
#           [,1]      [,2]         [,3]      [,4] [,5] [,6]
#[1,] 0.47268609 1.0020464 1.133941e-12 0.9999401    7    8
#[2,] 0.00128352 0.3350822           NA 0.9999074    4   14
#[3,]         NA 0.3346515 0.000000e+00 0.9999072    4    5  
#[4,]         NA 0.3346515           NA 0.9999072    4   14

#empiricalzip  ZIGP one stage procedure
#empiricalzip pickes the interactions on the Freq, sointeraction_snps_counts[1:50,]
interaction_snps_counts$interaction_snps[1:50]

sum(freqs[-c(1:7)]) #[1] 68
sum(freqs[-c(1:8)]) #[1] 57
sum(freqs[-c(1:9)]) #[1] 50

```
The ZIGP model selected by epiricalzip has

- $\eta= 0.472686094$      Zero-inflation parameter (proportion of structural zeros)   
- $\lambda= 1.0020464$    Mean parameter
- $\theta= 1.133941e-12$    Dispersion parameter (NA for models that don't use it)  
- $\pi= 0.9999401$       Estimated proportion of null hypotheses

See table \@ref(tab:zigp_comparison)

\begin{table}
\caption{\label{tab:zigp_comparison} Observed vs. Expected Counts Under ZIGP Model selected by mpiricalzip}
\center
\begin{tabular}{lrr}
\hline
\textbf{Count} & \textbf{Observed} & \textbf{Expected (ZIGP)} \\
\hline
0 & 4,463,369,228 & 4,463,369,228 \\
1 & 290,502 & 290,490 \\
2 & 418 & 427 \\
3 & 89 & 85 \\
4 & 29 & 21 \\
5 & 22 & 5 \\
6-10 & 41 & 3 \\
11-20 & 13 & $<$1 \\
$>$20 & 22 & $\approx$0 \\
\hline
\textbf{Total} & 4,463,660,370 & 4,463,660,370 \\
\hline
\end{tabular}
\label{tab:zigp_comparison}
\end{table}

```{R, echo=FALSE, eval=FALSE}
# protein example (move this)

the protein exaple from @gauranEmpiricalNullEstimation2018
library(empiricalzip)
data(protein, package="empiricalzip")
k<-1
x <- protein[,(k+1)][!is.na(protein[,(k+1)])]
M <- length(x)
cf <- 10^(-10)
count <- c(length(x[which(x==0)]), tabulate(x, nbins = max(x)))
names(count) <- c(0:max(x))
aa <- my.RealDataFDR(x, 0.1) #
##     ZIGP ZIP   GP  P         ZIGP ZIP   GP  P        ZIGP  ZIP     GP  P
##      one stage procedure      two stage             Storey's FDR
## [1,]  191  212  212  212      201  212  223  212     200   211   211   211
## [2,]  143  205  212  212      143  205  229  212     162   204   211   211

## [[2]]
## #        eta      lambda    theta      pi    C0   AD   
## [1,] 0.3245463 1.9167862 0.1415783 0.4576160    6    7
## [2,] 0.2288849 1.0948534        NA 0.3985044    3    6
## [3,]        NA 0.8082199 0.0000000 0.3944174    3    4
## [4,]        NA 0.8082199        NA 0.3944174    3    6

## [[3]]
##           [,1]      [,2]      [,3]      [,4] [,5] [,6]
## [1,] 0.2251642 2.1452087 0.5741445 0.6141275    4   36
## [2,] 0.2759650 1.3856234        NA 0.4244176    2    6
## [3,]        NA 0.7994067 0.0000000 0.3929166    2    3
## [4,]        NA 0.7994067        NA 0.3929166    2    6

# ZIGP parameters
eta <- 0.3245463
lambda <- 1.9167862 
theta <- 0.1415783

# Generate the plots and analysis
results <- plot_zigp_comparison(
  observed_counts = count,
  eta = eta,
  lambda = lambda, 
  theta = theta,
  max_display = 30,
  y_min = -15  # Extended y-axis
)

results$log_scale_tail
abline(v=6)

plot_normalized_mixture(
   eta = eta,
   lambda =  lambda, 
   theta = theta,
   pi0 = 1,
   n_binom = 15,
   p_binom = 0.2,
   shift = 7
)

```

```{R, echo=FALSE, eval=FALSE}

# Create a named vector with the observed counts
observed_data <- freqs
names(observed_data) <- as.character(counts)

# ZIGP parameters
eta <- 0.9993963  
lambda <- 1.0020464
theta <- 1.133941e-12

# Generate the plots and analysis
results <- plot_zigp_comparison(
  observed_counts = observed_data,
  eta = eta,
  lambda = lambda, 
  theta = theta,
  max_display = 30,
  y_min = -15  # Extended y-axis
)

# Display the plots
library(gridExtra)
grid.arrange(results$log_scale, results$linear_scale, ncol = 1)

# Testing separation between null and alternative distributions
# First, examine where the residuals cross from positive to negative
crossover_point <- min(which(results$separation_analysis$Residual < 0))
print(paste("Crossover point where observed becomes less than expected:", 
            results$separation_analysis$Count[crossover_point]))

# Calculate patterns of overdispersion
counts_with_excess <- sum(results$separation_analysis$Observed > results$separation_analysis$Fitted)
print(paste("Number of count values with excess observations:", counts_with_excess))

# Calculate the count where maximum deviation occurs
max_positive_deviation <- which.max(results$separation_analysis$RelativeDiff)
print(paste("Count with maximum positive deviation:", 
            results$separation_analysis$Count[max_positive_deviation]))

# Save the improved plot
#ggsave("./script3/zigp_log_scale_extended.png", results$log_scale, width = 10, height = 6)
#ggsave("zigp_ratio_plot.png", results$ratio_plot, width = 10, height = 6)

```

```{r zigplogscaleextended, eval=TRUE, echo=FALSE, fig.cap='empiricalzip fitted ZIPG',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("./script3/zigp_log_scale_extended.png")

```

\clearpage
\FloatBarrier
## Local FDR Calculation
The local FDR (false discovery rate) at a particular count value is calculated as:

- fdr(t) = ($\pi_0 \times f_0(t)) / f(t)$

Where:

-  $t4 is the mutation count
-  $\pi_0$ is the estimated proportion of nulls
-  $ f_0(t)$ is the null distribution probability at count $t$
-  $f(t)$ is the observed mixture distribution probability at count $t$

In the my.RealDataFDR function, this calculation happens here:

```{verbatim}
LFDR <- pi[model, CO[method, model]] * gpdMixture(x = sort(x), 
    estimate[model, 1, CO[method, model]], estimate[model, 
    2, CO[method, model]], estimate[model, 3, CO[method, 
    model]]) * f12(sort(x), length(x))
```
Breaking this down:

-  pi[model, CO[method, model]] is the estimated $\pi_0$  (proportion of nulls)
-  gpdMixture() calculates  $ f_0(t)$ , the probability under the null distribution (ZIGP, ZIP, GP, or P) using the estimated parameters
-  f12(sort(x), length(x)) is a correction factor related to f(t). Looking at the f12 function in the RcppExports.R, this factor is calculated as:

```{verbatim}
   M / length(x[which(x==x[i])])
```
where M is the total number of positions. This is effectively equivalent to $1/f(t)$, where $f(t)$ is the empirical probability of observing count $t$.

For the one-stage procedure, positions are declared significant if:

```{verbatim}
sum(LFDR[!is.na(LFDR)] < FDRLevel)
```
meaning their local FDR is below the specified threshold (in your case, 0.1).

Why Some Positions with Counts >7 Aren't Called Significant

With the ZIGP model estimated parameters for your data, the local FDR for positions with counts of 8, 9, etc. depends on:

-   How rare such counts are in the data (affecting $f(t)$)
-  How probable they are under the null distribution (affecting $f_0(t)$)

When a position has a count that is high but not extremely high relative to the estimated null distribution, its local
FDR might still exceed the threshold, especially when $\pi_0$ is estimated to be very close to 1.0.

This is why 7 positions with counts >7 aren't called significant by the one-stage procedure - their local FDR values
exceed your threshold of 0.1, indicating there's still a substantial probability they come from the null distribution
despite their relatively high counts.

The two-stage procedure, by contrast, automatically declares positions with counts $\geq$ AD (in this case 8) as significant,
bypassing the local FDR calculation for these high counts.

1. Stage 1 (Screening Step): We reject the $H_0 j \textrm{if } j \leq D_N(\hat{C})$ and do not reject if $j \leq \hat{C}$.
2. Stage 2 (Testing step): For $\hat{C} + 1 \leq j < D_N(\hat{C})$ , we calculate the local FDR

## Simplified two-stage procedure for empiricalzip

-  "Estimated C: 1"
-  "D_N threshold: 3"
-  "Significant counts:"   3  4  5  7 27
- used x_sample so some counts missing
- could use this value

```{R, echo=FALSE, eval=FALSE}
# Simplified two-stage procedure for extreme zero-inflation
simplified_two_stage <- function(data, fixed_C = 1, alpha = 0.05) {
  # Extract the counts and their frequencies
  tab <- table(data)
  counts <- as.numeric(names(tab))
  freqs <- as.numeric(tab)
  N <- sum(freqs)

  # Given extreme zero-inflation, set C directly
  C_hat <- fixed_C

  # Estimate parameters for the null distribution (ZIGP)
  # Given the extreme zeros, simplify parameter estimation
  eta_est <- freqs[counts == 0] / N  # Zero-inflation parameter

  # For non-zero values, estimate lambda and theta
  non_zero_counts <- rep(counts[counts > 0], freqs[counts > 0])
  lambda_est <- mean(non_zero_counts)
  theta_est <- min(0.5, var(non_zero_counts) / lambda_est - 1)

  # Calculate D_N (direct approach for computational stability)
  # With extreme zeros, a lower D_N is appropriate
  D_N <- max(3, C_hat + 1)  # Simple heuristic

  # Create results structure
  results <- list(
    C_hat = C_hat,
    D_N = D_N,
    parameters = c(eta = eta_est, lambda = lambda_est, theta = theta_est),
    significant_counts = counts[counts >= D_N]
  )

  return(results)
}

# Load necessary libraries
library(VGAM)
library(stats)

# Function to implement the two-stage procedure from Gauran et al.
two_stage_zigp_procedure <- function(data, alpha = 0.05) {
  # Extract the counts and their frequencies
  counts <- as.numeric(names(table(data)))
  freqs <- as.numeric(table(data))
  N <- sum(freqs)

  # Calculate the empirical probabilities
  f_hat <- freqs / N

  # Define function to estimate C using the likelihood approach (C1 method)
  estimate_C1 <- function(counts, freqs, N) {
    # Initialize variables
    max_S <- -Inf
    best_C <- 1

    # Try different values of C
    for (v in 1:(max(counts)-1)) {
      # Subset data for estimation
      subset_counts <- counts[counts <= v]
      subset_freqs <- freqs[counts <= v ]

      # Fit ZIGP to data up to C using MLE or method of moments
      # This is a simplified approach - in practice, use EM algorithm
      p_zeros <- subset_freqs[subset_counts == 0] / sum(subset_freqs)
      lambda_est <- mean(rep(subset_counts, subset_freqs))

      # Due to extreme zero-inflation, we'll use high eta
      eta_est <- max(0.95, p_zeros - (1-p_zeros)*exp(-lambda_est))
      theta_est <- min(0.5, var(rep(subset_counts, subset_freqs)) / lambda_est - 1)

      # Calculate pi_0 estimate
      pi0_est <- min(1, sum(freqs[counts <= v]) / (N * sum(zigp_pmf(0:v, eta_est, lambda_est, theta_est))))

      # Calculate S_v statistic
      S_v <- sum(freqs[counts <= v] * log(zigp_pmf(counts[counts <= v], eta_est, lambda_est, theta_est) / f_hat[counts <= v]))

      # Update if this is the best C so far
      if (S_v > max_S) {
        max_S <- S_v
        best_C <- v
        best_params <- c(eta_est, lambda_est, theta_est, pi0_est)
      }
    }

    return(list(C = best_C, params = best_params))
  }

  # Define ZIGP PMF function (simplified)
  zigp_pmf <- function(x, eta, lambda, theta) {
    result <- numeric(length(x))

    # Handle zeros separately due to zero-inflation
    result[x == 0] <- eta + (1 - eta) * exp(-lambda)

    # Handle positive values
    for (i in which(x > 0)) {
      val <- x[i]
      prob <- (1 - eta) * (lambda * (lambda + theta * val)^(val - 1) * exp(-lambda - theta * val)) / factorial(val)
      result[i] <- prob
    }

    return(result)
  }

  # Calculate D_N (truncation point)
  calculate_D_N <- function(C, params) {
    eta <- params[1]
    lambda <- params[2]
    theta <- params[3]

    # Calculate D_N based on the paper's formula
    # Simplified version for extreme zero-inflation
    D1 <- ceiling(log(N) / (-log(theta)))
    D2 <- ceiling(C + 1)

    return(max(D1, D2))
  }

  # Stage 1: Estimate C and calculate D_N
  C_result <- estimate_C1(counts, freqs, N)
  C_hat <- C_result$C
  params <- C_result$params
  D_N <- calculate_D_N(C_hat, params)

  # Stage 2: Apply local FDR to values between C+1 and D_N-1
  local_fdr_result <- numeric(max(counts))

  for (j in (C_hat+1):(D_N-1)) {
    if (j %in% counts) {
      # Calculate local FDR
      null_density <- zigp_pmf(j, params[1], params[2], params[3])
      mixture_density <- f_hat[counts == j]
      local_fdr_result[j] <- params[4] * null_density / mixture_density
    }
  }

  # Determine significant counts
  significant_counts <- numeric(0)

  # Reject all counts \geq D_N
  if (any(counts >= D_N)) {
    significant_counts <- c(significant_counts, counts[counts >= D_N])
  }

  # Reject counts between C+1 and D_N-1 if local FDR \leq alpha
  for (j in (C_hat+1):(D_N-1)) {
    if (j %in% counts && local_fdr_result[j] <= alpha) {
      significant_counts <- c(significant_counts, j)
    }
  }

  return(list(
    C_hat = C_hat,
    D_N = D_N,
    parameters = params,
    local_fdr = local_fdr_result,
    significant_counts = significant_counts
  ))
} #end two_stage_zigp_procedure

## Apply the simplified two-stage procedure
results <- simplified_two_stage(x_sample, fixed_C = 1)

# Display results
print(paste("Estimated C:", results$C_hat))
print(paste("D_N threshold:", results$D_N))
print("Significant counts:")
print(results$significant_counts)

```

# hurdle model implementation

```{R, echo=FALSE, eval=FALSE}

library(pscl)  # For hurdle model implementation

# Fit a hurdle model
hurdle_model <- hurdle(
  x ~ 1,               # Formula (just intercept in this simple case)
  data = data.frame(x = x_sample),
  dist = "negbin",     # Distribution for non-zero counts
  zero.dist = "binomial" # Distribution for zero vs non-zero process
)

summary(hurdle_model)

```

```{R, echo=FALSE, eval=FALSE}
# what is this? remove it
# Custom function for extreme zero-inflation
extreme_zero_model <- function(x, min_count = 3) {
  # Calculate basic statistics
  zeros <- sum(x == 0)
  n <- length(x)
  p_zero <- zeros/n

  # Estimate significance threshold
  # For extreme zero-inflation, we can use empirical approach
  unique_counts <- sort(unique(x))
  freq_table <- table(x)

  # Simple rule: counts are significant when they occur with 
  # frequency less than 0.1% of non-zero observations
  rarity_threshold <- 0.001 * (n - zeros)
  significant <- unique_counts[unique_counts >= min_count & 
                              freq_table[as.character(unique_counts)] < rarity_threshold]

  return(list(
    p_zero = p_zero,
    significant_counts = significant
  ))
}

```

# what is this 
```{R, echo=FALSE,eval=FALSE}

length(intersect(interaction_snps_counts$interaction_snps[1:50],lm_results$interaction_snps[1:88])) #22
# lm_results (has been ordered by lm_results$int_Pvalue)

t1 <- setdiff(interaction_snps_counts$interaction_snps[1:50],lm_results$interaction_snps[1:88])
match(t1,interaction_snps_counts$interaction_snps)  #3  4  5  7

t1[1]
#1] "X6_32222493 X6_32771829"
interaction_snps_counts[3,]
#              interaction_snps Freq
#280884 X6_32222493 X6_32771829  113

lm_results[grep("X6_32771829", lm_results$interaction_snps),] #301
lm_results[grep("X6_32222493 X6_32771829", lm_results$interaction_snps),] 
#              interaction_snps Freq SNP1_effect  SNP1_Pvalue SNP2_effect  SNP2_Pvalue int_effect int_Pvalue        Ch2 correlations splits_SNP1 splits_SNP2
#280884 X6_32222493 X6_32771829  113   0.3937617 6.972158e-09  -0.6516203 2.576182e-27 -0.1303979 0.08982739 0.03193039   -0.3742329          18          14
#       adj_p_value_bonferrroni adj_p_value_fdr adj_p_value_Ch2
#280884                57.48953       0.2521143       0.1011656

# so this is not significant for any of the tests except empiricalzip
# how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113

chisq.test(table( dt[,"X6_32222493",], dt[,"X6_32771829",]))

aa<-rep(NA, 74)
for ( i in 1:74){
    aa[i] <- length(grep(imp[i],interaction_snps_counts$interaction_snps[1:50]))
    }
aa
# [1] 21 13  8  7  5  5  5  8  5  3  2  2  2  1  2  2  1  1  0  0  1  0  0  0  0  0  1  0  0  0  1  0  0  1  1  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  1  0  0
#[56]  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0

index <- match(selectedSNPs,colnames(dt))
index <- index[order(index)]

data4interactions <- dt[,index]
saveRDS(data4interactions,paste0(dirOut,"/data4interactions.rds"))

SNPs <- lm_results$interaction_snps

SNPs <- as.character(SNPs)
SNPs <- gsub("X","" ,SNPs)
#SNPs <- read.table("Results/SNPs.txt",header=T)

```

4. what is the intesection of imp and interacting transcripts

   - see table at end of section

- "X6_32222493 X6_32771829" see below, how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113

   - it doesnt t2["X6_32222493"] =202 t2["X6_32771829"]= 1532 -- lm_result has a problem

# compare some distributions

- modified compare_distributions function and related plotting functions to work with a frequency table input rather than requiring the full vector. 
This will be much more memory-efficient when dealing with large datasets that have many zeros.

```{R, echo=FALSE, eval=FALSE}
# Then run the comparison
results <- compare_distributions_from_freq(counts, freqs, n_zeros)

# View results
print(results$comparison)
#    Model LogLikelihood Parameters     AIC     BIC AIC_Ratio
#2 ZI-Zipf      -3102980          2 6205964 6206005         1
#3    ZINB          -Inf          3     Inf     Inf       Inf
#1    ZIGP           NaN          3     NaN     NaN       NaN

print(paste("Best model:", results$best_model))

#png("./script3/ZI_Zipf_plot.png")
results$plots$"ZI_Zipf"
#dev.off()
#png("./script3/ZIGP_plot.png")
results$plots$"ZIGP"
#dev.off()
#png("./script3/ZINB_plot.png")
results$plots$"ZINB"
#dev.off()

results$parameters
#$ZIGP $ZIGP$eta [1] 0.9999 $ZIGP$lambda [1]  0.9752822  $ZIGP$theta  [1]  0.03459262
#$ZI_Zipf $ZI_Zipf$p0 [1] 0.9999348 $ZI_Zipf$alpha [1] 8.288837 
#$ZINB $ZINB$p0 [1] 0.9999348 $ZINB$pi_0  [1] 0.9998953 $ZINB$size  [1] 13.84857 $ZINB$mu [1] 1.010229

```

```{R, zipfplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Power Law (Zipf) Distribution", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/ZI_Zipf_plot.png")
```

```{R, zigpplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Generalized Poisson", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/ZIGP_plot.png")
```

```{R, zinbplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Negative Binomial", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/ZINB_plot.png")
```

```{R, echo=FALSE, eval=FALSE}
# Modified functions to work with frequency data instead of raw vectors

# ZIGP probability mass function
zigp_pmf <- function(x, eta, lambda, theta) {
  result <- numeric(length(x))

  # Handle zeros
  result[x == 0] <- eta + (1 - eta) * exp(-lambda)

  # Handle positive values
  for (i in which(x > 0)) {
    val <- x[i]
    prob <- (1 - eta) * lambda * (lambda + theta * val)^(val - 1) * exp(-lambda - theta * val) / factorial(val)
    result[i] <- prob
  }

  return(result)
}

# Modifications to handle extreme zero-inflation
# For the ZIGP model:
estimate_zigp_params_from_freq <- function(counts, freqs, n_zeros) {
  # Calculate basic statistics
  n <- sum(freqs) + n_zeros
  p_zeros <- n_zeros / n

  # If extreme zero-inflation
  if (p_zeros > 0.9999) {
    # Use a simpler model with high but not exactly 1 eta
    return(list(eta = 0.9999, lambda = mean(counts[counts > 0]), theta = 0))
  }

    # Rest of the function as before...
  # Calculate statistics for positive values
  if (sum(freqs[counts > 0]) == 0) {
    return(list(eta = 1, lambda = 1, theta = 0))  # All zeros case
  }

  # Calculate mean and variance of positive values
  # Weighted calculations using frequencies
  mean_pos <- sum(counts[counts > 0] * freqs[counts > 0]) / sum(freqs[counts > 0])

  # Calculate variance
  var_pos <- sum(freqs[counts > 0] * (counts[counts > 0] - mean_pos)^2) / sum(freqs[counts > 0])

  # Estimate parameters using method of moments
  # For zero-inflated data, first estimate eta
  eta_est <- max(0, (p_zeros - exp(-mean_pos)) / (1 - exp(-mean_pos)))

  # Then estimate lambda and theta from positive values
  lambda_est <- mean_pos
 # Estimate theta from variance relationship
  # In GP: Var(X) = lambda / (1-theta)^2
  # Solving for theta
  theta_est <- max(0, min(0.999, 1 - sqrt(lambda_est / var_pos)))

  return(list(eta = eta_est, lambda = lambda_est, theta = theta_est))
}
aa<- estimate_zigp_params_from_freq(counts, freqs, n_zeros)

```

```{R, echo=FALSE, eval=FALSE}
# Improved function to estimate ZIGP parameters from frequency data
# This handles zero-inflation properly by treating zeros separately
improved_estimate_zigp_params <- function(counts, freqs, n_zeros) {
  # Calculate basic statistics
  n <- sum(freqs) + n_zeros
  p_zeros <- n_zeros / n

  # Focus only on non-zero counts for fitting Generalized Poisson
  pos_idx <- counts > 0
  pos_counts <- counts[pos_idx]
  pos_freqs <- freqs[pos_idx]

  # Calculate weighted statistics for positive counts
  mean_pos <- sum(pos_counts * pos_freqs) / sum(pos_freqs)
  var_pos <- sum(pos_freqs * (pos_counts - mean_pos)^2) / sum(pos_freqs)

  # Check if the data is actually highly overdispersed
  if (var_pos < mean_pos) {
    # If variance is less than mean (underdispersed), default to Poisson (theta = 0)
    lambda_est <- mean_pos
    theta_est <- 0
  } else {
    # For overdispersed data, estimate theta using variance relationship
    # In GP: Var(X) = lambda * (1 + theta)^2 / (1 - theta)^2
    # Solving for theta (simplified version)
    dispersion_ratio <- var_pos / mean_pos

    if (dispersion_ratio <= 1) {
      # No overdispersion
      theta_est <- 0
    } else if (dispersion_ratio < 10) {
      # Moderate overdispersion - using approximation
      theta_est <- 1 - 1/sqrt(dispersion_ratio)
      theta_est <- max(0, min(0.8, theta_est))  # Cap at 0.8 for stability
    } else {
      # Extreme overdispersion - use caution
      theta_est <- 0.8  # Cap at 0.8 for numerical stability
    }

    # Adjust lambda based on theta estimate
    lambda_est <- mean_pos * (1 - theta_est)
  }

  # Calculate eta (zero-inflation parameter) using the fitted GP distribution
  p_zero_gp <- exp(-lambda_est)  # Probability of zero under GP

  # Handle extreme zero-inflation
  if (p_zeros > 0.9999) {
    eta_est <- 0.9999  # Cap at 0.9999 for numerical stability
  } else if (p_zeros <= p_zero_gp) {
    # If observed zeros are fewer than expected from GP alone
    eta_est <- 0  # No zero-inflation needed
  } else {
    # Calculate eta using the zero-inflation formula
    eta_est <- (p_zeros - p_zero_gp) / (1 - p_zero_gp)
    eta_est <- max(0, min(0.9999, eta_est))  # Ensure it's in [0, 0.9999]
  }

  # Print debugging information
  cat("Estimated parameters:\n")
  cat("Mean of positive counts:", mean_pos, "\n")
  cat("Variance of positive counts:", var_pos, "\n")
  cat("Dispersion ratio (var/mean):", var_pos/mean_pos, "\n")
  cat("eta:", eta_est, "\n")
  cat("lambda:", lambda_est, "\n")
  cat("theta:", theta_est, "\n")

  # Return estimates
  return(list(eta = eta_est, lambda = lambda_est, theta = theta_est))
}

# Function to calculate ZIGP probabilities
zigp_pmf <- function(x, eta, lambda, theta) {
  result <- numeric(length(x))

  # Handle zeros separately due to zero-inflation
  result[x == 0] <- eta + (1 - eta) * exp(-lambda)

  # Handle positive values
  for (i in which(x > 0)) {
    val <- x[i]
    prob <- (1 - eta) * lambda * (lambda + theta * val)^(val - 1) * 
            exp(-lambda - theta * val) / factorial(val)
    result[i] <- prob
  }

  return(result)
}

# Function to compare observed vs expected frequencies
compare_observed_expected <- function(counts, freqs, n_zeros, params) {
  # Create full data frame for comparison
  all_counts <- c(0, counts)
  all_freqs <- c(n_zeros, freqs)
  total_n <- sum(all_freqs)

  # Calculate ZIGP probabilities
  expected_probs <- zigp_pmf(all_counts, params$eta, params$lambda, params$theta)
  expected_freqs <- expected_probs * total_n

  comparison <- data.frame(
    Count = all_counts,
    Observed = all_freqs,
    Expected = expected_freqs,
    ObsProb = all_freqs / total_n,
    ExpProb = expected_probs,
    Ratio = all_freqs / expected_freqs
  )

  return(comparison)
}

# Example usage:
# counts <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
# freqs <- c(290502, 418, 89, 29, 22, 14, 11, 7, 5, 4)
# n_zeros <- 4463369228
# 
# # Estimate improved parameters
params <- improved_estimate_zigp_params(counts, freqs, n_zeros)
# 
# # Compare observed vs expected
comparison <- compare_observed_expected(counts, freqs, n_zeros, params)
print(comparison)
estimate_zigp_params_from_freq<- improved_estimate_zigp_params

```

```{R, echo=FALSE, eval=FALSE}
# Power law PMF (Zipf distribution)
zipf_pmf <- function(x, alpha, xmin = 1) {
  result <- numeric(length(x))
  # Use a more efficient approach for normalization
  denom <- zeta(alpha, xmin)  # Riemann zeta function starting at xmin

  result[x < xmin] <- 0
  result[x >= xmin] <- (x[x >= xmin]^(-alpha)) / denom

  return(result)
}

# Helper function: approximate Riemann zeta function
zeta <- function(alpha, xmin = 1) {
  # For alpha > 1, we can use a finite sum for approximation
  max_k <- 10000  # This should be enough for good approximation with alpha > 2
  sum((xmin:max_k)^(-alpha))
}

# For zero-inflated data, combine with point mass at zero
zi_zipf_pmf <- function(x, p0, alpha, xmin = 1) {
  result <- numeric(length(x))
  result[x == 0] <- p0
  result[x > 0] <- (1 - p0) * zipf_pmf(x[x > 0], alpha, xmin)
  return(result)
}

# Zero-inflated negative binomial
zinb_pmf <- function(x, p0, size, mu) {
  result <- numeric(length(x))
  result[x == 0] <- p0 + (1 - p0) * dnbinom(0, size = size, mu = mu)
  result[x > 0] <- (1 - p0) * dnbinom(x[x > 0], size = size, mu = mu)
  return(result)
}

# Fit Zipf distribution from frequency data
fit_zipf_from_freq <- function(counts, freqs) {
  # Skip 0 counts
  pos_idx <- counts > 0
  if(sum(pos_idx) == 0) return(list(alpha = 1))

  pos_counts <- counts[pos_idx]
  pos_freqs <- freqs[pos_idx]

  # MLE for alpha (approximate)
  alpha_start <- 1.5  # Starting value

  neg_ll <- function(alpha) {
    # Calculate log likelihood weighted by frequencies
    -sum(pos_freqs * log(zipf_pmf(pos_counts, alpha)))
  }

  result <- optim(alpha_start, neg_ll, method = "BFGS")
  return(list(alpha = result$par))
}

# Fit Negative Binomial from frequency data
fit_nb_from_freq <- function(counts, freqs) {
  # Skip 0 counts
  pos_idx <- counts > 0
  if(sum(pos_idx) == 0) return(list(size = 1, mu = 1))

  pos_counts <- counts[pos_idx]
  pos_freqs <- freqs[pos_idx]

  # Calculate mean and variance using frequencies as weights
  mu_est <- sum(pos_counts * pos_freqs) / sum(pos_freqs)

  # Calculate variance
  var_est <- sum(pos_freqs * (pos_counts - mu_est)^2) / sum(pos_freqs)

  if(var_est <= mu_est) var_est <- mu_est * 1.1  # Ensure overdispersion

  size_est <- mu_est^2 / (var_est - mu_est)

  return(list(size = size_est, mu = mu_est))
}

# Improved plotting function for frequency data
plot_distribution_from_freq <- function(counts, freqs, n_zeros, model_name, model_probs, p0 = NULL) {
  # Add the zero count
  all_counts <- c(0, counts)
  all_freqs <- c(n_zeros, freqs)

  # Calculate observed probabilities
  total_n <- sum(all_freqs)
  obs_probs <- all_freqs / total_n

  # Focus on counts with significant probability
  max_count <- min(50, max(all_counts))

  # Create a sequence of counts to display
  display_counts <- 0:max_count

  # Create data for plotting
  plot_data <- data.frame(
    Count = factor(display_counts),
    Observed = NA,
    Fitted = NA
  )

  # Fill in observed probabilities
  for(i in 1:length(all_counts)) {
    if(all_counts[i] <= max_count) {
      idx <- which(display_counts == all_counts[i])
      if(length(idx) > 0) {
        plot_data$Observed[idx] <- obs_probs[i]
      }
    }
  }

  # Fill in fitted probabilities
  model_counts <- c(0, counts) # Include zero
  if(length(model_probs) < length(model_counts)) {
    # Handle case where model_probs might not have same length
    model_probs <- c(model_probs[1], model_probs)  # Duplicate first value for zero
  }

  for(i in 1:length(model_counts)) {
    if(model_counts[i] <= max_count) {
      idx <- which(display_counts == model_counts[i])
      if(length(idx) > 0) {
        plot_data$Fitted[idx] <- model_probs[i]
      }
    }
  }

  # Convert to long format for plotting
  library(tidyr)
  plot_long <- pivot_longer(plot_data, 
                           cols = c("Observed", "Fitted"),
                           names_to = "Type", 
                           values_to = "Probability")

  # Remove missing values
  plot_long <- plot_long[!is.na(plot_long$Probability),]

  # Create the plot
  library(ggplot2)
  model_label <- paste("Fitted", model_name)
  fill_values <- c("red", "blue")
  names(fill_values) <- c("Observed", "Fitted")

  p <- ggplot(plot_long, aes(x = Count, y = Probability, fill = Type)) +
    geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
    scale_y_log10(limits = c(min(plot_long$Probability[plot_long$Probability > 0]) * 0.5, 1)) +
    theme_minimal() +
    labs(title = paste("Observed vs Fitted", model_name, "Probabilities (log scale)"),
         subtitle = if(!is.null(p0)) paste0("p0 = ", round(p0, 4)) else "",
         x = "Count Value", y = "Probability (log scale)") +
    scale_fill_manual(values = fill_values) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  return(p)
}

# Main function to compare distributions based on frequency data
compare_distributions_from_freq <- function(counts, freqs, n_zeros) {
  # Calculate total number of observations
    n <- sum(freqs) + n_zeros
    # For the log-likelihood calculations:
    # Add small epsilon to avoid log(0)
    epsilon <- 1e-10

  # Calculate empirical probability of zero
  p0_emp <- n_zeros / n

  # Fit ZIGP (as baseline)
  params_zigp <- estimate_zigp_params_from_freq(counts, freqs, n_zeros)

  # Calculate ZIGP probabilities and log-likelihood
  unique_counts <- sort(unique(c(0, counts)))
  probs_zigp <- zigp_pmf(unique_counts, params_zigp$eta, params_zigp$lambda, params_zigp$theta)

  # Calculate log likelihood using frequencies
  count_map <- match(c(0, counts), unique_counts)
  #ll_zigp <- n_zeros * log(probs_zigp[1]) + sum(freqs * log(probs_zigp[count_map[-1]]))
  ll_zigp <- n_zeros * log(max(probs_zigp[1], epsilon)) + sum(freqs * log(pmax(probs_zigp[count_map[-1]], epsilon)))

  # Fit power law (Zipf)
  params_zipf <- fit_zipf_from_freq(counts, freqs)
  p0_est <- p0_emp

  # Calculate Zipf probabilities and log-likelihood
  probs_zipf_pos <- zipf_pmf(counts, params_zipf$alpha)
  probs_zipf <- c(p0_est, (1-p0_est) * probs_zipf_pos)
  #ll_zipf <- n_zeros * log(p0_est) + sum(freqs * log((1-p0_est) * probs_zipf_pos))
  ll_zipf <- n_zeros * log(max(p0_est, epsilon)) +   sum(freqs * log(pmax((1-p0_est) * probs_zipf_pos, epsilon)))

  # Fit Negative Binomial
  params_nb <- fit_nb_from_freq(counts, freqs)

    # Calculate NB probabilities and log-likelihood
    # For the ZINB model:
    # Ensure the parameters are reasonable for extreme zero-inflation
    params_nb$size <- pmin(params_nb$size, 1000)  # Cap the size parameter

  probs_nb_pos <- dnbinom(counts, size = params_nb$size, mu = params_nb$mu)
  probs_nb_0 <- dnbinom(0, size = params_nb$size, mu = params_nb$mu)

  # For zero-inflated model
  pi_0 <- (p0_est - probs_nb_0) / (1 - probs_nb_0)
  pi_0 <- max(0, min(1, pi_0))  # Ensure valid probability

  probs_zinb <- c(p0_est, (1-pi_0) * probs_nb_pos)
  ll_nb <- n_zeros * log(p0_est) + sum(freqs * log((1-pi_0) * probs_nb_pos))

  # Calculate AIC and BIC
  aic_zigp <- -2 * ll_zigp + 2 * 3  # 3 parameters
  aic_zipf <- -2 * ll_zipf + 2 * 2  # 2 parameters
  aic_nb <- -2 * ll_nb + 2 * 3      # 3 parameters

  bic_zigp <- -2 * ll_zigp + log(n) * 3
  bic_zipf <- -2 * ll_zipf + log(n) * 2
  bic_nb <- -2 * ll_nb + log(n) * 3

  # Create comparison table
  results <- data.frame(
    Model = c("ZIGP", "ZI-Zipf", "ZINB"),
    LogLikelihood = c(ll_zigp, ll_zipf, ll_nb),
    Parameters = c(3, 2, 3),
    AIC = c(aic_zigp, aic_zipf, aic_nb),
    BIC = c(bic_zigp, bic_zipf, bic_nb),
    AIC_Ratio = c(aic_zigp/aic_zipf, 1, aic_nb/aic_zipf)  # Add ratio to better interpret differences
  )

  # Sort by AIC
  results <- results[order(results$AIC),]

  # Create plots for each model
  p_zigp <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZIGP", probs_zigp, params_zigp$eta)
  p_zipf <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZI-Zipf", probs_zipf, p0_est)
  p_nb <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZINB", probs_zinb, p0_est)

  # Return results
  return(list(
    comparison = results,
    best_model = results$Model[1],
    plots = list(ZIGP = p_zigp, ZI_Zipf = p_zipf, ZINB = p_nb),
    parameters = list(
      ZIGP = params_zigp,
      ZI_Zipf = c(list(p0 = p0_est), params_zipf),
      ZINB = c(list(p0 = p0_est, pi_0 = pi_0), params_nb)
    )
  ))
}

```

\begin{table}[h]
\centering
\caption{Model Comparison for Rheumatoid Arthritis Interaction Data}
\begin{tabular}{lrrrrr}
\hline
\textbf{Model} & \textbf{LogLikelihood} & \textbf{Parameters} & \textbf{AIC} & \textbf{BIC} & \textbf{AIC Ratio} \\
\hline
ZI-Zipf & -3,102,980 & 2 & 6,205,964 & 6,206,005  & 1.000 \\
ZINB    & -$\infty$  & 3 & $\infty$ & $\infty$    & $\infty$ \\
ZIGP    & NaN        & 3 & NaN      & NaN         & NaN \\
\hline
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Estimated Parameters for Distribution Models}
\begin{tabular}{lrrr}
\hline
\textbf{Model} & \textbf{Parameter 1} & \textbf{Parameter 2} & \textbf{Parameter 3} \\
\hline
ZIGP & $\eta = 0.9999$ & $\lambda = 46.5641$ & $\theta = 0$ \\
ZI-Zipf & $p_0 = 0.9999348$ & $\alpha = 8.288837$ & -- \\
ZINB & $p_0 = 0.9999348$ & $\pi_0 = 0.9998953$ & size = $13.84857$, $\mu = 1.010229$ \\
\hline
\end{tabular}
  \end{table}

- For ZIGP Model (NaN result):

  -  With eta = 0.9999 (extremely high zero-inflation) and theta = 0.03459262, the ZIGP model is not far from a Zero-Inflated Poisson with probability of zero very close to 1.
  -  When calculating log-likelihood for this model, you're likely getting extremely small probability values for non-zero counts, which when logged are approaching negative infinity.
  -  The combination of multiple very small probabilities in the log-likelihood calculation is causing arithmetic underflow, resulting in NaN.

- For ZINB Model (Inf result):

  -   With extremely high zero-inflation ($p_0 = 0.9999348 and pi_0 = 0.9998953), the model is struggling to explain the non-zero counts.
  -  The parameter size = 13.84857 with mu = 1.010229 means the distribution has a relatively narrow shape.
  -  The log-likelihood is likely resulting in -Inf because some observed counts have essentially zero probability under this model, and log(0) is -Inf.

The Core Issue: Extreme Zero-Inflation

- Your data has extreme zero-inflation (over 99.99% zeros). This causes numerical instabilities when fitting these models:

  -  When eta or p0 are extremely close to 1, the probability assigned to any non-zero count becomes vanishingly small.
  -  This leads to arithmetic underflow, where the computer can't represent such small numbers accurately.
  -  The ZI-Zipf model handles this better because power-law distributions naturally accommodate heavy tails better than Poisson or Negative Binomial.

# Power law

- A true power law would show a straight line. Deviations suggest a different distribution.
```{R, echo=FALSE, eval=FALSE}
# Plot log(frequency) vs log(count)
png("./script3/power_law.png")
plot(log(counts[counts > 0]), log(freqs[counts > 0]), 
     xlab="log(count)", ylab="log(frequency)")
abline(lm(log(freqs[counts > 0]) ~ log(counts[counts > 0])))
dev.off()

```

```{R, powerlaw, echo=FALSE, eval=TRUE, fig.cap="Plot of log(frequency) vs log(count). A power law would show a straight line", out.width = '50%',fig.align = 'center'}
knitr::include_graphics("./script3/power_law.png")
```

# thresholding

```{R, echo= TRUE, eval=FALSE}
# Function to calculate meaningful thresholds for each model
calculate_thresholds <- function(counts, freqs, n_zeros, model_params) {
  # Total count
  n_total <- sum(freqs) + n_zeros

  # Calculate mean and SD of positive counts
  mean_pos <- sum(counts * freqs) / sum(freqs)
  var_pos <- sum(freqs * (counts - mean_pos)^2) / sum(freqs)
  sd_pos <- sqrt(var_pos)

  # Calculate various thresholds
  thresholds <- list()

  # 1. Empirical approaches
  thresholds$empirical <- list(
    mean_plus_2sd = ceiling(mean_pos + 2 * sd_pos),
    mean_plus_3sd = ceiling(mean_pos + 3 * sd_pos),
    mean_plus_5sd = ceiling(mean_pos + 5 * sd_pos),
    percentile_99 = as.numeric(quantile(rep(counts, freqs), 0.99)),
    percentile_999 = as.numeric(quantile(rep(counts, freqs), 0.999))
  )

  # 2. Model-based approaches
  # a) For Zipf
  if (!is.null(model_params$ZI_Zipf)) {
    alpha <- model_params$ZI_Zipf$alpha
    p0 <- model_params$ZI_Zipf$p0

    # Find where probability becomes small
    find_zipf_threshold <- function(p_threshold) {
      count <- 1
      while(count < 100) {  # Reasonable upper limit
        prob <- (1-p0) * zipf_pmf(count, alpha)
        if (prob < p_threshold) break
        count <- count + 1
      }
      return(count)
    }

    thresholds$zipf <- list(
      prob_0001 = find_zipf_threshold(0.0001),
      prob_00001 = find_zipf_threshold(0.00001)
    )
  }

  # b) For ZIGP
  if (!is.null(model_params$ZIGP)) {
    eta <- model_params$ZIGP$eta
    lambda <- model_params$ZIGP$lambda
    theta <- model_params$ZIGP$theta

    # Find where probability becomes small
    find_zigp_threshold <- function(p_threshold) {
      count <- 1
      while(count < 100) {  # Reasonable upper limit
        prob <- (1-eta) * (lambda * (lambda + theta * count)^(count - 1) * 
                 exp(-lambda - theta * count)) / factorial(count)
        if (prob < p_threshold) break
        count <- count + 1
      }
      return(count)
    }

    thresholds$zigp <- list(
      prob_0001 = find_zigp_threshold(0.0001),
      prob_00001 = find_zigp_threshold(0.00001)
    )
  }

  # 3. Count-based approaches
  # Find natural break points in the distribution
  natural_breaks <- find_natural_breaks(counts, freqs)
  thresholds$breaks <- natural_breaks

  # Return all thresholds
  return(thresholds)
}

# Helper function to identify natural break points in count distribution
find_natural_breaks <- function(counts, freqs) {
  # Calculate rate of change between consecutive counts
  rate_change <- numeric(length(counts) - 1)
  for (i in 1:(length(counts)-1)) {
    if (counts[i+1] - counts[i] == 1) {  # Only consider consecutive integers
      rate_change[i] <- freqs[i] / freqs[i+1]
    } else {
      rate_change[i] <- NA  # Mark non-consecutive values
    }
  }

  # Find points where rate of change is significantly higher than average
  valid_rates <- rate_change[!is.na(rate_change)]
  if (length(valid_rates) < 3) return(list(break_points = NULL))

  mean_rate <- mean(valid_rates)
  sd_rate <- sd(valid_rates)

  # Points where rate of change is > mean + 1.5*SD
  break_indices <- which(rate_change > mean_rate + 1.5*sd_rate)
  break_points <- counts[break_indices]

  # Return the identified break points
  return(list(
    break_points = break_points,
    rate_change = rate_change,
    suggested_threshold = if(length(break_points) > 0) min(break_points) + 1 else NULL
  ))
}

thresholds <- calculate_thresholds(counts, freqs, n_zeros, results$parameters)
print(thresholds)
## empirical
## $empirical$mean_plus_2sd
## [1] 4

## $empirical$mean_plus_3sd
## [1] 5

## $empirical$mean_plus_5sd
## [1] 7

## $empirical$percentile_99
## [1] 1

## $empirical$percentile_999
## [1] 2

## $zipf
## $zipf$prob_0001
## [1] 1

## $zipf$prob_00001
## [1] 2

## $zigp
## $zigp$prob_0001
## [1] 1

## $zigp$prob_00001
## [1] 1

## $breaks
## $breaks$break_points
## [1] 1

## $breaks$rate_change
##  [1] 694.980861   4.696629   3.068966   1.318182   1.571429   1.272727   1.571429   1.400000   1.250000   1.333333   1.000000   1.000000   0.750000   4.000000   1.000000   1.000000
## [17]   1.000000         NA   2.000000         NA   1.000000   0.500000         NA         NA         NA         NA   1.000000         NA         NA   0.500000   2.000000         NA
## [33]   1.000000         NA         NA         NA         NA         NA

## $breaks$suggested_threshold
## [1] 2

find_natural_breaks (counts, freqs)
## $break_points
## [1] 1

## $rate_change
##  [1] 694.980861   4.696629   3.068966   1.318182   1.571429   1.272727   1.571429   1.400000   1.250000   1.333333   1.000000   1.000000   0.750000   4.000000   1.000000   1.000000
## [17]   1.000000         NA   2.000000         NA   1.000000   0.500000         NA         NA         NA         NA   1.000000         NA         NA   0.500000   2.000000         NA
## [33]   1.000000         NA         NA         NA         NA         NA

## $suggested_threshold
## [1] 2

```

```{R, echo=FALSE, eval=FALSE}
# more refined threshholding

# Robust Empirical Thresholding for Interaction Counts
# This approach works for both well-fitted distributions and those that don't follow 
# standard theoretical distributions

empirical_threshold <- function(counts, freqs, n_zeros = 0) {
  # Create a data frame for easier analysis
  df <- data.frame(
    count = counts,
    freq = freqs,
    prop = freqs / sum(freqs)
  )

  # Order by count
  df <- df[order(df$count), ]

  # Calculate cumulative proportion (excluding zeros)
  df$cum_prop <- cumsum(df$prop)

  # Calculate several thresholds
  thresholds <- list()

  # 1. Basic statistics on positive counts
  mean_pos <- sum(df$count * df$freq) / sum(df$freq)
  var_pos <- sum(df$freq * (df$count - mean_pos)^2) / sum(df$freq)
  sd_pos <- sqrt(var_pos)

  # 2. Compute empirical percentiles
  weighted_data <- rep(df$count, times = df$freq)

  thresholds$basic <- list(
    mean = mean_pos,
    sd = sd_pos,
    threshold_2sd = ceiling(mean_pos + 2 * sd_pos),
    threshold_3sd = ceiling(mean_pos + 3 * sd_pos),
    p95 = as.numeric(quantile(weighted_data, 0.95)),
    p99 = as.numeric(quantile(weighted_data, 0.99)),
    p999 = as.numeric(quantile(weighted_data, 0.999))
  )

  # 3. Detect "elbows" or breakpoints in the distribution
  # Calculate rate of change in frequency between consecutive counts
  df$rate_change <- NA
  for (i in 1:(nrow(df)-1)) {
    if (df$count[i+1] - df$count[i] == 1) {  # Only for consecutive counts
      df$rate_change[i] <- df$freq[i] / df$freq[i+1]
    }
  }

  valid_rates <- df$rate_change[!is.na(df$rate_change)]
  if (length(valid_rates) >= 3) {
    mean_rate <- mean(valid_rates)
    sd_rate <- sd(valid_rates)

    # Find significant jumps in rate of change
    # Points where rate of change is > mean + 1.5*SD
    break_indices <- which(df$rate_change > mean_rate + 1.5*sd_rate)
    break_points <- df$count[break_indices]

    thresholds$breakpoints <- list(
      points = break_points,
      suggested = if(length(break_points) > 0) min(break_points) + 1 else NULL
    )
  } else {
    thresholds$breakpoints <- list(
      points = NULL,
      suggested = NULL
    )
  }

  # 4. Entropy-based thresholding
  # Calculate entropy of frequency distribution
  p <- df$prop
  entropy <- -sum(p * log(p))

  # Find threshold that maximizes entropy difference between below and above threshold
  max_diff <- -Inf
  best_threshold <- 1

  for (i in 1:max(df$count)) {
    below <- df[df$count <= i, ]
    above <- df[df$count > i, ]

    # Skip if either group is empty
    if (nrow(below) == 0 || nrow(above) == 0) next

    # Normalize probabilities within each group
    p_below <- below$freq / sum(below$freq)
    p_above <- above$freq / sum(above$freq)

    # Calculate entropies
    entropy_below <- -sum(p_below * log(p_below))
    entropy_above <- -sum(p_above * log(p_above))

    # Calculate weighted entropy difference
    entropy_diff <- (sum(below$freq) * entropy_below + sum(above$freq) * entropy_above) / sum(df$freq)

    if (entropy - entropy_diff > max_diff) {
      max_diff <- entropy - entropy_diff
      best_threshold <- i
    }
  }

  thresholds$entropy <- list(
    threshold = best_threshold + 1  # Add 1 for exclusive threshold
  )

  # 5. Exponential decay test
  # Test if counts follow exponential decay by fitting log(frequency) ~ count
  if (nrow(df) > 5) {
    log_freq <- log(df$freq)
    model <- lm(log_freq ~ df$count)
    r_squared <- summary(model)$r.squared

    # If strong exponential decay, use where predicted frequency < 1
    if (r_squared > 0.8) {
      coefs <- coef(model)
      threshold_exp <- ceiling(-coefs[1] / coefs[2])  # Where predicted freq < 1
      thresholds$exponential <- list(
        r_squared = r_squared,
        threshold = threshold_exp
      )
    } else {
      thresholds$exponential <- list(
        r_squared = r_squared,
        threshold = NULL
      )
    }
  }

  # 6. Rarity-based threshold
  # Define thresholds based on observation frequency
  total_observations <- sum(df$freq)
  rare_threshold <- ceiling(total_observations * 0.001)  # 0.1% of observations

  rare_counts <- df[df$freq <= rare_threshold, "count"]
  if (length(rare_counts) > 0) {
    thresholds$rarity <- list(
      threshold = min(rare_counts)
    )
  } else {
    thresholds$rarity <- list(
      threshold = NULL
    )
  }

  # 7. Add zero-inflation context if provided
  if (n_zeros > 0) {
    # Calculate proportion of zeros
    p_zero <- n_zeros / (n_zeros + sum(df$freq))
    thresholds$zero_inflation <- list(
      p_zero = p_zero
    )

    # For extreme zero-inflation (>99.9%), more aggressive thresholds
    if (p_zero > 0.999) {
      # With extreme zeros, any occurrence might be significant
      thresholds$extreme_zero_inflation <- list(
        threshold = 2  # Consider anything geq 2 significant when zeros dominate
      )
    }
  }

  # 8. Final recommended threshold based on all methods
  # Logic for combining thresholds
  candidates <- c(
    thresholds$basic$threshold_2sd,
    thresholds$breakpoints$suggested,
    thresholds$entropy$threshold,
    thresholds$rarity$threshold
  )

  # Add exponential threshold if available
  if (!is.null(thresholds$exponential$threshold)) {
    candidates <- c(candidates, thresholds$exponential$threshold)
  }

  # Add extreme zero-inflation threshold if applicable
  if (!is.null(thresholds$extreme_zero_inflation$threshold)) {
    candidates <- c(candidates, thresholds$extreme_zero_inflation$threshold)
  }

  # Remove NAs
  candidates <- candidates[!is.na(candidates)]

  # For final recommendation, use median of available thresholds
  if (length(candidates) > 0) {
    thresholds$recommended <- median(candidates, na.rm = TRUE)
  } else {
    # Default if no thresholds available
    thresholds$recommended <- max(2, ceiling(mean_pos + 2 * sd_pos))
  }

  # Provide both results and visualization
  results <- list(
    thresholds = thresholds,
    data = df
  )

  # Return everything
  return(results)
}

# Function to visualize the thresholds on a count distribution
plot_thresholds <- function(threshold_results, log_scale = TRUE) {
  df <- threshold_results$data
  thresholds <- threshold_results$thresholds

  # Setup plot
  par(mfrow = c(1, 1))

  # Create bar plot of frequencies
  if (log_scale) {
    barplot(log(df$freq), names.arg = df$count, xlab = "Count", ylab = "Log Frequency", 
            main = "Distribution of Counts with Thresholds", col = "lightblue")
  } else {
    barplot(df$freq, names.arg = df$count, xlab = "Count", ylab = "Frequency", 
            main = "Distribution of Counts with Thresholds", col = "lightblue")
  }

  # Add recommended threshold line
  abline(v = which(df$count == ceiling(thresholds$recommended)), col = "red", lwd = 2)

  # Add other threshold lines
  # Add 2SD threshold
  if (!is.null(thresholds$basic$threshold_2sd)) {
    pos <- which(df$count >= thresholds$basic$threshold_2sd)[1]
    if (!is.na(pos)) {
      abline(v = pos, col = "blue", lty = 2)
    }
  }

  # Add breakpoint threshold
  if (!is.null(thresholds$breakpoints$suggested)) {
    pos <- which(df$count >= thresholds$breakpoints$suggested)[1]
    if (!is.na(pos)) {
      abline(v = pos, col = "green", lty = 2)
    }
  }

  # Add entropy threshold
  if (!is.null(thresholds$entropy$threshold)) {
    pos <- which(df$count >= thresholds$entropy$threshold)[1]
    if (!is.na(pos)) {
      abline(v = pos, col = "purple", lty = 2)
    }
  }

  # Add legend
  legend("topright", 
         legend = c("Recommended", "2SD", "Breakpoint", "Entropy"),
         col = c("red", "blue", "green", "purple"),
         lty = c(1, 2, 2, 2),
         lwd = c(2, 1, 1, 1))

  # Return the plot
  invisible(threshold_results)
}

rheumatoid_thresholds <- empirical_threshold(counts, freqs, n_zeros)
names(rheumatoid_thresholds$thresholds)
#[1] "basic"                  "breakpoints"            "entropy"                "exponential"            "rarity"                 "zero_inflation"         "extreme_zero_inflation"
#[8] "recommended"           
rheumatoid_thresholds$thresholds$basic$p95$threshold_2sd
rheumatoid_thresholds$thresholds$basic$threshold_2sd
rheumatoid_thresholds$thresholds$breakpoints$suggested
rheumatoid_thresholds$thresholds$entropy
rheumatoid_thresholds$thresholds$exponential$threshold
rheumatoid_thresholds$thresholds$rarity$threshold
rheumatoid_thresholds$thresholds$zero_inflation
rheumatoid_thresholds$thresholds$extreme_zero_inflation$threshold
rheumatoid_thresholds$thresholds$recommended

plot_thresholds(rheumatoid_thresholds)
print(rheumatoid_thresholds$thresholds$recommended)

```

# plink preprocessing

```{R, echo=FALSE, eval=FALSE}
maps <- read.table("Cleaned_CaseControl_RAauto_pruned200kb_.3.bim",header=FALSE)
#.bim (PLINK extended MAP file)
#Extended variant information file accompanying a .bed binary genotype table. (--make-just-bim can be used to update just this file.)
#A text file with no header line, and one line per variant with the following six fields:
#    Chromosome code (either an integer, or 'X'/'Y'/'XY'/'MT'; '0' indicates unknown) or name
#    Variant identifier
#    Position in morgans or centimorgans (safe to use dummy value of '0')
#    Base-pair coordinate (1-based; limited to 231-2)
#    Allele 1 (corresponding to clear bits in .bed; usually minor)
#    Allele 2 (corresponding to set bits in .bed; usually major)

## .fam (PLINK sample information file)
## Sample information file accompanying a .bed binary genotype table. (--make-just-fam can be used to update just this file.) Also generated by "--recode lgen" and "--recode rlist".
## A text file with no header line, and one line per sample with the following six fields:
##     Family ID ('FID')
##     Within-family ID ('IID'; cannot be '0')
##     Within-family ID of father ('0' if father isn't in dataset)
##     Within-family ID of mother ('0' if mother isn't in dataset)
##     Sex code ('1' = male, '2' = female, '0' = unknown)
##     Phenotype value ('1' = control, '2' = case, '-9'/'0'/non-numeric = missing data if case/control)
## With the use of additional loading flag(s), PLINK can also correctly interpret some .fam files missing one or more of these fields.

maps$chrNpos <- paste0(maps$V1,"_",maps$V4)
lm_results$SNP1ID <- ""
lm_results$SNP2ID <- ""

for(s in 1:length(SNPs)){
  tmp1 <- strsplit(SNPs[s]," ")[[1]][1]
  tmp2 <- strsplit(SNPs[s]," ")[[1]][2]

  index <- which(maps$chrNpos==tmp1)
  lm_results$SNP1ID[s] <- as.character(maps[index,2])
  index <- which(maps$chrNpos==tmp2)
  lm_results$SNP2ID[s] <- as.character(maps[index,2])

}

index <- which(lm_results$adj_p_value<0.05)
allSNPs <- union(lm_results$SNP1ID,lm_results$SNP2ID)

for(p in 1:length(index)){
  plinkCommand <- paste0("plink --bfile Cleaned_CaseControl_RAauto_pruned200kb_.3 --ld   ",lm_results$SNP1ID[p]," " ,lm_results$SNP2ID[p]," --out  ",lm_results$SNP1ID[p],"_" ,lm_results$SNP2ID[p])
  system(plinkCommand)
}

## PLINK v1.90p 64-bit (30 Nov 2019)              www.cog-genomics.org/plink/1.9/
## (C) 2005-2019 Shaun Purcell, Christopher Chang   GNU General Public License v3
## Logging to rs3134926_rs9268831.log.
## Options in effect:
##   --bfile Cleaned_CaseControl_RAauto_pruned200kb_.3
##   --ld rs3134926 rs9268831
##   --out rs3134926_rs9268831

## 515398 MB RAM detected; reserving 257699 MB for main workspace.
## Error: Failed to open Cleaned_CaseControl_RAauto_pruned200kb_.3.bed.
## PLINK v1.90p 64-bit (30 Nov 2019)              www.cog-genomics.org/plink/1.9/
## (C) 2005-2019 Shaun Purcell, Christopher Chang   GNU General Public License v3
## Logging to _.log.
## Options in effect:
##   --bfile Cleaned_CaseControl_RAauto_pruned200kb_.3
##   --ld
##   --out _

## Error: --ld requires at least 2 parameters.
## For more information, try "plink --help <flag name>" or "plink --help | more".

lm_results$LD <- ""
lm_results$Rsq <- ""
lm_results$Dprime <- ""
library(reader)
for(r in 1:length(index)){
  file <- paste0(lm_results$SNP1ID[r],"_",lm_results$SNP2ID[r],".log")
  lds <- n.readLines(file,n=24, skip = 23)[1]
  tmpR <- strsplit(lds,"D'")[[1]][1]
  tmpD <- strsplit(lds,"D'")[[1]][2]
  lm_results$Rsq[r] <- as.numeric(strsplit(tmpR,"R-sq = ")[[1]][2])
  lm_results$Dprime[r] <- as.numeric(strsplit(tmpD," <- ")[[1]][2])
  lm_results$LD[r]=lds
}

write.csv(lm_results[1:length(index),],paste0(dirOut,"/interacting_SNPs.csv"))

# counts=countMatrix#readRDS(paste0(dirOut,"/countMatrix_Nit",t,"_",".rds"))
# rm(countMatrix)
# gc()
# 
# counts[lower.tri(counts, diag = T)] =0# counts[upper.tri(counts)]
# index = which(counts>0)
# rowIndex = index%%ncol(counts)
# colIndex = floor(index/ncol(counts))+1
# length(index)
# 
# interactions = data.frame(SNP1=colnames(counts)[rowIndex],SNP2=colnames(counts)[colIndex],Index=index,counts=counts[index])
# interactions = interactions[order(interactions$counts,decreasing = T),]
# write.csv(interactions,paste0(dirOut,"/interactions",t,".csv"),row.names = F)
# 
# ######################## Interaction modelling and code here
# #interactions=read.csv(paste0(dirOut,"/interactions",t,".csv"))
# index=which(interactions$counts>=2)
# 
# lm_results = data.frame(interactions[index,],SNP1_effect=0,SNP1_Pvalue=1,SNP2_effect=0,
#                         SNP2_Pvalue=1,int_effect=0,int_Pvalue=1)
# dim(lm_results)
# 
# for(l in 1:nrow(lm_results)){
#   ind1 = which(colnames(dt)==lm_results$SNP1[l])
#   ind2 = which(colnames(dt)==lm_results$SNP2[l])
#   s = summary(lm(dt$pheno ~ dt[,ind1]*dt[,ind2]), data= dt)$coeff
#   if(nrow(s)==4){
#     lm_results$SNP1_effect[l] = s[2,1]
#     lm_results$SNP1_Pvalue[l] = s[2,4]
#     lm_results$SNP2_effect[l] = s[3,1]
#     lm_results$SNP2_Pvalue[l] = s[3,4]
#     lm_results$int_effect[l] = s[4,1]
#     lm_results$int_Pvalue[l] = s[4,4]
#   }
# }
# 
# 
# 
# lm_results = lm_results[order(lm_results$int_Pvalue,decreasing = F),]
# lm_results$adj_p = lm_results$int_Pvalue*(ncol(dt)-2)*(ncol(dt)-3)*.5
# # lm_results$adj_int_Pvalue = lm_results$int_Pvalue*(0.5*ncol(geno)*(ncol(geno)-1))
# write.csv(lm_results,paste0(dirOut,"/lm_results",t,".csv"),row.names = F)

```

### 2025-05-02 manuscript.tex
added Supplementary_File_Rheumatoid_Data.Rmd, other changes to manuscript.tex

@@
 %% Use the option review to obtain double line spacing
-\documentclass[authoryear,preprint,12pt,3p]{elsarticle}
+\documentclass[preprint,12pt,3p]{elsarticle}
+%%if you are using elsarticle.cls, do not use it with options
+%%    \documentclass[authoryear]{elsarticle}
+%% and the style
+%% \bibliographystyle{elsarticle-num}
+%% https://tex.stackexchange.com/questions/54480/package-natbib-error-bibliography-not-compatible-with-author-year-citations
 
@@
 \usepackage[figuresleft]{rotating}
-%\usepackage{rotating}
 
@@
 \usepackage{amsmath}
-
-% \newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
-% \newcommand{\DrawBox}[4][]{%
-%     \tikz[overlay,remember picture]{%
-%         \coordinate (TopLeft)     at ($(#2)+(-0.2em,0.9em)$);
-%         \coordinate (BottomRight) at ($(#3)+(0.2em,-0.3em)$);
-%         %
-%         \path (TopLeft); \pgfgetlastxy{\XCoord}{\IgnoreCoord};
-%         \path (BottomRight); \pgfgetlastxy{\IgnoreCoord}{\YCoord};
-%         \coordinate (LabelPoint) at ($(\XCoord,\YCoord)!0.5!(BottomRight)$);
-%         %
-%         \draw [red,#1] (TopLeft) rectangle (BottomRight);
-%         \node [below, #1, fill=none, fill opacity=1] at (LabelPoint) {#4};
-%     }
-% }
 
@@
 \begin{document}
-
-\tableofcontents 
-\clearpage
 
@@
 \author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
-\ead{hawlader.almamun@uwa.edu.au}
+\ead{halmamun@intergrain.com}
 %% \ead[url]{home page}
@@
 \author{Ross L. Tellam}
-\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
+\affiliation{organization={CSIRO Agriculture and Food},%Department and Organization
   addressline={306 Carmody Rd}, 
@@
 \author{Klara Verbyla}
-\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
-  addressline={2 - 40 Clunies Ross Street}, 
-  city={Acton},
-  postcode={2601}, 
-  state={ACT},
-  country={Australia}}
+\affiliation{organization={The Center for Aquaculture Technologies},%Department and Organization
+  addressline={8445 Camino Santa Fe, Suite 104}, 
+  city={San Diego},
+  postcode={92121}, 
+  state={California},
+  country={USA}}
 
@@
 
-%\usepackage{caption}
+%\usepackage{capton}
 %\usepackage{natbib}
@@
 Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forests with more
-trees and we will observe more counts.
-There is a trade-off here and as we are sampling a space with typically more than $10^{9}$ possible interactions,
-we have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
-distribution. Because of this we resort to a threshholding stratey that is both computationally easier and controls the FDR rate by testing
-each variable.
+trees and we will observe more counts.  There is a trade-off here and as we are sampling a space with typically more
+than $10^{9}$ possible interactions, we have to increase the number of trees a lot to get a meaningful difference in the
+mean of the contaminating distribution. We have found the thresholding to give a larger set of significant variables
+than the empirical Bayes method. However it controls the FDR rate by testing each variable and computationally easier.
 
@@
 \begin{itemize}
-\item threshold the observed count data. We do this a number of ways to do this (list them) and then take the median. In
+\item threshold the observed count data. We do this a number of ways to do this (breakpoints in the distribution,
+  entropy tests and otehrs. See the supplementary informaton) and then take the median. In
   the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
@@
 various autoimmune diseases i.e., rs9273363, rheumatoid arthritis; type 1 diabetes; rs6457617, multiple sclerosis,
-Graves disease, systemic sclerosis; rs615672, rheumatoid arthritis ("UCSC Genome Browser Home").
-\marginpar{is UCSC Genome Browser Home a reference?}
+Graves disease, systemic sclerosis; rs615672, rheumatoid arthritis \cite[]{UCSC.2022}.
 The results fromusing the proposed method on the human rheumatoid arthritis data revealed a number of promising results supporting the
@@
 \section{Acknowledgements}
-The authors thank the anonymous reviewers for their valuable suggestions. This work is supported by CSIRO MLAI Future
+This work is supported by CSIRO MLAI Future
 Science Platform. This study makes use of data generated by the Wellcome Trust Case Control Consortium. A full list of
@@
 
-\bibliographystyle{elsarticle-harv}
+\bibliographystyle{elsarticle-num}
 \bibliography{./manuscript.bib}

### 2025-05-03 Supplementary_Files/Supplementary_File_Rheumatoid_Data.Rmd
added Supplementary_Files

--- 2025-05-02 Supplementary_File_Rheumatoid_Data.Rmd
@@
 ---
-title: "The Rheumatoid Arthritis data  \nSupplementary File for "Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests"
+title: "The Rheumatoid Arthritis data  \nSupplementary File for  \n`Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests`"
 author: "Hawlader Al-Mamun,  Rob Dunne"
@@
 ```
-\clearpage
-# Notes
-
-- is the data available? What about the preprocessing? Can we include this in a vignette?
-- dim(dt)   4798 94485 but the paper says 94,486
-- looked at variable importance. Variables that have more splits as they are important are likely to have more consecutive splits. Should we take this into account?
-- I removed splits from AllInteractions where the split is on the same variable again. This does not matter for the regression method but might for other methods (Empirical Bayes)
-- There is a hotspot in chromosome 6. Did we know about this? Does it get dropped when you filter on LD
-- we are looking at the interaction as the product of the two SNP variables, so it takes the values $\{0,1,2,4\}$. so we are threating it as a
-  regression but it does not take the value 3.  If we treat them as factors we get levels(d1:d2) "0:0" "0:1" "0:2" "1:0" "1:1" "1:2" "2:0" "2:1" "2:2"
-  9 levels. Which is a more reasonable model? 
-- I look at using the **empiricalzip** (Empirical Bayes zero inflated Poisson) package to model the interaction counts and set a significance
-level. The agreement with the regression testing is not great.
-- we have the file "Cleaned_CaseControl_RAauto_pruned200kb_.3.bim"
-
 # load data
+
+We have
 
@@
 
-```{r, echo=FALSE, eval=FALSE}
+```{r, echo=TRUE, eval=FALSE}
 library(lmtest)
 library(ranger)
-#library(RCy3)
-
-#Noise <- 0.1
+
 numIte  <-  1
@@
 #dt <- dt[which(dt$treatment=="short"),-2]
-
-# quick load of results etc
+```
+
+```{r, echo=FALSE, eval=FALSE}
+# quick load of results etc if we have a
 dirOut  <-  paste0("Results")
@@
 ```{r, echo=FALSE, eval=FALSE}
-##Number of trees:                  1000 
-#set.seed(sample(1:1000,1))
 set.seed(100)
-#system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
-#     user    system   elapsed 
-#89141.264     3.601  1568.711 
-#save(model.rf, file="model.rf.Rdata")
-
-##Number of trees:                  5000 
-#system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = 5000, mtry = ncol(dt)/3))
-# user     system    elapsed 
-#443201.110      8.132   7185.540 
-#Number of trees:                  5000 
-#Sample size:                      4798 
-#Number of independent variables:  94484 
-#Mtry:                             31495 
-#Target node size:                 1 
-#Variable importance mode:         none 
-#Splitrule:                        gini 
-#OOB prediction error:             33.18 % 
-#save(model.rf, file="model.rf_temp2.Rdata")
-
-##Number of trees:   1000, importance = 'impurity'
-set.seed(100)
-#petrichor-i1: salloc -A OD-217714 --mem=100GB --nodes=1 --ntasks-per-node=64    -J interactive -t 6:00:00 srun --pty /bin/bash -l
-#etrichor-i1: sacct -j 60755241  --format="jobid,ntasks,Elapsed,TotalCPU,MaxRSS,AveVMSize,MaxVMSize"
-#JobID          NTasks    Elapsed   TotalCPU     MaxRSS  AveVMSize  MaxVMSize 
-#60755241                00:13:56  00:06.682                                  
-#60755241.ex+        1   00:13:56  00:00.001       995K    124080K    124080K 
-#60755241.0         64   00:13:53  00:06.680 104545713K 3274946112 192422568K
-
 system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
@@
 
-```{r RFlocalfdr, eval=TRUE, echo=FALSE, fig.cap='RFlocalfdr',  fig.align = 'center', out.width = '50%'}
-knitr::include_graphics("./script3/RFlocalfdr.png")
-
-```
-
-```{R, echo=FALSE, eval=FALSE}
+```{r RFlocalfdr, eval=TRUE, echo=FALSE, fig.cap='RFlocalfdr analysis',  fig.align = 'center', out.width = '50%'}
+knitr::include_graphics("./supp_figs/RFlocalfdr.png")
+
+```
+
+```{R, echo=TRUE, eval=FALSE}
 library(RFlocalfdr)
 vignette(package="RFlocalfdr")
-#no vignettes found -- why is this?
 imp<-log(var_importance)  #6275 var_importance vales =0 
@@
 res.con2 <- determine_cutoff(imp,t2 ,cutoff=cutoffs,plot=c(4,5,6))
-plot(c(0,1,2,3,4,5,6),c(res.con[,3],res.con2[,3]),pch=15,col="red",cex=1.5,ylab="max(abs(y - t1))")
+plot(c(0,1,2,3,4,5,6),c(res.con[,3],res.con2[,3]),pch=15,col="red",cex=1.5,
+     ylab="max(abs(y - t1))")
 
@@
 
-png("./script3/RFlocalfdr.png")
+png("./supp_figs/RFlocalfdr.png")
 aa<-significant.genes(ppp,temp,cutoff=0.05,debug.flag=0,do.plot=TRUE,use_95_q=TRUE)
@@
 ```{r hotspot, eval=TRUE, echo=FALSE, fig.cap='There is a hotspot in chromosome 6',  fig.align = 'center', out.width = '50%'}
-knitr::include_graphics("./figs/hotspot.png")
+knitr::include_graphics("./supp_figs/hotspot.png")
 
@@
 
-# count the interactions in actual data
-
-```{R, echo=FALSE,eval=FALSE}
-#t2 <- system.time({
-
-  # countMatrix <- matrix(0,ncol = ncol(dt)-1,nrow=ncol(dt)-1)
-  # row.names(countMatrix) <- colnames(dt)[-1]
-  # colnames(countMatrix) <- colnames(dt)[-1]
-  #
-
+# count the interactions in the data
+
+```{R, echo=TRUE, eval=FALSE}
 for(i in 1:ntree){
@@
 
-#saveRDS(All_Interactions,paste0(dirOut,"/All_Interactions.rds"))
-#All_Interactions <- readRDS(paste0(dirOut,"/All_Interactions.rds"))
 #save(All_Interactions,file=paste0(dirOut,"/All_Interactions.Rdata"))
@@
 ```
+
 ```{R, echo=FALSE, eval=FALSE}
@@
   - save **interaction_snps_counts_fixed.Rdata**
-
-b
 
@@
 
-#saveRDS(interaction_snps_counts,paste0(dirOut,"/interaction_snps_counts_fixed.rds"))
-#interaction_snps_counts <- readRDS("./Results/interaction_snps_counts_fixed.rds")
 #save(interaction_snps_counts, file="./Results/interaction_snps_counts_fixed.Rdata")
 #load(file="./Results/interaction_snps_counts_fixed.Rdata")
+
+```
+
+# Null distribution
+
+## data set up
+
+```{R, echo=FALSE, eval=FALSE}
+
+# Assuming your count data is in a vector 'x'
+x <- interaction_snps_counts$Freq
+
+dim(interaction_snps_counts)[1] #[1] 291142
+sum(interaction_snps_counts$Freq) # 294120
+
+94485*(94485 -1)/2 - 291142
+x2<-  c(rep(0,4463369228),x)
+
+#table(x2) long run time!
+#    0         1         2      3      4       5      6      7      8      9     10     11     12     13     14     15     16     17     18     22     23     25     26     27     31     34     41  51     52     54     62     63     64     78     79     90    100    113    144    466
+#4463369228  290502    418     89     29     22     14     11      7      5      4      3      3      3      4     1      1      1      1      2      1      1      1      2      1      1      1   1      1      1      1      2      1      1      1      1      1      1      1      1
+
+# First, prepare your data
+counts <- as.numeric(names(table(x)))  # Unique count values (1, 2, 3, ...)
+freqs <- as.numeric(table(x))          # Frequencies of each count
+n_zeros <- 4463369228                   # Number of zeros
+
+# Recreate the data based on your table x2
+counts0 <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14,15,  16,  17,  18,  22,  23,  25,  26,  27,  31,  34,  41, 51,  52,  54,  62,  63,  64,  78,  79,  90, 100, 113, 144, 466) 
+freqs0 <- c(4463369228, 290502, 418, 89, 29, 22, 14, 11, 7, 5, 4, 3, 3, 3, 4, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1, 1, 1, 1,   1,   1,   1,   2,   1,   1,   1,   1,   1,   1,   1,   1)
+# Recreate the data based on your table x
+counts <- c( 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14,15,  16,  17,  18,  22,  23,  25,  26,  27,  31,  34,  41, 51,  52,  54,  62,  63,  64,  78,  79,  90, 100, 113, 144, 466) 
+freqs <- c( 290502, 418, 89, 29, 22, 14, 11, 7, 5, 4, 3, 3, 3, 4, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1, 1, 1, 1,   1,   1,   1,   2,   1,   1,   1,   1,   1,   1,   1,   1)
+
+# Create a representative sample of this distribution (for computational efficiency)
+set.seed(123)
+sample_size <- 100000000
+sample_probs <- freqs / sum(freqs)
+x_sample <- sample(counts, size = sample_size, replace = TRUE, prob = sample_probs)
+
+```
+
+## empiricalzip
+
+- There are  4463613128 potential interactions. We have non-zero values for 294120, so there are a lot of 0 values.
+- what to do to handle this? 
+- **gpEMAlgorithm** fails -- set which estimates to calculate in the function call
+- better name than **RealDataFDR**, what other data would we be running it on
+- docuement output better
+- "X6_32222493 X6_32771829" see below, how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113
+
+   - it doesnt t2["X6_32222493"] =202 t2["X6_32771829"]= 1532 -- lm_result has a problem
+
+## some investigations of empiricalzip. 
+```{R, echo=FALSE, eval=FALSE}
+#devtools:::install_github("parsifal9/empiricalzip")
+estimate<-  zigpEMAlgorithm(x2) #works
+estimate<- zipEMAlgorithm(x2) #works
+system.time(estimate<- pEMAlgorithm(x2)) ##works
+estimate<- gpEMAlgorithm(x2) #Error in while (((abs(hold.lambda - lambdaEstimate)) > epsilon) || ((abs(hold.theta -  :   missing value where TRUE/FALSE needed
+
+aa <- my.RealDataFDR(x2, 0.1) #has gpEMAlgorithm commented out, accepts data ratehr than a file
+# Rejection counts
+#       ZIGP ZIP   GP  P         ZIGP ZIP   GP  P        ZIGP  ZIP     GP  P
+#   one stage procedure           two stage             Storey's FDR
+#[1,]   50   82   82   82      57   82  104   82        35    53    53    53
+#[2,]   50   82   82   82      57   82  104   82        35    53    53    53
+
+#[[2]]Parameter estimates using cutoff method C_0
+
+    ## eta: Zero-inflation parameter (proportion of structural zeros)
+    ## lambda: Mean parameter
+    ## theta: Dispersion parameter (NA for models that don't use it)
+    ## pi: Estimated proportion of null hypotheses ($\pi_0$)
+    ## C: The estimated cutoff value (\hat{C}) - positions with counts $\leq$ C are assumed to be from null
+    ## AD: The truncation point (D_N) - positions with counts $\geq$ AD are automatically considered significant
+
+#        eta         lambda    theta      pi          C0   AD
+#[1,] 0.472686094 1.0020464 1.133941e-12 0.9999401    7    8   zigp
+#[2,] 0.001277773 0.3350806           NA 0.9999075    5   14   zip
+#[3,]          NA 0.3346515 0.000000e+00 0.9999072    4    5   gp
+#[4,]          NA 0.3346515           NA 0.9999072    4   14    p
+#
+#[[3]]
+#           [,1]      [,2]         [,3]      [,4] [,5] [,6]
+#[1,] 0.47268609 1.0020464 1.133941e-12 0.9999401    7    8
+#[2,] 0.00128352 0.3350822           NA 0.9999074    4   14
+#[3,]         NA 0.3346515 0.000000e+00 0.9999072    4    5  
+#[4,]         NA 0.3346515           NA 0.9999072    4   14
+
+#empiricalzip  ZIGP one stage procedure
+#empiricalzip pickes the interactions on the Freq, sointeraction_snps_counts[1:50,]
+interaction_snps_counts$interaction_snps[1:50]
+
+sum(freqs[-c(1:7)]) #[1] 68
+sum(freqs[-c(1:8)]) #[1] 57
+sum(freqs[-c(1:9)]) #[1] 50
+
+```
+The ZIGP model selected by epiricalzip has
+
+- $\eta= 0.472686094$      Zero-inflation parameter (proportion of structural zeros)   
+- $\lambda= 1.0020464$    Mean parameter
+- $\theta= 1.133941e-12$    Dispersion parameter (NA for models that don't use it)  
+- $\pi= 0.9999401$       Estimated proportion of null hypotheses
+
+See table \@ref(tab:zigp_comparison)
+
+\begin{table}
+\caption{\label{tab:zigp_comparison} Observed vs. Expected Counts Under ZIGP Model selected by mpiricalzip}
+\center
+\begin{tabular}{lrr}
+\hline
+\textbf{Count} & \textbf{Observed} & \textbf{Expected (ZIGP)} \\
+\hline
+0 & 4,463,369,228 & 4,463,369,228 \\
+1 & 290,502 & 290,490 \\
+2 & 418 & 427 \\
+3 & 89 & 85 \\
+4 & 29 & 21 \\
+5 & 22 & 5 \\
+6-10 & 41 & 3 \\
+11-20 & 13 & $<$1 \\
+$>$20 & 22 & $\approx$0 \\
+\hline
+\textbf{Total} & 4,463,660,370 & 4,463,660,370 \\
+\hline
+\end{tabular}
+\label{tab:zigp_comparison}
+\end{table}
+
+```{R, echo=FALSE, eval=FALSE}
+
+# Create a named vector with the observed counts
+observed_data <- freqs
+names(observed_data) <- as.character(counts)
+
+# ZIGP parameters
+eta <- 0.9993963  
+lambda <- 1.0020464
+theta <- 1.133941e-12
+
+# Generate the plots and analysis
+results <- plot_zigp_comparison(
+  observed_counts = observed_data,
+  eta = eta,
+  lambda = lambda, 
+  theta = theta,
+  max_display = 30,
+  y_min = -15  # Extended y-axis
+)
+
+# Display the plots
+library(gridExtra)
+grid.arrange(results$log_scale, results$linear_scale, ncol = 1)
+
+# Testing separation between null and alternative distributions
+# First, examine where the residuals cross from positive to negative
+crossover_point <- min(which(results$separation_analysis$Residual < 0))
+print(paste("Crossover point where observed becomes less than expected:", 
+            results$separation_analysis$Count[crossover_point]))
+
+# Calculate patterns of overdispersion
+counts_with_excess <- sum(results$separation_analysis$Observed > results$separation_analysis$Fitted)
+print(paste("Number of count values with excess observations:", counts_with_excess))
+
+# Calculate the count where maximum deviation occurs
+max_positive_deviation <- which.max(results$separation_analysis$RelativeDiff)
+print(paste("Count with maximum positive deviation:", 
+            results$separation_analysis$Count[max_positive_deviation]))
+
+# Save the improved plot
+#ggsave("./supp_figs/zigp_log_scale_extended.png", results$log_scale, width = 10, height = 6)
+#ggsave("zigp_ratio_plot.png", results$ratio_plot, width = 10, height = 6)
+
+```
+
+```{r zigplogscaleextended, eval=TRUE, echo=FALSE, fig.cap='empiricalzip fitted ZIPG',  fig.align = 'center', out.width = '50%'}
+knitr::include_graphics("./supp_figs/zigp_log_scale_extended.png")
+
+```
+
+## Local FDR Calculation
+The local FDR (false discovery rate) at a particular count value is calculated as:
+
+- fdr(t) = ($\pi_0 \times f_0(t)) / f(t)$
+
+Where:
+
+-  $t4 is the mutation count
+-  $\pi_0$ is the estimated proportion of nulls
+-  $ f_0(t)$ is the null distribution probability at count $t$
+-  $f(t)$ is the observed mixture distribution probability at count $t$
+
+In the my.RealDataFDR function, this calculation happens here:
+
+```{verbatim}
+LFDR <- pi[model, CO[method, model]] * gpdMixture(x = sort(x), 
+    estimate[model, 1, CO[method, model]], estimate[model, 
+    2, CO[method, model]], estimate[model, 3, CO[method, 
+    model]]) * f12(sort(x), length(x))
+```
+Breaking this down:
+
+-  pi[model, CO[method, model]] is the estimated $\pi_0$  (proportion of nulls)
+-  gpdMixture() calculates  $ f_0(t)$ , the probability under the null distribution (ZIGP, ZIP, GP, or P) using the estimated parameters
+-  f12(sort(x), length(x)) is a correction factor related to f(t). Looking at the f12 function in the RcppExports.R, this factor is calculated as:
+
+```{verbatim}
+   M / length(x[which(x==x[i])])
+```
+where M is the total number of positions. This is effectively equivalent to $1/f(t)$, where $f(t)$ is the empirical probability of observing count $t$.
+
+For the one-stage procedure, positions are declared significant if:
+
+```{verbatim}
+sum(LFDR[!is.na(LFDR)] < FDRLevel)
+```
+meaning their local FDR is below the specified threshold (in your case, 0.1).
+
+Why Some Positions with Counts >7 Aren't Called Significant
+
+With the ZIGP model estimated parameters for your data, the local FDR for positions with counts of 8, 9, etc. depends on:
+
+-   How rare such counts are in the data (affecting $f(t)$)
+-  How probable they are under the null distribution (affecting $f_0(t)$)
+
+When a position has a count that is high but not extremely high relative to the estimated null distribution, its local
+FDR might still exceed the threshold, especially when $\pi_0$ is estimated to be very close to 1.0.
+
+This is why 7 positions with counts >7 aren't called significant by the one-stage procedure - their local FDR values
+exceed your threshold of 0.1, indicating there's still a substantial probability they come from the null distribution
+despite their relatively high counts.
+
+The two-stage procedure, by contrast, automatically declares positions with counts $\geq$ AD (in this case 8) as significant,
+bypassing the local FDR calculation for these high counts.
+
+1. Stage 1 (Screening Step): We reject the $H_0 j \textrm{if } j \leq D_N(\hat{C})$ and do not reject if $j \leq \hat{C}$.
+2. Stage 2 (Testing step): For $\hat{C} + 1 \leq j < D_N(\hat{C})$ , we calculate the local FDR
+
+## Simplified two-stage procedure for empiricalzip
+
+-  "Estimated C: 1"
+-  "D_N threshold: 3"
+-  "Significant counts:"   3  4  5  7 27
+- used x_sample so some counts missing
+- could use this value
+
+```{R, echo=FALSE, eval=FALSE}
+# Simplified two-stage procedure for extreme zero-inflation
+simplified_two_stage <- function(data, fixed_C = 1, alpha = 0.05) {
+  # Extract the counts and their frequencies
+  tab <- table(data)
+  counts <- as.numeric(names(tab))
+  freqs <- as.numeric(tab)
+  N <- sum(freqs)
+
+  # Given extreme zero-inflation, set C directly
+  C_hat <- fixed_C
+
+  # Estimate parameters for the null distribution (ZIGP)
+  # Given the extreme zeros, simplify parameter estimation
+  eta_est <- freqs[counts == 0] / N  # Zero-inflation parameter
+
+  # For non-zero values, estimate lambda and theta
+  non_zero_counts <- rep(counts[counts > 0], freqs[counts > 0])
+  lambda_est <- mean(non_zero_counts)
+  theta_est <- min(0.5, var(non_zero_counts) / lambda_est - 1)
+
+  # Calculate D_N (direct approach for computational stability)
+  # With extreme zeros, a lower D_N is appropriate
+  D_N <- max(3, C_hat + 1)  # Simple heuristic
+
+  # Create results structure
+  results <- list(
+    C_hat = C_hat,
+    D_N = D_N,
+    parameters = c(eta = eta_est, lambda = lambda_est, theta = theta_est),
+    significant_counts = counts[counts >= D_N]
+  )
+
+  return(results)
+}
+
+# Load necessary libraries
+library(VGAM)
+library(stats)
+
+# Function to implement the two-stage procedure from Gauran et al.
+two_stage_zigp_procedure <- function(data, alpha = 0.05) {
+  # Extract the counts and their frequencies
+  counts <- as.numeric(names(table(data)))
+  freqs <- as.numeric(table(data))
+  N <- sum(freqs)
+
+  # Calculate the empirical probabilities
+  f_hat <- freqs / N
+
+  # Define function to estimate C using the likelihood approach (C1 method)
+  estimate_C1 <- function(counts, freqs, N) {
+    # Initialize variables
+    max_S <- -Inf
+    best_C <- 1
+
+    # Try different values of C
+    for (v in 1:(max(counts)-1)) {
+      # Subset data for estimation
+      subset_counts <- counts[counts <= v]
+      subset_freqs <- freqs[counts <= v ]
+
+      # Fit ZIGP to data up to C using MLE or method of moments
+      # This is a simplified approach - in practice, use EM algorithm
+      p_zeros <- subset_freqs[subset_counts == 0] / sum(subset_freqs)
+      lambda_est <- mean(rep(subset_counts, subset_freqs))
+
+      # Due to extreme zero-inflation, we'll use high eta
+      eta_est <- max(0.95, p_zeros - (1-p_zeros)*exp(-lambda_est))
+      theta_est <- min(0.5, var(rep(subset_counts, subset_freqs)) / lambda_est - 1)
+
+      # Calculate pi_0 estimate
+      pi0_est <- min(1, sum(freqs[counts <= v]) / (N * sum(zigp_pmf(0:v, eta_est, lambda_est, theta_est))))
+
+      # Calculate S_v statistic
+      S_v <- sum(freqs[counts <= v] * log(zigp_pmf(counts[counts <= v], eta_est, lambda_est, theta_est) / f_hat[counts <= v]))
+
+      # Update if this is the best C so far
+      if (S_v > max_S) {
+        max_S <- S_v
+        best_C <- v
+        best_params <- c(eta_est, lambda_est, theta_est, pi0_est)
+      }
+    }
+
+    return(list(C = best_C, params = best_params))
+  }
+
+  # Define ZIGP PMF function (simplified)
+  zigp_pmf <- function(x, eta, lambda, theta) {
+    result <- numeric(length(x))
+
+    # Handle zeros separately due to zero-inflation
+    result[x == 0] <- eta + (1 - eta) * exp(-lambda)
+
+    # Handle positive values
+    for (i in which(x > 0)) {
+      val <- x[i]
+      prob <- (1 - eta) * (lambda * (lambda + theta * val)^(val - 1) * exp(-lambda - theta * val)) / factorial(val)
+      result[i] <- prob
+    }
+
+    return(result)
+  }
+
+  # Calculate D_N (truncation point)
+  calculate_D_N <- function(C, params) {
+    eta <- params[1]
+    lambda <- params[2]
+    theta <- params[3]
+
+    # Calculate D_N based on the paper's formula
+    # Simplified version for extreme zero-inflation
+    D1 <- ceiling(log(N) / (-log(theta)))
+    D2 <- ceiling(C + 1)
+
+    return(max(D1, D2))
+  }
+
+  # Stage 1: Estimate C and calculate D_N
+  C_result <- estimate_C1(counts, freqs, N)
+  C_hat <- C_result$C
+  params <- C_result$params
+  D_N <- calculate_D_N(C_hat, params)
+
+  # Stage 2: Apply local FDR to values between C+1 and D_N-1
+  local_fdr_result <- numeric(max(counts))
+
+  for (j in (C_hat+1):(D_N-1)) {
+    if (j %in% counts) {
+      # Calculate local FDR
+      null_density <- zigp_pmf(j, params[1], params[2], params[3])
+      mixture_density <- f_hat[counts == j]
+      local_fdr_result[j] <- params[4] * null_density / mixture_density
+    }
+  }
+
+  # Determine significant counts
+  significant_counts <- numeric(0)
+
+  # Reject all counts \geq D_N
+  if (any(counts >= D_N)) {
+    significant_counts <- c(significant_counts, counts[counts >= D_N])
+  }
+
+  # Reject counts between C+1 and D_N-1 if local FDR \leq alpha
+  for (j in (C_hat+1):(D_N-1)) {
+    if (j %in% counts && local_fdr_result[j] <= alpha) {
+      significant_counts <- c(significant_counts, j)
+    }
+  }
+
+  return(list(
+    C_hat = C_hat,
+    D_N = D_N,
+    parameters = params,
+    local_fdr = local_fdr_result,
+    significant_counts = significant_counts
+  ))
+} #end two_stage_zigp_procedure
+
+## Apply the simplified two-stage procedure
+results <- simplified_two_stage(x_sample, fixed_C = 1)
+
+# Display results
+print(paste("Estimated C:", results$C_hat))
+print(paste("D_N threshold:", results$D_N))
+print("Significant counts:")
+print(results$significant_counts)
+
+```
+
+# what is this 
+```{R, echo=FALSE,eval=FALSE}
+
+length(intersect(interaction_snps_counts$interaction_snps[1:50],lm_results$interaction_snps[1:88])) #22
+# lm_results (has been ordered by lm_results$int_Pvalue)
+
+t1 <- setdiff(interaction_snps_counts$interaction_snps[1:50],lm_results$interaction_snps[1:88])
+match(t1,interaction_snps_counts$interaction_snps)  #3  4  5  7
+
+t1[1]
+#1] "X6_32222493 X6_32771829"
+interaction_snps_counts[3,]
+#              interaction_snps Freq
+#280884 X6_32222493 X6_32771829  113
+
+lm_results[grep("X6_32771829", lm_results$interaction_snps),] #301
+lm_results[grep("X6_32222493 X6_32771829", lm_results$interaction_snps),] 
+#              interaction_snps Freq SNP1_effect  SNP1_Pvalue SNP2_effect  SNP2_Pvalue int_effect int_Pvalue        Ch2 correlations splits_SNP1 splits_SNP2
+#280884 X6_32222493 X6_32771829  113   0.3937617 6.972158e-09  -0.6516203 2.576182e-27 -0.1303979 0.08982739 0.03193039   -0.3742329          18          14
+#       adj_p_value_bonferrroni adj_p_value_fdr adj_p_value_Ch2
+#280884                57.48953       0.2521143       0.1011656
+
+# so this is not significant for any of the tests except empiricalzip
+# how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113
+
+chisq.test(table( dt[,"X6_32222493",], dt[,"X6_32771829",]))
+
+aa<-rep(NA, 74)
+for ( i in 1:74){
+    aa[i] <- length(grep(imp[i],interaction_snps_counts$interaction_snps[1:50]))
+    }
+aa
+# [1] 21 13  8  7  5  5  5  8  5  3  2  2  2  1  2  2  1  1  0  0  1  0  0  0  0  0  1  0  0  0  1  0  0  1  1  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  1  0  0
+#[56]  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0
+
+index <- match(selectedSNPs,colnames(dt))
+index <- index[order(index)]
+
+data4interactions <- dt[,index]
+saveRDS(data4interactions,paste0(dirOut,"/data4interactions.rds"))
+
+SNPs <- lm_results$interaction_snps
+
+SNPs <- as.character(SNPs)
+SNPs <- gsub("X","" ,SNPs)
+#SNPs <- read.table("Results/SNPs.txt",header=T)
+
+```
+
+4. what is the intesection of imp and interacting transcripts
+
+   - see table at end of section
+
+- "X6_32222493 X6_32771829" see below, how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113
+
+   - it doesnt t2["X6_32222493"] =202 t2["X6_32771829"]= 1532 -- lm_result has a problem
+
+# compare some distributions
+
+- modified compare_distributions function and related plotting functions to work with a frequency table input rather than requiring the full vector. 
+This will be much more memory-efficient when dealing with large datasets that have many zeros.
+
+```{R, echo=FALSE, eval=FALSE}
+# Then run the comparison
+results <- compare_distributions_from_freq(counts, freqs, n_zeros)
+
+# View results
+print(results$comparison)
+#    Model LogLikelihood Parameters     AIC     BIC AIC_Ratio
+#2 ZI-Zipf      -3102980          2 6205964 6206005         1
+#3    ZINB          -Inf          3     Inf     Inf       Inf
+#1    ZIGP           NaN          3     NaN     NaN       NaN
+
+print(paste("Best model:", results$best_model))
+
+#png("./supp_figs/ZI_Zipf_plot.png")
+results$plots$"ZI_Zipf"
+#dev.off()
+#png("./supp_figs/ZIGP_plot.png")
+results$plots$"ZIGP"
+#dev.off()
+#png("./supp_figs/ZINB_plot.png")
+results$plots$"ZINB"
+#dev.off()
+
+results$parameters
+#$ZIGP $ZIGP$eta [1] 0.9999 $ZIGP$lambda [1]  0.9752822  $ZIGP$theta  [1]  0.03459262
+#$ZI_Zipf $ZI_Zipf$p0 [1] 0.9999348 $ZI_Zipf$alpha [1] 8.288837 
+#$ZINB $ZINB$p0 [1] 0.9999348 $ZINB$pi_0  [1] 0.9998953 $ZINB$size  [1] 13.84857 $ZINB$mu [1] 1.010229
+
+```
+
+```{R, zipfplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Power Law (Zipf) Distribution", out.width = '50%',fig.align = 'center'}
+knitr::include_graphics("./supp_figs/ZI_Zipf_plot.png")
+```
+
+```{R, zigpplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Generalized Poisson", out.width = '50%',fig.align = 'center'}
+knitr::include_graphics("./supp_figs/ZIGP_plot.png")
+```
+
+```{R, zinbplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Negative Binomial", out.width = '50%',fig.align = 'center'}
+knitr::include_graphics("./supp_figs/ZINB_plot.png")
+```
+
+```{R, echo=FALSE, eval=FALSE}
+# Modified functions to work with frequency data instead of raw vectors
+
+# ZIGP probability mass function
+zigp_pmf <- function(x, eta, lambda, theta) {
+  result <- numeric(length(x))
+
+  # Handle zeros
+  result[x == 0] <- eta + (1 - eta) * exp(-lambda)
+
+  # Handle positive values
+  for (i in which(x > 0)) {
+    val <- x[i]
+    prob <- (1 - eta) * lambda * (lambda + theta * val)^(val - 1) * exp(-lambda - theta * val) / factorial(val)
+    result[i] <- prob
+  }
+
+  return(result)
+}
+
+# Modifications to handle extreme zero-inflation
+# For the ZIGP model:
+estimate_zigp_params_from_freq <- function(counts, freqs, n_zeros) {
+  # Calculate basic statistics
+  n <- sum(freqs) + n_zeros
+  p_zeros <- n_zeros / n
+
+  # If extreme zero-inflation
+  if (p_zeros > 0.9999) {
+    # Use a simpler model with high but not exactly 1 eta
+    return(list(eta = 0.9999, lambda = mean(counts[counts > 0]), theta = 0))
+  }
+
+    # Rest of the function as before...
+  # Calculate statistics for positive values
+  if (sum(freqs[counts > 0]) == 0) {
+    return(list(eta = 1, lambda = 1, theta = 0))  # All zeros case
+  }
+
+  # Calculate mean and variance of positive values
+  # Weighted calculations using frequencies
+  mean_pos <- sum(counts[counts > 0] * freqs[counts > 0]) / sum(freqs[counts > 0])
+
+  # Calculate variance
+  var_pos <- sum(freqs[counts > 0] * (counts[counts > 0] - mean_pos)^2) / sum(freqs[counts > 0])
+
+  # Estimate parameters using method of moments
+  # For zero-inflated data, first estimate eta
+  eta_est <- max(0, (p_zeros - exp(-mean_pos)) / (1 - exp(-mean_pos)))
+
+  # Then estimate lambda and theta from positive values
+  lambda_est <- mean_pos
+ # Estimate theta from variance relationship
+  # In GP: Var(X) = lambda / (1-theta)^2
+  # Solving for theta
+  theta_est <- max(0, min(0.999, 1 - sqrt(lambda_est / var_pos)))
+
+  return(list(eta = eta_est, lambda = lambda_est, theta = theta_est))
+}
+aa<- estimate_zigp_params_from_freq(counts, freqs, n_zeros)
+
+```
+
+```{R, echo=FALSE, eval=FALSE}
+# Improved function to estimate ZIGP parameters from frequency data
+# This handles zero-inflation properly by treating zeros separately
+improved_estimate_zigp_params <- function(counts, freqs, n_zeros) {
+  # Calculate basic statistics
+  n <- sum(freqs) + n_zeros
+  p_zeros <- n_zeros / n
+
+  # Focus only on non-zero counts for fitting Generalized Poisson
+  pos_idx <- counts > 0
+  pos_counts <- counts[pos_idx]
+  pos_freqs <- freqs[pos_idx]
+
+  # Calculate weighted statistics for positive counts
+  mean_pos <- sum(pos_counts * pos_freqs) / sum(pos_freqs)
+  var_pos <- sum(pos_freqs * (pos_counts - mean_pos)^2) / sum(pos_freqs)
+
+  # Check if the data is actually highly overdispersed
+  if (var_pos < mean_pos) {
+    # If variance is less than mean (underdispersed), default to Poisson (theta = 0)
+    lambda_est <- mean_pos
+    theta_est <- 0
+  } else {
+    # For overdispersed data, estimate theta using variance relationship
+    # In GP: Var(X) = lambda * (1 + theta)^2 / (1 - theta)^2
+    # Solving for theta (simplified version)
+    dispersion_ratio <- var_pos / mean_pos
+
+    if (dispersion_ratio <= 1) {
+      # No overdispersion
+      theta_est <- 0
+    } else if (dispersion_ratio < 10) {
+      # Moderate overdispersion - using approximation
+      theta_est <- 1 - 1/sqrt(dispersion_ratio)
+      theta_est <- max(0, min(0.8, theta_est))  # Cap at 0.8 for stability
+    } else {
+      # Extreme overdispersion - use caution
+      theta_est <- 0.8  # Cap at 0.8 for numerical stability
+    }
+
+    # Adjust lambda based on theta estimate
+    lambda_est <- mean_pos * (1 - theta_est)
+  }
+
+  # Calculate eta (zero-inflation parameter) using the fitted GP distribution
+  p_zero_gp <- exp(-lambda_est)  # Probability of zero under GP
+
+  # Handle extreme zero-inflation
+  if (p_zeros > 0.9999) {
+    eta_est <- 0.9999  # Cap at 0.9999 for numerical stability
+  } else if (p_zeros <= p_zero_gp) {
+    # If observed zeros are fewer than expected from GP alone
+    eta_est <- 0  # No zero-inflation needed
+  } else {
+    # Calculate eta using the zero-inflation formula
+    eta_est <- (p_zeros - p_zero_gp) / (1 - p_zero_gp)
+    eta_est <- max(0, min(0.9999, eta_est))  # Ensure it's in [0, 0.9999]
+  }
+
+  # Print debugging information
+  cat("Estimated parameters:\n")
+  cat("Mean of positive counts:", mean_pos, "\n")
+  cat("Variance of positive counts:", var_pos, "\n")
+  cat("Dispersion ratio (var/mean):", var_pos/mean_pos, "\n")
+  cat("eta:", eta_est, "\n")
+  cat("lambda:", lambda_est, "\n")
+  cat("theta:", theta_est, "\n")
+
+  # Return estimates
+  return(list(eta = eta_est, lambda = lambda_est, theta = theta_est))
+}
+
+# Function to calculate ZIGP probabilities
+zigp_pmf <- function(x, eta, lambda, theta) {
+  result <- numeric(length(x))
+
+  # Handle zeros separately due to zero-inflation
+  result[x == 0] <- eta + (1 - eta) * exp(-lambda)
+
+  # Handle positive values
+  for (i in which(x > 0)) {
+    val <- x[i]
+    prob <- (1 - eta) * lambda * (lambda + theta * val)^(val - 1) * 
+            exp(-lambda - theta * val) / factorial(val)
+    result[i] <- prob
+  }
+
+  return(result)
+}
+
+# Function to compare observed vs expected frequencies
+compare_observed_expected <- function(counts, freqs, n_zeros, params) {
+  # Create full data frame for comparison
+  all_counts <- c(0, counts)
+  all_freqs <- c(n_zeros, freqs)
+  total_n <- sum(all_freqs)
+
+  # Calculate ZIGP probabilities
+  expected_probs <- zigp_pmf(all_counts, params$eta, params$lambda, params$theta)
+  expected_freqs <- expected_probs * total_n
+
+  comparison <- data.frame(
+    Count = all_counts,
+    Observed = all_freqs,
+    Expected = expected_freqs,
+    ObsProb = all_freqs / total_n,
+    ExpProb = expected_probs,
+    Ratio = all_freqs / expected_freqs
+  )
+
+  return(comparison)
+}
+
+# Example usage:
+# counts <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
+# freqs <- c(290502, 418, 89, 29, 22, 14, 11, 7, 5, 4)
+# n_zeros <- 4463369228
+# 
+# # Estimate improved parameters
+params <- improved_estimate_zigp_params(counts, freqs, n_zeros)
+# 
+# # Compare observed vs expected
+comparison <- compare_observed_expected(counts, freqs, n_zeros, params)
+print(comparison)
+estimate_zigp_params_from_freq<- improved_estimate_zigp_params
+
+```
+
+```{R, echo=FALSE, eval=FALSE}
+# Power law PMF (Zipf distribution)
+zipf_pmf <- function(x, alpha, xmin = 1) {
+  result <- numeric(length(x))
+  # Use a more efficient approach for normalization
+  denom <- zeta(alpha, xmin)  # Riemann zeta function starting at xmin
+
+  result[x < xmin] <- 0
+  result[x >= xmin] <- (x[x >= xmin]^(-alpha)) / denom
+
+  return(result)
+}
+
+# Helper function: approximate Riemann zeta function
+zeta <- function(alpha, xmin = 1) {
+  # For alpha > 1, we can use a finite sum for approximation
+  max_k <- 10000  # This should be enough for good approximation with alpha > 2
+  sum((xmin:max_k)^(-alpha))
+}
+
+# For zero-inflated data, combine with point mass at zero
+zi_zipf_pmf <- function(x, p0, alpha, xmin = 1) {
+  result <- numeric(length(x))
+  result[x == 0] <- p0
+  result[x > 0] <- (1 - p0) * zipf_pmf(x[x > 0], alpha, xmin)
+  return(result)
+}
+
+# Zero-inflated negative binomial
+zinb_pmf <- function(x, p0, size, mu) {
+  result <- numeric(length(x))
+  result[x == 0] <- p0 + (1 - p0) * dnbinom(0, size = size, mu = mu)
+  result[x > 0] <- (1 - p0) * dnbinom(x[x > 0], size = size, mu = mu)
+  return(result)
+}
+
+# Fit Zipf distribution from frequency data
+fit_zipf_from_freq <- function(counts, freqs) {
+  # Skip 0 counts
+  pos_idx <- counts > 0
+  if(sum(pos_idx) == 0) return(list(alpha = 1))
+
+  pos_counts <- counts[pos_idx]
+  pos_freqs <- freqs[pos_idx]
+
+  # MLE for alpha (approximate)
+  alpha_start <- 1.5  # Starting value
+
+  neg_ll <- function(alpha) {
+    # Calculate log likelihood weighted by frequencies
+    -sum(pos_freqs * log(zipf_pmf(pos_counts, alpha)))
+  }
+
+  result <- optim(alpha_start, neg_ll, method = "BFGS")
+  return(list(alpha = result$par))
+}
+
+# Fit Negative Binomial from frequency data
+fit_nb_from_freq <- function(counts, freqs) {
+  # Skip 0 counts
+  pos_idx <- counts > 0
+  if(sum(pos_idx) == 0) return(list(size = 1, mu = 1))
+
+  pos_counts <- counts[pos_idx]
+  pos_freqs <- freqs[pos_idx]
+
+  # Calculate mean and variance using frequencies as weights
+  mu_est <- sum(pos_counts * pos_freqs) / sum(pos_freqs)
+
+  # Calculate variance
+  var_est <- sum(pos_freqs * (pos_counts - mu_est)^2) / sum(pos_freqs)
+
+  if(var_est <= mu_est) var_est <- mu_est * 1.1  # Ensure overdispersion
+
+  size_est <- mu_est^2 / (var_est - mu_est)
+
+  return(list(size = size_est, mu = mu_est))
+}
+
+# Improved plotting function for frequency data
+plot_distribution_from_freq <- function(counts, freqs, n_zeros, model_name, model_probs, p0 = NULL) {
+  # Add the zero count
+  all_counts <- c(0, counts)
+  all_freqs <- c(n_zeros, freqs)
+
+  # Calculate observed probabilities
+  total_n <- sum(all_freqs)
+  obs_probs <- all_freqs / total_n
+
+  # Focus on counts with significant probability
+  max_count <- min(50, max(all_counts))
+
+  # Create a sequence of counts to display
+  display_counts <- 0:max_count
+
+  # Create data for plotting
+  plot_data <- data.frame(
+    Count = factor(display_counts),
+    Observed = NA,
+    Fitted = NA
+  )
+
+  # Fill in observed probabilities
+  for(i in 1:length(all_counts)) {
+    if(all_counts[i] <= max_count) {
+      idx <- which(display_counts == all_counts[i])
+      if(length(idx) > 0) {
+        plot_data$Observed[idx] <- obs_probs[i]
+      }
+    }
+  }
+
+  # Fill in fitted probabilities
+  model_counts <- c(0, counts) # Include zero
+  if(length(model_probs) < length(model_counts)) {
+    # Handle case where model_probs might not have same length
+    model_probs <- c(model_probs[1], model_probs)  # Duplicate first value for zero
+  }
+
+  for(i in 1:length(model_counts)) {
+    if(model_counts[i] <= max_count) {
+      idx <- which(display_counts == model_counts[i])
+      if(length(idx) > 0) {
+        plot_data$Fitted[idx] <- model_probs[i]
+      }
+    }
+  }
+
+  # Convert to long format for plotting
+  library(tidyr)
+  plot_long <- pivot_longer(plot_data, 
+                           cols = c("Observed", "Fitted"),
+                           names_to = "Type", 
+                           values_to = "Probability")
+
+  # Remove missing values
+  plot_long <- plot_long[!is.na(plot_long$Probability),]
+
+  # Create the plot
+  library(ggplot2)
+  model_label <- paste("Fitted", model_name)
+  fill_values <- c("red", "blue")
+  names(fill_values) <- c("Observed", "Fitted")
+
+  p <- ggplot(plot_long, aes(x = Count, y = Probability, fill = Type)) +
+    geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
+    scale_y_log10(limits = c(min(plot_long$Probability[plot_long$Probability > 0]) * 0.5, 1)) +
+    theme_minimal() +
+    labs(title = paste("Observed vs Fitted", model_name, "Probabilities (log scale)"),
+         subtitle = if(!is.null(p0)) paste0("p0 = ", round(p0, 4)) else "",
+         x = "Count Value", y = "Probability (log scale)") +
+    scale_fill_manual(values = fill_values) +
+    theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+  return(p)
+}
+
+# Main function to compare distributions based on frequency data
+compare_distributions_from_freq <- function(counts, freqs, n_zeros) {
+  # Calculate total number of observations
+    n <- sum(freqs) + n_zeros
+    # For the log-likelihood calculations:
+    # Add small epsilon to avoid log(0)
+    epsilon <- 1e-10
+
+  # Calculate empirical probability of zero
+  p0_emp <- n_zeros / n
+
+  # Fit ZIGP (as baseline)
+  params_zigp <- estimate_zigp_params_from_freq(counts, freqs, n_zeros)
+
+  # Calculate ZIGP probabilities and log-likelihood
+  unique_counts <- sort(unique(c(0, counts)))
+  probs_zigp <- zigp_pmf(unique_counts, params_zigp$eta, params_zigp$lambda, params_zigp$theta)
+
+  # Calculate log likelihood using frequencies
+  count_map <- match(c(0, counts), unique_counts)
+  #ll_zigp <- n_zeros * log(probs_zigp[1]) + sum(freqs * log(probs_zigp[count_map[-1]]))
+  ll_zigp <- n_zeros * log(max(probs_zigp[1], epsilon)) + sum(freqs * log(pmax(probs_zigp[count_map[-1]], epsilon)))
+
+  # Fit power law (Zipf)
+  params_zipf <- fit_zipf_from_freq(counts, freqs)
+  p0_est <- p0_emp
+
+  # Calculate Zipf probabilities and log-likelihood
+  probs_zipf_pos <- zipf_pmf(counts, params_zipf$alpha)
+  probs_zipf <- c(p0_est, (1-p0_est) * probs_zipf_pos)
+  #ll_zipf <- n_zeros * log(p0_est) + sum(freqs * log((1-p0_est) * probs_zipf_pos))
+  ll_zipf <- n_zeros * log(max(p0_est, epsilon)) +   sum(freqs * log(pmax((1-p0_est) * probs_zipf_pos, epsilon)))
+
+  # Fit Negative Binomial
+  params_nb <- fit_nb_from_freq(counts, freqs)
+
+    # Calculate NB probabilities and log-likelihood
+    # For the ZINB model:
+    # Ensure the parameters are reasonable for extreme zero-inflation
+    params_nb$size <- pmin(params_nb$size, 1000)  # Cap the size parameter
+
+  probs_nb_pos <- dnbinom(counts, size = params_nb$size, mu = params_nb$mu)
+  probs_nb_0 <- dnbinom(0, size = params_nb$size, mu = params_nb$mu)
+
+  # For zero-inflated model
+  pi_0 <- (p0_est - probs_nb_0) / (1 - probs_nb_0)
+  pi_0 <- max(0, min(1, pi_0))  # Ensure valid probability
+
+  probs_zinb <- c(p0_est, (1-pi_0) * probs_nb_pos)
+  ll_nb <- n_zeros * log(p0_est) + sum(freqs * log((1-pi_0) * probs_nb_pos))
+
+  # Calculate AIC and BIC
+  aic_zigp <- -2 * ll_zigp + 2 * 3  # 3 parameters
+  aic_zipf <- -2 * ll_zipf + 2 * 2  # 2 parameters
+  aic_nb <- -2 * ll_nb + 2 * 3      # 3 parameters
+
+  bic_zigp <- -2 * ll_zigp + log(n) * 3
+  bic_zipf <- -2 * ll_zipf + log(n) * 2
+  bic_nb <- -2 * ll_nb + log(n) * 3
+
+  # Create comparison table
+  results <- data.frame(
+    Model = c("ZIGP", "ZI-Zipf", "ZINB"),
+    LogLikelihood = c(ll_zigp, ll_zipf, ll_nb),
+    Parameters = c(3, 2, 3),
+    AIC = c(aic_zigp, aic_zipf, aic_nb),
+    BIC = c(bic_zigp, bic_zipf, bic_nb),
+    AIC_Ratio = c(aic_zigp/aic_zipf, 1, aic_nb/aic_zipf)  # Add ratio to better interpret differences
+  )
+
+  # Sort by AIC
+  results <- results[order(results$AIC),]
+
+  # Create plots for each model
+  p_zigp <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZIGP", probs_zigp, params_zigp$eta)
+  p_zipf <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZI-Zipf", probs_zipf, p0_est)
+  p_nb <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZINB", probs_zinb, p0_est)
+
+  # Return results
+  return(list(
+    comparison = results,
+    best_model = results$Model[1],
+    plots = list(ZIGP = p_zigp, ZI_Zipf = p_zipf, ZINB = p_nb),
+    parameters = list(
+      ZIGP = params_zigp,
+      ZI_Zipf = c(list(p0 = p0_est), params_zipf),
+      ZINB = c(list(p0 = p0_est, pi_0 = pi_0), params_nb)
+    )
+  ))
+}
+
+```
+
+\begin{table}[h]
+\centering
+\caption{Model Comparison for Rheumatoid Arthritis Interaction Data}
+\begin{tabular}{lrrrrr}
+\hline
+\textbf{Model} & \textbf{LogLikelihood} & \textbf{Parameters} & \textbf{AIC} & \textbf{BIC} & \textbf{AIC Ratio} \\
+\hline
+ZI-Zipf & -3,102,980 & 2 & 6,205,964 & 6,206,005  & 1.000 \\
+ZINB    & -$\infty$  & 3 & $\infty$ & $\infty$    & $\infty$ \\
+ZIGP    & NaN        & 3 & NaN      & NaN         & NaN \\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[h]
+\centering
+\caption{Estimated Parameters for Distribution Models}
+\begin{tabular}{lrrr}
+\hline
+\textbf{Model} & \textbf{Parameter 1} & \textbf{Parameter 2} & \textbf{Parameter 3} \\
+\hline
+ZIGP & $\eta = 0.9999$ & $\lambda = 46.5641$ & $\theta = 0$ \\
+ZI-Zipf & $p_0 = 0.9999348$ & $\alpha = 8.288837$ & -- \\
+ZINB & $p_0 = 0.9999348$ & $\pi_0 = 0.9998953$ & size = $13.84857$, $\mu = 1.010229$ \\
+\hline
+\end{tabular}
+  \end{table}
+
+- For ZIGP Model (NaN result):
+
+  -  With eta = 0.9999 (extremely high zero-inflation) and theta = 0.03459262, the ZIGP model is not far from a Zero-Inflated Poisson with probability of zero very close to 1.
+  -  When calculating log-likelihood for this model, you're likely getting extremely small probability values for non-zero counts, which when logged are approaching negative infinity.
+  -  The combination of multiple very small probabilities in the log-likelihood calculation is causing arithmetic underflow, resulting in NaN.
+
+- For ZINB Model (Inf result):
+
+  -   With extremely high zero-inflation ($p_0 = 0.9999348 and pi_0 = 0.9998953), the model is struggling to explain the non-zero counts.
+  -  The parameter size = 13.84857 with mu = 1.010229 means the distribution has a relatively narrow shape.
+  -  The log-likelihood is likely resulting in -Inf because some observed counts have essentially zero probability under this model, and log(0) is -Inf.
+
+The Core Issue: Extreme Zero-Inflation
+
+- Your data has extreme zero-inflation (over 99.99% zeros). This causes numerical instabilities when fitting these models:
+
+  -  When eta or p0 are extremely close to 1, the probability assigned to any non-zero count becomes vanishingly small.
+  -  This leads to arithmetic underflow, where the computer can't represent such small numbers accurately.
+  -  The ZI-Zipf model handles this better because power-law distributions naturally accommodate heavy tails better than Poisson or Negative Binomial.
+
+# Power law
+
+- A true power law would show a straight line. Deviations suggest a different distribution.
+```{R, echo=FALSE, eval=FALSE}
+# Plot log(frequency) vs log(count)
+png("./supp_figs/power_law.png")
+plot(log(counts[counts > 0]), log(freqs[counts > 0]), 
+     xlab="log(count)", ylab="log(frequency)")
+abline(lm(log(freqs[counts > 0]) ~ log(counts[counts > 0])))
+dev.off()
+
+```
+
+```{R, powerlaw, echo=FALSE, eval=TRUE, fig.cap="Plot of log(frequency) vs log(count). A power law would show a straight line", out.width = '50%',fig.align = 'center'}
+knitr::include_graphics("./supp_figs/power_law.png")
+```
+
+# thresholding
+
+```{R, echo= TRUE, eval=FALSE}
+# Function to calculate meaningful thresholds for each model
+calculate_thresholds <- function(counts, freqs, n_zeros, model_params) {
+  # Total count
+  n_total <- sum(freqs) + n_zeros
+
+  # Calculate mean and SD of positive counts
+  mean_pos <- sum(counts * freqs) / sum(freqs)
+  var_pos <- sum(freqs * (counts - mean_pos)^2) / sum(freqs)
+  sd_pos <- sqrt(var_pos)
+
+  # Calculate various thresholds
+  thresholds <- list()
+
+  # 1. Empirical approaches
+  thresholds$empirical <- list(
+    mean_plus_2sd = ceiling(mean_pos + 2 * sd_pos),
+    mean_plus_3sd = ceiling(mean_pos + 3 * sd_pos),
+    mean_plus_5sd = ceiling(mean_pos + 5 * sd_pos),
+    percentile_99 = as.numeric(quantile(rep(counts, freqs), 0.99)),
+    percentile_999 = as.numeric(quantile(rep(counts, freqs), 0.999))
+  )
+
+  # 2. Model-based approaches
+  # a) For Zipf
+  if (!is.null(model_params$ZI_Zipf)) {
+    alpha <- model_params$ZI_Zipf$alpha
+    p0 <- model_params$ZI_Zipf$p0
+
+    # Find where probability becomes small
+    find_zipf_threshold <- function(p_threshold) {
+      count <- 1
+      while(count < 100) {  # Reasonable upper limit
+        prob <- (1-p0) * zipf_pmf(count, alpha)
+        if (prob < p_threshold) break
+        count <- count + 1
+      }
+      return(count)
+    }
+
+    thresholds$zipf <- list(
+      prob_0001 = find_zipf_threshold(0.0001),
+      prob_00001 = find_zipf_threshold(0.00001)
+    )
+  }
+
+  # b) For ZIGP
+  if (!is.null(model_params$ZIGP)) {
+    eta <- model_params$ZIGP$eta
+    lambda <- model_params$ZIGP$lambda
+    theta <- model_params$ZIGP$theta
+
+    # Find where probability becomes small
+    find_zigp_threshold <- function(p_threshold) {
+      count <- 1
+      while(count < 100) {  # Reasonable upper limit
+        prob <- (1-eta) * (lambda * (lambda + theta * count)^(count - 1) * 
+                 exp(-lambda - theta * count)) / factorial(count)
+        if (prob < p_threshold) break
+        count <- count + 1
+      }
+      return(count)
+    }
+
+    thresholds$zigp <- list(
+      prob_0001 = find_zigp_threshold(0.0001),
+      prob_00001 = find_zigp_threshold(0.00001)
+    )
+  }
+
+  # 3. Count-based approaches
+  # Find natural break points in the distribution
+  natural_breaks <- find_natural_breaks(counts, freqs)
+  thresholds$breaks <- natural_breaks
+
+  # Return all thresholds
+  return(thresholds)
+}
+
+# Helper function to identify natural break points in count distribution
+find_natural_breaks <- function(counts, freqs) {
+  # Calculate rate of change between consecutive counts
+  rate_change <- numeric(length(counts) - 1)
+  for (i in 1:(length(counts)-1)) {
+    if (counts[i+1] - counts[i] == 1) {  # Only consider consecutive integers
+      rate_change[i] <- freqs[i] / freqs[i+1]
+    } else {
+      rate_change[i] <- NA  # Mark non-consecutive values
+    }
+  }
+
+  # Find points where rate of change is significantly higher than average
+  valid_rates <- rate_change[!is.na(rate_change)]
+  if (length(valid_rates) < 3) return(list(break_points = NULL))
+
+  mean_rate <- mean(valid_rates)
+  sd_rate <- sd(valid_rates)
+
+  # Points where rate of change is > mean + 1.5*SD
+  break_indices <- which(rate_change > mean_rate + 1.5*sd_rate)
+  break_points <- counts[break_indices]
+
+  # Return the identified break points
+  return(list(
+    break_points = break_points,
+    rate_change = rate_change,
+    suggested_threshold = if(length(break_points) > 0) min(break_points) + 1 else NULL
+  ))
+}
+
+thresholds <- calculate_thresholds(counts, freqs, n_zeros, results$parameters)
+print(thresholds)
+## empirical
+## $empirical$mean_plus_2sd
+## [1] 4
+
+## $empirical$mean_plus_3sd
+## [1] 5
+
+## $empirical$mean_plus_5sd
+## [1] 7
+
+## $empirical$percentile_99
+## [1] 1
+
+## $empirical$percentile_999
+## [1] 2
+
+## $zipf
+## $zipf$prob_0001
+## [1] 1
+
+## $zipf$prob_00001
+## [1] 2
+
+## $zigp
+## $zigp$prob_0001
+## [1] 1
+
+## $zigp$prob_00001
+## [1] 1
+
+## $breaks
+## $breaks$break_points
+## [1] 1
+
+## $breaks$rate_change
+##  [1] 694.980861   4.696629   3.068966   1.318182   1.571429   1.272727   1.571429   1.400000   1.250000   1.333333   1.000000   1.000000   0.750000   4.000000   1.000000   1.000000
+## [17]   1.000000         NA   2.000000         NA   1.000000   0.500000         NA         NA         NA         NA   1.000000         NA         NA   0.500000   2.000000         NA
+## [33]   1.000000         NA         NA         NA         NA         NA
+
+## $breaks$suggested_threshold
+## [1] 2
+
+find_natural_breaks (counts, freqs)
+## $break_points
+## [1] 1
+
+## $rate_change
+##  [1] 694.980861   4.696629   3.068966   1.318182   1.571429   1.272727   1.571429   1.400000   1.250000   1.333333   1.000000   1.000000   0.750000   4.000000   1.000000   1.000000
+## [17]   1.000000         NA   2.000000         NA   1.000000   0.500000         NA         NA         NA         NA   1.000000         NA         NA   0.500000   2.000000         NA
+## [33]   1.000000         NA         NA         NA         NA         NA
+
+## $suggested_threshold
+## [1] 2
+
+```
+
+```{R, echo=FALSE, eval=FALSE}
+# more refined threshholding
+
+# Robust Empirical Thresholding for Interaction Counts
+# This approach works for both well-fitted distributions and those that don't follow 
+# standard theoretical distributions
+
+empirical_threshold <- function(counts, freqs, n_zeros = 0) {
+  # Create a data frame for easier analysis
+  df <- data.frame(
+    count = counts,
+    freq = freqs,
+    prop = freqs / sum(freqs)
+  )
+
+  # Order by count
+  df <- df[order(df$count), ]
+
+  # Calculate cumulative proportion (excluding zeros)
+  df$cum_prop <- cumsum(df$prop)
+
+  # Calculate several thresholds
+  thresholds <- list()
+
+  # 1. Basic statistics on positive counts
+  mean_pos <- sum(df$count * df$freq) / sum(df$freq)
+  var_pos <- sum(df$freq * (df$count - mean_pos)^2) / sum(df$freq)
+  sd_pos <- sqrt(var_pos)
+
+  # 2. Compute empirical percentiles
+  weighted_data <- rep(df$count, times = df$freq)
+
+  thresholds$basic <- list(
+    mean = mean_pos,
+    sd = sd_pos,
+    threshold_2sd = ceiling(mean_pos + 2 * sd_pos),
+    threshold_3sd = ceiling(mean_pos + 3 * sd_pos),
+    p95 = as.numeric(quantile(weighted_data, 0.95)),
+    p99 = as.numeric(quantile(weighted_data, 0.99)),
+    p999 = as.numeric(quantile(weighted_data, 0.999))
+  )
+
+  # 3. Detect "elbows" or breakpoints in the distribution
+  # Calculate rate of change in frequency between consecutive counts
+  df$rate_change <- NA
+  for (i in 1:(nrow(df)-1)) {
+    if (df$count[i+1] - df$count[i] == 1) {  # Only for consecutive counts
+      df$rate_change[i] <- df$freq[i] / df$freq[i+1]
+    }
+  }
+
+  valid_rates <- df$rate_change[!is.na(df$rate_change)]
+  if (length(valid_rates) >= 3) {
+    mean_rate <- mean(valid_rates)
+    sd_rate <- sd(valid_rates)
+
+    # Find significant jumps in rate of change
+    # Points where rate of change is > mean + 1.5*SD
+    break_indices <- which(df$rate_change > mean_rate + 1.5*sd_rate)
+    break_points <- df$count[break_indices]
+
+    thresholds$breakpoints <- list(
+      points = break_points,
+      suggested = if(length(break_points) > 0) min(break_points) + 1 else NULL
+    )
+  } else {
+    thresholds$breakpoints <- list(
+      points = NULL,
+      suggested = NULL
+    )
+  }
+
+  # 4. Entropy-based thresholding
+  # Calculate entropy of frequency distribution
+  p <- df$prop
+  entropy <- -sum(p * log(p))
+
+  # Find threshold that maximizes entropy difference between below and above threshold
+  max_diff <- -Inf
+  best_threshold <- 1
+
+  for (i in 1:max(df$count)) {
+    below <- df[df$count <= i, ]
+    above <- df[df$count > i, ]
+
+    # Skip if either group is empty
+    if (nrow(below) == 0 || nrow(above) == 0) next
+
+    # Normalize probabilities within each group
+    p_below <- below$freq / sum(below$freq)
+    p_above <- above$freq / sum(above$freq)
+
+    # Calculate entropies
+    entropy_below <- -sum(p_below * log(p_below))
+    entropy_above <- -sum(p_above * log(p_above))
+
+    # Calculate weighted entropy difference
+    entropy_diff <- (sum(below$freq) * entropy_below + sum(above$freq) * entropy_above) / sum(df$freq)
+
+    if (entropy - entropy_diff > max_diff) {
+      max_diff <- entropy - entropy_diff
+      best_threshold <- i
+    }
+  }
+
+  thresholds$entropy <- list(
+    threshold = best_threshold + 1  # Add 1 for exclusive threshold
+  )
+
+  # 5. Exponential decay test
+  # Test if counts follow exponential decay by fitting log(frequency) ~ count
+  if (nrow(df) > 5) {
+    log_freq <- log(df$freq)
+    model <- lm(log_freq ~ df$count)
+    r_squared <- summary(model)$r.squared
+
+    # If strong exponential decay, use where predicted frequency < 1
+    if (r_squared > 0.8) {
+      coefs <- coef(model)
+      threshold_exp <- ceiling(-coefs[1] / coefs[2])  # Where predicted freq < 1
+      thresholds$exponential <- list(
+        r_squared = r_squared,
+        threshold = threshold_exp
+      )
+    } else {
+      thresholds$exponential <- list(
+        r_squared = r_squared,
+        threshold = NULL
+      )
+    }
+  }
+
+  # 6. Rarity-based threshold
+  # Define thresholds based on observation frequency
+  total_observations <- sum(df$freq)
+  rare_threshold <- ceiling(total_observations * 0.001)  # 0.1% of observations
+
+  rare_counts <- df[df$freq <= rare_threshold, "count"]
+  if (length(rare_counts) > 0) {
+    thresholds$rarity <- list(
+      threshold = min(rare_counts)
+    )
+  } else {
+    thresholds$rarity <- list(
+      threshold = NULL
+    )
+  }
+
+  # 7. Add zero-inflation context if provided
+  if (n_zeros > 0) {
+    # Calculate proportion of zeros
+    p_zero <- n_zeros / (n_zeros + sum(df$freq))
+    thresholds$zero_inflation <- list(
+      p_zero = p_zero
+    )
+
+    # For extreme zero-inflation (>99.9%), more aggressive thresholds
+    if (p_zero > 0.999) {
+      # With extreme zeros, any occurrence might be significant
+      thresholds$extreme_zero_inflation <- list(
+        threshold = 2  # Consider anything geq 2 significant when zeros dominate
+      )
+    }
+  }
+
+  # 8. Final recommended threshold based on all methods
+  # Logic for combining thresholds
+  candidates <- c(
+    thresholds$basic$threshold_2sd,
+    thresholds$breakpoints$suggested,
+    thresholds$entropy$threshold,
+    thresholds$rarity$threshold
+  )
+
+  # Add exponential threshold if available
+  if (!is.null(thresholds$exponential$threshold)) {
+    candidates <- c(candidates, thresholds$exponential$threshold)
+  }
+
+  # Add extreme zero-inflation threshold if applicable
+  if (!is.null(thresholds$extreme_zero_inflation$threshold)) {
+    candidates <- c(candidates, thresholds$extreme_zero_inflation$threshold)
+  }
+
+  # Remove NAs
+  candidates <- candidates[!is.na(candidates)]
+
+  # For final recommendation, use median of available thresholds
+  if (length(candidates) > 0) {
+    thresholds$recommended <- median(candidates, na.rm = TRUE)
+  } else {
+    # Default if no thresholds available
+    thresholds$recommended <- max(2, ceiling(mean_pos + 2 * sd_pos))
+  }
+
+  # Provide both results and visualization
+  results <- list(
+    thresholds = thresholds,
+    data = df
+  )
+
+  # Return everything
+  return(results)
+}
+
+# Function to visualize the thresholds on a count distribution
+plot_thresholds <- function(threshold_results, log_scale = TRUE) {
+  df <- threshold_results$data
+  thresholds <- threshold_results$thresholds
+
+  # Setup plot
+  par(mfrow = c(1, 1))
+
+  # Create bar plot of frequencies
+  if (log_scale) {
+    barplot(log(df$freq), names.arg = df$count, xlab = "Count", ylab = "Log Frequency", 
+            main = "Distribution of Counts with Thresholds", col = "lightblue")
+  } else {
+    barplot(df$freq, names.arg = df$count, xlab = "Count", ylab = "Frequency", 
+            main = "Distribution of Counts with Thresholds", col = "lightblue")
+  }
+
+  # Add recommended threshold line
+  abline(v = which(df$count == ceiling(thresholds$recommended)), col = "red", lwd = 2)
+
+  # Add other threshold lines
+  # Add 2SD threshold
+  if (!is.null(thresholds$basic$threshold_2sd)) {
+    pos <- which(df$count >= thresholds$basic$threshold_2sd)[1]
+    if (!is.na(pos)) {
+      abline(v = pos, col = "blue", lty = 2)
+    }
+  }
+
+  # Add breakpoint threshold
+  if (!is.null(thresholds$breakpoints$suggested)) {
+    pos <- which(df$count >= thresholds$breakpoints$suggested)[1]
+    if (!is.na(pos)) {
+      abline(v = pos, col = "green", lty = 2)
+    }
+  }
+
+  # Add entropy threshold
+  if (!is.null(thresholds$entropy$threshold)) {
+    pos <- which(df$count >= thresholds$entropy$threshold)[1]
+    if (!is.na(pos)) {
+      abline(v = pos, col = "purple", lty = 2)
+    }
+  }
+
+  # Add legend
+  legend("topright", 
+         legend = c("Recommended", "2SD", "Breakpoint", "Entropy"),
+         col = c("red", "blue", "green", "purple"),
+         lty = c(1, 2, 2, 2),
+         lwd = c(2, 1, 1, 1))
+
+  # Return the plot
+  invisible(threshold_results)
+}
+
+rheumatoid_thresholds <- empirical_threshold(counts, freqs, n_zeros)
+names(rheumatoid_thresholds$thresholds)
+#[1] "basic"                  "breakpoints"            "entropy"                "exponential"            "rarity"                 "zero_inflation"         "extreme_zero_inflation"
+#[8] "recommended"           
+rheumatoid_thresholds$thresholds$basic$p95$threshold_2sd
+rheumatoid_thresholds$thresholds$basic$threshold_2sd
+rheumatoid_thresholds$thresholds$breakpoints$suggested
+rheumatoid_thresholds$thresholds$entropy
+rheumatoid_thresholds$thresholds$exponential$threshold
+rheumatoid_thresholds$thresholds$rarity$threshold
+rheumatoid_thresholds$thresholds$zero_inflation
+rheumatoid_thresholds$thresholds$extreme_zero_inflation$threshold
+rheumatoid_thresholds$thresholds$recommended
+
+plot_thresholds(rheumatoid_thresholds)
+print(rheumatoid_thresholds$thresholds$recommended)
 
@@
 #[51]  0  0  0  0  0  1  0  0  0  0  0  0  0  0  0  1  0  0  0  0  0  0  0  0
-
-```
-
-# data set up
-
-```{R, echo=FALSE, eval=FALSE}
-
-# Assuming your count data is in a vector 'x'
-x <- interaction_snps_counts$Freq
-
-dim(interaction_snps_counts)[1] #[1] 291142
-sum(interaction_snps_counts$Freq) # 294120
-
-94485*(94485 -1)/2 - 291142
-x2<-  c(rep(0,4463369228),x)
-
-#table(x2) long run time!
-#    0         1         2      3      4       5      6      7      8      9     10     11     12     13     14     15     16     17     18     22     23     25     26     27     31     34     41  51     52     54     62     63     64     78     79     90    100    113    144    466
-#4463369228  290502    418     89     29     22     14     11      7      5      4      3      3      3      4     1      1      1      1      2      1      1      1      2      1      1      1   1      1      1      1      2      1      1      1      1      1      1      1      1
-
-# First, prepare your data
-counts <- as.numeric(names(table(x)))  # Unique count values (1, 2, 3, ...)
-freqs <- as.numeric(table(x))          # Frequencies of each count
-n_zeros <- 4463369228                   # Number of zeros
-
-# Recreate the data based on your table x2
-counts0 <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14,15,  16,  17,  18,  22,  23,  25,  26,  27,  31,  34,  41, 51,  52,  54,  62,  63,  64,  78,  79,  90, 100, 113, 144, 466) 
-freqs0 <- c(4463369228, 290502, 418, 89, 29, 22, 14, 11, 7, 5, 4, 3, 3, 3, 4, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1, 1, 1, 1,   1,   1,   1,   2,   1,   1,   1,   1,   1,   1,   1,   1)
-# Recreate the data based on your table x
-counts <- c( 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14,15,  16,  17,  18,  22,  23,  25,  26,  27,  31,  34,  41, 51,  52,  54,  62,  63,  64,  78,  79,  90, 100, 113, 144, 466) 
-freqs <- c( 290502, 418, 89, 29, 22, 14, 11, 7, 5, 4, 3, 3, 3, 4, 1, 1, 1, 1, 2, 1, 1, 1, 2, 1, 1, 1, 1,   1,   1,   1,   2,   1,   1,   1,   1,   1,   1,   1,   1)
-
-# Create a representative sample of this distribution (for computational efficiency)
-set.seed(123)
-sample_size <- 100000000
-sample_probs <- freqs / sum(freqs)
-x_sample <- sample(counts, size = sample_size, replace = TRUE, prob = sample_probs)
-
-```
-
-# Null distribution
-
-What would the distribution of counts look like if 291142 interactions were assigned randomly?
-Under a random assignment, this follows a binomial distribution that can be approximated by a Poisson distribution with:
-$\lambda= 291,142 / 4,463,565,886 \approx 0.000065$
-
--  Count = 1: approximately 290,849 pairs
--  Count = 2: approximately 9.5 pairs
--  Count $\geq 3$: effectively 0 pairs
-
-The observed data shows:
-
-- Count = 1: 290502  pairs
--  Count = 2: 418   pairs
--  Count = 3: 89  pairs
--  Count $\geq 4$: 133 pairs
-
-\clearpage
-\FloatBarrier
-
-# empiricalzip
-
-- There are  4463613128 potential interactions. We have non-zero values for 294120, so there are a lot of 0 values.
-- what to do to handle this? 
-- **gpEMAlgorithm** fails -- set which estimates to calculate in the function call
-- better name than **RealDataFDR**, what other data would we be running it on
-- docuement output better
-- "X6_32222493 X6_32771829" see below, how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113
-
-   - it doesnt t2["X6_32222493"] =202 t2["X6_32771829"]= 1532 -- lm_result has a problem
-
-## some investigations of empiricalzip. 
-```{R, echo=FALSE, eval=FALSE}
-#devtools:::install_github("parsifal9/empiricalzip")
-estimate<-  zigpEMAlgorithm(x2) #works
-estimate<- zipEMAlgorithm(x2) #works
-system.time(estimate<- pEMAlgorithm(x2)) ##works
-estimate<- gpEMAlgorithm(x2) #Error in while (((abs(hold.lambda - lambdaEstimate)) > epsilon) || ((abs(hold.theta -  :   missing value where TRUE/FALSE needed
-
-aa <- my.RealDataFDR(x2, 0.1) #has gpEMAlgorithm commented out, accepts data ratehr than a file
-# Rejection counts
-#       ZIGP ZIP   GP  P         ZIGP ZIP   GP  P        ZIGP  ZIP     GP  P
-#   one stage procedure           two stage             Storey's FDR
-#[1,]   50   82   82   82      57   82  104   82        35    53    53    53
-#[2,]   50   82   82   82      57   82  104   82        35    53    53    53
-
-#[[2]]Parameter estimates using cutoff method C_0
-
-    ## eta: Zero-inflation parameter (proportion of structural zeros)
-    ## lambda: Mean parameter
-    ## theta: Dispersion parameter (NA for models that don't use it)
-    ## pi: Estimated proportion of null hypotheses ($\pi_0$)
-    ## C: The estimated cutoff value (\hat{C}) - positions with counts $\leq$ C are assumed to be from null
-    ## AD: The truncation point (D_N) - positions with counts $\geq$ AD are automatically considered significant
-
-#        eta         lambda    theta      pi          C0   AD
-#[1,] 0.472686094 1.0020464 1.133941e-12 0.9999401    7    8   zigp
-#[2,] 0.001277773 0.3350806           NA 0.9999075    5   14   zip
-#[3,]          NA 0.3346515 0.000000e+00 0.9999072    4    5   gp
-#[4,]          NA 0.3346515           NA 0.9999072    4   14    p
-#
-#[[3]]
-#           [,1]      [,2]         [,3]      [,4] [,5] [,6]
-#[1,] 0.47268609 1.0020464 1.133941e-12 0.9999401    7    8
-#[2,] 0.00128352 0.3350822           NA 0.9999074    4   14
-#[3,]         NA 0.3346515 0.000000e+00 0.9999072    4    5  
-#[4,]         NA 0.3346515           NA 0.9999072    4   14
-
-#empiricalzip  ZIGP one stage procedure
-#empiricalzip pickes the interactions on the Freq, sointeraction_snps_counts[1:50,]
-interaction_snps_counts$interaction_snps[1:50]
-
-sum(freqs[-c(1:7)]) #[1] 68
-sum(freqs[-c(1:8)]) #[1] 57
-sum(freqs[-c(1:9)]) #[1] 50
-
-```
-The ZIGP model selected by epiricalzip has
-
-- $\eta= 0.472686094$      Zero-inflation parameter (proportion of structural zeros)   
-- $\lambda= 1.0020464$    Mean parameter
-- $\theta= 1.133941e-12$    Dispersion parameter (NA for models that don't use it)  
-- $\pi= 0.9999401$       Estimated proportion of null hypotheses
-
-See table \@ref(tab:zigp_comparison)
-
-\begin{table}
-\caption{\label{tab:zigp_comparison} Observed vs. Expected Counts Under ZIGP Model selected by mpiricalzip}
-\center
-\begin{tabular}{lrr}
-\hline
-\textbf{Count} & \textbf{Observed} & \textbf{Expected (ZIGP)} \\
-\hline
-0 & 4,463,369,228 & 4,463,369,228 \\
-1 & 290,502 & 290,490 \\
-2 & 418 & 427 \\
-3 & 89 & 85 \\
-4 & 29 & 21 \\
-5 & 22 & 5 \\
-6-10 & 41 & 3 \\
-11-20 & 13 & $<$1 \\
-$>$20 & 22 & $\approx$0 \\
-\hline
-\textbf{Total} & 4,463,660,370 & 4,463,660,370 \\
-\hline
-\end{tabular}
-\label{tab:zigp_comparison}
-\end{table}
-
-```{R, echo=FALSE, eval=FALSE}
-# protein example (move this)
-
-the protein exaple from @gauranEmpiricalNullEstimation2018
-library(empiricalzip)
-data(protein, package="empiricalzip")
-k<-1
-x <- protein[,(k+1)][!is.na(protein[,(k+1)])]
-M <- length(x)
-cf <- 10^(-10)
-count <- c(length(x[which(x==0)]), tabulate(x, nbins = max(x)))
-names(count) <- c(0:max(x))
-aa <- my.RealDataFDR(x, 0.1) #
-##     ZIGP ZIP   GP  P         ZIGP ZIP   GP  P        ZIGP  ZIP     GP  P
-##      one stage procedure      two stage             Storey's FDR
-## [1,]  191  212  212  212      201  212  223  212     200   211   211   211
-## [2,]  143  205  212  212      143  205  229  212     162   204   211   211
-
-## [[2]]
-## #        eta      lambda    theta      pi    C0   AD   
-## [1,] 0.3245463 1.9167862 0.1415783 0.4576160    6    7
-## [2,] 0.2288849 1.0948534        NA 0.3985044    3    6
-## [3,]        NA 0.8082199 0.0000000 0.3944174    3    4
-## [4,]        NA 0.8082199        NA 0.3944174    3    6
-
-## [[3]]
-##           [,1]      [,2]      [,3]      [,4] [,5] [,6]
-## [1,] 0.2251642 2.1452087 0.5741445 0.6141275    4   36
-## [2,] 0.2759650 1.3856234        NA 0.4244176    2    6
-## [3,]        NA 0.7994067 0.0000000 0.3929166    2    3
-## [4,]        NA 0.7994067        NA 0.3929166    2    6
-
-# ZIGP parameters
-eta <- 0.3245463
-lambda <- 1.9167862 
-theta <- 0.1415783
-
-# Generate the plots and analysis
-results <- plot_zigp_comparison(
-  observed_counts = count,
-  eta = eta,
-  lambda = lambda, 
-  theta = theta,
-  max_display = 30,
-  y_min = -15  # Extended y-axis
-)
-
-results$log_scale_tail
-abline(v=6)
-
-plot_normalized_mixture(
-   eta = eta,
-   lambda =  lambda, 
-   theta = theta,
-   pi0 = 1,
-   n_binom = 15,
-   p_binom = 0.2,
-   shift = 7
-)
-
-```
-
-```{R, echo=FALSE, eval=FALSE}
-
-# Create a named vector with the observed counts
-observed_data <- freqs
-names(observed_data) <- as.character(counts)
-
-# ZIGP parameters
-eta <- 0.9993963  
-lambda <- 1.0020464
-theta <- 1.133941e-12
-
-# Generate the plots and analysis
-results <- plot_zigp_comparison(
-  observed_counts = observed_data,
-  eta = eta,
-  lambda = lambda, 
-  theta = theta,
-  max_display = 30,
-  y_min = -15  # Extended y-axis
-)
-
-# Display the plots
-library(gridExtra)
-grid.arrange(results$log_scale, results$linear_scale, ncol = 1)
-
-# Testing separation between null and alternative distributions
-# First, examine where the residuals cross from positive to negative
-crossover_point <- min(which(results$separation_analysis$Residual < 0))
-print(paste("Crossover point where observed becomes less than expected:", 
-            results$separation_analysis$Count[crossover_point]))
-
-# Calculate patterns of overdispersion
-counts_with_excess <- sum(results$separation_analysis$Observed > results$separation_analysis$Fitted)
-print(paste("Number of count values with excess observations:", counts_with_excess))
-
-# Calculate the count where maximum deviation occurs
-max_positive_deviation <- which.max(results$separation_analysis$RelativeDiff)
-print(paste("Count with maximum positive deviation:", 
-            results$separation_analysis$Count[max_positive_deviation]))
-
-# Save the improved plot
-#ggsave("./script3/zigp_log_scale_extended.png", results$log_scale, width = 10, height = 6)
-#ggsave("zigp_ratio_plot.png", results$ratio_plot, width = 10, height = 6)
-
-```
-
-```{r zigplogscaleextended, eval=TRUE, echo=FALSE, fig.cap='empiricalzip fitted ZIPG',  fig.align = 'center', out.width = '50%'}
-knitr::include_graphics("./script3/zigp_log_scale_extended.png")
-
-```
-
-\clearpage
-\FloatBarrier
-## Local FDR Calculation
-The local FDR (false discovery rate) at a particular count value is calculated as:
-
-- fdr(t) = ($\pi_0 \times f_0(t)) / f(t)$
-
-Where:
-
--  $t4 is the mutation count
--  $\pi_0$ is the estimated proportion of nulls
--  $ f_0(t)$ is the null distribution probability at count $t$
--  $f(t)$ is the observed mixture distribution probability at count $t$
-
-In the my.RealDataFDR function, this calculation happens here:
-
-```{verbatim}
-LFDR <- pi[model, CO[method, model]] * gpdMixture(x = sort(x), 
-    estimate[model, 1, CO[method, model]], estimate[model, 
-    2, CO[method, model]], estimate[model, 3, CO[method, 
-    model]]) * f12(sort(x), length(x))
-```
-Breaking this down:
-
--  pi[model, CO[method, model]] is the estimated $\pi_0$  (proportion of nulls)
--  gpdMixture() calculates  $ f_0(t)$ , the probability under the null distribution (ZIGP, ZIP, GP, or P) using the estimated parameters
--  f12(sort(x), length(x)) is a correction factor related to f(t). Looking at the f12 function in the RcppExports.R, this factor is calculated as:
-
-```{verbatim}
-   M / length(x[which(x==x[i])])
-```
-where M is the total number of positions. This is effectively equivalent to $1/f(t)$, where $f(t)$ is the empirical probability of observing count $t$.
-
-For the one-stage procedure, positions are declared significant if:
-
-```{verbatim}
-sum(LFDR[!is.na(LFDR)] < FDRLevel)
-```
-meaning their local FDR is below the specified threshold (in your case, 0.1).
-
-Why Some Positions with Counts >7 Aren't Called Significant
-
-With the ZIGP model estimated parameters for your data, the local FDR for positions with counts of 8, 9, etc. depends on:
-
--   How rare such counts are in the data (affecting $f(t)$)
--  How probable they are under the null distribution (affecting $f_0(t)$)
-
-When a position has a count that is high but not extremely high relative to the estimated null distribution, its local
-FDR might still exceed the threshold, especially when $\pi_0$ is estimated to be very close to 1.0.
-
-This is why 7 positions with counts >7 aren't called significant by the one-stage procedure - their local FDR values
-exceed your threshold of 0.1, indicating there's still a substantial probability they come from the null distribution
-despite their relatively high counts.
-
-The two-stage procedure, by contrast, automatically declares positions with counts $\geq$ AD (in this case 8) as significant,
-bypassing the local FDR calculation for these high counts.
-
-1. Stage 1 (Screening Step): We reject the $H_0 j \textrm{if } j \leq D_N(\hat{C})$ and do not reject if $j \leq \hat{C}$.
-2. Stage 2 (Testing step): For $\hat{C} + 1 \leq j < D_N(\hat{C})$ , we calculate the local FDR
-
-## Simplified two-stage procedure for empiricalzip
-
--  "Estimated C: 1"
--  "D_N threshold: 3"
--  "Significant counts:"   3  4  5  7 27
-- used x_sample so some counts missing
-- could use this value
-
-```{R, echo=FALSE, eval=FALSE}
-# Simplified two-stage procedure for extreme zero-inflation
-simplified_two_stage <- function(data, fixed_C = 1, alpha = 0.05) {
-  # Extract the counts and their frequencies
-  tab <- table(data)
-  counts <- as.numeric(names(tab))
-  freqs <- as.numeric(tab)
-  N <- sum(freqs)
-
-  # Given extreme zero-inflation, set C directly
-  C_hat <- fixed_C
-
-  # Estimate parameters for the null distribution (ZIGP)
-  # Given the extreme zeros, simplify parameter estimation
-  eta_est <- freqs[counts == 0] / N  # Zero-inflation parameter
-
-  # For non-zero values, estimate lambda and theta
-  non_zero_counts <- rep(counts[counts > 0], freqs[counts > 0])
-  lambda_est <- mean(non_zero_counts)
-  theta_est <- min(0.5, var(non_zero_counts) / lambda_est - 1)
-
-  # Calculate D_N (direct approach for computational stability)
-  # With extreme zeros, a lower D_N is appropriate
-  D_N <- max(3, C_hat + 1)  # Simple heuristic
-
-  # Create results structure
-  results <- list(
-    C_hat = C_hat,
-    D_N = D_N,
-    parameters = c(eta = eta_est, lambda = lambda_est, theta = theta_est),
-    significant_counts = counts[counts >= D_N]
-  )
-
-  return(results)
-}
-
-# Load necessary libraries
-library(VGAM)
-library(stats)
-
-# Function to implement the two-stage procedure from Gauran et al.
-two_stage_zigp_procedure <- function(data, alpha = 0.05) {
-  # Extract the counts and their frequencies
-  counts <- as.numeric(names(table(data)))
-  freqs <- as.numeric(table(data))
-  N <- sum(freqs)
-
-  # Calculate the empirical probabilities
-  f_hat <- freqs / N
-
-  # Define function to estimate C using the likelihood approach (C1 method)
-  estimate_C1 <- function(counts, freqs, N) {
-    # Initialize variables
-    max_S <- -Inf
-    best_C <- 1
-
-    # Try different values of C
-    for (v in 1:(max(counts)-1)) {
-      # Subset data for estimation
-      subset_counts <- counts[counts <= v]
-      subset_freqs <- freqs[counts <= v ]
-
-      # Fit ZIGP to data up to C using MLE or method of moments
-      # This is a simplified approach - in practice, use EM algorithm
-      p_zeros <- subset_freqs[subset_counts == 0] / sum(subset_freqs)
-      lambda_est <- mean(rep(subset_counts, subset_freqs))
-
-      # Due to extreme zero-inflation, we'll use high eta
-      eta_est <- max(0.95, p_zeros - (1-p_zeros)*exp(-lambda_est))
-      theta_est <- min(0.5, var(rep(subset_counts, subset_freqs)) / lambda_est - 1)
-
-      # Calculate pi_0 estimate
-      pi0_est <- min(1, sum(freqs[counts <= v]) / (N * sum(zigp_pmf(0:v, eta_est, lambda_est, theta_est))))
-
-      # Calculate S_v statistic
-      S_v <- sum(freqs[counts <= v] * log(zigp_pmf(counts[counts <= v], eta_est, lambda_est, theta_est) / f_hat[counts <= v]))
-
-      # Update if this is the best C so far
-      if (S_v > max_S) {
-        max_S <- S_v
-        best_C <- v
-        best_params <- c(eta_est, lambda_est, theta_est, pi0_est)
-      }
-    }
-
-    return(list(C = best_C, params = best_params))
-  }
-
-  # Define ZIGP PMF function (simplified)
-  zigp_pmf <- function(x, eta, lambda, theta) {
-    result <- numeric(length(x))
-
-    # Handle zeros separately due to zero-inflation
-    result[x == 0] <- eta + (1 - eta) * exp(-lambda)
-
-    # Handle positive values
-    for (i in which(x > 0)) {
-      val <- x[i]
-      prob <- (1 - eta) * (lambda * (lambda + theta * val)^(val - 1) * exp(-lambda - theta * val)) / factorial(val)
-      result[i] <- prob
-    }
-
-    return(result)
-  }
-
-  # Calculate D_N (truncation point)
-  calculate_D_N <- function(C, params) {
-    eta <- params[1]
-    lambda <- params[2]
-    theta <- params[3]
-
-    # Calculate D_N based on the paper's formula
-    # Simplified version for extreme zero-inflation
-    D1 <- ceiling(log(N) / (-log(theta)))
-    D2 <- ceiling(C + 1)
-
-    return(max(D1, D2))
-  }
-
-  # Stage 1: Estimate C and calculate D_N
-  C_result <- estimate_C1(counts, freqs, N)
-  C_hat <- C_result$C
-  params <- C_result$params
-  D_N <- calculate_D_N(C_hat, params)
-
-  # Stage 2: Apply local FDR to values between C+1 and D_N-1
-  local_fdr_result <- numeric(max(counts))
-
-  for (j in (C_hat+1):(D_N-1)) {
-    if (j %in% counts) {
-      # Calculate local FDR
-      null_density <- zigp_pmf(j, params[1], params[2], params[3])
-      mixture_density <- f_hat[counts == j]
-      local_fdr_result[j] <- params[4] * null_density / mixture_density
-    }
-  }
-
-  # Determine significant counts
-  significant_counts <- numeric(0)
-
-  # Reject all counts \geq D_N
-  if (any(counts >= D_N)) {
-    significant_counts <- c(significant_counts, counts[counts >= D_N])
-  }
-
-  # Reject counts between C+1 and D_N-1 if local FDR \leq alpha
-  for (j in (C_hat+1):(D_N-1)) {
-    if (j %in% counts && local_fdr_result[j] <= alpha) {
-      significant_counts <- c(significant_counts, j)
-    }
-  }
-
-  return(list(
-    C_hat = C_hat,
-    D_N = D_N,
-    parameters = params,
-    local_fdr = local_fdr_result,
-    significant_counts = significant_counts
-  ))
-} #end two_stage_zigp_procedure
-
-## Apply the simplified two-stage procedure
-results <- simplified_two_stage(x_sample, fixed_C = 1)
-
-# Display results
-print(paste("Estimated C:", results$C_hat))
-print(paste("D_N threshold:", results$D_N))
-print("Significant counts:")
-print(results$significant_counts)
-
-```
-
-# hurdle model implementation
-
-```{R, echo=FALSE, eval=FALSE}
-
-library(pscl)  # For hurdle model implementation
-
-# Fit a hurdle model
-hurdle_model <- hurdle(
-  x ~ 1,               # Formula (just intercept in this simple case)
-  data = data.frame(x = x_sample),
-  dist = "negbin",     # Distribution for non-zero counts
-  zero.dist = "binomial" # Distribution for zero vs non-zero process
-)
-
-summary(hurdle_model)
-
-```
-
-```{R, echo=FALSE, eval=FALSE}
-# what is this? remove it
-# Custom function for extreme zero-inflation
-extreme_zero_model <- function(x, min_count = 3) {
-  # Calculate basic statistics
-  zeros <- sum(x == 0)
-  n <- length(x)
-  p_zero <- zeros/n
-
-  # Estimate significance threshold
-  # For extreme zero-inflation, we can use empirical approach
-  unique_counts <- sort(unique(x))
-  freq_table <- table(x)
-
-  # Simple rule: counts are significant when they occur with 
-  # frequency less than 0.1% of non-zero observations
-  rarity_threshold <- 0.001 * (n - zeros)
-  significant <- unique_counts[unique_counts >= min_count & 
-                              freq_table[as.character(unique_counts)] < rarity_threshold]
-
-  return(list(
-    p_zero = p_zero,
-    significant_counts = significant
-  ))
-}
-
-```
-
-# what is this 
-```{R, echo=FALSE,eval=FALSE}
-
-length(intersect(interaction_snps_counts$interaction_snps[1:50],lm_results$interaction_snps[1:88])) #22
-# lm_results (has been ordered by lm_results$int_Pvalue)
-
-t1 <- setdiff(interaction_snps_counts$interaction_snps[1:50],lm_results$interaction_snps[1:88])
-match(t1,interaction_snps_counts$interaction_snps)  #3  4  5  7
-
-t1[1]
-#1] "X6_32222493 X6_32771829"
-interaction_snps_counts[3,]
-#              interaction_snps Freq
-#280884 X6_32222493 X6_32771829  113
-
-lm_results[grep("X6_32771829", lm_results$interaction_snps),] #301
-lm_results[grep("X6_32222493 X6_32771829", lm_results$interaction_snps),] 
-#              interaction_snps Freq SNP1_effect  SNP1_Pvalue SNP2_effect  SNP2_Pvalue int_effect int_Pvalue        Ch2 correlations splits_SNP1 splits_SNP2
-#280884 X6_32222493 X6_32771829  113   0.3937617 6.972158e-09  -0.6516203 2.576182e-27 -0.1303979 0.08982739 0.03193039   -0.3742329          18          14
-#       adj_p_value_bonferrroni adj_p_value_fdr adj_p_value_Ch2
-#280884                57.48953       0.2521143       0.1011656
-
-# so this is not significant for any of the tests except empiricalzip
-# how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113
-
-chisq.test(table( dt[,"X6_32222493",], dt[,"X6_32771829",]))
-
-aa<-rep(NA, 74)
-for ( i in 1:74){
-    aa[i] <- length(grep(imp[i],interaction_snps_counts$interaction_snps[1:50]))
-    }
-aa
-# [1] 21 13  8  7  5  5  5  8  5  3  2  2  2  1  2  2  1  1  0  0  1  0  0  0  0  0  1  0  0  0  1  0  0  1  1  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  1  0  0
-#[56]  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0  0
-
-index <- match(selectedSNPs,colnames(dt))
-index <- index[order(index)]
-
-data4interactions <- dt[,index]
-saveRDS(data4interactions,paste0(dirOut,"/data4interactions.rds"))
-
-SNPs <- lm_results$interaction_snps
-
-SNPs <- as.character(SNPs)
-SNPs <- gsub("X","" ,SNPs)
-#SNPs <- read.table("Results/SNPs.txt",header=T)
-
-```
-
-4. what is the intesection of imp and interacting transcripts
-
-   - see table at end of section
-
-- "X6_32222493 X6_32771829" see below, how does it have  splits_SNP1=18,  splits_SNP2=14 but Freq=113
-
-   - it doesnt t2["X6_32222493"] =202 t2["X6_32771829"]= 1532 -- lm_result has a problem
-
-# compare some distributions
-
-- modified compare_distributions function and related plotting functions to work with a frequency table input rather than requiring the full vector. 
-This will be much more memory-efficient when dealing with large datasets that have many zeros.
-
-```{R, echo=FALSE, eval=FALSE}
-# Then run the comparison
-results <- compare_distributions_from_freq(counts, freqs, n_zeros)
-
-# View results
-print(results$comparison)
-#    Model LogLikelihood Parameters     AIC     BIC AIC_Ratio
-#2 ZI-Zipf      -3102980          2 6205964 6206005         1
-#3    ZINB          -Inf          3     Inf     Inf       Inf
-#1    ZIGP           NaN          3     NaN     NaN       NaN
-
-print(paste("Best model:", results$best_model))
-
-#png("./script3/ZI_Zipf_plot.png")
-results$plots$"ZI_Zipf"
-#dev.off()
-#png("./script3/ZIGP_plot.png")
-results$plots$"ZIGP"
-#dev.off()
-#png("./script3/ZINB_plot.png")
-results$plots$"ZINB"
-#dev.off()
-
-results$parameters
-#$ZIGP $ZIGP$eta [1] 0.9999 $ZIGP$lambda [1]  0.9752822  $ZIGP$theta  [1]  0.03459262
-#$ZI_Zipf $ZI_Zipf$p0 [1] 0.9999348 $ZI_Zipf$alpha [1] 8.288837 
-#$ZINB $ZINB$p0 [1] 0.9999348 $ZINB$pi_0  [1] 0.9998953 $ZINB$size  [1] 13.84857 $ZINB$mu [1] 1.010229
-
-```
-
-```{R, zipfplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Power Law (Zipf) Distribution", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./script3/ZI_Zipf_plot.png")
-```
-
-```{R, zigpplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Generalized Poisson", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./script3/ZIGP_plot.png")
-```
-
-```{R, zinbplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Negative Binomial", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./script3/ZINB_plot.png")
-```
-
-```{R, echo=FALSE, eval=FALSE}
-# Modified functions to work with frequency data instead of raw vectors
-
-# ZIGP probability mass function
-zigp_pmf <- function(x, eta, lambda, theta) {
-  result <- numeric(length(x))
-
-  # Handle zeros
-  result[x == 0] <- eta + (1 - eta) * exp(-lambda)
-
-  # Handle positive values
-  for (i in which(x > 0)) {
-    val <- x[i]
-    prob <- (1 - eta) * lambda * (lambda + theta * val)^(val - 1) * exp(-lambda - theta * val) / factorial(val)
-    result[i] <- prob
-  }
-
-  return(result)
-}
-
-# Modifications to handle extreme zero-inflation
-# For the ZIGP model:
-estimate_zigp_params_from_freq <- function(counts, freqs, n_zeros) {
-  # Calculate basic statistics
-  n <- sum(freqs) + n_zeros
-  p_zeros <- n_zeros / n
-
-  # If extreme zero-inflation
-  if (p_zeros > 0.9999) {
-    # Use a simpler model with high but not exactly 1 eta
-    return(list(eta = 0.9999, lambda = mean(counts[counts > 0]), theta = 0))
-  }
-
-    # Rest of the function as before...
-  # Calculate statistics for positive values
-  if (sum(freqs[counts > 0]) == 0) {
-    return(list(eta = 1, lambda = 1, theta = 0))  # All zeros case
-  }
-
-  # Calculate mean and variance of positive values
-  # Weighted calculations using frequencies
-  mean_pos <- sum(counts[counts > 0] * freqs[counts > 0]) / sum(freqs[counts > 0])
-
-  # Calculate variance
-  var_pos <- sum(freqs[counts > 0] * (counts[counts > 0] - mean_pos)^2) / sum(freqs[counts > 0])
-
-  # Estimate parameters using method of moments
-  # For zero-inflated data, first estimate eta
-  eta_est <- max(0, (p_zeros - exp(-mean_pos)) / (1 - exp(-mean_pos)))
-
-  # Then estimate lambda and theta from positive values
-  lambda_est <- mean_pos
- # Estimate theta from variance relationship
-  # In GP: Var(X) = lambda / (1-theta)^2
-  # Solving for theta
-  theta_est <- max(0, min(0.999, 1 - sqrt(lambda_est / var_pos)))
-
-  return(list(eta = eta_est, lambda = lambda_est, theta = theta_est))
-}
-aa<- estimate_zigp_params_from_freq(counts, freqs, n_zeros)
-
-```
-
-```{R, echo=FALSE, eval=FALSE}
-# Improved function to estimate ZIGP parameters from frequency data
-# This handles zero-inflation properly by treating zeros separately
-improved_estimate_zigp_params <- function(counts, freqs, n_zeros) {
-  # Calculate basic statistics
-  n <- sum(freqs) + n_zeros
-  p_zeros <- n_zeros / n
-
-  # Focus only on non-zero counts for fitting Generalized Poisson
-  pos_idx <- counts > 0
-  pos_counts <- counts[pos_idx]
-  pos_freqs <- freqs[pos_idx]
-
-  # Calculate weighted statistics for positive counts
-  mean_pos <- sum(pos_counts * pos_freqs) / sum(pos_freqs)
-  var_pos <- sum(pos_freqs * (pos_counts - mean_pos)^2) / sum(pos_freqs)
-
-  # Check if the data is actually highly overdispersed
-  if (var_pos < mean_pos) {
-    # If variance is less than mean (underdispersed), default to Poisson (theta = 0)
-    lambda_est <- mean_pos
-    theta_est <- 0
-  } else {
-    # For overdispersed data, estimate theta using variance relationship
-    # In GP: Var(X) = lambda * (1 + theta)^2 / (1 - theta)^2
-    # Solving for theta (simplified version)
-    dispersion_ratio <- var_pos / mean_pos
-
-    if (dispersion_ratio <= 1) {
-      # No overdispersion
-      theta_est <- 0
-    } else if (dispersion_ratio < 10) {
-      # Moderate overdispersion - using approximation
-      theta_est <- 1 - 1/sqrt(dispersion_ratio)
-      theta_est <- max(0, min(0.8, theta_est))  # Cap at 0.8 for stability
-    } else {
-      # Extreme overdispersion - use caution
-      theta_est <- 0.8  # Cap at 0.8 for numerical stability
-    }
-
-    # Adjust lambda based on theta estimate
-    lambda_est <- mean_pos * (1 - theta_est)
-  }
-
-  # Calculate eta (zero-inflation parameter) using the fitted GP distribution
-  p_zero_gp <- exp(-lambda_est)  # Probability of zero under GP
-
-  # Handle extreme zero-inflation
-  if (p_zeros > 0.9999) {
-    eta_est <- 0.9999  # Cap at 0.9999 for numerical stability
-  } else if (p_zeros <= p_zero_gp) {
-    # If observed zeros are fewer than expected from GP alone
-    eta_est <- 0  # No zero-inflation needed
-  } else {
-    # Calculate eta using the zero-inflation formula
-    eta_est <- (p_zeros - p_zero_gp) / (1 - p_zero_gp)
-    eta_est <- max(0, min(0.9999, eta_est))  # Ensure it's in [0, 0.9999]
-  }
-
-  # Print debugging information
-  cat("Estimated parameters:\n")
-  cat("Mean of positive counts:", mean_pos, "\n")
-  cat("Variance of positive counts:", var_pos, "\n")
-  cat("Dispersion ratio (var/mean):", var_pos/mean_pos, "\n")
-  cat("eta:", eta_est, "\n")
-  cat("lambda:", lambda_est, "\n")
-  cat("theta:", theta_est, "\n")
-
-  # Return estimates
-  return(list(eta = eta_est, lambda = lambda_est, theta = theta_est))
-}
-
-# Function to calculate ZIGP probabilities
-zigp_pmf <- function(x, eta, lambda, theta) {
-  result <- numeric(length(x))
-
-  # Handle zeros separately due to zero-inflation
-  result[x == 0] <- eta + (1 - eta) * exp(-lambda)
-
-  # Handle positive values
-  for (i in which(x > 0)) {
-    val <- x[i]
-    prob <- (1 - eta) * lambda * (lambda + theta * val)^(val - 1) * 
-            exp(-lambda - theta * val) / factorial(val)
-    result[i] <- prob
-  }
-
-  return(result)
-}
-
-# Function to compare observed vs expected frequencies
-compare_observed_expected <- function(counts, freqs, n_zeros, params) {
-  # Create full data frame for comparison
-  all_counts <- c(0, counts)
-  all_freqs <- c(n_zeros, freqs)
-  total_n <- sum(all_freqs)
-
-  # Calculate ZIGP probabilities
-  expected_probs <- zigp_pmf(all_counts, params$eta, params$lambda, params$theta)
-  expected_freqs <- expected_probs * total_n
-
-  comparison <- data.frame(
-    Count = all_counts,
-    Observed = all_freqs,
-    Expected = expected_freqs,
-    ObsProb = all_freqs / total_n,
-    ExpProb = expected_probs,
-    Ratio = all_freqs / expected_freqs
-  )
-
-  return(comparison)
-}
-
-# Example usage:
-# counts <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
-# freqs <- c(290502, 418, 89, 29, 22, 14, 11, 7, 5, 4)
-# n_zeros <- 4463369228
-# 
-# # Estimate improved parameters
-params <- improved_estimate_zigp_params(counts, freqs, n_zeros)
-# 
-# # Compare observed vs expected
-comparison <- compare_observed_expected(counts, freqs, n_zeros, params)
-print(comparison)
-estimate_zigp_params_from_freq<- improved_estimate_zigp_params
-
-```
-
-```{R, echo=FALSE, eval=FALSE}
-# Power law PMF (Zipf distribution)
-zipf_pmf <- function(x, alpha, xmin = 1) {
-  result <- numeric(length(x))
-  # Use a more efficient approach for normalization
-  denom <- zeta(alpha, xmin)  # Riemann zeta function starting at xmin
-
-  result[x < xmin] <- 0
-  result[x >= xmin] <- (x[x >= xmin]^(-alpha)) / denom
-
-  return(result)
-}
-
-# Helper function: approximate Riemann zeta function
-zeta <- function(alpha, xmin = 1) {
-  # For alpha > 1, we can use a finite sum for approximation
-  max_k <- 10000  # This should be enough for good approximation with alpha > 2
-  sum((xmin:max_k)^(-alpha))
-}
-
-# For zero-inflated data, combine with point mass at zero
-zi_zipf_pmf <- function(x, p0, alpha, xmin = 1) {
-  result <- numeric(length(x))
-  result[x == 0] <- p0
-  result[x > 0] <- (1 - p0) * zipf_pmf(x[x > 0], alpha, xmin)
-  return(result)
-}
-
-# Zero-inflated negative binomial
-zinb_pmf <- function(x, p0, size, mu) {
-  result <- numeric(length(x))
-  result[x == 0] <- p0 + (1 - p0) * dnbinom(0, size = size, mu = mu)
-  result[x > 0] <- (1 - p0) * dnbinom(x[x > 0], size = size, mu = mu)
-  return(result)
-}
-
-# Fit Zipf distribution from frequency data
-fit_zipf_from_freq <- function(counts, freqs) {
-  # Skip 0 counts
-  pos_idx <- counts > 0
-  if(sum(pos_idx) == 0) return(list(alpha = 1))
-
-  pos_counts <- counts[pos_idx]
-  pos_freqs <- freqs[pos_idx]
-
-  # MLE for alpha (approximate)
-  alpha_start <- 1.5  # Starting value
-
-  neg_ll <- function(alpha) {
-    # Calculate log likelihood weighted by frequencies
-    -sum(pos_freqs * log(zipf_pmf(pos_counts, alpha)))
-  }
-
-  result <- optim(alpha_start, neg_ll, method = "BFGS")
-  return(list(alpha = result$par))
-}
-
-# Fit Negative Binomial from frequency data
-fit_nb_from_freq <- function(counts, freqs) {
-  # Skip 0 counts
-  pos_idx <- counts > 0
-  if(sum(pos_idx) == 0) return(list(size = 1, mu = 1))
-
-  pos_counts <- counts[pos_idx]
-  pos_freqs <- freqs[pos_idx]
-
-  # Calculate mean and variance using frequencies as weights
-  mu_est <- sum(pos_counts * pos_freqs) / sum(pos_freqs)
-
-  # Calculate variance
-  var_est <- sum(pos_freqs * (pos_counts - mu_est)^2) / sum(pos_freqs)
-
-  if(var_est <= mu_est) var_est <- mu_est * 1.1  # Ensure overdispersion
-
-  size_est <- mu_est^2 / (var_est - mu_est)
-
-  return(list(size = size_est, mu = mu_est))
-}
-
-# Improved plotting function for frequency data
-plot_distribution_from_freq <- function(counts, freqs, n_zeros, model_name, model_probs, p0 = NULL) {
-  # Add the zero count
-  all_counts <- c(0, counts)
-  all_freqs <- c(n_zeros, freqs)
-
-  # Calculate observed probabilities
-  total_n <- sum(all_freqs)
-  obs_probs <- all_freqs / total_n
-
-  # Focus on counts with significant probability
-  max_count <- min(50, max(all_counts))
-
-  # Create a sequence of counts to display
-  display_counts <- 0:max_count
-
-  # Create data for plotting
-  plot_data <- data.frame(
-    Count = factor(display_counts),
-    Observed = NA,
-    Fitted = NA
-  )
-
-  # Fill in observed probabilities
-  for(i in 1:length(all_counts)) {
-    if(all_counts[i] <= max_count) {
-      idx <- which(display_counts == all_counts[i])
-      if(length(idx) > 0) {
-        plot_data$Observed[idx] <- obs_probs[i]
-      }
-    }
-  }
-
-  # Fill in fitted probabilities
-  model_counts <- c(0, counts) # Include zero
-  if(length(model_probs) < length(model_counts)) {
-    # Handle case where model_probs might not have same length
-    model_probs <- c(model_probs[1], model_probs)  # Duplicate first value for zero
-  }
-
-  for(i in 1:length(model_counts)) {
-    if(model_counts[i] <= max_count) {
-      idx <- which(display_counts == model_counts[i])
-      if(length(idx) > 0) {
-        plot_data$Fitted[idx] <- model_probs[i]
-      }
-    }
-  }
-
-  # Convert to long format for plotting
-  library(tidyr)
-  plot_long <- pivot_longer(plot_data, 
-                           cols = c("Observed", "Fitted"),
-                           names_to = "Type", 
-                           values_to = "Probability")
-
-  # Remove missing values
-  plot_long <- plot_long[!is.na(plot_long$Probability),]
-
-  # Create the plot
-  library(ggplot2)
-  model_label <- paste("Fitted", model_name)
-  fill_values <- c("red", "blue")
-  names(fill_values) <- c("Observed", "Fitted")
-
-  p <- ggplot(plot_long, aes(x = Count, y = Probability, fill = Type)) +
-    geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
-    scale_y_log10(limits = c(min(plot_long$Probability[plot_long$Probability > 0]) * 0.5, 1)) +
-    theme_minimal() +
-    labs(title = paste("Observed vs Fitted", model_name, "Probabilities (log scale)"),
-         subtitle = if(!is.null(p0)) paste0("p0 = ", round(p0, 4)) else "",
-         x = "Count Value", y = "Probability (log scale)") +
-    scale_fill_manual(values = fill_values) +
-    theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-  return(p)
-}
-
-# Main function to compare distributions based on frequency data
-compare_distributions_from_freq <- function(counts, freqs, n_zeros) {
-  # Calculate total number of observations
-    n <- sum(freqs) + n_zeros
-    # For the log-likelihood calculations:
-    # Add small epsilon to avoid log(0)
-    epsilon <- 1e-10
-
-  # Calculate empirical probability of zero
-  p0_emp <- n_zeros / n
-
-  # Fit ZIGP (as baseline)
-  params_zigp <- estimate_zigp_params_from_freq(counts, freqs, n_zeros)
-
-  # Calculate ZIGP probabilities and log-likelihood
-  unique_counts <- sort(unique(c(0, counts)))
-  probs_zigp <- zigp_pmf(unique_counts, params_zigp$eta, params_zigp$lambda, params_zigp$theta)
-
-  # Calculate log likelihood using frequencies
-  count_map <- match(c(0, counts), unique_counts)
-  #ll_zigp <- n_zeros * log(probs_zigp[1]) + sum(freqs * log(probs_zigp[count_map[-1]]))
-  ll_zigp <- n_zeros * log(max(probs_zigp[1], epsilon)) + sum(freqs * log(pmax(probs_zigp[count_map[-1]], epsilon)))
-
-  # Fit power law (Zipf)
-  params_zipf <- fit_zipf_from_freq(counts, freqs)
-  p0_est <- p0_emp
-
-  # Calculate Zipf probabilities and log-likelihood
-  probs_zipf_pos <- zipf_pmf(counts, params_zipf$alpha)
-  probs_zipf <- c(p0_est, (1-p0_est) * probs_zipf_pos)
-  #ll_zipf <- n_zeros * log(p0_est) + sum(freqs * log((1-p0_est) * probs_zipf_pos))
-  ll_zipf <- n_zeros * log(max(p0_est, epsilon)) +   sum(freqs * log(pmax((1-p0_est) * probs_zipf_pos, epsilon)))
-
-  # Fit Negative Binomial
-  params_nb <- fit_nb_from_freq(counts, freqs)
-
-    # Calculate NB probabilities and log-likelihood
-    # For the ZINB model:
-    # Ensure the parameters are reasonable for extreme zero-inflation
-    params_nb$size <- pmin(params_nb$size, 1000)  # Cap the size parameter
-
-  probs_nb_pos <- dnbinom(counts, size = params_nb$size, mu = params_nb$mu)
-  probs_nb_0 <- dnbinom(0, size = params_nb$size, mu = params_nb$mu)
-
-  # For zero-inflated model
-  pi_0 <- (p0_est - probs_nb_0) / (1 - probs_nb_0)
-  pi_0 <- max(0, min(1, pi_0))  # Ensure valid probability
-
-  probs_zinb <- c(p0_est, (1-pi_0) * probs_nb_pos)
-  ll_nb <- n_zeros * log(p0_est) + sum(freqs * log((1-pi_0) * probs_nb_pos))
-
-  # Calculate AIC and BIC
-  aic_zigp <- -2 * ll_zigp + 2 * 3  # 3 parameters
-  aic_zipf <- -2 * ll_zipf + 2 * 2  # 2 parameters
-  aic_nb <- -2 * ll_nb + 2 * 3      # 3 parameters
-
-  bic_zigp <- -2 * ll_zigp + log(n) * 3
-  bic_zipf <- -2 * ll_zipf + log(n) * 2
-  bic_nb <- -2 * ll_nb + log(n) * 3
-
-  # Create comparison table
-  results <- data.frame(
-    Model = c("ZIGP", "ZI-Zipf", "ZINB"),
-    LogLikelihood = c(ll_zigp, ll_zipf, ll_nb),
-    Parameters = c(3, 2, 3),
-    AIC = c(aic_zigp, aic_zipf, aic_nb),
-    BIC = c(bic_zigp, bic_zipf, bic_nb),
-    AIC_Ratio = c(aic_zigp/aic_zipf, 1, aic_nb/aic_zipf)  # Add ratio to better interpret differences
-  )
-
-  # Sort by AIC
-  results <- results[order(results$AIC),]
-
-  # Create plots for each model
-  p_zigp <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZIGP", probs_zigp, params_zigp$eta)
-  p_zipf <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZI-Zipf", probs_zipf, p0_est)
-  p_nb <- plot_distribution_from_freq(counts, freqs, n_zeros, "ZINB", probs_zinb, p0_est)
-
-  # Return results
-  return(list(
-    comparison = results,
-    best_model = results$Model[1],
-    plots = list(ZIGP = p_zigp, ZI_Zipf = p_zipf, ZINB = p_nb),
-    parameters = list(
-      ZIGP = params_zigp,
-      ZI_Zipf = c(list(p0 = p0_est), params_zipf),
-      ZINB = c(list(p0 = p0_est, pi_0 = pi_0), params_nb)
-    )
-  ))
-}
-
-```
-
-\begin{table}[h]
-\centering
-\caption{Model Comparison for Rheumatoid Arthritis Interaction Data}
-\begin{tabular}{lrrrrr}
-\hline
-\textbf{Model} & \textbf{LogLikelihood} & \textbf{Parameters} & \textbf{AIC} & \textbf{BIC} & \textbf{AIC Ratio} \\
-\hline
-ZI-Zipf & -3,102,980 & 2 & 6,205,964 & 6,206,005  & 1.000 \\
-ZINB    & -$\infty$  & 3 & $\infty$ & $\infty$    & $\infty$ \\
-ZIGP    & NaN        & 3 & NaN      & NaN         & NaN \\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}[h]
-\centering
-\caption{Estimated Parameters for Distribution Models}
-\begin{tabular}{lrrr}
-\hline
-\textbf{Model} & \textbf{Parameter 1} & \textbf{Parameter 2} & \textbf{Parameter 3} \\
-\hline
-ZIGP & $\eta = 0.9999$ & $\lambda = 46.5641$ & $\theta = 0$ \\
-ZI-Zipf & $p_0 = 0.9999348$ & $\alpha = 8.288837$ & -- \\
-ZINB & $p_0 = 0.9999348$ & $\pi_0 = 0.9998953$ & size = $13.84857$, $\mu = 1.010229$ \\
-\hline
-\end{tabular}
-  \end{table}
-
-- For ZIGP Model (NaN result):
-
-  -  With eta = 0.9999 (extremely high zero-inflation) and theta = 0.03459262, the ZIGP model is not far from a Zero-Inflated Poisson with probability of zero very close to 1.
-  -  When calculating log-likelihood for this model, you're likely getting extremely small probability values for non-zero counts, which when logged are approaching negative infinity.
-  -  The combination of multiple very small probabilities in the log-likelihood calculation is causing arithmetic underflow, resulting in NaN.
-
-- For ZINB Model (Inf result):
-
-  -   With extremely high zero-inflation ($p_0 = 0.9999348 and pi_0 = 0.9998953), the model is struggling to explain the non-zero counts.
-  -  The parameter size = 13.84857 with mu = 1.010229 means the distribution has a relatively narrow shape.
-  -  The log-likelihood is likely resulting in -Inf because some observed counts have essentially zero probability under this model, and log(0) is -Inf.
-
-The Core Issue: Extreme Zero-Inflation
-
-- Your data has extreme zero-inflation (over 99.99% zeros). This causes numerical instabilities when fitting these models:
-
-  -  When eta or p0 are extremely close to 1, the probability assigned to any non-zero count becomes vanishingly small.
-  -  This leads to arithmetic underflow, where the computer can't represent such small numbers accurately.
-  -  The ZI-Zipf model handles this better because power-law distributions naturally accommodate heavy tails better than Poisson or Negative Binomial.
-
-# Power law
-
-- A true power law would show a straight line. Deviations suggest a different distribution.
-```{R, echo=FALSE, eval=FALSE}
-# Plot log(frequency) vs log(count)
-png("./script3/power_law.png")
-plot(log(counts[counts > 0]), log(freqs[counts > 0]), 
-     xlab="log(count)", ylab="log(frequency)")
-abline(lm(log(freqs[counts > 0]) ~ log(counts[counts > 0])))
-dev.off()
-
-```
-
-```{R, powerlaw, echo=FALSE, eval=TRUE, fig.cap="Plot of log(frequency) vs log(count). A power law would show a straight line", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./script3/power_law.png")
-```
-
-# thresholding
-
-```{R, echo= TRUE, eval=FALSE}
-# Function to calculate meaningful thresholds for each model
-calculate_thresholds <- function(counts, freqs, n_zeros, model_params) {
-  # Total count
-  n_total <- sum(freqs) + n_zeros
-
-  # Calculate mean and SD of positive counts
-  mean_pos <- sum(counts * freqs) / sum(freqs)
-  var_pos <- sum(freqs * (counts - mean_pos)^2) / sum(freqs)
-  sd_pos <- sqrt(var_pos)
-
-  # Calculate various thresholds
-  thresholds <- list()
-
-  # 1. Empirical approaches
-  thresholds$empirical <- list(
-    mean_plus_2sd = ceiling(mean_pos + 2 * sd_pos),
-    mean_plus_3sd = ceiling(mean_pos + 3 * sd_pos),
-    mean_plus_5sd = ceiling(mean_pos + 5 * sd_pos),
-    percentile_99 = as.numeric(quantile(rep(counts, freqs), 0.99)),
-    percentile_999 = as.numeric(quantile(rep(counts, freqs), 0.999))
-  )
-
-  # 2. Model-based approaches
-  # a) For Zipf
-  if (!is.null(model_params$ZI_Zipf)) {
-    alpha <- model_params$ZI_Zipf$alpha
-    p0 <- model_params$ZI_Zipf$p0
-
-    # Find where probability becomes small
-    find_zipf_threshold <- function(p_threshold) {
-      count <- 1
-      while(count < 100) {  # Reasonable upper limit
-        prob <- (1-p0) * zipf_pmf(count, alpha)
-        if (prob < p_threshold) break
-        count <- count + 1
-      }
-      return(count)
-    }
-
-    thresholds$zipf <- list(
-      prob_0001 = find_zipf_threshold(0.0001),
-      prob_00001 = find_zipf_threshold(0.00001)
-    )
-  }
-
-  # b) For ZIGP
-  if (!is.null(model_params$ZIGP)) {
-    eta <- model_params$ZIGP$eta
-    lambda <- model_params$ZIGP$lambda
-    theta <- model_params$ZIGP$theta
-
-    # Find where probability becomes small
-    find_zigp_threshold <- function(p_threshold) {
-      count <- 1
-      while(count < 100) {  # Reasonable upper limit
-        prob <- (1-eta) * (lambda * (lambda + theta * count)^(count - 1) * 
-                 exp(-lambda - theta * count)) / factorial(count)
-        if (prob < p_threshold) break
-        count <- count + 1
-      }
-      return(count)
-    }
-
-    thresholds$zigp <- list(
-      prob_0001 = find_zigp_threshold(0.0001),
-      prob_00001 = find_zigp_threshold(0.00001)
-    )
-  }
-
-  # 3. Count-based approaches
-  # Find natural break points in the distribution
-  natural_breaks <- find_natural_breaks(counts, freqs)
-  thresholds$breaks <- natural_breaks
-
-  # Return all thresholds
-  return(thresholds)
-}
-
-# Helper function to identify natural break points in count distribution
-find_natural_breaks <- function(counts, freqs) {
-  # Calculate rate of change between consecutive counts
-  rate_change <- numeric(length(counts) - 1)
-  for (i in 1:(length(counts)-1)) {
-    if (counts[i+1] - counts[i] == 1) {  # Only consider consecutive integers
-      rate_change[i] <- freqs[i] / freqs[i+1]
-    } else {
-      rate_change[i] <- NA  # Mark non-consecutive values
-    }
-  }
-
-  # Find points where rate of change is significantly higher than average
-  valid_rates <- rate_change[!is.na(rate_change)]
-  if (length(valid_rates) < 3) return(list(break_points = NULL))
-
-  mean_rate <- mean(valid_rates)
-  sd_rate <- sd(valid_rates)
-
-  # Points where rate of change is > mean + 1.5*SD
-  break_indices <- which(rate_change > mean_rate + 1.5*sd_rate)
-  break_points <- counts[break_indices]
-
-  # Return the identified break points
-  return(list(
-    break_points = break_points,
-    rate_change = rate_change,
-    suggested_threshold = if(length(break_points) > 0) min(break_points) + 1 else NULL
-  ))
-}
-
-thresholds <- calculate_thresholds(counts, freqs, n_zeros, results$parameters)
-print(thresholds)
-## empirical
-## $empirical$mean_plus_2sd
-## [1] 4
-
-## $empirical$mean_plus_3sd
-## [1] 5
-
-## $empirical$mean_plus_5sd
-## [1] 7
-
-## $empirical$percentile_99
-## [1] 1
-
-## $empirical$percentile_999
-## [1] 2
-
-## $zipf
-## $zipf$prob_0001
-## [1] 1
-
-## $zipf$prob_00001
-## [1] 2
-
-## $zigp
-## $zigp$prob_0001
-## [1] 1
-
-## $zigp$prob_00001
-## [1] 1
-
-## $breaks
-## $breaks$break_points
-## [1] 1
-
-## $breaks$rate_change
-##  [1] 694.980861   4.696629   3.068966   1.318182   1.571429   1.272727   1.571429   1.400000   1.250000   1.333333   1.000000   1.000000   0.750000   4.000000   1.000000   1.000000
-## [17]   1.000000         NA   2.000000         NA   1.000000   0.500000         NA         NA         NA         NA   1.000000         NA         NA   0.500000   2.000000         NA
-## [33]   1.000000         NA         NA         NA         NA         NA
-
-## $breaks$suggested_threshold
-## [1] 2
-
-find_natural_breaks (counts, freqs)
-## $break_points
-## [1] 1
-
-## $rate_change
-##  [1] 694.980861   4.696629   3.068966   1.318182   1.571429   1.272727   1.571429   1.400000   1.250000   1.333333   1.000000   1.000000   0.750000   4.000000   1.000000   1.000000
-## [17]   1.000000         NA   2.000000         NA   1.000000   0.500000         NA         NA         NA         NA   1.000000         NA         NA   0.500000   2.000000         NA
-## [33]   1.000000         NA         NA         NA         NA         NA
-
-## $suggested_threshold
-## [1] 2
-
-```
-
-```{R, echo=FALSE, eval=FALSE}
-# more refined threshholding
-
-# Robust Empirical Thresholding for Interaction Counts
-# This approach works for both well-fitted distributions and those that don't follow 
-# standard theoretical distributions
-
-empirical_threshold <- function(counts, freqs, n_zeros = 0) {
-  # Create a data frame for easier analysis
-  df <- data.frame(
-    count = counts,
-    freq = freqs,
-    prop = freqs / sum(freqs)
-  )
-
-  # Order by count
-  df <- df[order(df$count), ]
-
-  # Calculate cumulative proportion (excluding zeros)
-  df$cum_prop <- cumsum(df$prop)
-
-  # Calculate several thresholds
-  thresholds <- list()
-
-  # 1. Basic statistics on positive counts
-  mean_pos <- sum(df$count * df$freq) / sum(df$freq)
-  var_pos <- sum(df$freq * (df$count - mean_pos)^2) / sum(df$freq)
-  sd_pos <- sqrt(var_pos)
-
-  # 2. Compute empirical percentiles
-  weighted_data <- rep(df$count, times = df$freq)
-
-  thresholds$basic <- list(
-    mean = mean_pos,
-    sd = sd_pos,
-    threshold_2sd = ceiling(mean_pos + 2 * sd_pos),
-    threshold_3sd = ceiling(mean_pos + 3 * sd_pos),
-    p95 = as.numeric(quantile(weighted_data, 0.95)),
-    p99 = as.numeric(quantile(weighted_data, 0.99)),
-    p999 = as.numeric(quantile(weighted_data, 0.999))
-  )
-
-  # 3. Detect "elbows" or breakpoints in the distribution
-  # Calculate rate of change in frequency between consecutive counts
-  df$rate_change <- NA
-  for (i in 1:(nrow(df)-1)) {
-    if (df$count[i+1] - df$count[i] == 1) {  # Only for consecutive counts
-      df$rate_change[i] <- df$freq[i] / df$freq[i+1]
-    }
-  }
-
-  valid_rates <- df$rate_change[!is.na(df$rate_change)]
-  if (length(valid_rates) >= 3) {
-    mean_rate <- mean(valid_rates)
-    sd_rate <- sd(valid_rates)
-
-    # Find significant jumps in rate of change
-    # Points where rate of change is > mean + 1.5*SD
-    break_indices <- which(df$rate_change > mean_rate + 1.5*sd_rate)
-    break_points <- df$count[break_indices]
-
-    thresholds$breakpoints <- list(
-      points = break_points,
-      suggested = if(length(break_points) > 0) min(break_points) + 1 else NULL
-    )
-  } else {
-    thresholds$breakpoints <- list(
-      points = NULL,
-      suggested = NULL
-    )
-  }
-
-  # 4. Entropy-based thresholding
-  # Calculate entropy of frequency distribution
-  p <- df$prop
-  entropy <- -sum(p * log(p))
-
-  # Find threshold that maximizes entropy difference between below and above threshold
-  max_diff <- -Inf
-  best_threshold <- 1
-
-  for (i in 1:max(df$count)) {
-    below <- df[df$count <= i, ]
-    above <- df[df$count > i, ]
-
-    # Skip if either group is empty
-    if (nrow(below) == 0 || nrow(above) == 0) next
-
-    # Normalize probabilities within each group
-    p_below <- below$freq / sum(below$freq)
-    p_above <- above$freq / sum(above$freq)
-
-    # Calculate entropies
-    entropy_below <- -sum(p_below * log(p_below))
-    entropy_above <- -sum(p_above * log(p_above))
-
-    # Calculate weighted entropy difference
-    entropy_diff <- (sum(below$freq) * entropy_below + sum(above$freq) * entropy_above) / sum(df$freq)
-
-    if (entropy - entropy_diff > max_diff) {
-      max_diff <- entropy - entropy_diff
-      best_threshold <- i
-    }
-  }
-
-  thresholds$entropy <- list(
-    threshold = best_threshold + 1  # Add 1 for exclusive threshold
-  )
-
-  # 5. Exponential decay test
-  # Test if counts follow exponential decay by fitting log(frequency) ~ count
-  if (nrow(df) > 5) {
-    log_freq <- log(df$freq)
-    model <- lm(log_freq ~ df$count)
-    r_squared <- summary(model)$r.squared
-
-    # If strong exponential decay, use where predicted frequency < 1
-    if (r_squared > 0.8) {
-      coefs <- coef(model)
-      threshold_exp <- ceiling(-coefs[1] / coefs[2])  # Where predicted freq < 1
-      thresholds$exponential <- list(
-        r_squared = r_squared,
-        threshold = threshold_exp
-      )
-    } else {
-      thresholds$exponential <- list(
-        r_squared = r_squared,
-        threshold = NULL
-      )
-    }
-  }
-
-  # 6. Rarity-based threshold
-  # Define thresholds based on observation frequency
-  total_observations <- sum(df$freq)
-  rare_threshold <- ceiling(total_observations * 0.001)  # 0.1% of observations
-
-  rare_counts <- df[df$freq <= rare_threshold, "count"]
-  if (length(rare_counts) > 0) {
-    thresholds$rarity <- list(
-      threshold = min(rare_counts)
-    )
-  } else {
-    thresholds$rarity <- list(
-      threshold = NULL
-    )
-  }
-
-  # 7. Add zero-inflation context if provided
-  if (n_zeros > 0) {
-    # Calculate proportion of zeros
-    p_zero <- n_zeros / (n_zeros + sum(df$freq))
-    thresholds$zero_inflation <- list(
-      p_zero = p_zero
-    )
-
-    # For extreme zero-inflation (>99.9%), more aggressive thresholds
-    if (p_zero > 0.999) {
-      # With extreme zeros, any occurrence might be significant
-      thresholds$extreme_zero_inflation <- list(
-        threshold = 2  # Consider anything geq 2 significant when zeros dominate
-      )
-    }
-  }
-
-  # 8. Final recommended threshold based on all methods
-  # Logic for combining thresholds
-  candidates <- c(
-    thresholds$basic$threshold_2sd,
-    thresholds$breakpoints$suggested,
-    thresholds$entropy$threshold,
-    thresholds$rarity$threshold
-  )
-
-  # Add exponential threshold if available
-  if (!is.null(thresholds$exponential$threshold)) {
-    candidates <- c(candidates, thresholds$exponential$threshold)
-  }
-
-  # Add extreme zero-inflation threshold if applicable
-  if (!is.null(thresholds$extreme_zero_inflation$threshold)) {
-    candidates <- c(candidates, thresholds$extreme_zero_inflation$threshold)
-  }
-
-  # Remove NAs
-  candidates <- candidates[!is.na(candidates)]
-
-  # For final recommendation, use median of available thresholds
-  if (length(candidates) > 0) {
-    thresholds$recommended <- median(candidates, na.rm = TRUE)
-  } else {
-    # Default if no thresholds available
-    thresholds$recommended <- max(2, ceiling(mean_pos + 2 * sd_pos))
-  }
-
-  # Provide both results and visualization
-  results <- list(
-    thresholds = thresholds,
-    data = df
-  )
-
-  # Return everything
-  return(results)
-}
-
-# Function to visualize the thresholds on a count distribution
-plot_thresholds <- function(threshold_results, log_scale = TRUE) {
-  df <- threshold_results$data
-  thresholds <- threshold_results$thresholds
-
-  # Setup plot
-  par(mfrow = c(1, 1))
-
-  # Create bar plot of frequencies
-  if (log_scale) {
-    barplot(log(df$freq), names.arg = df$count, xlab = "Count", ylab = "Log Frequency", 
-            main = "Distribution of Counts with Thresholds", col = "lightblue")
-  } else {
-    barplot(df$freq, names.arg = df$count, xlab = "Count", ylab = "Frequency", 
-            main = "Distribution of Counts with Thresholds", col = "lightblue")
-  }
-
-  # Add recommended threshold line
-  abline(v = which(df$count == ceiling(thresholds$recommended)), col = "red", lwd = 2)
-
-  # Add other threshold lines
-  # Add 2SD threshold
-  if (!is.null(thresholds$basic$threshold_2sd)) {
-    pos <- which(df$count >= thresholds$basic$threshold_2sd)[1]
-    if (!is.na(pos)) {
-      abline(v = pos, col = "blue", lty = 2)
-    }
-  }
-
-  # Add breakpoint threshold
-  if (!is.null(thresholds$breakpoints$suggested)) {
-    pos <- which(df$count >= thresholds$breakpoints$suggested)[1]
-    if (!is.na(pos)) {
-      abline(v = pos, col = "green", lty = 2)
-    }
-  }
-
-  # Add entropy threshold
-  if (!is.null(thresholds$entropy$threshold)) {
-    pos <- which(df$count >= thresholds$entropy$threshold)[1]
-    if (!is.na(pos)) {
-      abline(v = pos, col = "purple", lty = 2)
-    }
-  }
-
-  # Add legend
-  legend("topright", 
-         legend = c("Recommended", "2SD", "Breakpoint", "Entropy"),
-         col = c("red", "blue", "green", "purple"),
-         lty = c(1, 2, 2, 2),
-         lwd = c(2, 1, 1, 1))
-
-  # Return the plot
-  invisible(threshold_results)
-}
-
-rheumatoid_thresholds <- empirical_threshold(counts, freqs, n_zeros)
-names(rheumatoid_thresholds$thresholds)
-#[1] "basic"                  "breakpoints"            "entropy"                "exponential"            "rarity"                 "zero_inflation"         "extreme_zero_inflation"
-#[8] "recommended"           
-rheumatoid_thresholds$thresholds$basic$p95$threshold_2sd
-rheumatoid_thresholds$thresholds$basic$threshold_2sd
-rheumatoid_thresholds$thresholds$breakpoints$suggested
-rheumatoid_thresholds$thresholds$entropy
-rheumatoid_thresholds$thresholds$exponential$threshold
-rheumatoid_thresholds$thresholds$rarity$threshold
-rheumatoid_thresholds$thresholds$zero_inflation
-rheumatoid_thresholds$thresholds$extreme_zero_inflation$threshold
-rheumatoid_thresholds$thresholds$recommended
-
-plot_thresholds(rheumatoid_thresholds)
-print(rheumatoid_thresholds$thresholds$recommended)
 

### 2025-05-03 manuscript.tex
removed all the commented out text from manuscript.tex

@@
 
-%% \fntext[label3]{}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{}
-%% \affiliation[label1]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-%%
-%% \affiliation[label2]{organization={},
-%%             addressline={},
-%%             city={},
-%%             postcode={},
-%%             state={},
-%%             country={}}
-
 \author{Rob Dunne} %% Author name
@@
 
-% %%Graphical abstract
-% \begin{graphicalabstract}
-% %\includegraphics{grabs}
-% \end{graphicalabstract}
-
 %%Research highlights
@@
 Epistatic interactions \sep  Random Forests \sep SNP data
-
-%% keywords here, in the form: keyword \sep keyword
-
-%% PACS codes here, in the form: \PACS code \sep code
-
-%% MSC codes here, in the form: \MSC code \sep code
-%% or \MSC[2008] code \sep code (2000 is the default)
-
 \end{keyword}
@@
 \end{frontmatter}
-
-%% Use a table environment to create tables.
-%% Refer following link for more details.
-%% https://en.wikibooks.org/wiki/LaTeX/Tables
-%\begin{table}[t]%% placement specifier
-%% Use tabular environment to tag the tabular data.
-%% https://en.wikibooks.org/wiki/LaTeX/Tables#The_tabular_environment
-%\centering%% For centre alignment of tabular.
-%\begin{tabular}{l c r}%% Table column specifiers
-%% Tabular cells are separated by &
-%  1 & 2 & 3 \\ %% A tabular row ends with \\
-%  4 & 5 & 6 \\
-%  7 & 8 & 9 \\
-%\end{tabular}
-%% Use \caption command for table caption and label.
-%\caption{Table Caption}\label{fig1}
-%\end{table}
-
-%% Use figure environment to create figures
-%% Refer following link for more details.
-%% https://en.wikibooks.org/wiki/LaTeX/Floats,_Figures_and_Captions
-%\begin{figure}[t]%% placement specifier
-%% Use \includegraphics command to insert graphic files. Place graphics files in 
-%% working directory.
-%\centering%% For centre alignment of image.
-%\includegraphics{example-image-a}
-%% Use \caption command for figure caption and label.
-%\caption{Figure Caption}\label{fig1}
-%% https://en.wikibooks.org/wiki/LaTeX/Importing_Graphics#Importing_external_graphics
-%\end{figure}
-
-%% For citations use: 
-%%       \cite{<label>} ==> [1]
-
-%%
-%%Example citation, See \cite{lamport94}.
-
-%% If you have bib database file and want bibtex to generate the
-%% bibitems, please use
-%%
-%%  \bibliographystyle{elsarticle-num} 
-%%  \bibliography{<your bibdatabase>}
-
-%% else use the following coding to input the bibitems directly in the
-%% TeX file.
-
-%% Refer following link for more details about bibliography and citations.
-%% https://en.wikibooks.org/wiki/LaTeX/Bibliography_Management
-
-%\endinput
-%%
-%% End of file `elsarticle-template-num.tex'.
-
-%\onecolumn % for one column layouts
-
-%\usepackage{showframe}
-
-%\theoremstyle{thmstyleone}%
-%\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-%\newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-%\theoremstyle{thmstyletwo}%
-%\newtheorem{example}{Example}%
-%\newtheorem{remark}{Remark}%
-%\theoremstyle{thmstylethree}%
-%\newtheorem{definition}{Definition}
-
-%\usepackage{capton}
-%\usepackage{natbib}
-%\usepackage[backend=biber,style=apa,defernumbers=true]{biblatex}
-%\addbibresource{./manuscript.bib}
-%\journal{Nuclear Physics B}
-
-%\begin{document}
-
-%\journaltitle{Briefings in Bioinformatics}
-%\DOI{DOI HERE}
-%\copyrightyear{2019}
-%\pubyear{2019}
-%\access{Advance Access Publication Date: Day Month Year}
-%\appnotes{Paper}
-
-%\firstpage{1}
-
-%\subtitle{Subject Section}
-
-%\corresp[$\ast$]{Corresponding author. \href{email:hawlader.almamun@data61.csiro.au}{hawlader.almamun@.csiro.au}}
-
-%\received{Date}{0}{Year}
-%\revised{Date}{0}{Year}
-%\accepted{Date}{0}{Year}
-
-%\editor{Associate Editor: Name}
-
-%\abstract{
-%\textbf{Motivation:} .\\
-%\textbf{Results:} .\\
-%\textbf{Availability:} .\\
-%\textbf{Contact:} \href{name@bio.com}{name@bio.com}\\
-%\textbf{Supplementary information:} Supplementary data are available at \textit{Briefings in Bioinformatics}
-%online.}
-
-% \abstract{Detecting epistatic interactions is very important because it will help us to understand the genetic mechanism
-%   on phenotypic variation. Finding epistatic interactions from high density SNP data is very challenging due to the
-%   computational burden induced by the combinatorial number of tests that have to be performed. Random Forests models were
-%   widely used in a variety of disciplines including epistatic interactions. Although Random Forests models account for
-%   variable interactions, it does not explicitly detect variable interactions. And most of the Random Forests based
-%   methods to detect interactions relay on some kind of variable importance measures that suffer when the interacting
-%   variables have very small or no marginal effects. We proposed a Random Forests based method that is not based on
-%   variable importance and can find epistatic interactions from high density SNP markers in reasonable time. The proposed
-%   method was first applied on several simulation datasets representing different data structures with respect to SNP
-%   markers and heritibilities. Once the method was validated for finding epistatic interactions from simulated SNP data,
-%   it was applied on two different traits (case-control data rheumatoid arthritis, and quantitative traits of mouse body
-%   weight) in human and mouse, respectively. In both cases, the method identified several genes closely positioned to the
-%   interacting SNPs that potentially have effects on the respective traits.}  \keywords{Epistatic interactions, Random
-%   Forest, SNP data}
-
-% \boxedtext{
-% \begin{itemize}
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \end{itemize}}
-
-%\maketitle : Missing \endcsname inserted.
 
@@
 economically important traits such as wheat yield or carcass weight of cattle.
-
-% The most direct approach to detect epistatic interactions is to evaluate pair-wise linear models (LM) for all
-% possible locus pairs (usually SNP pairs) to test for non-additive interactions associated with a complex trait
-% \cite[]{Zhang.et.al.2008}. This type of exhaustive search is only feasible when the number of candidate SNPs is
-% relatively small. However, as the number of SNP markers in a dataset grows, this approach suffers from the computational
-% burden induced by the combinatorial number of tests that have to be performed. For example, The Welcome Trust Case
-% Control studies \cite[]{burtonGenomewideAssociationStudy2007brief} has 17,000 individuals (2,000 cases for each of the
-% seven diseases and 3,000 combined controls) genotyped with the GeneChip 500K Mapping Array Set (Affymetrix chip), which
-% interrogates 500,568 SNPs. Thus more than $1.25\times 10^{13}$ pairwise tests would need to be performed, which is not
-% computationally feasible. Moreover, recent technological advancements and reduced costs have led to experimental designs
-% utilising more than 1 million SNP markers for hundreds of individuals, which massively confound the ability to identify
-% interacting pairs of SNPs \cite[]{Ha.et.al2014, Gholami.et.el2014,1000GenomeProject.2015.brief,utah.edu}. Consequently,
-% two step-based methods have been proposed to detect epistatic interactions from genomic data, where in the first step
-% SNPs are pre-selected or weighted and then in the second step interaction testing is performed on the pre-selected SNPs
-% only. The preselection can be performed based on statistical tests (e.g., GWAS \cite[]{Pecanka.et.al.2017}) or machine
-% learning methods with feature selection \cite[]{Jiang.et.al.2009, Meng.et.al.2007}.
 
@@
 
-% Random Forests are an ensemble learning method first proposed by Leo Breiman \cite[]{Breiman2001}. Ensemble learning
-% combines multiple learning algorithms/models to improve prediction or classification performance beyond what could be
-% obtained from the constituent models.  In a Random Forests, \textit{m} random samples (with replacement) are selected by
-% bootstrap sampling. Then multiple decision trees are created based on the bootstrap samples. To break
-% down the correlation between features, only a random subset of features is considered at any node of the tree building
-% process. For classification problems, the class of the validation data is predicted based on majority voting and for the
-% regression problem, the model outputs the mean of the individual trees prediction. There are two main parameters in the
-% implementation of Random Forests algorithm namely the number of decision trees (\texttt{num.trees}) and the number of
-% randomly selected features (\texttt{mtry}) from which the best split is selected at any given node.
-
 Random Forests (RF) algorithms are an ensemble learning method first proposed by Leo Breiman \cite[]{Breiman2001}. Ensemble
@@
 not interacting''}.
-
-% This observation was used to narrow down the vast search space of pairwise tests to detect SNP
-% interactions. First, we built a Random Forests and for each pair of SNPs we count the total number of times any SNP pair
-% occurred as parent-child nodes in this forest. Then we selected only the SNP pairs that occurred more times that would
-% be expected by chance and exceeded a set threshold of \textit{t}. In our study, we selected all the SNP pairs those have
-% at least two (\textit{t}=2) appearance as parent-child nodes in the decision trees of the Random Forests.  Finally, using
-% a linear model, each selected SNP pair was tested to determine whether the interaction was significant.
 
@@
 
-% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model was created using all avaiable data. Then for each pair 
-% of SNP, how many times they appeared as parent-child node in each of the decision trees were recorded. Then the SNP pairs were ranked based on
-% the count and only the pairs having count >= 2 were retained. Finally, using the linier equation the interaction term was tested. In our
-% studies, we used the ranger R package \cite[]{Wright.and.Ziegler.2017} and we selected \texttt{num.tree} = 1000 and \texttt{mtry} =
-% \#SNPs/3.
-% For a linear model, an interaction occurs when an independent variable has a different effect on the outcome depending
-% on the values of another independent variable. So our variables in the model can be divided between main effects and
-% interaction effects. A multiplicative epistasis model was used, where the interaction between two SNPs, $i$ and $j$,  were modelled as the
-% product of the two genotype values, using the following linear regression model:
-
 Then the marker pairs were ranked based on the count and only the pairs having count
@@
 through a logit function as follows:
-
-%Here $y$ is the qualitative measure of phenotype, and $x$ are the genotype values \{0, 1, 2\} for SNP $i$ and $j$
-%respectively, $\beta_0$ is the mean phenotype, and $\{\beta_i, \beta_j\}$ are the marginal effects for SNP $i$ and $j$,
-%$\beta_{i,j}$ is the interaction effect between SNP $i$ and $j$, and $\epsilon$ is the residual with N(0,1).  For
-%case-control disease data, the interaction was modelled through a logit function as follows:
 
@@
 \subsection{Simulation data}
-% We considered two-way interactions between two SNPs based on three interaction models. In the ``base model'' neither the
-%  SNPs nor their interaction have any effect on the phenotype; in the interaction only model both SNP have no marginal
-% effects but have interaction effect, and in the ``no interaction'' model both SNPs have their marginal effect but there is
-% no additional interaction effect. We used publicly available genotype data of 3,534 animals from a single PIC nucleus
-% pig line genotyped using the Illumina ProcineSNP60 chip \cite[]{clevelandCommonDatasetGenomic2012}.
-
-% The original dataset contained 52,843 SNPs, but in our studies, the non-informative SNPs were removed leaving 48,015
-% SNPs. From these genotype data, the simulated data was created by randomly selecting 5,000, 10,000 and 48,000
-% SNPs. Using these genotypes, simulated phenotypes were then created having different heritabilities (h2 = 0.33, 0.50,
-% 0.66 and 1.00). Data were generated using 3 different models:
-% \begin{itemize}
-% \item \textbf{Base model} phenotype = sum of ten QTL effects + noise (depending on heritability);
-% \item \textbf{no-interaction model} we added the effect of another 6 SNPs marginal effects on the top of 10 QTL
-%   effects, but there were no interaction effects between these 3 SNP pairs. phenotype = sum of ten QTL effects + sum of
-%   marginal effect of epistatic SNPs + noise (depending on heritability);
-% \item \textbf{interaction model} where phenotypes were the sum of 10 QTL effects plus pairwise interaction effects of
-%   the 6 SNPs. phenotype = sum of ten QTL effects + sum of interaction effect + noise (depending on
-%   heritability).
-% \end{itemize}
 
@@
 
-% \subsection{The Proposed Method -- count co-occurrences}
-% To overcome the challenge of a large number of pairwise interaction tests, we proposed an effective two-step process
-% based on our observations from using Random Forests. Using different simulation data, we observed that ``if two SNPs are
-% interacting, they appear as parent-child nodes in a decision tree of a Random Forests more frequently than if they are
-% not interacting''.  So, we utilised this observation to narrow down the vast search space of pairwise tests to detect
-% interactions. First, we build a Random Forests and for each pair of SNPs we count the total number of times any SNP pair
-% occurred as parent-child nodes in this forest. Then we select only the SNP pairs that occur more times that would be
-% expected by chance and exceed the set threshold. Finally, using a linear model, each selected SNP pair is tested to see
-% if the interaction is significant.
-
 \section{Results and Discussion}
 \subsection{Simulated data}
-% Based on publicly available pig genotype data, simulated phenotypes using different scenarios with various combinations
-% of number of SNPs, number of interacting SNP pairs, heritability, QTL effect size and interaction effect size were used.
-% The performance of the method was assessed based on the proportion of time the method selected the interacting SNP
-% pairs for pairwise interaction tests. Table \ref{table:ranking10qtl} shows the importance ranking of three interacting
-% SNP pairs based on the number of times they appear as parent-child in the Random Forests decision trees. Higher ranking,
-% represented by a smaller number, means a higher count. For example, a value of 1 for pair P3 means that SNP pair 3
-% appeared as parent-child more than any other SNP pair in that Random Forests. Since in the base model three pairs of
-% interacting SNPs do not have any marginal effects or any interaction effects, they did not appear as parent-child in any
-% decision tree of the Random Forests. In the no interaction model, the three interacting SNP pairs appeared as
-% parent-child in some of the scenarios. This is because in the no interaction model, although the three pairs do not
-% have any interaction effect, they have their own marginal effects. Hence, they were selected at different nodes of the
-% decision trees. For the interaction model, rankings of the three interacting SNP pairs are relatively higher than the
-% other two models. This ranking table demonstrates that if we rank the SNP pairs based on the count of their appearance
-% as parent-child, we only need to test pairwise interaction for a small amount of SNP pairs from the vast number of total
-% possible combinations of SNP.
-
 The method's performance was assessed by determining the number of interactions that were correctly identified and
@@
 
-% In the proposed two step method for detecting SNP-SNP interactions, a random forests model was built and the count
-% determined for how many times a given pair of SNPs appears as parent-child. In the second step, only those SNP pairs
-% having the counts higher than a certain number t were tested. In this simulation, we set t = 2, meaning that if any of
-% the SNP pairs appeared as parent-child in the decision trees of the Random Forests for at least 2 times, we selected them
-% to test for pairwise interaction using Equation 1. The third column of Table
-% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of pair-wise interaction tests performed
-% with the interaction model for different combinations of heritability and number of SNPs. The fourth column of Table
-% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of possible pairwise tests for a given
-% number of SNPs. From the last column of the Table \ref{table:Number.of.tests.performed.to.detect.interactions} we can
-% see that, using our method we were able to narrow down the search space of pairwise test significantly. For all
-% scenarios, less than one percentage of SNP pairs were tested for interaction.
-
 \begin{table}
@@
 \end{table}
-
-% \subsection{Effect of heritability and interactions p-value on the proposed method to detect candidate SNP pairs}
-% Figure \ref{figure:Eff_Trait_h2_simulated_data.png} shows the pairwise counts for each of the interacting SNP pairs in
-% each Random Forests at different heritabilities (h = 0.33, 0.50, 0.66, and 1.00) for 10,000 SNP with 10 QTL and 3 pairs
-% of interacting SNPs. At low heritability, the pairwise count is smaller and as heritability increases, the pairwise
-% count also increases. Although the three interacting SNP pairs have very similar interaction effects (ranged between 0.3
-% to 0.4), they have very different numbers of counts. There is a direct relationship between the $p$-value of the
-% interaction term with the pairwise count. Lower $p$-values of the interaction terms have higher pairwise count
-% (correlation coefficient = 0.84). This is particularly interesting as even very small interaction effects with high
-% significance will not be overlooked because of higher pairwise count.
 
@@
 \subsection{Repeatability and false discovery rate}
-% To test repeatability of our proposed method, we ran the process at different heritabilities (0.75, 0.66, 0.50, 0.33,
-% and 0.25) with ten repeats for each heritability. We evaluated the performance of the process based on how many times it
-% was successful to select any interacting SNP pair. Table \ref{table:Repeatability_of_the_proposed_method} shows the
-% results for 10,000 SNPs with 10 QTLs and 3 pairs of interacting SNPs. In most of the scenarios tested, the method was
-% able to successfully select the interacting SNP pairs except for interacting pair one at low heritability.  This is
-% expected as at low heritability, there was more noise induced in the data.
 
@@
 
-% \begin{table}
-% \begin{center}
-%   \begin{tabular}{|cccc|}
-%     \hline
-%     &P& P2& P\\
-%     H2=0.75&10& 10& 10\\
-%     H2=0.66&10& 10& 10\\
-%     H2=0.50&9&10 &10\\
-%     H2=0.33&7&10 &10\\
-%     H2=0.25&4&10 &10\\
-%     \hline          
-%   \end{tabular}
-% \end{center}
-% \caption{Repeatability of the proposed method. The data contains 10,000 with 10 QTLS and 3 pairs of interacting SNP. The
-%   process was repeated 10 times for each heritability times for each heritability.}
-% \label{table:Repeatability_of_the_proposed_method}
-% \end{table}
-
 \begin{table}
@@
 \section{Results for real datasets}
-% Once the method demonstrated that pairwise SNP-SNP interactions could be identified in the simulated data, we then applied
-% the method to two different real datasets.
-
 To evaluate the method's performance on real data, two real datasets were assessed for the presence of interacting SNP
@@
 \subsubsection{Rheumatoid arthritis data}
-% After pre-processing and filtering, the RA dataset contained 94,486 SNPs (~4.46 billion possible pairwise interaction
-% tests) and 4,798 individuals.  Using Random Forests, we have selected only 654 SNP pairs (out of more than 4.46 billion
-% possible pairs) for pairwise interaction tests (Equation 2). Among the 654 SNP pairs, 26 pairs have significant
-% Bonferroni corrected interaction $p$-values ($p < 0.05$). In detail, 24 SNPs were involved in 26 significant interactions
-% among which 22 SNPs were from chromosome 6 and two other from chromosome 1 and chromosome 7. Furthermore, several SNPs
-% from chromosome 6 were involved in multiple interactions. All the 22 SNPs in chromosome 6 are in the HLA (human
-% leukocyte antigen) region spanning from 31.35 MB to 32.80 MB. The HLA region in human genome is known to contain very
-% high linkage disequilibrium (LD). If interacting SNP pairs are in the same haplotype or LD block, the significant
-% interaction between the pairs might be a haplotype effect on RA, rather than true interaction. Out of 26 pairs of
-% interacting SNPs, 12 pairs (all are in chromosome 6) have moderate to high LD (D’ = (0.26 to 0.55)) between the
-% pairs. Hence, we retained only 14 SNP pairs which has low LD (D’ $\leq$ 0.2).
-
-% Table 10 lists 14 interacting SNP pairs that are associated with rheumatoid arthritis in humans. Genes
-% containing any of the interacting SNPs as well as flanking genes for intergenic SNP are also listed. All 14 significant
-% SNP pair interactions had at least one SNP of each pair positioned in the major histocompatibility complex (MHC) class
-% II centred on the HLA-DQ to HLA-DR regions located in chromosome 6p21. The HLA region in human chromosome is a 3.78 Mb
-% region that is the most gene dense region withing the human genome encoding about 253 genes including several key immune
-% response genes \cite[]{shiinaHLAGenomicLoci2009}. The genes in this region predominantly have immune response
-% functions. The dominance of the HLA-DQ to HLA-DR region in the interacting SNPs may be due to several factors including
-% the clustering of genes with key roles in immune response and its regulation, high gene density including some
-% overlapping and anti-sense genes, and a high rate of polymorphism and therefore high linkage disequilibrium. After
-% filtering of the SNP pairs for LD less than xx, eight SNP pairs remained. This result suggests that the dominance of the
-% identified SNP interactions in this subregion of the MHC class II region reflects strong association of SNPs with the
-% rheumatoid arthritis phenotype and not a consequence of the relatively high level of LD in the broader region. Hence,
-% all 14 SNP pair interactions were retained for subsequent analyses of gene and potentially SNP functions. Future fine
-% mapping of this region in chromosome 6p21 to localise causal SNP interaction pairs associated with rheumatoid arthritis
-% is likely to be difficult due to the reasons listed above. Therefore, other approaches are required to understand the
-% abundance of interactive SNPs in this subregion of the MHC class II region that are associated with rheumatoid
-% arthritis.  The normal function of the MHC class II cell surface receptors encoded by the HLA-DQ and HLA-DR genes is to
-% bind and present foreign antigens to T-cells, which then proliferate and signal to B-cells to produce antibodies against
-% potential threats. The antigens can be foreign and self-antigens. In the latter instance, immune tolerance to
-% self-antigens is normally developed at a very young age. Thus, activation of the MHC class II receptors can elicit or
-% suppress T-(helper)-cell responses. In some autoimmune conditions, however, self-antigens activate T-cells and generate
-% an inappropriate immune response resulting in chronic inflammation. The HLA-DQ and HLA-DR genes and genes within the
-% wider MHC class II region are associated with a variety of inflammatory autoimmune diseases including rheumatoid
-% arthritis, type 1 diabetes, systemic sclerosis, psoriasis, amkylosing spondylitis, systemic lupus erythematosus, Crohn’s
-% disease, dermatomyositis and Graves’ \cite{Kennedy.et.al.2017};
-% \cite{liuGenomewideScreenGenegene2011}; \cite{matzarakiMHCLocusGenetic2017}.
-
-% The current investigation has identified 14 pairs of interacting SNPs in the HLA-DQ - HLA-DR region on chromosome
-% 6p21. This finding suggests extensive cis-acting positive and negative interactions between SNPs in this region and
-% hence potential for positive and negative interactions between genes in this gene-dense region. How these interactions
-% occur is unclear. The identified SNPs may simply be genetic markers of closely linked causal genetic variants. However,
-% it is also possible that some of the interacting SNPs may be causal. None of the unique SNPs are present in gene coding
-% sequences, 6 are intronic and 14 are intergenic. Examination of the UCSC human genome (hg38 assembly) browser 
-% and its Regulation tracks for this region identified several of the interacting SNPs located
-% within potential long range gene regulatory elements defined by a variety signatures (i.e.,
-% \begin{itemize}
-% \item rs2734573, H3K4me1 and H3K27Ac (ENCODE Regulation track);
-% \item rs9296009, distal enhancer-like signature (enhD) (ENCODE Cis Regulatory Elements track);
-% \item   rs427037, enhD; rs910050, enhD; rs9268831, enhD;
-%   \item rs615672, regulatory elements and gene-gene interactions (GeneHancer Regulatory Elements and Gene Interactions
-%     track);
-%   \item rs9273363, transcription factor cluster and H3K4me1
-% (ENCODE Transcription Factor ChIP-seq Clusters track and ENCODE Regulation track, respectively)
-% \end{itemize}
-
-% % rs2734573, H3K4me1 and
-% % H3K27Ac (ENCODE Regulation track); rs9296009, distal enhancer-like signature (enhD) (ENCODE Cis Regulatory Elements
-% % track); rs427037, enhD; rs910050, enhD; rs9268831, enhD; rs615672, regulatory elements and gene-gene interactions
-% % (GeneHancer Regulatory Elements and Gene Interactions track); rs9273363, transcription factor cluster and H3K4me1
-% % (ENCODE Transcription Factor ChIP-seq Clusters track and ENCODE Regulation track, respectively).
-
-% Two pairs of interactive SNP (rs615672/rs9268831 and rs92688831/rs615672) contained both interactive SNP in potential
-% genome regulatory elements. Collectively, this information suggests that local gene-gene interactions in this region
-% could be mediated by cis-acting regulatory mechanisms involving short and long range DNA loops. Some of identified
-% interactive SNPs were located near or within introns of genes unrelated to the HLA-DQ and HLA-DR genes but interspersed
-% within the same region on chromosome 6p21. These genes included NOTCH4 (notch receptor 4), TNXB (tenascin XB) and BTNL2
-% (butryophilin-like 2). NOTCH4 regulates cell fate determination, proliferation and apoptosis programs and has a
-% regulatory role in branching morphogenesis in the developing vascular system. It is also implicated in the regulation of
-% inflammation \cite{Harb.et.el.2020}\cite{Zheng.et.el.2018}, a characteristic feature of autoimmune diseases such as
-% rheumatoid arthritis. TNXB (tenascin XB) encodes a tenascin family member of extracellular matrix proteins. It has
-% anti-adhesive effects and is involved in extracellular matrix maturation following wound healing \cite{GeneCards}. Thus,
-% the protein encoded by TNXB may potentially be involved in tissue repair after damage caused by chronic
-% inflammation. BTNL2 belongs to a family of immune regulators. It decreases T-cell proliferation and cytokine release
-% during an inflammatory response and has been implicated in a number autoimmune diseases
-% \cite{arnettBTNL2ButyrophilinB7Like2007}; \cite{GeneCards};
-% % Arnett et al., 2007; "GeneCards - Human Genes | Gene Database | GeneSearch," ;
-% \cite{lebrero-fernandezAlteredExpressionButyrophilin2016}. The SNP interaction partner to the SNP identified in BTNL2 on
-% 6p21 is located near NRF1 (nuclear respiratory facto 1) on chromosome 7. NRF1 encodes a transcription factor activating
-% key metabolic and cell growth genes as well as mitochondrial DNA transcription and replication. NRF1 has also been
-% implicated in the regulation of inflammation, \cite{GeneCards}; \cite{vantienenProlongedNrf1Overexpression2010};
-% \cite{yangNglycanaseNGLY1Regulates2018}. One possibility is that NRF1 acts in trans directly or indirectly via its
-% encoded transcription factor to regulate BTNL2 expression on chromosome 6p21 and thus modulate T-cell mediated immune
-% activation and inflammation.
-
-% Notably, multiple genome wide association studies (GWAS) identified some of the interactive SNPs as associated with
-% various autoimmune diseases i.e., rs9273363, rheumatoid arthritis; type 1 diabetes; rs6457617, multiple sclerosis,
-% Graves disease, systemic sclerosis; rs615672, rheumatoid arthritis ("UCSC Genome Browser Home,").
-
 After pre-processing and filtering, the Rheumatoid arthritis (RA) dataset contained 94,486 SNPs (4.46 billion possible
@@
 \subsection{Mouse  data}
-% As the curated mouse data contains 9,265 SNPs there were nearly 43 million possible pairwise comparisons to test for 
-% interactions. Using our method, 273,391 pairs of SNP (~0.0064\%) were tested for interactions. The approach identified 
-% 59 pairs of interacting SNPs after multiple testing using the Bonferroni correction. These 59 pairs of interactions were 
-% from 63 unique SNPs mainly located on MMU chromosomes 2, 8 and 13. We then identified the genes associated with each of 
-% these SNPs using the Mouse Genome Informatics \cite[]{mgi}.
-
 The curated mouse data contained 9,265 SNPs resulting in nearly 43 million possible pairwise comparisons to test for the
@@
 
-% \subsection{Gene functions potentially affected by interacting SNPs}
-% Supplementary Table S2 lists significant interacting pairs of SNPs that are associated with murine growth traits. The
-% nearest genes (mean distance = 36.7 Kbp; range = 42 bp – 286 Kbp) associated with these SNPs are also listed. The
-% associated MGI ID (Mouse Genome Informatics ID), Entrez ID and GO terms (Gene Ontology) are listed in Supplementary
-% Table S3. Using GSEA (C1 positional gene sets; \cite{Liberzon.et.al.2011}), genes with official gene symbols from Table
-% S2 were enriched at two human cytogenic bands equivalent to two syntenic murine cytogenetic band locations (mmu Chr8qA2,
-% FDR q = 1.5E-4, Tm2d2, Tacc1, Slc20a2, Fgfr1; mmu Chr2qA3-2qB boundary, FDR q = 8.2E-3, Mrps2, Ralgds, Ddx31, Ttf1).
-
-% The four genes located in mmu Chr8qA2 and two functionally anonymous genes (miR8108 and Gm31784) associated with two
-% additional SNPs are all present in a 3.8 Mbp interval. Tm2d2, Tacc1 and Fgfr1 are contained within a small region of
-% ~0.5 Mbp of this larger segment. The microRNA miR8108 is located within an intron of Tacc1 and Gm31784 is a poorly
-% defined ncRNA located 0.8 Mbp downstream of Fgfr1. Using syntenic human genes to the murine genes near all significant
-% SNPs in conjunction with the GSEA CGP database, there was enrichment (q = 2.1E-4) for TM2D2, TACC1, SLC20A2, FGFR1 in
-% the Nikolsky Breast Cancer dataset for genes in an amplicon from the human 8p12\_P11 chromosomal region susceptible to
-% copy number variation (CNV) in breast cancer samples.  Thus, while the six SNPs in the mmu Chr8qA2 region likely reflect
-% SNPs in linkage disequilibrium, another possibility is that the identified SNPs could be contained within or near a CNV
-% in this chromosomal region. Notably, a 96,111 bp CNV (1810011O10Rik) has been reported in this region in some mouse
-% strains, although it only incorporates Fgf1 \cite{lockeGenomicCopyNumber2015}.
-
-% The four identified and well characterised genes in the mmu Chr8qA2 region have a variety of functions that potentially impact a phenotype associated
-% with growth and development. Fgfr1 (fibroblast growth factor receptor 1) has wide ranging biological functions
-% associated with regulation of cell division, cell growth and maturation, embryonic development, wound healing, and
-% formation of blood vessels. In humans, the cellular signalling mediated by this receptor has roles in the development
-% and growth of bones, particularly bones in the head, face, hand, feet, and long bones of the arms and legs.  Mutations
-% in the gene are associated with human dwarfism (Genetics Home Reference, NIH US National Library of Medicine
-% \cite[]{medline}. Tacc1 (transforming acidic coiled-coil containing protein 1) is involved in processes
-% that promote cell division before the formation of differentiated tissues \cite{GeneCards}. Tacc1 encodes a protein that
-% regulates the localization of nuclear receptors, including T3 thyroid hormone and retinoic pathways involved in
-% regulation cell growth and differentiation. Tm2d2 (TM2 domain containing 2) encodes a G-coupled receptor that regulates
-% cell death and proliferation cascades \cite{GeneCards}. Slc20a2 (solute carrier family 20 member 2) encodes an inorganic
-% phosphate transporter important in maintaining phosphate homeostasis. This function is essential for all intracellular
-% function. The encoded protein also has roles in extracellular matrix and cartilage calcification, which is important for
-% tissue function \cite{GeneCards}. It is noted that growing bones have high demands for phosphate
-% \cite{Goretti.et.el.2012}.
-
-% The four genes enriched in a localised 0.25 Mbp region on the mmu chr 2qA3-2qB boundary
-% (Mrps2 (mammalian mitochondrial ribosomal protein), Ralgds (ral guanine nucleotide dissociation stimulator), Ddx31
-% (dead-box helicase 31), Ttf1 (transcription termination factor 1) have a variety of functions \cite{GeneCards} including
-% mitochondrial protein synthesis, intracellular signalling, cell growth and division in embryogenesis, and transcription,
-% respectively. At present, only Ddx31 is directly linked to growth related traits, although the molecular functions of
-% the remaining three genes may have this potential.
-
-% The remaining genes listed in Tables S2 and S3 map to various
-% chromosomal regions. Notable gene functions or mutational effects that could influence growth traits include: regulation
-% of cell proliferation, differentiation and survival (Klf7, Kruppel like factor 7); osteogenesis (Stmn2, Stathmin 2);
-% short stature and skeletal abnormalities (Wipi2, WD40 repeat protein interacting with phosphoinositides 2); increased
-% blood testosterone, precocious sexual maturation, increased fecundity, and clearance of luteinizing hormone (Chst8,
-% carbohydrate sulfo-transferase 8) \cite{GeneCards}). In all, there were several genes identified close to interacting
-% SNPs listed in Tables S2 and S3 that have effects on murine growth and developmental traits, particularly osteogenesis
-% and body stature.
-
 \section{Conclusion}
-% In this study, we proposed a simple and easy method to select candidate SNP pairs to test for interactions in large
-% genomic data. We used a Random Forests algorithm to select a subset of SNP pairs from a vast search space so that the
-% selected SNP pairs can be tested for interaction in a reasonable computational time. The method was validated with
-% several simulation datasets representing different data structures (with respect to SNP marker density and
-% heritabilities). We then used the method to identify SNP interactions for two different traits (case-control data
-% rheumatoid arthritis, and quantitative traits of mouse body weight) in human and mouse, respectively. In both cases, the
-% method identified several genes closely positioned to the interacting SNPs that potentially have effects on the
-% respective traits. Although the approach requires further biological validation, candidate SNP pair markers identified
-% by this method will provide valuable information and a genetic basis for understanding a complex trait that is the
-% result of effects from multiple genes and their complex interactions. We only used SNP data to demonstrate our proposed
-% method. However, the method is equally applicable when the datasets contain both SNP data and environmental data.
 
@@
 
+%%%%%%%%%%%%%%
+\appendix
+
 \section{Data availability}
@@
 
-%%%%%%%%%%%%%%
-\appendix
-
 \FloatBarrier
-\section{Section title of first appendix}
-\label{sec11}
-
 \section{Competing interests}
@@
 \end{document}
-
-% \begin{sidewaystable}
-% \begin{center}
-%   \begin{tabularx}{\textwidth}{|c|c|c|c|c|c|c|c|c|c|c|c|}
-%     \hline
-%                 SNP pair & D’         & r$^2$ & SNP1 location & \makecell{SNP1 \\$p$-value$^d$} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} & SNP2 location & \makecell{SNP2 odds \\ ratio (95\% CI) $^b$ } & \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} & \makecell{Interaction \\odds ratio} \\
-% \hline
-% rs3134926:rs910050       & 0.10       & 0.01  & 6:32232370    & 0.61                            & 0.55,0.6                                    & 2.35E-20      & 6:32347877                                    & 0.79                            & 0.72,0.86                                         & 8.40E-05                               & 1.77 \\
-% rs3134926:rs4530903      & 0.14       & 0.01  & 6:32232370    & 0.61                            & 0.55,0.67                                   & 2.35E-20      & 6:32614112                                    & 1.41                            & 1.24,1.59                                         & 7.78E-05                               & 1.64 \\
-% rs9431614:rs9296009      & 0.01       & 0.00  & 1:230944623   & 1.00                            & 0.89,1.11                                   & 1             & 6:32146738                                    & 1.80                            & 1.64,1.97                                         & 2.96E-33                               & 1.62 \\
-% rs3134926:rs9391858      & 0.06       & 0.00  & 6:32232370    & 0.61                            & 0.55,0.67                                   & 2.35E-20      & 6:32373621                                    & 1.24                            & 1.11,1.39                                         & 0.07                                   & 1.55 \\
-% rs10456058:rs9273363$^a$ & 0.08       & 0.00  & 6:31526961    & 1.11                            & 1.03,1.21                                   & 1             & 6:32658495                                    & 1.34                            & 1.23,1.46                                         & 4.01E-08                               & 1.50 \\
-% rs910050:rs6457617       & 0.17       & 0.02  & 6:32347877    & 0.79                            & 0.72,0.86                                   & 8.40E-05      & 6:32696074                                    & 0.44                            & 0.4,0.48                                          & 1.25E-69                               & 1.39 \\
-% rs9268831:rs615672       & 0.15       & 0.01  & 6:32459971    & 0.69                            & 0.63,0.75                                   & 3.98E-16      & 6:32606394                                    & 0.60                            & 0.54,0.66                                         & 9.19E-24                               & 1.34 \\
-% \hline
-% \end{tabularx}
-% \end{center}
-% \caption{Eight pairs of interacting SNPs have significant odds ratio after pairwise LD evaluation}
-% \label{table:eight.pairs}
-% \begin{tablenotes}%
-% \item[$^{a}$] rs10456058 has merged into rs2734573
-% \item[$^{b}$]  Odds ratio (OR) of having at least one copy of the risk allele compared to having 0 copies of the same allele for the given SNP
-% \item[$^{c}$]  Odds ratio (OR) of having at least one copy of each risk allele compared to having 0 copies of the same allele for the given SNP pair.
-% \item[$^{d}$]  Bonferroni adjusted p-value
-% \end{tablenotes}
-% \end{sidewaystable}
-
-%   \begin{tabularx}{\textwidth}{|c|c|c|c|c|c|c|c|c|c|c|c|}
-%     \hline
-%                 SNP pair & D’         & r$^2$ & SNP1 location & \makecell{SNP1 \\$p$-value$^d$} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} & SNP2 location & \makecell{SNP2 odds \\ ratio (95\% CI) $^b$ } & \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} & \makecell{Interaction \\odds ratio} \\
-% \hline
-%   \begin{tabular}{|c|p{1cm}|p{1cm}|p{2cm}|p{2cm}|p{1.5cm}| p{2cm}|p{2cm}|p{1.5cm}| p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
-%     \hline
-%     \hline
-%                 SNP pair & D’ & r$^2$   & SNP1 location & SNP1 odds ratio & SNP1 odds ratio (95\% CI)$^b$ & SNP1 p  value$^d$ &  SNP2 location & SNP2 odds ratio & SNP2 odds ratio (95\% CI) $^b$ & SNP2 pvalue$^d$   & Odds ratio for interaction (95\% CI) $^c$ & interaction p value $^d$   \\
-%                SNP pair & D’         & r$^2$ & SNP1 location & \makecell{SNP1 \\$p$-value$^d$} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} & SNP2 location & \makecell{SNP2 odds \\ ratio (95\% CI) $^b$ } & \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} & \makecell{Interaction \\odds ratio} \\
-
-%\onecolumn

### 2025-05-04 manuscript.tex
rearranged supp files for submission
changes to mauscript.tex

@@
 %% and the form \ead[url] for the home page:
-\title{Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests}
+\title{Detection of Ecstatic Interactions in High-Dimensional Genomic Data Using Random Forests}
 
-\author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
+\author[1,2,3]{Hawlader A. Al-Mamun\corref{cor1}}
 \ead{halmamun@intergrain.com}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
+\author[4]{Rob Dunne} %% Author name
+\author[5]{Ross L. Tellam}
+\author[6]{Klara Verbyla}
+
+\cortext[cor1]{Corresponding author}
+
+\affiliation[1]{organization={Agriculture \& Food, CSIRO},%Department and Organization
   addressline={2 - 40 Clunies Ross Street}, 
@@
 
-\affiliation{organization={InterGrain Pty Ltd.},%Department and Organization
+\affiliation[2]{organization={InterGrain Pty Ltd.},%Department and Organization
   addressline={19 Ambitious Link}, 
@@
 
-\affiliation{organization={School of Biological Sciences},%Department and Organization
+\affiliation[3]{organization={School of Biological Sciences},%Department and Organization
   addressline={The University of Western Australia}, 
@@
 
-\author{Rob Dunne} %% Author name
-\affiliation{organization={Data61, CSIRO},%Department and Organization
+\affiliation[4]{organization={Data61, CSIRO},%Department and Organization
   addressline={13 Garden Street}, 
@@
 
-\author{Ross L. Tellam}
-\affiliation{organization={CSIRO Agriculture and Food},%Department and Organization
+\affiliation[5]{organization={CSIRO Agriculture and Food},%Department and Organization
   addressline={306 Carmody Rd}, 
@@
 
-\author{Klara Verbyla}
-\affiliation{organization={The Center for Aquaculture Technologies},%Department and Organization
+\affiliation[6]{organization={The Center for Aquaculture Technologies},%Department and Organization
   addressline={8445 Camino Santa Fe, Suite 104}, 
@@
 There are several options described in  \cite{Gauran.et.al.2018}.
-Using the ``one stage procedure'' and $C_1$ constant (see supplementary information)
+Using the ``one stage procedure'' and $C_1$ constant (see supplementary file Supplementary_File_Rheumatoid_Data.Rmd)
 we get a $C$ value (equation \ref{eqn:C}) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.
@@
 $\mu=50.$ It would appear that these examples all have $\lambda$ and $\mu$ (the mean of the contaminating distribution) with a
-greater separation than we have in the Rheumatoid data. So we may have a more inteactable problem here.
+greater separation than we have in the Rheumatoid data. We may have a more intractable problem here.
 
-Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forests with more
-trees and we will observe more counts.  There is a trade-off here and as we are sampling a space with typically more
-than $10^{9}$ possible interactions, we have to increase the number of trees a lot to get a meaningful difference in the
-mean of the contaminating distribution. We have found the thresholding to give a larger set of significant variables
-than the empirical Bayes method. However it controls the FDR rate by testing each variable and computationally easier.
+Uniquely amongst estimation problems, we can move the distribution of our data. If we fit a Random Forests with more
+trees we will get a better estimate of the null distribution and also shift the comtaminating distribution to the
+right. However, as we are sampling a space with possibly more than $10^{10}$ possible interactions, we may have to increase the
+number of trees a lot to get a meaningful difference in the mean of the contaminating distribution.
+So while the Empirical Bayes method presented here can estimate the significane level, we have found that
+thresholding the counts can be computationally easier.
 
@@
 \begin{itemize}
-\item threshold the observed count data. We do this a number of ways to do this (breakpoints in the distribution,
-  entropy tests and otehrs. See the supplementary informaton) and then take the median. In
-  the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
-\item test each interaction with a linear model (see below).
+\item threshold the observed count data. We do this in number of ways  (breakpoints in the distribution,
+  entropy tests and others. See the supplementary information file Supplementary_File_Rheumatoid_Data.Rmd) and then take the median. In
+  the case of the Rheumatoid data, the selected breakpoint was 2 which gave us 222 interactions;
+\item test each interaction with a linear model as described below.
 \end{itemize}
@@
     \centering
-    \includegraphics[totalheight=8cm]{./figs/plot_zigp_comparison_fixed.png}
+    \includegraphics[totalheight=8cm]{plot_zigp_comparison_fixed.png}
     \caption{The observed and expected values. The 0 counts are not shown, and the missing bars are count values that
@@
     \centering
-    \includegraphics[totalheight=5cm]{./figs/Figure_1.png}
+    \includegraphics[totalheight=5cm]{Figure_1.png}
     \caption{The stages of the proposed method. (a) A Random Forests model is trained with all genomic markers. (b) A
@@
 demonstrated that if the SNP pairs are ranked based on the count of their appearance as parent-child nodes, it should be
-possible to only test pairwise interaction for a subset of marker pairs from the vast number of total possible combina-
-tions of markers.  To quantify the efficiency of the method in reducing the search space, the number of tests performed
+possible to only test pairwise interaction for a subset of marker pairs from the vast number of total possible combinations-
+ of markers.  To quantify the efficiency of the method in reducing the search space, the number of tests performed
 was compared to the number of markers and potential pairwise combinations. For this test, a threshold of $t = 2$ was
@@
   \centering
-\includegraphics[totalheight=5cm]{./figs/Figure_2.png}
+\includegraphics[totalheight=5cm]{Figure_2.png}
 \caption{ The effect of trait heritability on pairwise counts. P1, P2 and P3 are interacting SNP pair 1, 2 and 3 respectively.
@@
 when compared to pairs P2 and P3. Pair P1 appeared on the candidate lists in all instances but did not pass multiple
-test corrections several times when the heritabilities were low (ie 0.25, 0.33).  To assess the false discovery rate of
+test corrections several times when the heritabilities were low (i.e. 0.25, 0.33).  To assess the false discovery rate of
 the method, the number of false positive interactions were counted across the 10 replicates at the different

### 2025-05-04 epistatic_interactions.tex
changed names to epistatic_interactions.tex

--- 2025-05-04 manuscript.tex
@@
 \usepackage{placeins}
-\usepackage{tikz}
-\usetikzlibrary{calc}
+%\usepackage{tikz}
+%\usetikzlibrary{calc}
 \usepackage[figuresleft]{rotating}
-
+\usepackage{longtable, lscape, booktabs, array}
 %usepackage[nodisplayskipstretch]{setspace}\setstretch{1.5}
@@
 
-\author[1,2,3]{Hawlader A. Al-Mamun\corref{cor1}}
+\author[1]{Hawlader A. Al-Mamun\corref{cor1}\fnref{fn1}}
 \ead{halmamun@intergrain.com}
@@
 \author[5]{Ross L. Tellam}
-\author[6]{Klara Verbyla}
+\author[1]{Klara Verbyla, \fnref{fn2}}
 
@@
 
-\affiliation[2]{organization={InterGrain Pty Ltd.},%Department and Organization
-  addressline={19 Ambitious Link}, 
-  city={Bibra Lake},
-  postcode={6163}, 
-  state={WA},
-  country={Australia}}
+% \affiliation{organization={InterGrain Pty Ltd.},
+%   addressline={19 Ambitious Link}, 
+%   city={Bibra Lake},
+%   postcode={6163}, 
+%   state={WA},
+%   country={Australia}}
+% and
+% \affiliation{organization={School of Biological Sciences},%Department and Organization
+%   addressline={The University of Western Australia}, 
+%   city={Crawley},
+%   postcode={6009}, 
+%   state={WA},
+%   country={Australia}}
 
-\affiliation[3]{organization={School of Biological Sciences},%Department and Organization
-  addressline={The University of Western Australia}, 
-  city={Crawley},
-  postcode={6009}, 
-  state={WA},
-  country={Australia}}
+\fntext[fn1]{Currently at InterGrain Pty Ltd, 19 Ambitious Link, Bibra Lake, 6163, WA and
+  School of Biological Sciences, The University of Western Australia, Crawley, 6009, WA Australia}
 
-\affiliation[4]{organization={Data61, CSIRO},%Department and Organization
+\affiliation[4]{organization={Data61, CSIRO},
   addressline={13 Garden Street}, 
@@
 
-\affiliation[6]{organization={The Center for Aquaculture Technologies},%Department and Organization
-  addressline={8445 Camino Santa Fe, Suite 104}, 
-  city={San Diego},
-  postcode={92121}, 
-  state={California},
-  country={USA}}
+% \affiliation[6]{organization={The Center for Aquaculture Technologies},%Department and Organization
+%   addressline={8445 Camino Santa Fe, Suite 104}, 
+%   city={San Diego},
+%   postcode={92121}, 
+%   state={California},
+%   country={USA}}
+\fntext[fn2]{Currently at The Center for Aquaculture Technologies, 8445 Camino Santa Fe, Suite 104, San Diego, 92121,
+  California, USA}
+
+%%Research highlights
+\begin{highlights}
+\item The method detects epistatic interactions in high dimensional genomic data.
+\item Computationally efective and produces reliable results in the case of simulated and interpretable results in several real-world
+  data sets
+\end{highlights}
 
@@
 There are several options described in  \cite{Gauran.et.al.2018}.
-Using the ``one stage procedure'' and $C_1$ constant (see supplementary file Supplementary_File_Rheumatoid_Data.Rmd)
+Using the ``one stage procedure'' and $C_1$ constant (see supplementary file Supplementary\_File\_Rheumatoid\_Data.Rmd)
 we get a $C$ value (equation \ref{eqn:C}) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.
@@
 \item threshold the observed count data. We do this in number of ways  (breakpoints in the distribution,
-  entropy tests and others. See the supplementary information file Supplementary_File_Rheumatoid_Data.Rmd) and then take the median. In
+  entropy tests and others. See the supplementary information file Supplementary\_File\_Rheumatoid\_Data.Rmd) and then take the median. In
   the case of the Rheumatoid data, the selected breakpoint was 2 which gave us 222 interactions;
@@
  \begin{center}
-   \begin{tabularx}{\textwidth}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c||}
+   \begin{tabularx}{\textwidth}{|c|c|c|p{1.16cm}|p{0.8cm}|p{1.16cm}|c|c|p{0.8cm} |c|c|c|c|p{1cm}||}
  \hline
-  SNP pair & D'         & r$^2$ & SNP1 location & \makecell{SNP1 odds\\ ratio} & \makecell{SNP1 odds \\ ratio (95\% CI)$^b$} &
-  \makecell{SNP1 \\$p$-value$^d$} & SNP2 location & \makecell{SNP2 odds\\ ratio} & \makecell{SNP2 odds \\ ratio (95\% CI)$^b$ } &
-  \makecell{SNP2 \\$p$-value$^d$} & \makecell{interaction\\ Odds ratio} &\makecell{interaction Odds \\ratio (95\% CI)$^c$} & \makecell{interaction\\ $p$-value$^d$} \\
+ SNP pair & D' & r$^2$ & 
+      \rotatebox{60}{SNP1 location} & 
+      \rotatebox{60}{\makecell{SNP1 odds\\ ratio}} & 
+      \rotatebox{60}{\makecell{SNP1 odds ratio \\(95\% CI)\textsuperscript{b}}} &
+      \rotatebox{60}{SNP1 $p$-value\textsuperscript{d}} & 
+      \rotatebox{60}{SNP2 location} & 
+      \rotatebox{60}{\makecell{SNP2 odds\\ ratio}} & 
+      \rotatebox{60}{\makecell{SNP2 odds ratio \\ (95\% CI)\textsuperscript{b}}} &
+      \rotatebox{60}{SNP2 $p$-value\textsuperscript{d}} & 
+      \rotatebox{60}{\makecell{Interaction \\Odds ratio}} &
+      \rotatebox{60}{\makecell{Interaction Odds ratio\\ (95\% CI)\textsuperscript{c}}} & 
+      \rotatebox{60}{\makecell{Interaction\\ $p$-value\textsuperscript{d}}} \\
+      \hline
    rs3134926:rs910050     & 0.10 & 0.01 & 6:32232370  & 0.61 & 0.55,0.6  & 1.99E-20 & 6:32347877 & 0.79 & 0.72,0.86 &  7.11E-05& 1.77 & 1.53,2.06 & 3.38E-11 \\
@@
  rs3134926:rs9391858      & 0.06 & 0.00 & 6:32232370  & 0.61 & 0.55,0.67 & 1.99E-20 & 6:32373621 & 1.24 & 1.11,1.39 &  5.76E-02& 1.55 & 1.28,1.88 & 4.22E-03 \\
- rs10456058:rs9273363$^a$ & 0.08 & 0.00 & 6:31526961  & 1.11 & 1.03,1.21 & 5.97E+00 & 6:32658495 & 1.34 & 1.23,1.46 &  3.39E-08& 1.50 & 1.32,1.7  & 2.73E-07 \\
+ rs10456058:rs9273363$^a$ & 0.08 & 0.00 & 6:31526961  & 1.11 & 1.03,1.21 & 5.97   & 6:32658495 & 1.34 & 1.23,1.46 &  3.39E-08& 1.50 & 1.32,1.7  & 2.73E-07 \\
  rs910050:rs6457617       & 0.17 & 0.02 & 6:32347877  & 0.79 & 0.72,0.86 & 7.11E-05 & 6:32696074 & 0.44 & 0.4,0.48  &  1.06E-69& 1.39 & 1.22,1.59 & 4.32E-04 \\
  rs9268831:rs615672       & 0.15 & 0.01 & 6:32459971  & 0.69 & 0.63,0.75 & 3.37E-16 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 1.34 & 1.17,1.53 & 9.12E-03 \\
- rs2076533:rs1270689      & 0.03 & 0.00 & 6:32395750  & 0.49 & 0.45,0.54 & 4.60E-55 & 7:12961515 & 1.06 & 0.98,1.16 &  8.23E+01& 1.31 & 1.16,1.48 & 1.48E-02 \\
+ rs2076533:rs1270689      & 0.03 & 0.00 & 6:32395750  & 0.49 & 0.45,0.54 & 4.60E-55 & 7:12961515 & 1.06 & 0.98,1.16 &  8.23E+1& 1.31 & 1.16,1.48 & 1.48E-02 \\
  rs17421624:rs615672      & 0.18 & 0.01 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32606394 & 0.60 & 0.54,0.66 &  7.78E-24& 0.73 & 0.64,0.84 & 7.58E-03 \\
@@
  rs17421624:rs3134926     & 0.08 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.72 & 0.62,0.83 & 7.12E-03 \\
- rs17421624:rs427037      & 0.03 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:3224448  & 0.70 & 0.62,0.78 &  2.70E-07& 0.69 & 0.59,0.82 & 1.11E-0  \\
-rs9296009:rs4530903       & 0.10 & 0.00 & 6:32146738  & 1.80 & 1.64,1.97 & 2.51E-33 & 6:3261411  & 1.41 & 1.24,1.59 &  6.59E-05& 0.67 & 0.55,0.81 & 3.16E-0  \\
-rs206015:rs3134926        & 0.06 & 0.00 & 6:32214982  & 1.49 & 1.33,1.67 & 2.90E-09 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.67 & 0.55,0.82 & 4.66E-0  \\
+ rs17421624:rs427037      & 0.03 & 0.00 & 6:32098400  & 1.70 & 1.56,1.86 & 7.24E-31 & 6:3224448  & 0.70 & 0.62,0.78 &  2.70E-07& 0.69 & 0.59,0.82 & 1.11E-03  \\
+rs9296009:rs4530903       & 0.10 & 0.00 & 6:32146738  & 1.80 & 1.64,1.97 & 2.51E-33 & 6:3261411  & 1.41 & 1.24,1.59 &  6.59E-05& 0.67 & 0.55,0.81 & 3.16E-03  \\
+rs206015:rs3134926        & 0.06 & 0.00 & 6:32214982  & 1.49 & 1.33,1.67 & 2.90E-09 & 6:32232370 & 0.61 & 0.55,0.67 &  1.99E-20& 0.67 & 0.55,0.82 & 4.66E-03  \\
 \hline
@@
  \begin{tablenotes}%
- \item[$^{a}$] rs10456058 has merged into rs2734573.
- \item[$^{b}$]  Odds ratio (OR) of having at least one copy of the risk allele compared to having 0 copies of the same allele for the given SNP.
- \item[$^{c}$]  Odds ratio (OR) of having at least one copy of each risk allele compared to having 0 copies of the same allele for the given SNP pair.
- \item[$^{d}$]  Bonferroni adjusted p-value.
- \item[$^e$]   The pair of SNPs used as an example in section \label{section:A_linear_model_example}.
+ \item $^{a}$ rs10456058 has merged into rs2734573.
+ \item $^{b}$  Odds ratio (OR) of having at least one copy of the risk allele compared to having 0 copies of the same allele for the given SNP.
+ \item $^{c}$  Odds ratio (OR) of having at least one copy of each risk allele compared to having 0 copies of the same allele for the given SNP pair.
+ \item $^{d}$  Bonferroni adjusted p-value.
+ \item $^e$   The pair of SNPs used as an example in section \label{section:A_linear_model_example}.
  \end{tablenotes}
@@
 \item The method produced reliable results in the case of simulated and interpretable results in several real-world
-  data sets
+  data sets.
 \end{itemize}
@@
 \bibliographystyle{elsarticle-num}
-\bibliography{./manuscript.bib}
+\bibliography{./epistatic_interactions.bib}
 

### 2025-05-04 cover_letter.tex
smale changes to paper

\documentclass{article}

\usepackage{charter} % Use the Charter font for the document text

\usepackage[margin=1in]{geometry}
\usepackage{fancyhdr,graphicx}

\fancypagestyle{firstpage}{%
  \fancyhf{}% Clear header/footer
  \renewcommand{\headrulewidth}{0pt}%
  \renewcommand{\footrulewidth}{1pt}%
  % Add more detail here if needed
}
\fancypagestyle{otherpages}{%
  \fancyhf{}% Clear header/footer
  \renewcommand{\headrulewidth}{1pt}%
  \renewcommand{\footrulewidth}{1pt}%
  % Add more detail here if needed
}

\AtBeginDocument{\thispagestyle{firstpage}}
\pagestyle{otherpages}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1ex}

\begin{document}

\vspace*{\dimexpr-\headsep-\headheight-1pt}

\noindent
\begin{minipage}[t]{0.5\textwidth}
  \includegraphics[width=4cm]{CSIRO_logo.png}
\end{minipage}%
\begin{minipage}[t]{0.5\textwidth}
  \raggedleft
  \raisebox{1cm}{\begin{tabular}{ l @{} }
  \today \\[12pt] % Date
   Robert Dunne \\
   13 Garden Street\\
    Eveleigh NSW \\
    Phone: (+61) 439 905  075 \\
    Email: rob.dunne@csiro.au
  \end{tabular}}
\end{minipage}

\rule{\linewidth}{1pt}

\bigskip\bigskip

Dr. Christiana Gousiadi

Research Networks AS, Oslo, Norway

\noindent
\begin{minipage}[t]{\textwidth}
  Dr. Christiana Gousiadi \\ 
  Research Networks AS, Oslo, Norway \\
  Editor-in-Chief,  Computational and Structural Biotechnology Reports \\ 
\end{minipage}

\bigskip\bigskip

% Start of the letter content
\noindent
Dear Dr. Christiana Gousiadi,

\bigskip

I am sending you the manuscript of our research paper entitled ``Detection of Ecstatic Interactions in High-Dimensional
Genomic Data Using Random Forests''. We would like to have the manuscript considered for publication in  Computational and Structural Biotechnology Reports.

Epistatic interactions can play an important role in the genetic mechanisms that control phenotypic variation. However,
identifying these interactions in high dimensional genomic data can be very challenging due to the large computational
burden.

Random Forests Decision Trees are widely used in a variety of disciplines and handle interactions well.
In fact, in a Random Forest model there are no ``main effects'', they are all interactions.

However, Random Forests models do not explicitly detect variable interactions. We have addressed this problem with a
computationally feasible approach.

We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.
All authors have approved the manuscript and agree with its submission to [insert the name of the target journal].

We believe that this paper is well aligned with the aims and scope of the journal Computational and Structural
Biotechnology Reports.

Robert Dunne

\end{document}

### 2025-07-02 epistatic_interactions.tex
smale changes to paper
changes to epistatic_interactions.tex

@@
 %% and the form \ead[url] for the home page:
-\title{Detection of Ecstatic Interactions in High-Dimensional Genomic Data Using Random Forests}
+\title{Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests}
 
@@
 
-%%Research highlights
-\begin{highlights}
-\item The method detects epistatic interactions in high dimensional genomic data.
-\item Computationally efective and produces reliable results in the case of simulated and interpretable results in several real-world
-  data sets
-\end{highlights}
-
 %% Abstract
@@
 \begin{highlights}
-\item Research highlight 1
-\item Research highlight 2
+\item The method detects epistatic interactions in high dimensional genomic data.
+\item Computationally efective and produces reliable results in the case of simulated and interpretable results in several real-world
+  data sets
 \end{highlights}
@@
 guanine nucleotide dissociation stimulator), Ddx31 (dead-box helicase 31) and Ttf1 (transcription termination factor
-1). These genes have a variety of functions
-
-including mitochondrial protein synthesis, intracellular signalling, cell growth and division in embryogenesis, and
+1). These genes have a variety of functions including mitochondrial protein synthesis, intracellular signalling, cell growth and division in embryogenesis, and
 transcription, respectively \cite[]{stelzerGeneCardsSuiteGene2016}. At present, only Ddx31 is directly linked to

### 2025-07-02 Supplementary_Files/Supplementary_File_Comparison_with_BOOST.Rmd
added Supplementary_Files and bib items

---
title: "Comparison with BOOST"
author: "Rob Dunne"
date: "Monday, June 16, 2025"
output: 
    bookdown::pdf_document2:
     toc: true
     toc_depth: 3
     fig_caption: yes
     citation_package: biblatex
bibliography: 
    ../epistaticInteractions.bib 
header-includes:
  \usepackage{dcolumn}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage{biblatex}
  \usepackage{caption}
  \usepackage{subcaption}
---

# load data

- 4798 samples
- 94485 SNPs

```{r, echo=FALSE, eval=FALSE}
library(lmtest)
library(ranger)
#library(RCy3)

#Noise <- 0.1
numIte  <-  1
t <- numIte
ntree <- 1000

dirOut  <-  paste0("Results")
if(!dir.exists(dirOut)){
  dir.create(dirOut)
}
#dirOut  <-  paste0("Results/pruned200K.3r2_geno_Missing_Averaged")
dt  <-  readRDS("pruned200K.3r2_geno_Missing_Averaged.rds")
pheno <- read.table("Cleaned_CaseControl_RAauto.fam",header=F)
pheno$ID <- paste0(pheno$V1,"_",pheno$V2)
all.equal(row.names(dt),pheno$ID)
dt <- data.frame(pheno=as.factor(pheno$V6),dt)
dim(dt) #[1]  4798 94485

#94,486 SNPs (4.46 billion possible pairwise interaction tests) and 4,798 individuals.

dt$pheno <- as.factor(dt$pheno)
#dt <- dt[which(dt$treatment=="short"),-2]

# quick load of results etc
#dirOut  <-  paste0("Results")
#load(file="Results/var_importance_gini.Rdata")
#load(file=paste0(dirOut,"/All_Interactions.Rdata"))
#load(file="./Results/interaction_snps_counts_fixed.Rdata")
#load(file=paste0(dirOut,"/lm_results_fixed.Rdata"))

```

# BOOST

To demonstrate the computational advantage of our Random Forest approach, we attempted exhaustive pairwise epistasis
testing using PLINK v1.9 on the full rheumatoid arthritis dataset (94,484 SNPs, requiring 4.46 billion tests). Despite
allocating 257GB of RAM, the analysis resulted in segmentation faults due to computational and memory limitations. This
demonstrates the practical impossibility of exhaustive approaches on genome-wide datasets.

In contrast, our Random Forest method completed analysis of the same dataset in [X] minutes, identifying [Y] candidate
interactions for follow-up testing. To enable direct comparison, we implemented Boolean operation screening (similar to
BOOST) on random subsets of 50,000 SNP pairs, showing [Z]% overlap in top-ranked interactions while requiring [W]-fold
less computation time.

We have used the BOOST (BOOlean operation-based Screening and Testing) @wanBOOSTFastApproach2010 as a comparitor.

- Stage 1 (Screening): Uses Boolean operations to quickly identify potentially interacting SNP pairs

    - For each SNP pair, creates a 3×3 contingency table (0,1,2 genotypes for each SNP)
    -    Applies Boolean functions (AND, OR, XOR) to identify patterns
    -    Ranks pairs by how well Boolean operations separate cases from controls
- Stage 2 (Testing): Applies statistical tests to top-ranked pairs

   - Uses logistic regression or other statistical tests
   -  Applies multiple testing corrections

However, even with over 500GB of RAM, the algorithm could not run on this data set.
We implemented the core BOOT functionality in R and and ran it on 500000 randomly selected SNPS.
To process the whole data set would have taken over 4 days with 64 cores.

```{R, echo=FALSE, eval=FALSE}
# Convert your data to PLINK format
library(snpStats)
# Assuming your data is in the format you used for RF
temp<-as.matrix(dt[,-1])
ptemp<- dt[,1]
temp_snpmatrix <- temp + 1  # Convert 0,1,2 to 1,2,3
temp_snpmatrix[is.na(temp)] <- 0  # Set missing to 0

# Create SnpMatrix object
snp_matrix <- new("SnpMatrix", temp_snpmatrix)

write.plink("boost_input", snps =snp_matrix,
            phenotype = ptemp, 
            id = rownames(dt)
            )
#Writing FAM file to boost_input.fam 
#Writing extended MAP file to boost_input.bim 
#Writing BED file to boost_input.bed (SNP-major mode)
#NULL

```
```{verbatim}
> plink --bfile boost_input --epistasis --allow-no-sex --out epistasis_results
PLINK v1.90p 64-bit (30 Nov 2019)              www.cog-genomics.org/plink/1.9/
(C) 2005-2019 Shaun Purcell, Christopher Chang   GNU General Public License v3
Logging to epistasis_results.log.
Options in effect:
  --allow-no-sex
  --bfile boost_input
  --epistasis
  --out epistasis_results
515555 MB RAM detected; reserving 257777 MB for main workspace.
94484 variants loaded from .bim file.
4798 people (0 males, 0 females, 4798 ambiguous) loaded from .fam.
Ambiguous sex IDs written to epistasis_results.nosex .
4798 phenotype values loaded from .fam.
Using up to 63 threads (change this with --threads).
Before main variant filters, 4798 founders and 0 nonfounders present.
Calculating allele frequencies... 0%Segmentation fault (core dumped)
```

## Boost in R

```{R, echo=FALSE, eval=FALSE}
# Use shared memory approach (no data copying)
library(parallel)

# This approach doesn't copy data to each worker
boolean_shared_memory <- function(snp_data, phenotype, max_pairs = 100000, n_cores = 20) {

  # Use mclapply instead of makeCluster (shares memory)
  n_snps <- ncol(snp_data)

  # Generate random pairs
  set.seed(123)
  random_pairs <- replicate(max_pairs, {
    i <- sample(1:(n_snps-1), 1)
    j <- sample((i+1):n_snps, 1)
    c(i, j)
  }, simplify = FALSE)

  # Split into chunks
  chunk_size <- ceiling(length(random_pairs) / n_cores)
  chunks <- split(random_pairs, ceiling(seq_along(random_pairs) / chunk_size))

  cat("Using mclapply with", n_cores, "cores (shared memory)\n")

  # mclapply shares memory instead of copying
  results_list <- mclapply(chunks, function(chunk) {
    chunk_results <- data.frame()

    for(pair in chunk) {
      i <- pair[1]
      j <- pair[2]

      snp1 <- snp_data[, i]  # References original data, no copy
      snp2 <- snp_data[, j]

      if (length(unique(snp1)) < 2 || length(unique(snp2)) < 2) next

      tryCatch({
        xor_op <- (snp1 >= 1) != (snp2 >= 1)
        test_result <- fisher.test(table(xor_op, phenotype))

        chunk_results <- rbind(chunk_results, data.frame(
          SNP1_idx = i, SNP2_idx = j,
          p_value = test_result$p.value,
          odds_ratio = as.numeric(test_result$estimate)
        ))
      }, error = function(e) {})
    }

    return(chunk_results)
  }, mc.cores = n_cores)

  final_results <- do.call(rbind, results_list)
  return(final_results[order(final_results$p_value), ])
}
# This should use much less memory since it doesn't copy data to workers
cat("Trying shared memory approach...\n")

boost_time_shared <- system.time({
  boost_results_shared <- boolean_shared_memory(
    dt[,-1],           # SNP data  
    dt[,1],            # phenotype
    max_pairs = 500000,  # Start with 500k pairs
    n_cores = 60       # Conservative core count
  )
})

# max_pairs = 10000000 failed -- many processes in "D" state, I eventually killed it after about 1 hour
# max_pairs = 5000000 also  failed -- many processes in "D" state, I eventually killed it

cat("Shared memory approach completed in", boost_time_shared[3], "seconds\n")
cat("Found", nrow(boost_results_shared), "tested pairs\n")
print(head(boost_results_shared, 5))

#save(boost_results_shared, file="./script3/boost_results_shared_500000.Rdata")
#load(file="./script3/boost_results_shared_500000.Rdata")

```

```{R, echo=TRUE, eval=FALSE}
boost_time_shared 
#    user   system  elapsed 
#1593.886   99.380   42.000

((choose(94485,2)/500000)*42)/(60*60*24)
#[1] 4.33967 days run time to process the whole data set

print(head(boost_results_shared, 5))
#        SNP1_idx SNP2_idx      p_value odds_ratio
#5.3801     34461    75705 1.031365e-33  0.4767794
#21.4869    34461    58455 4.896431e-32  0.4869738
#6.8174     34461    67197 2.060349e-27  0.5194040
#3.15611    34457    58152 3.514479e-17  0.6043620
#19.6946    17151    34442 9.914728e-15  1.5910823

```
# comparison of RF method and BOOST

We load the **interaction_snps_counts** (calculated in Supp1.Rmd),

```{R, echo=TRUE, eval=FALSE}
load(file="./Results/interaction_snps_counts_fixed.Rdata")
head(interaction_snps_counts)
#              interaction_snps Freq
#281122 X6_32471505 X6_32771829  466
#281468 X6_32682149 X6_32771829  144
#280884 X6_32222493 X6_32771829  113

```

```{R, echo=FALSE, eval=FALSE}

load(file="./script3/boost_results_shared_500000.Rdata")
load(file="./Results/interaction_snps_counts_fixed.Rdata")

# Function to compare RF and BOOST results
compare_rf_boost_results <- function(rf_results, boost_results, snp_names) {

  # Convert BOOST indices to SNP names
  boost_with_names <- boost_results
  boost_with_names$SNP1_name <- snp_names[boost_results$SNP1_idx]
  boost_with_names$SNP2_name <- snp_names[boost_results$SNP2_idx]

  # Create standardized pair identifiers (alphabetical order)
  boost_with_names$pair_id <- apply(boost_with_names[, c("SNP1_name", "SNP2_name")], 1, 
                                   function(x) paste(sort(x), collapse = " "))

  # Process RF results - split the interaction_snps column
  rf_processed <- rf_results
  rf_processed$pair_id <- rf_processed$nteraction_snps

  # Find overlapping pairs
  common_pairs <- intersect(rf_processed$pair_id, boost_with_names$pair_id)

  # Create comparison dataframe
  comparison_results <- data.frame()

  for(pair in common_pairs) {
    rf_row <- rf_processed[rf_processed$pair_id == pair, ]
    boost_row <- boost_with_names[boost_with_names$pair_id == pair, ]

    comparison_results <- rbind(comparison_results, data.frame(
      pair_id = pair,
      rf_frequency = rf_row$Freq[1],
      boost_p_value = boost_row$p_value[1],
      boost_odds_ratio = boost_row$odds_ratio[1],
      rf_rank = which(rf_processed$pair_id == pair)[1],
      boost_rank = which(boost_with_names$pair_id == pair)[1]
    ))
  }

  return(list(
    comparison = comparison_results,
    overlap_count = length(common_pairs),
    total_rf = nrow(rf_processed),
    total_boost = nrow(boost_with_names),
    overlap_percent = length(common_pairs) / min(nrow(rf_processed), nrow(boost_with_names)) * 100
  ))
}

# Apply the comparison
snp_names <- colnames(dt[,-1])  # Get SNP names from your data

comparison_result <- compare_rf_boost_results(
  interaction_snps_counts,  # Your RF results
  boost_results_shared,     # Your BOOST results  
  snp_names
)

# View results
cat("Overlap summary:\n")
cat("Common pairs found:", comparison_result$overlap_count, "\n") #Common pairs found: 28 
cat("RF total pairs:", comparison_result$total_rf, "\n")         #RF total pairs: 291142 
cat("BOOST total pairs:", comparison_result$total_boost, "\n")   #BOOST total pairs: 500000 
cat("Overlap percentage:", round(comparison_result$overlap_percent, 2), "%\n") #Overlap percentage: 0.01 %

print("Top overlapping pairs:")
print(head(comparison_result$comparison[order(comparison_result$comparison$rf_frequency, decreasing = TRUE), ], 10))

# Analyze correlation between RF frequency and BOOST significance
 if(nrow(comparison_result$comparison) > 0) {

  # Correlation between RF frequency and BOOST -log10(p-value)
  rf_freq <- comparison_result$comparison$rf_frequency
  boost_neg_log_p <- -log10(comparison_result$comparison$boost_p_value)

  correlation <- cor(rf_freq, boost_neg_log_p, use = "complete.obs")

  cat("Correlation between RF frequency and BOOST -log10(p-value):", round(correlation, 3), "\n")

  # Plot the relationship
  plot(rf_freq, boost_neg_log_p, 
       xlab = "RF Co-occurrence Frequency", 
       ylab = "BOOST -log10(p-value)",
       main = "RF vs BOOST: Method Comparison",
       pch = 19, col = "blue")

  # Add trend line
  if(!is.na(correlation) && abs(correlation) > 0.1) {
    abline(lm(boost_neg_log_p ~ rf_freq), col = "red", lwd = 2)
    legend("topright", paste("r =", round(correlation, 3)), bty = "n")
  }
}
# very little correlation

# Compare top-ranked pairs from each method
top_rf <- head(interaction_snps_counts[order(interaction_snps_counts$Freq, decreasing = TRUE), ], 20)
top_boost <- head(boost_results_shared[order(boost_results_shared$p_value), ], 20)

cat("Top 20 RF pairs:\n")
print(top_rf)

cat("\nTop 20 BOOST pairs (with names):\n")
top_boost_named <- top_boost
top_boost_named$SNP1_name <- snp_names[top_boost$SNP1_idx]
top_boost_named$SNP2_name <- snp_names[top_boost$SNP2_idx]
print(top_boost_named[, c("SNP1_name", "SNP2_name", "p_value", "odds_ratio")])

# Analyze what types of interactions each method finds
cat("Analysis of method differences:\n")
cat("RF finds", nrow(interaction_snps_counts), "co-occurrence based interactions\n")   #RF finds 291142 co-occurrence based interactions
cat("BOOST finds", sum(boost_results_shared$p_value < 0.05), "Boolean logic interactions\n") #BOOST finds 25436 Boolean logic interactions
cat("Methods are detecting different interaction mechanisms\n")

# Convert BOOST indices to SNP names
boost_with_names <-  boost_results_shared
boost_with_names$SNP1_name <- snp_names[boost_results_shared$SNP1_idx]
boost_with_names$SNP2_name <- snp_names[boost_results_shared$SNP2_idx]

  # Create standardized pair identifiers (alphabetical order)
boost_with_names$pair_id <- apply(boost_with_names[, c("SNP1_name", "SNP2_name")], 1, 
                                   function(x) paste(sort(x), collapse = " "))

# Look at the p-value distribution of overlapping vs non-overlapping
overlapping_pairs <- comparison_result$comparison$pair_id
length(overlapping_pairs) #[1] 28
boost_overlap <- boost_with_names$pair_id %in% overlapping_pairs

boost_non_overlap <- !boost_overlap

cat("BOOST p-values for overlapping pairs:\n")
summary(-log10(boost_results_shared$p_value[boost_overlap]))
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.0000  0.3034  0.4397  0.7352  0.8353  4.7758

cat("BOOST p-values for non-overlapping pairs:\n") 
summary(-log10(boost_results_shared$p_value[boost_non_overlap]))
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 0.0000  0.1178  0.2972  0.4342  0.6056 32.9866

```

In the paper we only consider the  640 interaction_snps_counts with Freq > 1

```{R, echo=FALSE, eval=FALSE}
# Add FDR correction and pair_id to boost_results_shared
library(stats)

# Add FDR corrected p-values
boost_results_shared$fdr_p_value <- p.adjust(boost_results_shared$p_value, method = "fdr")

# Add pair_id column
snp_names <- colnames(dt[,-1])
boost_results_shared$SNP1_name <- snp_names[boost_results_shared$SNP1_idx]
boost_results_shared$SNP2_name <- snp_names[boost_results_shared$SNP2_idx]

# Create standardized pair_id (alphabetical order)
boost_results_shared$pair_id <- apply(boost_results_shared[, c("SNP1_name", "SNP2_name")], 1, 
                                     function(x) paste(sort(x), collapse = " "))

# Check the results
cat("BOOST results with FDR correction:\n")
cat("Total pairs tested:", nrow(boost_results_shared), "\n") #500000 
cat("Significant at p < 0.05:", sum(boost_results_shared$p_value < 0.05), "\n")  # 25436 
cat("Significant at FDR < 0.05:", sum(boost_results_shared$fdr_p_value < 0.05), "\n") #55 
cat("Significant at FDR < 0.10:", sum(boost_results_shared$fdr_p_value < 0.10), "\n") #60

print(head(boost_results_shared[, c("pair_id", "p_value", "fdr_p_value", "odds_ratio")], 10))

#2) Focused Comparison: RF Freq > 1 vs Significant BOOST
# Filter to meaningful comparisons
rf_significant <- interaction_snps_counts[interaction_snps_counts$Freq > 1, ]
boost_significant <- boost_results_shared[boost_results_shared$fdr_p_value < 0.10, ]  # Adjust threshold as needed

cat("Focused comparison:\n")
cat("RF pairs with Freq > 1:", nrow(rf_significant), "\n")              #640 
cat("BOOST pairs with FDR < 0.10:", nrow(boost_significant), "\n")      #60

# Updated comparison function
compare_rf_boost_focused <- function(rf_results, boost_results) {

  # Find overlapping pairs
  common_pairs <- intersect(rf_results$interaction_snps, boost_results$pair_id)

  # Create detailed comparison for overlapping pairs
  comparison_detailed <- data.frame()

  for(pair in common_pairs) {
    rf_row <- rf_results[rf_results$interaction_snps == pair, ]
    boost_row <- boost_results[boost_results$pair_id == pair, ]

    comparison_detailed <- rbind(comparison_detailed, data.frame(
      pair_id = pair,
      rf_frequency = rf_row$Freq[1],
      boost_p_value = boost_row$p_value[1],
      boost_fdr_p_value = boost_row$fdr_p_value[1],
      boost_odds_ratio = boost_row$odds_ratio[1],
      rf_rank = which(rf_results$interaction_snps == pair)[1],
      boost_rank = which(boost_results$pair_id == pair)[1]
    ))
  }

  # Summary statistics
  overlap_stats <- list(
    total_overlap = length(common_pairs),
    rf_total = nrow(rf_results),
    boost_total = nrow(boost_results), 
    overlap_percent_rf = length(common_pairs) / nrow(rf_results) * 100,
    overlap_percent_boost = length(common_pairs) / nrow(boost_results) * 100,
    comparison_table = comparison_detailed
  )

  return(overlap_stats)
}

# Perform focused comparison
focused_comparison <- compare_rf_boost_focused(rf_significant, boost_significant)

cat("\n=== FOCUSED COMPARISON RESULTS ===\n")
cat("RF significant pairs (Freq > 1):", focused_comparison$rf_total, "\n")                         #640 
cat("BOOST significant pairs (FDR < 0.10):", focused_comparison$boost_total, "\n")                 #60 
cat("Overlapping pairs:", focused_comparison$total_overlap, "\n")                                  #Overlapping pairs: 0 
cat("Overlap as % of RF significant:", round(focused_comparison$overlap_percent_rf, 2), "%\n")
cat("Overlap as % of BOOST significant:", round(focused_comparison$overlap_percent_boost, 2), "%\n")

if(nrow(focused_comparison$comparison_table) > 0) {
  cat("\nOverlapping pairs details:\n")
  print(focused_comparison$comparison_table[order(focused_comparison$comparison_table$rf_frequency, decreasing = TRUE), ])
} else {
  cat("\nNo overlapping pairs found between RF (Freq > 1) and BOOST (FDR < 0.10)\n")
}
#No overlapping pairs found between RF (Freq > 1) and BOOST (FDR < 0.10)

#3) Alternative Comparison Strategy
#If there's still little overlap, let's try a more lenient approach:

# More lenient comparison
lenient_comparison <- function() {

  # Use top performers from each method
  top_rf <- head(rf_significant[order(rf_significant$Freq, decreasing = TRUE), ], 100)
  top_boost <- head(boost_results_shared[order(boost_results_shared$p_value), ], 100)

  common_pairs_lenient <- intersect(top_rf$interaction_snps, top_boost$pair_id)

  cat("=== LENIENT COMPARISON (Top 100 each) ===\n")
  cat("Top 100 RF pairs overlap with top 100 BOOST pairs:", length(common_pairs_lenient), "\n")

  if(length(common_pairs_lenient) > 0) {
    for(pair in common_pairs_lenient) {
      rf_info <- top_rf[top_rf$interaction_snps == pair, ]
      boost_info <- top_boost[top_boost$pair_id == pair, ]
      cat("Pair:", pair, "| RF Freq:", rf_info$Freq, "| BOOST p:", boost_info$p_value, "\n")
    }
  }

  return(common_pairs_lenient)
}

lenient_results <- lenient_comparison()
#> === LENIENT COMPARISON (Top 100 each) ===
#Top 100 RF pairs overlap with top 100 BOOST pairs: 0

```

```{R, echo=FALSE, eval=FALSE}
# Extract all unique SNPs from RF interaction pairs
extract_rf_snps <- function(interaction_snps_counts) {

  # Split the interaction pairs and collect all SNP names
  all_snps <- c()

  for(i in 1:nrow(interaction_snps_counts)) {
    pair <- interaction_snps_counts$interaction_snps[i]
    snp_pair <- strsplit(as.character(pair), " ")[[1]]
    all_snps <- c(all_snps, snp_pair[1], snp_pair[2])
  }

  # Get unique SNPs
  unique_snps <- unique(all_snps)

  cat("RF interaction analysis:\n")
  cat("Total interaction pairs:", nrow(interaction_snps_counts), "\n")
  cat("Total SNP mentions:", length(all_snps), "\n") 
  cat("Unique SNPs involved:", length(unique_snps), "\n")

  return(unique_snps)
}

# Extract the SNPs
rf_snps <- extract_rf_snps(interaction_snps_counts[interaction_snps_counts$Freq > 1,]) #533 SNPS

# Get their indices in the original data
snp_names <- colnames(dt[,-1])
rf_snp_indices <- which(snp_names %in% rf_snps)

cat("SNP indices for RF-identified SNPs:\n")
cat("Number of indices found:", length(rf_snp_indices), "\n")
cat("Sample indices:", head(rf_snp_indices, 10), "\n")

# Boolean screening focused on RF-identified SNPs only
boolean_rf_focused <- function(snp_data, phenotype, rf_snp_indices, n_cores = 30) {

  n_rf_snps <- length(rf_snp_indices)
  total_rf_pairs <- choose(n_rf_snps, 2)

  cat("Focused Boolean screening on RF SNPs:\n")
  cat("RF-identified SNPs:", n_rf_snps, "\n")
  cat("Possible pairs among RF SNPs:", total_rf_pairs, "\n")
  cat("Using", n_cores, "cores\n")

  # Generate ALL possible pairs among RF SNPs (since the number is manageable)
  rf_pairs <- t(combn(rf_snp_indices, 2))

  cat("Testing all", nrow(rf_pairs), "pairs among RF SNPs\n")

  # Split into chunks for parallel processing
  chunk_size <- ceiling(nrow(rf_pairs) / n_cores)
  pair_chunks <- split(1:nrow(rf_pairs), ceiling(seq_along(1:nrow(rf_pairs)) / chunk_size))

  # Process in parallel using mclapply (shared memory)
  library(parallel)

  results_list <- mclapply(pair_chunks, function(chunk_indices) {
    chunk_results <- data.frame()

    for(idx in chunk_indices) {
      i <- rf_pairs[idx, 1]  # These are already the correct indices
      j <- rf_pairs[idx, 2]

      snp1 <- snp_data[, i]
      snp2 <- snp_data[, j]

      # Skip if no variation
      if (length(unique(snp1)) < 2 || length(unique(snp2)) < 2) next

      tryCatch({
        # Boolean operations
        and_op <- (snp1 >= 1) & (snp2 >= 1)
        or_op <- (snp1 >= 1) | (snp2 >= 1)
        xor_op <- (snp1 >= 1) != (snp2 >= 1)

        # Test association with phenotype
        and_test <- fisher.test(table(and_op, phenotype))
        or_test <- fisher.test(table(or_op, phenotype))
        xor_test <- fisher.test(table(xor_op, phenotype))

        chunk_results <- rbind(chunk_results, data.frame(
          SNP1_idx = i,
          SNP2_idx = j,
          SNP1_name = colnames(snp_data)[i],
          SNP2_name = colnames(snp_data)[j],
          boolean_and_p = and_test$p.value,
          boolean_or_p = or_test$p.value,
          boolean_xor_p = xor_test$p.value,
          best_boolean_p = min(and_test$p.value, or_test$p.value, xor_test$p.value),
          and_odds_ratio = as.numeric(and_test$estimate),
          or_odds_ratio = as.numeric(or_test$estimate),
          xor_odds_ratio = as.numeric(xor_test$estimate)
        ))

      }, error = function(e) {
        # Skip problematic pairs
      })
    }

    return(chunk_results)
  }, mc.cores = n_cores)

  # Combine results
  final_results <- do.call(rbind, results_list)

  if(nrow(final_results) > 0) {
    # Add FDR correction
    final_results$fdr_p_value <- p.adjust(final_results$best_boolean_p, method = "fdr")

    # Add pair_id
    final_results$pair_id <- apply(final_results[, c("SNP1_name", "SNP2_name")], 1,
                                  function(x) paste(sort(x), collapse = " "))

    # Sort by p-value
    final_results <- final_results[order(final_results$best_boolean_p), ]
  }

  cat("Completed! Tested", nrow(final_results), "pairs\n")
  return(final_results)
}

# Run the focused Boolean screening
cat("Starting focused Boolean screening on RF-identified SNPs...\n")

boost_time_focused <- system.time({
  boost_rf_focused <- boolean_rf_focused(
    dt[,-1],              # SNP data
    dt[,1],               # phenotype  
    rf_snp_indices,       # Indices of RF-identified SNPs
    n_cores = 60
  )
})

cat("Focused Boolean screening completed in", boost_time_focused[3], "seconds\n")
#Focused Boolean screening completed in 47.242 seconds
cat("Results summary:\n")
cat("Total pairs tested:", nrow(boost_rf_focused), "\n") #Total pairs tested: 141778 
cat("Significant at p < 0.05:", sum(boost_rf_focused$best_boolean_p < 0.05), "\n") #Significant at p < 0.05: 61106 
cat("Significant at FDR < 0.05:", sum(boost_rf_focused$fdr_p_value < 0.05), "\n")  #Significant at FDR < 0.05: 41651

print(head(boost_rf_focused, 10))

# Now we have a perfect comparison - same SNPs, different methods
direct_comparison <- function(rf_results, boost_focused_results) {

  # Find overlapping pairs
  common_pairs <- intersect(rf_results$interaction_snps, boost_focused_results$pair_id)

  cat("=== DIRECT COMPARISON (Same SNPs) ===\n")
  cat("RF pairs:", nrow(rf_results), "\n")
  cat("BOOST pairs tested:", nrow(boost_focused_results), "\n")
  cat("BOOST significant (FDR < 0.05):", sum(boost_focused_results$fdr_p_value < 0.05), "\n")
  cat("Overlapping pairs:", length(common_pairs), "\n")
  cat("Overlap percentage:", round(length(common_pairs) / nrow(rf_results) * 100, 2), "%\n")

  # Detailed analysis of overlapping pairs
  if(length(common_pairs) > 0) {
    overlap_details <- data.frame()

    for(pair in common_pairs) {
      rf_row <- rf_results[rf_results$interaction_snps == pair, ]
      boost_row <- boost_focused_results[boost_focused_results$pair_id == pair, ]

      overlap_details <- rbind(overlap_details, data.frame(
        pair_id = pair,
        rf_frequency = rf_row$Freq[1],
        boost_p_value = boost_row$best_boolean_p[1],
        boost_fdr_p_value = boost_row$fdr_p_value[1],
        boost_odds_ratio = boost_row$and_odds_ratio[1]  # or whichever was best
      ))
    }

    print("Overlapping pairs:")
    print(overlap_details[order(overlap_details$rf_frequency, decreasing = TRUE), ])

    # Correlation analysis
    if(nrow(overlap_details) > 1) {
      correlation <- cor(overlap_details$rf_frequency, -log10(overlap_details$boost_p_value))
      cat("Correlation between RF frequency and BOOST -log10(p):", round(correlation, 3), "\n")
    }
  }

  return(list(
    common_pairs = common_pairs,
    overlap_details = if(exists("overlap_details")) overlap_details else data.frame()
  ))
}

# Perform direct comparison
direct_results <- direct_comparison(interaction_snps_counts[interaction_snps_counts$Freq > 1,], boost_rf_focused)
#=== DIRECT COMPARISON (Same SNPs) ===
#RF pairs: 640 
#BOOST pairs tested: 141778 
#BOOST significant (FDR < 0.05): 41651 
#Overlapping pairs: 640 
#Overlap percentage: 100 %

head(direct_results$overlap_details)

# Save all results
#save(rf_snps, rf_snp_indices, boost_rf_focused, direct_results, 
#     file = "./script3/focused_rf_boost_comparison.Rdata")
#load("./script3/focused_rf_boost_comparison.Rdata")

```

```{R, echo=FALSE, eval=FALSE}
# Analyze the relationship between RF and BOOST findings
analyze_rf_boost_relationship <- function(direct_results) {

  overlap_data <- direct_results$overlap_details

  cat("=== RF vs BOOST RELATIONSHIP ANALYSIS ===\n")

  # Summary statistics
  cat("RF frequency range:", range(overlap_data$rf_frequency), "\n")
  cat("BOOST p-value range:", range(overlap_data$boost_p_value), "\n")
  cat("Correlation (RF freq vs -log10(p)):", round(cor(overlap_data$rf_frequency, -log10(overlap_data$boost_p_value)), 3), "\n")

  # Categorize RF pairs by frequency
  high_freq_rf <- overlap_data[overlap_data$rf_frequency >= 50, ]
  med_freq_rf <- overlap_data[overlap_data$rf_frequency >= 10 & overlap_data$rf_frequency < 50, ]
  low_freq_rf <- overlap_data[overlap_data$rf_frequency < 10, ]

  cat("\n=== FREQUENCY CATEGORIES ===\n")
  cat("High frequency RF pairs (≥50):", nrow(high_freq_rf), "\n")
  cat("- Mean BOOST -log10(p):", round(mean(-log10(high_freq_rf$boost_p_value)), 2), "\n")
  cat("- BOOST FDR < 0.05:", sum(high_freq_rf$boost_fdr_p_value < 0.05), "/", nrow(high_freq_rf), "\n")

  cat("Medium frequency RF pairs (10-49):", nrow(med_freq_rf), "\n")
  cat("- Mean BOOST -log10(p):", round(mean(-log10(med_freq_rf$boost_p_value)), 2), "\n")
  cat("- BOOST FDR < 0.05:", sum(med_freq_rf$boost_fdr_p_value < 0.05), "/", nrow(med_freq_rf), "\n")

  cat("Low frequency RF pairs (2-9):", nrow(low_freq_rf), "\n")
  cat("- Mean BOOST -log10(p):", round(mean(-log10(low_freq_rf$boost_p_value)), 2), "\n")
  cat("- BOOST FDR < 0.05:", sum(low_freq_rf$boost_fdr_p_value < 0.05), "/", nrow(low_freq_rf), "\n")

  return(list(high = high_freq_rf, medium = med_freq_rf, low = low_freq_rf))
}

# Run the analysis
freq_analysis <- analyze_rf_boost_relationship(direct_results)
## === RF vs BOOST RELATIONSHIP ANALYSIS ===
## RF frequency range: 2 466 
## BOOST p-value range: 5.977923e-71 0.8822303 
## Correlation (RF freq vs -log10(p)): 0.149

## === FREQUENCY CATEGORIES ===
## High frequency RF pairs (≥50): 14 
## - Mean BOOST -log10(p): 33.3 
## - BOOST FDR < 0.05: 12 / 14 
## Medium frequency RF pairs (10-49): 31 
## - Mean BOOST -log10(p): 28.6 
## - BOOST FDR < 0.05: 30 / 31 
## Low frequency RF pairs (2-9): 595 
## - Mean BOOST -log10(p): 24.77 
## - BOOST FDR < 0.05: 537 / 595

# Compare what each method considers "significant"
method_comparison_analysis <- function(rf_data, boost_data) {

  cat("=== METHOD SENSITIVITY COMPARISON ===\n")

  # RF considers Freq > 1 as significant (your filter)
  rf_significant <- nrow(rf_data)  # All 640 pairs

  # BOOST considers FDR < 0.05 as significant  
  boost_significant <- sum(boost_data$fdr_p_value < 0.05)
  boost_very_significant <- sum(boost_data$fdr_p_value < 0.01)

  cat("RF significant pairs (Freq > 1):", rf_significant, "\n")
  cat("BOOST significant pairs (FDR < 0.05):", boost_significant, "\n")
  cat("BOOST very significant pairs (FDR < 0.01):", boost_very_significant, "\n")

  # How many RF pairs are also BOOST significant?
  rf_pairs_also_boost_sig <- sum(direct_results$overlap_details$boost_fdr_p_value < 0.05)
  rf_pairs_also_boost_very_sig <- sum(direct_results$overlap_details$boost_fdr_p_value < 0.01)

  cat("\nOverlap analysis:\n")
  cat("RF pairs also BOOST significant (FDR < 0.05):", rf_pairs_also_boost_sig, "/", rf_significant, 
      "(", round(rf_pairs_also_boost_sig/rf_significant*100, 1), "%)\n")
  cat("RF pairs also BOOST very significant (FDR < 0.01):", rf_pairs_also_boost_very_sig, "/", rf_significant,
      "(", round(rf_pairs_also_boost_very_sig/rf_significant*100, 1), "%)\n")

  # What about BOOST-only significant pairs?
  boost_only_pairs <- boost_data[!(boost_data$pair_id %in% rf_data$interaction_snps) & 
                                 boost_data$fdr_p_value < 0.05, ]

  cat("BOOST-only significant pairs (not in RF Freq > 1):", nrow(boost_only_pairs), "\n")

  return(list(
    rf_sig = rf_significant,
    boost_sig = boost_significant, 
    overlap_sig = rf_pairs_also_boost_sig,
    boost_only = boost_only_pairs
  ))
}

# Run method comparison
method_comp <- method_comparison_analysis(
  interaction_snps_counts[interaction_snps_counts$Freq > 1,], 
  boost_rf_focused
)
## === METHOD SENSITIVITY COMPARISON ===
## RF significant pairs (Freq > 1): 640 
## BOOST significant pairs (FDR < 0.05): 41651 
## BOOST very significant pairs (FDR < 0.01): 24334

## Overlap analysis:
## RF pairs also BOOST significant (FDR < 0.05): 579 / 640 ( 90.5 %)
## RF pairs also BOOST very significant (FDR < 0.01): 572 / 640 ( 89.4 %)
## BOOST-only significant pairs (not in RF Freq > 1): 41072

# Summary table 
create_final_summary <- function() {

  summary_stats <- data.frame(
    Metric = c(
      "Total unique SNPs tested",
      "Total possible pairs", 
      "RF significant pairs (Freq > 1)",
      "BOOST pairs tested",
      "BOOST significant (FDR < 0.05)",
      "RF-BOOST overlap",
      "Correlation (RF freq vs BOOST -log10p)"
    ),
    Value = c(
      "533",
      format(choose(533, 2), big.mark = ","),
      "640", 
      format(nrow(boost_rf_focused), big.mark = ","),
      format(sum(boost_rf_focused$fdr_p_value < 0.05), big.mark = ","),
      "640 (100%)",
      "0.149"
    )
  )

  print(summary_stats)

  cat("\n=== KEY FINDINGS FOR PAPER ===\n")
  cat("1. When testing the same SNPs, both methods detect interactions\n")
  cat("2. RF frequency correlates weakly with BOOST significance (r=0.149)\n") 
  cat("3. RF identifies", nrow(direct_results$overlap_details), "pairs based on co-occurrence\n")
  cat("4. BOOST identifies", sum(boost_rf_focused$fdr_p_value < 0.05), "pairs based on Boolean logic\n")
  cat("5. Methods capture different aspects of epistatic interactions\n")
}

create_final_summary()
create_final_summary()
#                                  Metric      Value
#1               Total unique SNPs tested        533
#2                   Total possible pairs    141,778
#3        RF significant pairs (Freq > 1)        640
#4                     BOOST pairs tested    141,778
#5         BOOST significant (FDR < 0.05)     41,651
#6                       RF-BOOST overlap 640 (100%)
#7 Correlation (RF freq vs BOOST -log10p)      0.149

#=== KEY FINDINGS FOR PAPER ===
#1. When testing the same SNPs, both methods detect interactions
#2. RF frequency correlates weakly with BOOST significance (r=0.149)
#3. RF identifies 640 pairs based on co-occurrence
#4. BOOST identifies 41651 pairs based on Boolean logic
#5. Methods capture different aspects of epistatic interactions

# Save final results
#save(direct_results, freq_analysis, method_comp, 
#     file = "./script3/final_rf_boost_comparison.Rdata")
#load( file = "./script3/final_rf_boost_comparison.Rdata")

```

- things to do for comparison with BOOST (first in the Supplementary and then in the paper)

   -   Methods section - Clean description of both RF and Boolean screening approaches
   -  Results section - Key findings table and interpretation
   -  Comparison figures - Correlation plots, overlap analysis
   -  Discussion points - Why methods detect different interaction types

### 2025-07-02 Supplementary_Files/Supplementary_File_Parameter_Optimization_for_RF.Rmd
added Supplementary_Files and bib items

---
title: "Parameter Optimization for Random Forest"
author: "Rob Dunne"
date: "Monday, June 16, 2025"
output: 
    bookdown::pdf_document2:
     toc: true
     toc_depth: 3
     fig_caption: yes
     citation_package: biblatex
bibliography: 
    ../epistaticInteractions.bib 
header-includes:
  \usepackage{dcolumn}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage{biblatex}
  \usepackage{caption}
  \usepackage{subcaption}
---

# load data

- 4798 samples
- 94485 SNPs

```{r, echo=FALSE, eval=FALSE}
library(lmtest)
library(ranger)
#library(RCy3)

#Noise <- 0.1
numIte  <-  1
t <- numIte
ntree <- 1000

dirOut  <-  paste0("Results")
if(!dir.exists(dirOut)){
  dir.create(dirOut)
}
#dirOut  <-  paste0("Results/pruned200K.3r2_geno_Missing_Averaged")
dt  <-  readRDS("pruned200K.3r2_geno_Missing_Averaged.rds")
pheno <- read.table("Cleaned_CaseControl_RAauto.fam",header=F)
pheno$ID <- paste0(pheno$V1,"_",pheno$V2)
all.equal(row.names(dt),pheno$ID)
dt <- data.frame(pheno=as.factor(pheno$V6),dt)
dim(dt) #[1]  4798 94485

#94,486 SNPs (4.46 billion possible pairwise interaction tests) and 4,798 individuals.

dt$pheno <- as.factor(dt$pheno)
#dt <- dt[which(dt$treatment=="short"),-2]

# quick load of results etc
#dirOut  <-  paste0("Results")
#load(file="Results/var_importance_gini.Rdata")
#load(file=paste0(dirOut,"/All_Interactions.Rdata"))
#load(file="./Results/interaction_snps_counts_fixed.Rdata")
#load(file=paste0(dirOut,"/lm_results_fixed.Rdata"))

```

# Random Forest Parameters

Key Parameter Effects:

-   **mtry**: Lower values of mtry lead to more different, less correlated trees, yielding better stability when
    aggregating. However, lower values of mtry also lead to trees that perform on average worse
-  **Sample size**  Similar trade-off between tree diversity and individual tree performance
- **Node size** Setting it lower leads to trees with a larger depth... computation time decreases approximately exponentially with increasing node size

Random forest is an algorithm which is known to provide good results in the default settings. @probstTunabilityImportanceHyperparameters2018,
measure the "tunability" of algorithms and hyperparameters of algorithms and conclude that random forest is far less
tunable than other algorithms such as support vector machines. Nevertheless, a small performance gain (e.g., an average
increase of the AUC of 0.010 based on the 38 considered datasets) can be achieved via tuning.

Tuning the parameter **mtry** provides the biggest average improvement of
the AUC, followed by the sample size, while the node size had only a
small effect. Changing the replace parameter from drawing with
replacement to drawing without replacement (see @stroblUnbiasedSplitSelection2007) also had a small positive effect.

@probstHyperparametersTuningStrategies2019 comments that the number of trees in a forest is a parameter that is not tunable in the 
classical sense but should be set sufficiently high.
The convergence rate, and thus the number of trees needed to obtain optimal performance, depends on the dataset's
properties. Out-of-bag
error curves (slightly) increasing with the number of trees are occasionally observed for certain error measures.

Using a  number of datasets, @oshiroHowManyTrees2012 and @probstTuneNotTune2017 show empirically
that the biggest performance gain can often be achieved when growing the first 100 trees.

It is generally thought that trees are required for stable variable importance estimates than for the simple prediction
purpose. @lunettaScreeningLargescaleAssociation2004 performed simulations with more noisy variables than truly
associated covariates and concluded that multiple thousands of trees must be trained in order to get stable estimates of
the variable importance.

Note that we are using the Gini impurity measure for variable importance. See arguments in ****

-    Much faster computation - no permutation required. 
-    Deterministic - same results every time (less affected by randomness)
-    Built into tree construction - no additional computational overhead
-  in the case of SNP random varaibles, the  
Limitations (from Probst paper): the Gini variable importance measure assigns higher importance values to variables with
more categories or continuous variables and to categorical variables with equally sized categories even if all variables
are independent of the response variable

```{R, echo=FALSE, eval=FALSE}
# Test different numbers of trees with impurity importance
tree_counts <- c(100, 250, 500, 1000, 2000, 5000)
results <- data.frame()

for(n_trees in tree_counts) {
  print(paste("Testing", n_trees, "trees"))

  # Time the training
  start_time <- Sys.time()

  rf_model <- ranger(
    dependent.variable.name = "pheno",
    data = dt,
    num.trees = n_trees,
    mtry = ncol(dt)/3,
    importance = 'impurity',
    seed = 123  # For reproducibility
  )

  end_time <- Sys.time()

  # Store results
  results <- rbind(results, data.frame(
    num_trees = n_trees,
    oob_error = rf_model$prediction.error,
    training_time = as.numeric(end_time - start_time, units = "secs"),
    top_var_importance = max(rf_model$variable.importance)
  ))
}

results<- structure(list(num_trees = c(100, 250, 500, 1000, 2000, 5000), 
    oob_error = c(0.342434347644852, 0.333055439766569, 0.334305960817007, 
    0.333680700291788, 0.330345977490621, 0.331804918716132), 
    training_time = c(217.293184995651, 437.686666965485, 794.375539779663, 
    1539.06033182144, 2960.24087452888, 7178.61296319962), top_var_importance = c(82.9423155959121, 
    86.0040477297634, 81.0779355241392, 83.3997545953136, 83.5774924785954, 
    83.5603383214029)), row.names = c(NA, -6L), class = "data.frame")

# Plot convergence
plot(results$num_trees, results$oob_error, type = "b", 
     xlab = "Number of Trees", ylab = "OOB Error",
     main = "OOB Error Convergence")

```

-  Error convergence: Should see diminishing returns after ~100-500 trees
-  Variable importance: Should stabilize but may need more trees than error
-  Computational time: Linear increase with number of trees

For Addressing Reviewer Concerns:

Since you're using impurity importance, you can argue:

    "We used impurity-based variable importance measures, which are deterministic and computationally efficient compared to permutation-based measures. While Probst et al. (2019) noted that permutation-based importance may require thousands of trees for stability in high-dimensional genomic data, impurity-based measures are less affected by this limitation as they are calculated during tree construction rather than through post-hoc permutation."

Both Gini AND Permutation Importance Have Bias Issues:

From the literature, both Gini and permutation importance suffer from bias toward variables with more categories:

We show that random forest variable importance measures are a sensible means for variable selection in many applications, but are not reliable in situations where potential predictor variables vary in their scale of measurement or their number of categories... the Gini importance shows a strong preference for variables with many categories and the continuous variable
How to Set the Number of Trees in Random Forest | Towards Data Science

And importantly: We now consider the more advanced permutation importance measure... We found that, even when the cforest function based on unbiased classification trees is used, variables with more categories are preferred when bootstrap sampling is conducted with replacement, while no bias occurs when subsampling is conducted without replacement
How to Set the Number of Trees in Random Forest | Towards Data Science
Strobl et al.'s Solution: Conditional Variable Importance

You're correct that Carolin Strobl and colleagues developed conditional variable importance to address these bias problems:

Based on these considerations we develop a new, conditional permutation scheme for the computation of the variable importance measure
11.6 Random Forest | Practitioner’s Guide to Data Science
The key insight was that We identify two mechanisms responsible for this finding: (i) A preference for the selection of correlated predictors in the tree building process and (ii) an additional advantage for correlated predictor variables induced by the unconditional permutation scheme

```{r, echo=FALSE, eval=FALSE}

set.seed(100)
#petrichor-i1: salloc -A OD-217714 --mem=100GB --nodes=1 --ntasks-per-node=64    -J interactive -t 6:00:00 srun --pty /bin/bash -l
#etrichor-i1: sacct -j 60755241  --format="jobid,ntasks,Elapsed,TotalCPU,MaxRSS,AveVMSize,MaxVMSize"
#JobID          NTasks    Elapsed   TotalCPU     MaxRSS  AveVMSize  MaxVMSize 
#60755241                00:13:56  00:06.682                                  
#60755241.ex+        1   00:13:56  00:00.001       995K    124080K    124080K 
#60755241.0         64   00:13:53  00:06.680 104545713K 3274946112 192422568K

# this is the model that was used in the main paper
system.time(model.rf <- ranger(dependent.variable.name="pheno",data = dt,num.trees = ntree, mtry = ncol(dt)/3, importance = 'impurity'))
#     user    system   elapsed 
#89397.171     4.807  1591.941 
#save(model.rf, file="model.rf_temp3.Rdata")
var_importance <- model.rf$variable.importance
head(rev(sort(var_importance)))
t2 <-count_variables(model.rf)
names(t2) <-names(var_importance)
#save(var_importance , t2, file="Results/var_importance_gini.Rdata")
#load(file="Results/var_importance_gini.Rdata")

set.seed(123)
system.time(rf_100 <-  ranger(dependent.variable.name="pheno",data = dt,num.trees = 100, mtry = ncol(dt)/3, importance = 'impurity')) 
#     user    system   elapsed 
#8897.449    3.043  224.81
#OOB prediction error:             34.66 % 
rf_100_variable.importance <- rf_100$variable.importance

set.seed(123)
system.time(rf_100 <-  ranger(dependent.variable.name="pheno",data = dt,num.trees = 100, mtry = ncol(dt)/3, importance = 'impurity')) 
#     user    system   elapsed 
#8897.449    3.043  224.81
#OOB prediction error:             34.66 % 
rf_100_variable.importance <- rf_100$variable.importance

set.seed(123)
system.time(rf_250 <-  ranger(dependent.variable.name="pheno",data = dt,num.trees = 250, mtry = ncol(dt)/3, importance = 'impurity')) 
#     user    system   elapsed 
#21996.459     3.480   414.998
#OOB prediction error:       33.62 % 
rf_250_variable.importance <- rf_250$variable.importance

set.seed(123)
system.time(rf_500 <-  ranger(dependent.variable.name="pheno",data = dt,num.trees = 500, mtry = ncol(dt)/3, importance = 'impurity')) 
#     user    system   elapsed 
#44385.171     6.711   813.847 
#OOB prediction error:             33.24 % 
#rf_500_variable.importance <- rf_500$variable.importance

system.time(rf_1000  <-  ranger(dependent.variable.name="pheno",data = dt,num.trees = 1000, mtry = ncol(dt)/3, importance = 'impurity'))
#     user    system   elapsed 
#90550.779     7.683  1691.032 
#OOB prediction error:             33.10 % 
rf_1000_variable.importance <- rf_1000$variable.importance

system.time(rf_2000  <-  ranger(dependent.variable.name="pheno",data = dt,num.trees = 2000, mtry = ncol(dt)/3, importance = 'impurity'))
#     user     system    elapsed 
#177425.239      9.037   3496.305 
#OOB prediction error:             33.08 % 
rf_2000_variable.importance <- rf_2000$variable.importance

system.time(rf_5000  <-  ranger(dependent.variable.name="pheno",data = dt,num.trees = 5000, mtry = ncol(dt)/3, importance = 'impurity'))
#      user     system    elapsed 
#441795.347      9.380   7199.468 
#OOB prediction error:             32.97 % 
#rf_5000_variable.importance <- rf_5000$variable.importance

save(rf_100_variable.importance , rf_250_variable.importance ,rf_500_variable.importance,rf_1000_variable.importance , rf_2000_variable.importance, 
rf_5000_variable.importance , file="./script4_temp.Rdata")
load(file="./script4_temp.Rdata")

# Compare variable importance stability
cor(rf_100_variable.importance, rf_250_variable.importance)   #0.9909538
cor(rf_250_variable.importance, rf_500_variable.importance)   #0.9970955
cor(rf_500_variable.importance, rf_1000_variable.importance)   #0.9958397
cor(rf_1000_variable.importance, var_importance)  #0.9971393

aa<-cbind(rf_100_variable.importance, rf_250_variable.importance,
          rf_500_variable.importance, rf_1000_variable.importance, var_importance)
cor(aa)
#                            rf_100_variable.importance rf_250_variable.importance rf_500_variable.importance rf_1000_variable.importance
#rf_100_variable.importance                   1.0000000                  0.9909538                  0.9866379                   0.9840345
#rf_250_variable.importance                   0.9909538                  1.0000000                  0.9970955                   0.9934666
#rf_500_variable.importance                   0.9866379                  0.9970955                  1.0000000                   0.9958397
#rf_1000_variable.importance                  0.9840345                  0.9934666                  0.9958397                   1.0000000

results2 <- data.frame(num.tree=c(100, 250, 500, 1000, 2000,5000),
               OOB_prediction_error=c( 34.66 ,33.62 , 33.24 , 33.10 ,33.08,  32.97  )/100  ,
               training_time = c(8897.449, 21996.459  , 44385.171   ,   90550.779 , 177425.239 ,441795.347      )
               )

# Plot convergence
#png("./script3/OOB_Error_Convergence.png")
plot(results$num_trees, results$oob_error, type = "b", 
      xlab = "Number of Trees", ylab = "OOB Error",
      main = "OOB Error Convergence",ylim=c(0.33,0.35))
points(results2$num.tree, results2$OOB_prediction_error, col="red", type = "b", pch=14)
#dev.off()

head(rev(sort(t2)))
#X6_32771829 X6_32471505 X6_32682149 X6_32222493 X6_32308125 X6_32406920 
#       1532         993         277         202         190         170

```

```{r OOBErrorConvergence, eval=TRUE, echo=FALSE, fig.cap='OOB Error Convergence',  fig.align = 'center', out.width = '50%'}
knitr::include_graphics("OOB_Error_Convergence.png")
```

- things to do for parameter optimization (in the Supplementarty information)

  - we are using Gini impurity

	  - show that this reasonable for SNP measurments
	  - do one timing run for permutation and one for cforest
  - do we need to optimize mtry etc

### 2025-07-02 Supplementary_Files/Supplementary_File_Rheumatoid_Data.Rmd
added Supplementary_Files and bib items

@@
 bibliography: 
+    ../epistaticInteractions.bib 
 header-includes:
@@
 ```{r RFlocalfdr, eval=TRUE, echo=FALSE, fig.cap='RFlocalfdr analysis',  fig.align = 'center', out.width = '50%'}
-knitr::include_graphics("./supp_figs/RFlocalfdr.png")
+knitr::include_graphics("RFlocalfdr.png")
 
@@
 ```{r hotspot, eval=TRUE, echo=FALSE, fig.cap='There is a hotspot in chromosome 6',  fig.align = 'center', out.width = '50%'}
-knitr::include_graphics("./supp_figs/hotspot.png")
+knitr::include_graphics("hotspot.png")
 
@@
 ```{r zigplogscaleextended, eval=TRUE, echo=FALSE, fig.cap='empiricalzip fitted ZIPG',  fig.align = 'center', out.width = '50%'}
-knitr::include_graphics("./supp_figs/zigp_log_scale_extended.png")
+knitr::include_graphics("zigp_log_scale_extended.png")
 
@@
 ```{R, zipfplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Power Law (Zipf) Distribution", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./supp_figs/ZI_Zipf_plot.png")
+knitr::include_graphics("ZI_Zipf_plot.png")
 ```
@@
 ```{R, zigpplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Generalized Poisson", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./supp_figs/ZIGP_plot.png")
+knitr::include_graphics("ZIGP_plot.png")
 ```
@@
 ```{R, zinbplot, echo=FALSE, eval=TRUE, fig.cap="Zero Inflated Negative Binomial", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./supp_figs/ZINB_plot.png")
+knitr::include_graphics("ZINB_plot.png")
 ```
@@
 ```{R, powerlaw, echo=FALSE, eval=TRUE, fig.cap="Plot of log(frequency) vs log(count). A power law would show a straight line", out.width = '50%',fig.align = 'center'}
-knitr::include_graphics("./supp_figs/power_law.png")
+knitr::include_graphics("power_law.png")
 ```
